---
title: "SI: LLM Simulation (Phase 1 Benchmark)"
format:
  html:
    toc: true
    code-fold: true
---

This notebook analyzes LLM reference game simulations to evaluate whether
AI agents can form stable conventions in Phase 1. Used as an AI detection
benchmark for the registered report.

**Prerequisites:** Run `process_llm_results.py` on a results directory first
to generate the CSVs this notebook reads.

```{r setup, message=FALSE, warning=FALSE}
source(here::here("analysis", "config.R"))
library(jsonlite)

# -- Find results directory (most recent llm_results_*) --
sim_dir <- here::here("analysis", "llm_simulation")
results_dirs <- sort(
  list.dirs(sim_dir, recursive = FALSE, full.names = TRUE)[
    grepl("llm_results_", basename(list.dirs(sim_dir, recursive = FALSE)))
  ],
  decreasing = TRUE
)
stopifnot("No llm_results_* directory found. Run the simulation first." = length(results_dirs) > 0)
results_dir <- results_dirs[1]
cat("Results directory:", basename(results_dir), "\n")

# Load CSVs produced by process_llm_results.py
rounds <- read_csv(file.path(results_dir, "llm_rounds.csv"), show_col_types = FALSE)
listener_trials <- read_csv(file.path(results_dir, "llm_listener_trials.csv"), show_col_types = FALSE)
exclusion <- read_csv(file.path(results_dir, "llm_exclusion.csv"), show_col_types = FALSE)
has_similarity <- file.exists(file.path(results_dir, "llm_similarity.csv"))
if (has_similarity) {
  similarity <- read_csv(file.path(results_dir, "llm_similarity.csv"), show_col_types = FALSE)
}

# Load config
config <- fromJSON(file.path(results_dir, "llm_config.json"))
cat("Model:", config$model, "\n")
cat("Temperature:", ifelse(is.null(config$temperature), "API default", config$temperature), "\n")
cat("Groups:", config$n_groups, "\n")
cat("Blocks:", config$n_blocks, "\n")
```

## Exclusion Criterion

The paper's exclusion criterion: groups pass Phase 1 if **at least 2 of 3 players
achieve at least 2/3 referential accuracy across the last 3 blocks**.

```{r exclusion-criterion}
group_summary <- exclusion %>%
  distinct(group_id, n_players_passing, group_passes)

n_groups <- nrow(group_summary)
n_passing <- sum(group_summary$group_passes)
pct_passing <- n_passing / n_groups * 100

cat(sprintf("Groups passing: %d/%d (%.0f%%)\n", n_passing, n_groups, pct_passing))
cat(sprintf("Threshold: >=66.7%% accuracy, >=2/3 players, last 3 blocks\n\n"))

exclusion %>%
  select(group_id, player_id, listener_accuracy, passes_threshold, group_passes) %>%
  knitr::kable(digits = 3)
```

## Accuracy Trajectory

```{r accuracy-trajectory, fig.width=6, fig.height=4}
# Per-group per-block accuracy
block_acc <- listener_trials %>%
  group_by(group_id, block_num) %>%
  summarise(accuracy = mean(correct), .groups = "drop")

# Aggregate across groups
acc_summary <- block_acc %>%
  group_by(block_num) %>%
  summarise(
    mean_acc = mean(accuracy),
    se = sd(accuracy) / sqrt(n()),
    .groups = "drop"
  )

p_acc <- ggplot(acc_summary, aes(x = block_num, y = mean_acc)) +
  geom_line(color = "steelblue") +
  geom_point(color = "steelblue", size = 2.5) +
  geom_errorbar(aes(ymin = mean_acc - se, ymax = mean_acc + se),
                width = 0.2, color = "steelblue") +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = 1/16, linetype = "dotted", color = "gray50", alpha = 0.5) +
  annotate("text", x = config$n_blocks - 1, y = 2/3 + 0.03,
           label = "Threshold (66.7%)", color = "red", size = 3, hjust = 1) +
  annotate("text", x = config$n_blocks - 1, y = 1/16 + 0.03,
           label = "Chance (6.25%)", color = "gray50", size = 3, hjust = 1) +
  scale_y_continuous(labels = scales::percent, limits = c(0, NA)) +
  labs(x = "Block", y = "Listener Accuracy",
       title = paste0("Accuracy Trajectory (", config$model, ")")) +
  theme(plot.title = element_text(size = 14))

p_acc
save_fig(p_acc, "SI_llm_accuracy_trajectory.pdf", width = 6, height = 4)

cat("Mean accuracy by block:\n")
for (i in seq_len(nrow(acc_summary))) {
  cat(sprintf("  Block %d: %.1f%% (SE=%.1f%%)\n",
              acc_summary$block_num[i],
              acc_summary$mean_acc[i] * 100,
              acc_summary$se[i] * 100))
}
```

## Word Count Trajectory

```{r wordcount-trajectory, fig.width=6, fig.height=4}
# Per-group per-block word count
block_wc <- rounds %>%
  group_by(group_id, block_num) %>%
  summarise(word_count = mean(word_count), .groups = "drop")

wc_summary <- block_wc %>%
  group_by(block_num) %>%
  summarise(
    mean_wc = mean(word_count),
    se = sd(word_count) / sqrt(n()),
    .groups = "drop"
  )

p_wc <- ggplot(wc_summary, aes(x = block_num, y = mean_wc)) +
  geom_line(color = "darkorange") +
  geom_point(color = "darkorange", size = 2.5) +
  geom_errorbar(aes(ymin = mean_wc - se, ymax = mean_wc + se),
                width = 0.2, color = "darkorange") +
  scale_y_continuous(limits = c(0, NA)) +
  labs(x = "Block", y = "Mean Word Count",
       title = paste0("Description Length (", config$model, ")")) +
  theme(plot.title = element_text(size = 14))

p_wc
save_fig(p_wc, "SI_llm_wordcount_trajectory.pdf", width = 6, height = 4)

cat("Mean word count by block:\n")
for (i in seq_len(nrow(wc_summary))) {
  cat(sprintf("  Block %d: %.1f (SE=%.1f)\n",
              wc_summary$block_num[i], wc_summary$mean_wc[i], wc_summary$se[i]))
}
```

## Description Similarity (SBERT)

For each group, we compute cosine similarity between descriptions of the same
tangram across blocks. Increasing similarity indicates convergence on stable
expressions.

```{r sbert-similarity, fig.width=6, fig.height=4}
if (has_similarity) {
  # Per-group per-block similarity
  block_sim <- similarity %>%
    group_by(group_id, block_num) %>%
    summarise(similarity = mean(similarity), .groups = "drop")

  sim_summary <- block_sim %>%
    group_by(block_num) %>%
    summarise(
      mean_sim = mean(similarity),
      se = sd(similarity) / sqrt(n()),
      .groups = "drop"
    )

  p_sim <- ggplot(sim_summary, aes(x = block_num, y = mean_sim)) +
    geom_line(color = "seagreen") +
    geom_point(color = "seagreen", size = 2.5) +
    geom_errorbar(aes(ymin = mean_sim - se, ymax = mean_sim + se),
                  width = 0.2, color = "seagreen") +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Block", y = "Cosine Similarity (adjacent descriptions)",
         title = paste0("Description Convergence (", config$model, ")")) +
    theme(plot.title = element_text(size = 14))

  p_sim
  save_fig(p_sim, "SI_llm_similarity_trajectory.pdf", width = 6, height = 4)

  cat("Mean adjacent similarity by block:\n")
  for (i in seq_len(nrow(sim_summary))) {
    cat(sprintf("  Block %d: %.3f (SE=%.3f)\n",
                sim_summary$block_num[i], sim_summary$mean_sim[i], sim_summary$se[i]))
  }

  overall_sim <- mean(similarity$similarity)
  overall_sim_sd <- similarity %>%
    group_by(group_id) %>%
    summarise(m = mean(similarity), .groups = "drop") %>%
    pull(m) %>%
    sd()
} else {
  cat("No similarity data. Run process_llm_results.py without --skip-similarity.\n")
  overall_sim <- NA
  overall_sim_sd <- NA
}
```

## Per-Group Accuracy Heatmap

```{r heatmap, fig.width=5, fig.height=max(3, n_groups * 0.5)}
heat_data <- exclusion %>%
  mutate(player_label = paste0("Player ", player_id))

p_heat <- ggplot(heat_data, aes(x = player_label, y = group_id, fill = listener_accuracy)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", listener_accuracy * 100)), size = 3.5) +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850",
                       midpoint = 0.33, limits = c(0, 1),
                       labels = scales::percent) +
  labs(x = NULL, y = NULL, fill = "Accuracy",
       title = paste0("Listener Accuracy (Last 3 Blocks)")) +
  theme(plot.title = element_text(size = 14))

p_heat
```

## Draft SI Paragraph

```{r draft-paragraph}
# Final stats
mean_last3 <- exclusion %>%
  summarise(m = mean(listener_accuracy) * 100) %>%
  pull(m)

sd_last3 <- exclusion %>%
  group_by(group_id) %>%
  summarise(m = mean(listener_accuracy) * 100, .groups = "drop") %>%
  pull(m) %>%
  sd()

block0_wc <- wc_summary$mean_wc[wc_summary$block_num == 0]
block_last_wc <- wc_summary$mean_wc[wc_summary$block_num == config$n_blocks - 1]
wc_direction <- ifelse(block_last_wc >= block0_wc * 0.85, "did not show", "showed")

temp_str <- ifelse(is.null(config$temperature),
                   "default temperature",
                   sprintf("temperature $= %s$", config$temperature))

if (!is.na(overall_sim)) {
  sim_str <- sprintf("%.2f (SD $= %.2f$)", overall_sim, overall_sim_sd)
  sim_interp <- ifelse(overall_sim < 0.5, "low", "moderate")
} else {
  sim_str <- "N/A"
  sim_interp <- "unknown"
}

paragraph <- sprintf(
"We simulated $N = %d$ independent groups of 3 \\texttt{%s} agents
playing the Phase~1 reference game (6 blocks $\\times$ 6 tangrams $= 36$ rounds
per group, %s). We evaluated each group against the same exclusion
criterion applied to human participants: at least 2 of 3 group members achieving
$\\geq$66.7\\%% referential accuracy across the last 3 blocks.
%d/%d groups (%.0f\\%%) met this criterion.
Mean listener accuracy was %.1f\\%% (SD $= %.1f$\\%%) across
the final 3 blocks, compared to the 66.7\\%% threshold. Description length
%s the expected reduction over blocks (block~1:
%.1f words; block~6: %.1f words). Cosine similarity
between adjacent descriptions of the same tangram was %s, indicating
%s convergence on stable expressions.",
  n_groups, config$model, temp_str,
  n_passing, n_groups, pct_passing,
  mean_last3, sd_last3,
  wc_direction, block0_wc, block_last_wc,
  sim_str, sim_interp
)

cat("============================================================\n")
cat("DRAFT SI PARAGRAPH (copy to LaTeX)\n")
cat("============================================================\n\n")
cat(paragraph, "\n")
```
