---
title: "SI: LLM Simulation (Phase 1 Benchmark)"
format:
  html:
    toc: true
    code-fold: true
---

This notebook analyzes LLM reference game simulations to evaluate whether
AI agents can form stable conventions in Phase 1. Used as an AI detection
benchmark for the registered report.

**Prerequisites:** Run `process_llm_results.py` on a results directory first
to generate the CSVs this notebook reads. Groups still marked `in_progress`
are automatically excluded by the processing script.

```{r setup, message=FALSE, warning=FALSE}
source(here::here("analysis", "config.R"))
library(jsonlite)
library(tidyboot)

# LLM plot color — distinct from condition palette (greens/oranges)
LLM_COLOR <- "#7B68AE"

# Override figures_dir for LLM outputs
figures_dir <- here::here("analysis", "llm_baseline_outputs")
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# -- Find results directory (most recent llm_results_*) --
sim_dir <- here::here("analysis", "llm_simulation")
results_dirs <- sort(
  list.dirs(sim_dir, recursive = FALSE, full.names = TRUE)[
    grepl("llm_results_", basename(list.dirs(sim_dir, recursive = FALSE)))
  ],
  decreasing = TRUE
)
stopifnot("No llm_results_* directory found. Run the simulation first." = length(results_dirs) > 0)
results_dir <- results_dirs[1]
cat("Results directory:", basename(results_dir), "\n")

# Load CSVs produced by process_llm_results.py
rounds <- read_csv(file.path(results_dir, "llm_rounds.csv"), show_col_types = FALSE)
listener_trials <- read_csv(file.path(results_dir, "llm_listener_trials.csv"), show_col_types = FALSE)
exclusion <- read_csv(file.path(results_dir, "llm_exclusion.csv"), show_col_types = FALSE)
has_similarity <- file.exists(file.path(results_dir, "llm_similarity.csv"))
if (has_similarity) {
  similarity <- read_csv(file.path(results_dir, "llm_similarity.csv"), show_col_types = FALSE)
}

# Load config
config <- fromJSON(file.path(results_dir, "llm_config.json"))

# -- Load pilot data for comparison overlay --
pilot_data_dir <- here::here("data", "pilots")
has_pilot <- file.exists(file.path(pilot_data_dir, "trials.csv"))
if (has_pilot) {
  pilot_utts_file <- if (file.exists(file.path(pilot_data_dir, "speaker_utterances_filtered.csv"))) {
    "speaker_utterances_filtered.csv"
  } else {
    "speaker_utterances.csv"
  }
  pilot_utts <- read_csv(file.path(pilot_data_dir, pilot_utts_file), show_col_types = FALSE) |>
    filter(phaseNum == 1)
  pilot_trials <- read_csv(file.path(pilot_data_dir, "trials.csv"), show_col_types = FALSE) |>
    filter(phaseNum == 1, role == "listener") |>
    mutate(correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"))
  pilot_has_adjacent <- file.exists(file.path(pilot_data_dir, "adjacent_similarities.csv"))
  if (pilot_has_adjacent) {
    pilot_adjacent <- read_csv(file.path(pilot_data_dir, "adjacent_similarities.csv"), show_col_types = FALSE) |>
      filter(phaseNum == 1)
  }
}
cat("Model:", config$model, "\n")
cat("Temperature:", ifelse(is.null(config$temperature), "API default", config$temperature), "\n")
cat("Groups:", config$n_groups, "\n")
cat("Blocks:", config$n_blocks, "\n")
```

## Exclusion criterion

The paper's exclusion criterion: groups pass Phase 1 if **at least 2 of 3 players
achieve at least 2/3 referential accuracy across the last 3 blocks**.

```{r exclusion-criterion}
group_summary <- exclusion |>
  distinct(group_id, n_players_passing, group_passes)

n_groups <- nrow(group_summary)
n_passing <- sum(group_summary$group_passes)
pct_passing <- n_passing / n_groups * 100

cat(sprintf("Groups passing: %d/%d (%.0f%%)\n", n_passing, n_groups, pct_passing))
cat(sprintf("Threshold: >=66.7%% accuracy, >=2/3 players, last 3 blocks\n\n"))

exclusion |>
  select(group_id, player_id, listener_accuracy, passes_threshold, group_passes) |>
  knitr::kable(digits = 3)
```

## Accuracy trajectory

```{r accuracy-trajectory, fig.width=6, fig.height=4}
# Per-group per-block accuracy
block_acc <- listener_trials |>
  group_by(group_id, block_num) |>
  summarise(accuracy = mean(correct), .groups = "drop")

# Bootstrap CIs across groups
acc_summary <- block_acc |>
  group_by(block_num) |>
  tidyboot_mean(accuracy, nboot = 1000)

p_acc <- ggplot(acc_summary, aes(x = block_num + 1, y = empirical_stat)) +
  geom_line(data = block_acc, aes(x = block_num + 1, y = accuracy, group = group_id),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
              fill = LLM_COLOR, alpha = 0.15) +
  geom_line(color = LLM_COLOR, linewidth = 1) +
  geom_point(color = LLM_COLOR, size = 2.5) +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = 1:6, labels = as.integer) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Block", y = "Proportion correct",
       title = "Listener accuracy")

cat("Mean accuracy by block:\n")
for (i in seq_len(nrow(acc_summary))) {
  cat(sprintf("  Block %d: %.1f%%\n",
              acc_summary$block_num[i] + 1,
              acc_summary$empirical_stat[i] * 100))
}
```

## Word count trajectory

```{r wordcount-trajectory, fig.width=6, fig.height=4}
# Per-group per-block word count
block_wc <- rounds |>
  group_by(group_id, block_num) |>
  summarise(word_count = mean(word_count), .groups = "drop")

# Bootstrap CIs across groups
wc_summary <- block_wc |>
  group_by(block_num) |>
  tidyboot_mean(word_count, nboot = 1000)

p_wc <- ggplot(wc_summary, aes(x = block_num + 1, y = empirical_stat)) +
  geom_line(data = block_wc, aes(x = block_num + 1, y = word_count, group = group_id),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
              fill = LLM_COLOR, alpha = 0.15) +
  geom_line(color = LLM_COLOR, linewidth = 1) +
  geom_point(color = LLM_COLOR, size = 2.5) +
  scale_x_continuous(breaks = 1:6, labels = as.integer) +
  labs(x = "Block", y = "Words per description",
       title = "Description length")

cat("Mean word count by block:\n")
for (i in seq_len(nrow(wc_summary))) {
  cat(sprintf("  Block %d: %.1f\n",
              wc_summary$block_num[i] + 1, wc_summary$empirical_stat[i]))
}
```

## Description similarity (SBERT)

For each group, we compute cosine similarity between descriptions of the same
tangram across blocks. Increasing similarity indicates convergence on stable
expressions.

```{r sbert-similarity, fig.width=6, fig.height=4}
if (has_similarity) {
  # Per-group per-block similarity
  block_sim <- similarity |>
    group_by(group_id, block_num) |>
    summarise(similarity = mean(similarity), .groups = "drop")

  # Bootstrap CIs across groups
  sim_summary <- block_sim |>
    group_by(block_num) |>
    tidyboot_mean(similarity, nboot = 1000)

  p_sim <- ggplot(sim_summary, aes(x = block_num + 1, y = empirical_stat)) +
    geom_line(data = block_sim, aes(x = block_num + 1, y = similarity, group = group_id),
              color = "gray70", alpha = 0.5, linewidth = 0.4) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                fill = LLM_COLOR, alpha = 0.15) +
    geom_line(color = LLM_COLOR, linewidth = 1) +
    geom_point(color = LLM_COLOR, size = 2.5) +
    scale_x_continuous(breaks = 1:6, labels = as.integer, limits = c(1, 6)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Block", y = "Cosine similarity",
         title = "Convention stability")

  cat("Mean adjacent similarity by block:\n")
  for (i in seq_len(nrow(sim_summary))) {
    cat(sprintf("  Block %d: %.3f\n",
                sim_summary$block_num[i] + 1, sim_summary$empirical_stat[i]))
  }

  overall_sim <- mean(similarity$similarity)
  overall_sim_sd <- similarity |>
    group_by(group_id) |>
    summarise(m = mean(similarity), .groups = "drop") |>
    pull(m) |>
    sd()
} else {
  cat("No similarity data. Run process_llm_results.py without --skip-similarity.\n")
  overall_sim <- NA
  overall_sim_sd <- NA
}
```

## Per-group accuracy heatmap

```{r heatmap, fig.width=5, fig.height=max(3, n_groups * 0.5)}
heat_data <- exclusion |>
  mutate(player_label = paste0("Player ", player_id))

p_heat <- ggplot(heat_data, aes(x = player_label, y = group_id, fill = listener_accuracy)) +
  geom_tile(color = "white", linewidth = 0.5) +
  geom_text(aes(label = sprintf("%.0f%%", listener_accuracy * 100)), size = 3.5) +
  scale_fill_gradient2(low = "#d73027", mid = "#fee08b", high = "#1a9850",
                       midpoint = 0.33, limits = c(0, 1),
                       labels = scales::percent) +
  labs(x = NULL, y = NULL, fill = "Accuracy",
       title = "Listener accuracy (last 3 blocks)")

p_heat
save_fig(p_heat, "SI_llm_exclusion_heatmap.pdf",
         width = 5, height = max(3, n_groups * 0.5))
```

## Combined panel figure

Three-panel figure matching the structure of the pilot outcome-neutral criteria
plots: description length, listener accuracy, and convention stability.

```{r panel-figure, fig.width=14, fig.height=4.5}
library(patchwork)

PILOT_COLOR <- "black"

# Compute pilot summaries if available
if (has_pilot) {
  pilot_wc_summary <- pilot_utts |>
    group_by(blockNum) |>
    tidyboot_mean(uttLength, nboot = 1000) |>
    mutate(block1 = blockNum + 1)

  pilot_acc_by_round <- pilot_trials |>
    group_by(gameId, roundId, blockNum, originalGroup) |>
    summarise(ref_accuracy = mean(correct, na.rm = TRUE), .groups = "drop")
  pilot_acc_summary <- pilot_acc_by_round |>
    group_by(blockNum) |>
    tidyboot_mean(ref_accuracy, nboot = 1000) |>
    mutate(block1 = blockNum + 1)

  if (pilot_has_adjacent) {
    pilot_sim_summary <- pilot_adjacent |>
      filter(!is.na(simAdjacent), blockNum > 0) |>
      group_by(blockNum) |>
      tidyboot_mean(simAdjacent, nboot = 1000) |>
      mutate(block1 = blockNum + 1)
  }
}

# Rebuild panels with pilot overlay
p_wc2 <- p_wc
p_acc2 <- p_acc
p_sim2 <- if (has_similarity) p_sim else NULL

legend_scale <- scale_color_manual(
  name = NULL,
  values = c("LLM (Gemini 3.1 Pro)" = LLM_COLOR, "Human (pilot)" = PILOT_COLOR),
  guide = guide_legend(override.aes = list(linewidth = 1, size = 2.5))
)

# Helper to add legend-mapped LLM line to a panel
add_llm_legend <- function(p, summary_df, x_col, y_col, group_df = NULL, group_y = NULL) {
  p <- p +
    geom_line(data = summary_df,
              aes(x = .data[[x_col]], y = .data[[y_col]],
                  color = "LLM (Gemini 3.1 Pro)"),
              linewidth = 1, inherit.aes = FALSE) +
    geom_point(data = summary_df,
               aes(x = .data[[x_col]], y = .data[[y_col]],
                   color = "LLM (Gemini 3.1 Pro)"),
               size = 2.5, inherit.aes = FALSE)
  p
}

# Rebuild panels: strip hardcoded LLM color, add via aes for legend
# Word count
llm_wc_for_legend <- wc_summary |> mutate(block1 = block_num + 1)
p_wc2 <- ggplot() +
  geom_line(data = block_wc, aes(x = block_num + 1, y = word_count, group = group_id),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(data = wc_summary, aes(x = block_num + 1, ymin = ci_lower, ymax = ci_upper),
              fill = LLM_COLOR, alpha = 0.15) +
  geom_line(data = llm_wc_for_legend,
            aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
            linewidth = 1) +
  geom_point(data = llm_wc_for_legend,
             aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
             size = 2.5) +
  scale_x_continuous(breaks = 1:6, labels = as.integer) +
  labs(x = "Block", y = "Words per description", title = "Description length") +
  legend_scale

# Accuracy
llm_acc_for_legend <- acc_summary |> mutate(block1 = block_num + 1)
p_acc2 <- ggplot() +
  geom_line(data = block_acc, aes(x = block_num + 1, y = accuracy, group = group_id),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(data = acc_summary, aes(x = block_num + 1, ymin = ci_lower, ymax = ci_upper),
              fill = LLM_COLOR, alpha = 0.15) +
  geom_line(data = llm_acc_for_legend,
            aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
            linewidth = 1) +
  geom_point(data = llm_acc_for_legend,
             aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
             size = 2.5) +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "red", alpha = 0.7) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  scale_x_continuous(breaks = 1:6, labels = as.integer) +
  scale_y_continuous(limits = c(0, 1)) +
  labs(x = "Block", y = "Proportion correct", title = "Listener accuracy") +
  legend_scale

# Similarity
p_sim2 <- NULL
if (has_similarity) {
  llm_sim_for_legend <- sim_summary |> mutate(block1 = block_num + 1)
  p_sim2 <- ggplot() +
    geom_line(data = block_sim, aes(x = block_num + 1, y = similarity, group = group_id),
              color = "gray70", alpha = 0.5, linewidth = 0.4) +
    geom_ribbon(data = sim_summary, aes(x = block_num + 1, ymin = ci_lower, ymax = ci_upper),
                fill = LLM_COLOR, alpha = 0.15) +
    geom_line(data = llm_sim_for_legend,
              aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
              linewidth = 1) +
    geom_point(data = llm_sim_for_legend,
               aes(x = block1, y = empirical_stat, color = "LLM (Gemini 3.1 Pro)"),
               size = 2.5) +
    scale_x_continuous(breaks = 1:6, labels = as.integer, limits = c(1, 6)) +
    scale_y_continuous(limits = c(0, 1)) +
    labs(x = "Block", y = "Cosine similarity", title = "Convention stability") +
    legend_scale
}

# Add pilot overlay
if (has_pilot) {
  p_wc2 <- p_wc2 +
    geom_ribbon(data = pilot_wc_summary,
                aes(x = block1, ymin = ci_lower, ymax = ci_upper),
                fill = PILOT_COLOR, alpha = 0.08) +
    geom_line(data = pilot_wc_summary,
              aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
              linewidth = 1) +
    geom_point(data = pilot_wc_summary,
               aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
               size = 2.5)

  p_acc2 <- p_acc2 +
    geom_ribbon(data = pilot_acc_summary,
                aes(x = block1, ymin = ci_lower, ymax = ci_upper),
                fill = PILOT_COLOR, alpha = 0.08) +
    geom_line(data = pilot_acc_summary,
              aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
              linewidth = 1) +
    geom_point(data = pilot_acc_summary,
               aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
               size = 2.5)

  if (!is.null(p_sim2) && pilot_has_adjacent) {
    p_sim2 <- p_sim2 +
      geom_ribbon(data = pilot_sim_summary,
                  aes(x = block1, ymin = ci_lower, ymax = ci_upper),
                  fill = PILOT_COLOR, alpha = 0.08) +
      geom_line(data = pilot_sim_summary,
                aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
                linewidth = 1) +
      geom_point(data = pilot_sim_summary,
                 aes(x = block1, y = empirical_stat, color = "Human (pilot)"),
                 size = 2.5)
  }
}

if (!is.null(p_sim2)) {
  p_combined <- p_wc2 + p_acc2 + p_sim2 +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 20),
          legend.position = "bottom")
} else {
  p_combined <- p_wc2 + p_acc2 +
    plot_layout(guides = "collect") +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 20),
          legend.position = "bottom")
}

p_combined
save_fig(p_combined, "SI_llm_combined.pdf", width = 14, height = 5)
```

## Example descriptions across blocks

To illustrate the lack of stable convention formation, we show how one group's
descriptions of the same tangram evolve (or fail to converge) across blocks.

```{r example-descriptions}
# Pick the group with median accuracy and show one tangram's descriptions
median_group <- block_acc |>
  group_by(group_id) |>
  summarise(mean_acc = mean(accuracy), .groups = "drop") |>
  arrange(mean_acc) |>
  slice(ceiling(n() / 2)) |>
  pull(group_id)

example_tangram <- rounds |>
  filter(group_id == median_group) |>
  count(target) |>
  slice_max(n, n = 1) |>
  pull(target)

rounds |>
  filter(group_id == median_group, target == example_tangram) |>
  select(block_num, speaker_id, word_count, description) |>
  knitr::kable(col.names = c("Block", "Speaker", "Words", "Description"),
               caption = sprintf("Descriptions of %s by %s across blocks",
                                 example_tangram, median_group))
```

## Draft SI paragraph

```{r draft-paragraph}
# Final stats
mean_last3 <- exclusion |>
  summarise(m = mean(listener_accuracy) * 100) |>
  pull(m)

sd_last3 <- exclusion |>
  group_by(group_id) |>
  summarise(m = mean(listener_accuracy) * 100, .groups = "drop") |>
  pull(m) |>
  sd()

block0_wc <- wc_summary$empirical_stat[wc_summary$block_num == 0]
block_last_wc <- wc_summary$empirical_stat[wc_summary$block_num == config$n_blocks - 1]
wc_direction <- ifelse(block_last_wc >= block0_wc * 0.85, "did not show", "showed")

temp_str <- ifelse(is.null(config$temperature),
                   "default temperature",
                   sprintf("temperature $= %s$", config$temperature))

if (!is.na(overall_sim)) {
  sim_str <- sprintf("%.2f (SD $= %.2f$)", overall_sim, overall_sim_sd)
  sim_interp <- ifelse(overall_sim < 0.5, "low", "moderate")
} else {
  sim_str <- "N/A"
  sim_interp <- "unknown"
}

paragraph <- sprintf(
"We simulated $N = %d$ independent groups of 3 \\texttt{%s} agents
playing the Phase~1 reference game (6 blocks $\\times$ 6 tangrams $= 36$ rounds
per group, %s). We evaluated each group against the same exclusion
criterion applied to human participants: at least 2 of 3 group members achieving
$\\geq$66.7\\%% referential accuracy across the last 3 blocks.
%d/%d groups (%.0f\\%%) met this criterion.
Mean listener accuracy was %.1f\\%% (SD $= %.1f$\\%%) across
the final 3 blocks, compared to the 66.7\\%% threshold. Description length
%s the expected reduction over blocks (block~1:
%.1f words; block~6: %.1f words). Cosine similarity
between adjacent descriptions of the same tangram was %s, indicating
%s convergence on stable expressions.",
  n_groups, config$model, temp_str,
  n_passing, n_groups, pct_passing,
  mean_last3, sd_last3,
  wc_direction, block0_wc, block_last_wc,
  sim_str, sim_interp
)

cat("============================================================\n")
cat("DRAFT SI PARAGRAPH (copy to LaTeX)\n")
cat("============================================================\n\n")
cat(paragraph, "\n")
```
