---
title: Power analysis plots
author: Alicia Chen
date: 2025-12-19
format:
  html:
    code-fold: false
---

```{r setup}
library(here)
library(tidyverse)
library(ggplot2)
library(tidyboot)

theme_set(theme_classic(base_size = 20))
```

# Load data

```{r}
pairwise_mixed <- read.csv(here("analysis/power_analysis/pairwise_mixed.csv"))
pairwise_separate <- read.csv(here("analysis/power_analysis/pairwise_separate.csv"))
results_mixed <- read.csv(here("analysis/power_analysis/results_mixed.csv"))
results_separate <- read.csv(here("analysis/power_analysis/results_separate.csv"))
```

Stack the pairwise dataframes, and the results dataframes, and add columns for condition

```{r}
pairwise_all <- bind_rows(
  pairwise_mixed %>% mutate(condition = "Mixed", gameId = as.character(gameId)),
  pairwise_separate %>% mutate(condition = "Separate", gameId = as.character(gameId))
)

results_all <- bind_rows(
  results_mixed %>% mutate(gameId = as.character(gameId)),
  results_separate %>% mutate(gameId = as.character(gameId))
)

head(pairwise_all)
head(results_all)
```

# Plot results

## Within-group vs Between-group similarity

```{r}
pairwise_all_summary <- pairwise_all |> 
  group_by(condition, same_group) |> 
  tidyboot_mean(similarity, na.rm = T)

pairwise_all_summary_game <- pairwise_all |> 
  group_by(gameId, condition, same_group) |> 
  tidyboot_mean(similarity, na.rm = T)
```

Compare similarity distributions for within-group vs between-group pairs by condition:

```{r}
ggplot(pairwise_all, aes(x = condition, y = similarity, fill = factor(same_group))) +
  geom_violin(alpha = 0.7, position = position_dodge(width = 0.6)) +
  geom_errorbar(data = pairwise_all_summary,
                aes(x = condition, y = mean, ymin = ci_lower, ymax = ci_upper, group = factor(same_group)),
                position = position_dodge(width = 0.6),
                width = 0.15,
                inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Set2", labels = c("Between group", "Within group")) +
  labs(
    title = "Pairwise similarities",
    x = "Condition",
    y = "Cosine similarity",
    fill = "Comparison"
  ) +
  theme(legend.position = "right")
```

## Group-specificity scores

Distribution of same_group coefficients across simulated games

```{r}
group_specificity_summary <- results_all |> 
  group_by(condition) |> 
  tidyboot_mean(coefficient)
```


```{r}
ggplot(results_all, aes(x = condition, y = coefficient)) +
  geom_point(data = group_specificity_summary, aes(x = condition, y = empirical_stat)) +
  geom_errorbar(data = group_specificity_summary, aes(x = condition, ymin = ci_lower, ymax = ci_upper), inherit.aes = F, width = 0.2) +
  geom_jitter(alpha = 0.3, width = 0.2, height = 0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = "Group-specificity scores",
    x = "Condition",
    y = "Coefficient"
  ) +
  theme(legend.position = "none")
```

## Individual game coefficients

Show coefficient for each simulated game:

```{r}
results_all %>%
  arrange(coefficient) %>%
  mutate(game_rank = row_number(),
         ci_lower = coefficient - std_error,
         ci_upper = coefficient + std_error) %>%
  ggplot(aes(x = game_rank, y = coefficient, color = condition)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  geom_point(size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Group-specificity scores by game",
    x = "Game (ranked by coefficient)",
    y = "Coefficient",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```
# Power analysis

