---
title: Power analysis plots
author: Alicia Chen
date: 2025-12-19
format:
  html:
    code-fold: false
---

```{r setup}
library(here)
library(tidyverse)
library(ggplot2)
library(tidyboot)
library(effectsize)

theme_set(theme_classic(base_size = 20))
```

# Load data

```{r}
pairwise_mixed <- read.csv(here("analysis/power_analysis/pairwise_mixed.csv"))
pairwise_separate <- read.csv(here("analysis/power_analysis/pairwise_separate.csv"))
results_mixed <- read.csv(here("analysis/power_analysis/results_mixed.csv"))
results_separate <- read.csv(here("analysis/power_analysis/results_separate.csv"))
```

Stack the pairwise dataframes, and the results dataframes, and add columns for condition

```{r}
pairwise_all <- bind_rows(
  pairwise_mixed %>% mutate(condition = "Mixed", gameId = as.character(gameId)),
  pairwise_separate %>% mutate(condition = "Separate", gameId = as.character(gameId))
)

results_all <- bind_rows(
  results_mixed %>% mutate(gameId = as.character(gameId)),
  results_separate %>% mutate(gameId = as.character(gameId))
)

head(pairwise_all)
head(results_all)
```

# Plot results

## Within-group vs Between-group similarity

```{r}
pairwise_all_summary <- pairwise_all |> 
  group_by(condition, same_group) |> 
  tidyboot_mean(similarity, na.rm = T)

pairwise_all_summary_game <- pairwise_all |> 
  group_by(gameId, condition, same_group) |> 
  tidyboot_mean(similarity, na.rm = T)
```

Compare similarity distributions for within-group vs between-group pairs by condition:

```{r}
ggplot(pairwise_all, aes(x = condition, y = similarity, fill = factor(same_group))) +
  geom_violin(alpha = 0.7, position = position_dodge(width = 0.6)) +
  geom_errorbar(data = pairwise_all_summary,
                aes(x = condition, y = mean, ymin = ci_lower, ymax = ci_upper, group = factor(same_group)),
                position = position_dodge(width = 0.6),
                width = 0.15,
                inherit.aes = FALSE) +
  scale_fill_brewer(palette = "Set2", labels = c("Between group", "Within group")) +
  labs(
    title = "Pairwise similarities",
    x = "Condition",
    y = "Cosine similarity",
    fill = "Comparison"
  ) +
  theme(legend.position = "right")
```

## Group-specificity scores

Distribution of same_group coefficients across simulated games

```{r}
group_specificity_summary <- results_all |> 
  group_by(condition) |> 
  tidyboot_mean(coefficient)
```


```{r}
ggplot(results_all, aes(x = condition, y = coefficient)) +
  geom_point(data = group_specificity_summary, aes(x = condition, y = empirical_stat)) +
  geom_errorbar(data = group_specificity_summary, aes(x = condition, ymin = ci_lower, ymax = ci_upper), inherit.aes = F, width = 0.2) +
  geom_jitter(alpha = 0.3, width = 0.2, height = 0) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray") +
  labs(
    title = "Group-specificity scores",
    x = "Condition",
    y = "Coefficient"
  ) +
  theme(legend.position = "none")
```

## Individual game coefficients

Show coefficient for each simulated game:

```{r}
results_all %>%
  arrange(coefficient) %>%
  mutate(game_rank = row_number(),
         ci_lower = coefficient - std_error,
         ci_upper = coefficient + std_error) %>%
  ggplot(aes(x = game_rank, y = coefficient, color = condition)) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.3) +
  geom_point(size = 2, alpha = 0.7) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_color_brewer(palette = "Set1") +
  labs(
    title = "Group-specificity scores by game",
    x = "Game (ranked by coefficient)",
    y = "Coefficient",
    color = "Condition"
  ) +
  theme(legend.position = "bottom")
```
# Power analysis

How many games do we need to detect an effect between 'mixed' and 'separate' for group-specificity?

What about for 75% and 50% of the effect above?

## Observed effect size

First, calculate the observed difference in group-specificity coefficients between conditions:

```{r}
# Calculate summary statistics by condition
effect_stats <- results_all %>%
  group_by(condition) %>%
  summarise(
    n = n(),
    mean_coef = mean(coefficient),
    sd_coef = sd(coefficient),
    .groups = "drop"
  )

effect_stats

d_result <- cohens_d(coefficient ~ condition, data = results_all)
d_result

mean_mixed <- effect_stats$mean_coef[effect_stats$condition == "Mixed"]
mean_separate <- effect_stats$mean_coef[effect_stats$condition == "Separate"]
observed_diff <- abs(mean_separate - mean_mixed)
cohens_d_value <- abs(as.numeric(d_result$Cohens_d))
pooled_sd <- observed_diff / cohens_d_value

cat("Observed difference |Separate - Mixed|:", round(observed_diff, 4), "\n")
cat("Pooled SD:", round(pooled_sd, 4), "\n")
```

## Power analysis

Using a two-sample t-test with alpha = 0.05 and power = 0.80:

```{r}
# Function to calculate required n per group for a given effect size
calc_required_n <- function(effect_size, pooled_sd, alpha = 0.05, power = 0.80) {
  result <- power.t.test(
    delta = effect_size,
    sd = pooled_sd,
    sig.level = alpha,
    power = power,
    type = "two.sample",
    alternative = "two.sided"
  )
  return(ceiling(result$n))
}

# Calculate for 100%, 75%, and 50% of observed effect
effect_proportions <- c(1.0, 0.75, 0.50)

power_results <- tibble(
  effect_proportion = effect_proportions,
  effect_size = observed_diff * effect_proportions,
  cohens_d = cohens_d_value * effect_proportions,
  n_per_condition = sapply(observed_diff * effect_proportions,
                            calc_required_n,
                            pooled_sd = pooled_sd)
)

power_results
```

## Power curve

Visualize power as a function of sample size:

```{r}
sample_sizes <- seq(2, 30, by = 1)

power_curves <- expand_grid(
  n = sample_sizes,
  effect_proportion = effect_proportions
) %>%
  mutate(
    effect_size = observed_diff * effect_proportion,
    power = pmap_dbl(list(n, effect_size), function(n, es) {
      power.t.test(
        n = n,
        delta = es,
        sd = pooled_sd,
        sig.level = 0.05,
        type = "two.sample",
        alternative = "two.sided"
      )$power
    }),
    effect_label = paste0(effect_proportion * 100, "% effect")
  )

ggplot(power_curves, aes(x = n, y = power, color = effect_label)) +
  geom_line(linewidth = 1.2) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray") +
  scale_y_continuous(limits = c(0, 1), breaks = seq(0, 1, 0.2)) +
  labs(
    title = "Power curves",
    x = "Games per condition",
    y = "Power",
    color = "Effect size"
  ) +
  theme(legend.position = "bottom")
```

