---
title: "00: Data overview"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(jsonlite)
source(here::here("analysis", "config.R"))
# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/overview.tex)
macros <- list()
```

# Load data

```{r load-data}
list2env(load_tables(), envir = globalenv())
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

# Game-level summary

```{r game-summary}
games |>
  group_by(condition) |>
  summarise(
    n_games = n(),
    mean_players = mean(numPlayers, na.rm = TRUE),
    mean_p1_blocks = mean(phase1Blocks, na.rm = TRUE),
    mean_p2_blocks = mean(phase2Blocks, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Games per condition")
```

# Player-level summary

```{r player-summary}
players |>
  group_by(gameId) |>
  summarise(
    n_players = n(),
    n_active = sum(isActive == TRUE | isActive == "true", na.rm = TRUE),
    n_inactive = sum(isActive == FALSE | isActive == "false", na.rm = TRUE),
    mean_score = mean(as.numeric(score), na.rm = TRUE),
    mean_bonus = mean(as.numeric(bonus), na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Players per game")
```

```{r player-demographics}
# Summarize exit survey demographics if available
demo_cols <- names(players)[str_detect(names(players), "^exitSurvey_")]
if (length(demo_cols) > 0) {
  cat("Exit survey fields:", paste(demo_cols, collapse = ", "), "\n")

  if ("exitSurvey_gender" %in% names(players)) {
    players |>
      count(exitSurvey_gender) |>
      knitr::kable(caption = "Gender distribution")
  }

  if ("exitSurvey_education" %in% names(players)) {
    players |>
      count(exitSurvey_education) |>
      knitr::kable(caption = "Education distribution")
  }
}
```

# Trial-level summary

```{r trial-summary}
cat("Total trial rows:", nrow(trials), "\n")
cat("Unique games:", n_distinct(trials$gameId), "\n")
cat("Unique players:", n_distinct(trials$playerId), "\n")

# Referential accuracy over eligible listener response opportunities: an
# active listener on a played trial whose speaker sent a message. Timeouts
# and late arrivals count as unsuccessful; trials the listener could not have
# answered leave the denominator instead of counting against them, and
# observations removed by a technical failure or a post-hoc participant
# exclusion are missing (the excluded_* columns).
ref_opportunities <- bind_rows(
  listener_trials(trials, games, phase = 1),
  listener_trials(trials, games, phase = 2)
)

response_opportunity_summary(ref_opportunities, by = "phaseNum") |>
  knitr::kable(
    digits = 3,
    caption = "Referential accuracy by phase, over eligible response opportunities"
  )
```

The table separates nonresponse from incorrect answers: `on_time_rate` is the
proportion of opportunities answered by the server's scoring cutoff, and
`accuracy_on_time` is accuracy among those answers. `accuracy` is the primary
outcome and counts an unanswered opportunity as unsuccessful.

```{r ref-accuracy-condition}
response_opportunity_summary(ref_opportunities) |>
  knitr::kable(
    digits = 3,
    caption = "Referential accuracy by condition and phase"
  )

no_opportunity <- trials |>
  filter(role == "listener", !responseOpportunity) |>
  nrow()
cat(no_opportunity, "listener trials had no speaker message and are excluded",
    "from the accuracy denominators.\n")
```

# Exclusions

Post-hoc participant exclusions (paper, "Exclusion criteria") are listed in
`data/<dataset>/participant_exclusions.csv` and applied by the pipeline: the
excluded participants' messages leave the corpus, and their own rows of
`trials.csv` and `social_guesses.csv`, plus the rows of the listeners they
spoke to, carry `excluded` and `exclusionReason`. `code_accuracy()` treats
those observations as missing, not unsuccessful, alongside the documented
technical failures in `technical_exclusions.csv`. The counts here are the
rows affected; the accuracy tables above report them as `excluded_participant`
and `excluded_technical`.

```{r participant-exclusions}
excl_file <- participant_exclusions()
if (nrow(excl_file) == 0) {
  cat(sprintf(
    "No post-hoc participant exclusions recorded (data/%s/participant_exclusions.csv absent or empty).\n",
    DATASET
  ))
} else {
  cat(sprintf("%d participant(s) excluded after the fact:\n", nrow(excl_file)))
  print(knitr::kable(excl_file |> mutate(playerId = substr(playerId, 1, 10))))
}

is_flagged <- function(x) (x == TRUE | x == "true") %in% TRUE
n_excluded_trial_rows <- 0L
if (!exclusion_columns_present(trials)) {
  cat("NOTE: trials.csv has no excluded/exclusionReason columns (regenerate with the current pipeline);",
      "no post-hoc exclusions are applied in this render.\n")
} else {
  flagged <- trials |> add_condition(games) |> filter(is_flagged(excluded))
  n_excluded_trial_rows <- nrow(flagged)
  if (nrow(flagged) == 0) {
    cat("No trial rows are flagged excluded.\n")
  } else {
    cat("\nTrial rows flagged excluded, by condition, role, and reason:\n")
    print(knitr::kable(flagged |> count(condition, role, exclusionReason, name = "rows")))
  }
  if (has_rows(social_guesses) && exclusion_columns_present(social_guesses)) {
    sg_flagged <- social_guesses |> add_condition(games) |> filter(is_flagged(excluded))
    cat(sprintf("\nSocial-guess opportunities flagged excluded: %d\n", nrow(sg_flagged)))
    if (nrow(sg_flagged) > 0) {
      print(knitr::kable(sg_flagged |> count(condition, exclusionReason, name = "rows")))
    }
  }
}

tech <- technical_exclusions()
cat(sprintf("\nDocumented technical failures (technical_exclusions.csv): %d\n", nrow(tech)))
if (nrow(tech) > 0) {
  print(knitr::kable(tech |> count(outcome, reason, name = "observations")))
}

macros <- c(macros, list(
  overviewNExcludedParticipants = nrow(excl_file),
  overviewNExcludedTrialRows = n_excluded_trial_rows,
  overviewNTechnicalExclusions = nrow(tech)
))
```

# Message summary

```{r message-summary}
cat("Total messages:", nrow(messages), "\n")
cat("Speaker messages:", sum(messages$senderRole == "speaker", na.rm = TRUE), "\n")
cat("Listener messages:", sum(messages$senderRole == "listener", na.rm = TRUE), "\n")

messages |>
  mutate(wordCount = str_count(text, "\\S+")) |>
  group_by(senderRole) |>
  summarise(
    n = n(),
    mean_words = mean(wordCount, na.rm = TRUE),
    median_words = median(wordCount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Message statistics by role")
```

# Speaker utterance summary

```{r utterance-summary}
cat("Total speaker utterances:", nrow(speaker_utts), "\n")

speaker_utts |>
  group_by(phase, phaseNum) |>
  summarise(
    n = n(),
    mean_length = mean(uttLength, na.rm = TRUE),
    sd_length = sd(uttLength, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Speaker utterance length by phase")
```

## Speaker trials without an analyzable description

A speaker trial has an analyzable description when it appears in the
(filtered) speaker-utterance table. The endpoint coverage rule in
`02_primary_analysis.qmd` counts such descriptions, so the proportion of
speaker trials that lack one, by condition and phase, is reported here with
its reasons: the speaker sent nothing (`silent`), the speaker was excluded
after the fact (`excluded`), or every message the speaker sent was classified
as non-referential (`non_referential_only`).

```{r description-coverage}
desc_cov <- speaker_description_coverage(trials, speaker_utts, games)
knitr::kable(
  desc_cov,
  digits = 3,
  caption = "Speaker trials without an analyzable description, by condition and phase"
)
macros <- c(macros, list(
  overviewPropNoDescription = fmt_stat(
    sum(desc_cov$without_description) / sum(desc_cov$speaker_trials), 3
  )
))
```

# Social guesses summary

```{r social-guess-summary}
if (nrow(social_guesses) > 0) {
  # Same denominator as referential accuracy: social_guesses.csv is an
  # opportunity frame, so a listener who never answered is a row with an
  # empty guess and counts as unsuccessful rather than disappearing.
  social_opportunities <- social_guess_trials(social_guesses, games)
  cat("Social-identification opportunities:", nrow(social_opportunities), "\n")
  cat("Never answered:", sum(social_opportunities$socialTimeout), "\n")

  response_opportunity_summary(social_opportunities) |>
    knitr::kable(
      digits = 3,
      caption = "Social-identification accuracy by condition, over eligible response opportunities"
    )
} else {
  cat("No social guess data (expected for refer_separated and refer_mixed conditions)\n")
}
```

The components of social-identification accuracy (paper,
"Social-identification components and opportunities"): in-group and
out-group opportunities, the on-time response rate, the in-group hit rate,
the out-group false-alarm rate, and balanced accuracy. The full table, with
the on-time accuracy and the answered counts, is in `02_primary_analysis.qmd`.

```{r social-components}
if (exists("social_opportunities") && has_rows(social_opportunities)) {
  social_components <- social_identification_components(social_opportunities, trials)
  knitr::kable(
    social_components |> select(all_of(SOCIAL_COMPONENT_COLUMNS_SHORT)),
    digits = 3,
    caption = "Social-identification components by condition (Phase 2)"
  )
} else {
  cat("No social-identification opportunities in this dataset.\n")
}
```

# Exclusion criteria checks

```{r exclusion-checks}
# Groups where <2/3 participants achieved <2/3 accuracy in last 3 blocks of
# Phase 1. This mirrors the live screen in experiment/server/src/accuracy.js,
# which is unchanged: it counts only trials where the speaker sent a message
# and treats a missing selection as incorrect, which is what the response
# opportunity rule below reproduces. (`p1_listener` rather than
# `listener_trials`, which is the function that builds it.)
p1_listener <- listener_trials(trials, games, phase = 1)

# Get phase1Blocks per game
game_p1 <- games |> select(gameId, phase1Blocks)

listener_accuracy <- p1_listener |>
  left_join(game_p1, by = "gameId") |>
  filter(blockNum >= phase1Blocks - 3) |>
  group_by(gameId, playerId, originalGroup) |>
  summarise(
    accuracy = mean(correct, na.rm = TRUE),
    n_trials = n(),
    .groups = "drop"
  )

# Flag players below 2/3 accuracy
listener_accuracy <- listener_accuracy |>
  mutate(meets_threshold = accuracy >= 2/3)

# Check groups
group_accuracy <- listener_accuracy |>
  group_by(gameId, originalGroup) |>
  summarise(
    n_players = n(),
    n_meeting_threshold = sum(meets_threshold),
    prop_meeting = n_meeting_threshold / n_players,
    group_meets = prop_meeting >= 2/3,
    .groups = "drop"
  )

cat("\nGroup accuracy check (last 3 blocks Phase 1):\n")
group_accuracy |>
  knitr::kable()

excluded_groups <- group_accuracy |> filter(!group_meets)
if (nrow(excluded_groups) > 0) {
  cat("\nGroups flagged for exclusion:\n")
  print(excluded_groups)
} else {
  cat("\nNo groups flagged for exclusion.\n")
}
```

# Attrition

The preregistration commits to reporting attrition and exclusions by
condition at the participant, group, and game levels, and to testing whether
dropout is differential across conditions (paper, "Dropout handling"). A
player counts as removed when the server took them out of the game. Two kinds
of removal are distinguished: for the player's own inactivity (idle for three
rounds) or their group's low Phase 1 accuracy, and through no fault of their
own, when their group disbanded, too few groups remained for the game to go
on (`insufficient groups`), or the researcher stopped the batch
(`game terminated`); the latter are compensated for their time. A game ran to
completion when at least two groups reached the end; a complete nine-player
game kept every player. The Phase 2 analyses are rerun on complete games in
`02_primary_analysis.qmd`. The last column counts players flagged for a
Phase 1 description-length increase (the AI-use trigger).

```{r attrition}
pa <- player_attrition(players, games)
status <- game_status(players, games)

cat("Players by condition:\n")
print(knitr::kable(attrition_by_condition(pa), digits = 2))

cat("\nGames by condition:\n")
print(knitr::kable(games_by_condition(status), digits = 2))

if (any(pa$removed)) {
  cat("\nRemoved players:\n")
  print(knitr::kable(pa |> filter(removed) |>
                       select(condition, gameId, originalGroup, reason, no_fault, idleRounds) |>
                       mutate(gameId = substr(gameId, 1, 10)) |> arrange(condition)))
}

tests <- differential_dropout_tests(pa, status)
if (length(tests) == 0) {
  cat("\nNo dropout to test: no player was removed, or every player was.\n")
} else {
  if (!is.null(tests$players)) {
    cat("\nPlayers removed by condition (chi-square; players are clustered in games, so treat as descriptive):\n")
    print(tests$players)
  }
  if (!is.null(tests$games)) {
    cat("\nComplete nine-player games by condition (Fisher's exact test; the game is the unit):\n")
    print(tests$games)
  }
}
```

## Original groups

One row per original group: an intact group had no removal; a lost group
ended with fewer than two active members (the server's `MIN_GROUP_SIZE`),
whether because it failed the Phase 1 accuracy screen (`low_accuracy`) or
because it disbanded.

```{r attrition-groups}
ga <- group_attrition(players, games)
cat("Original groups by condition:\n")
print(knitr::kable(groups_by_condition(ga), digits = 2))

groups_with_removal <- ga |> filter(removed > 0)
if (nrow(groups_with_removal) > 0) {
  cat("\nGroups with at least one removal:\n")
  print(knitr::kable(groups_with_removal |> mutate(gameId = substr(gameId, 1, 10))))
} else {
  cat("\nNo group lost a member.\n")
}
```

## Players who never played a real game

`data/<dataset>/dropouts.csv` (written by `combine_runs.py`) holds one row per
player record that never reached a real game: participants who failed the
comprehension quiz three times, waited in a lobby that timed out, or arrived
after the games were full. They have no condition, so the counts are for the
dataset as a whole.

```{r dropouts}
if (!has_rows(dropouts)) {
  cat(sprintf(
    "NOTE: data/%s/dropouts.csv is absent or empty (a dataset combined before September 2026 has no dropout records); quiz failures cannot be counted.\n",
    DATASET
  ))
} else {
  cat(sprintf("%d player record(s) never played a real game:\n", nrow(dropouts)))
  print(knitr::kable(dropout_summary(dropouts), digits = 2))
  cat(sprintf("\nFailed the comprehension quiz (three attempts): %d\n", quiz_failures(dropouts)))
}
```

# Reshuffle health

In Phase 2 of the mixed conditions the server reassigns players to groups on
every trial so that each speaker has exactly one listener from their original
group. With nine players that always holds; after a removal the reshuffle
maximizes the number of trios that still meet the rule and the remaining players
form a pair, and a block's designated speaker may have to be replaced.
`reshuffle_health()` (analysis/R/reshuffle.R) reports what each game actually
produced: the Phase 2 roster at the first and last round, group-trials by size,
the share of trios that met the rule (1 whenever nine players remained), and the
number of speaker reassignments across both phases. Games whose roster shrank
are the ones the roster-size check in `02_primary_analysis.qmd` sets aside.

```{r reshuffle-health}
health <- reshuffle_health(trials, games, players)
print(knitr::kable(
  health |>
    mutate(gameId = substr(gameId, 1, 10)) |>
    select(
      condition, gameId, activeGroups, phase2_players_start, phase2_players_end,
      trios, trios_one_in_group, prop_trios_one_in_group, pairs, speaker_reassignments
    ),
  digits = 2
))
```

# Statistics for the manuscript

```{r write-stats}
macros <- c(macros, list(
  overviewNGames = sum(!is.na(games$condition)),
  overviewNPlayers = nrow(pa),
  overviewNTrials = fmt_count(nrow(trials)),
  overviewNMessages = fmt_count(nrow(messages)),
  overviewNSpeakerUtts = fmt_count(nrow(speaker_utts)),
  overviewNRemoved = sum(pa$removed),
  overviewNRemovedNoFault = sum(pa$no_fault),
  overviewNGroupsLost = sum(ga$lost),
  overviewNGamesComplete = sum(status$complete_nine),
  overviewNGamesRanToCompletion = sum(status$ran_to_completion),
  overviewNDropouts = nrow(dropouts),
  overviewNQuizFailed = quiz_failures(dropouts),
  overviewNNoOpportunity = no_opportunity
))
write_stats_tex(macros, stats_file("overview"), source = "analysis/00_data_overview.qmd")
```
