---
title: "00: Data overview"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(jsonlite)
source(here::here("analysis", "config.R"))
```

# Load data

```{r load-data}
list2env(load_tables(), envir = globalenv())
```

# Game-level summary

```{r game-summary}
games |>
  group_by(condition) |>
  summarise(
    n_games = n(),
    mean_players = mean(numPlayers, na.rm = TRUE),
    mean_p1_blocks = mean(phase1Blocks, na.rm = TRUE),
    mean_p2_blocks = mean(phase2Blocks, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Games per condition")
```

# Player-level summary

```{r player-summary}
players |>
  group_by(gameId) |>
  summarise(
    n_players = n(),
    n_active = sum(isActive == TRUE | isActive == "true", na.rm = TRUE),
    n_inactive = sum(isActive == FALSE | isActive == "false", na.rm = TRUE),
    mean_score = mean(as.numeric(score), na.rm = TRUE),
    mean_bonus = mean(as.numeric(bonus), na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Players per game")
```

```{r player-demographics}
# Summarize exit survey demographics if available
demo_cols <- names(players)[str_detect(names(players), "^exitSurvey_")]
if (length(demo_cols) > 0) {
  cat("Exit survey fields:", paste(demo_cols, collapse = ", "), "\n")

  if ("exitSurvey_gender" %in% names(players)) {
    players |>
      count(exitSurvey_gender) |>
      knitr::kable(caption = "Gender distribution")
  }

  if ("exitSurvey_education" %in% names(players)) {
    players |>
      count(exitSurvey_education) |>
      knitr::kable(caption = "Education distribution")
  }
}
```

# Trial-level summary

```{r trial-summary}
cat("Total trial rows:", nrow(trials), "\n")
cat("Unique games:", n_distinct(trials$gameId), "\n")
cat("Unique players:", n_distinct(trials$playerId), "\n")

# Referential accuracy over eligible listener response opportunities: an
# active listener on a played trial whose speaker sent a message. Timeouts
# and late arrivals count as unsuccessful; trials the listener could not have
# answered leave the denominator instead of counting against them.
ref_opportunities <- bind_rows(
  listener_trials(trials, games, phase = 1),
  listener_trials(trials, games, phase = 2)
)

response_opportunity_summary(ref_opportunities, by = "phaseNum") |>
  knitr::kable(
    digits = 3,
    caption = "Referential accuracy by phase, over eligible response opportunities"
  )
```

The table separates nonresponse from incorrect answers: `on_time_rate` is the
proportion of opportunities answered by the server's scoring cutoff, and
`accuracy_on_time` is accuracy among those answers. `accuracy` is the primary
outcome and counts an unanswered opportunity as unsuccessful.

```{r ref-accuracy-condition}
response_opportunity_summary(ref_opportunities) |>
  knitr::kable(
    digits = 3,
    caption = "Referential accuracy by condition and phase"
  )

no_opportunity <- trials |>
  filter(role == "listener", !responseOpportunity) |>
  nrow()
cat(no_opportunity, "listener trials had no speaker message and are excluded",
    "from the accuracy denominators.\n")
```

# Message summary

```{r message-summary}
cat("Total messages:", nrow(messages), "\n")
cat("Speaker messages:", sum(messages$senderRole == "speaker", na.rm = TRUE), "\n")
cat("Listener messages:", sum(messages$senderRole == "listener", na.rm = TRUE), "\n")

messages |>
  mutate(wordCount = str_count(text, "\\S+")) |>
  group_by(senderRole) |>
  summarise(
    n = n(),
    mean_words = mean(wordCount, na.rm = TRUE),
    median_words = median(wordCount, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Message statistics by role")
```

# Speaker utterance summary

```{r utterance-summary}
cat("Total speaker utterances:", nrow(speaker_utts), "\n")

speaker_utts |>
  group_by(phase, phaseNum) |>
  summarise(
    n = n(),
    mean_length = mean(uttLength, na.rm = TRUE),
    sd_length = sd(uttLength, na.rm = TRUE),
    .groups = "drop"
  ) |>
  knitr::kable(caption = "Speaker utterance length by phase")
```

# Social guesses summary

```{r social-guess-summary}
if (nrow(social_guesses) > 0) {
  # Same denominator as referential accuracy: social_guesses.csv is an
  # opportunity frame, so a listener who never answered is a row with an
  # empty guess and counts as unsuccessful rather than disappearing.
  social_opportunities <- social_guess_trials(social_guesses, games)
  cat("Social-identification opportunities:", nrow(social_opportunities), "\n")
  cat("Never answered:", sum(social_opportunities$socialTimeout), "\n")

  response_opportunity_summary(social_opportunities) |>
    knitr::kable(
      digits = 3,
      caption = "Social-identification accuracy by condition, over eligible response opportunities"
    )
} else {
  cat("No social guess data (expected for refer_separated and refer_mixed conditions)\n")
}
```

# Exclusion criteria checks

```{r exclusion-checks}
# Groups where <2/3 participants achieved <2/3 accuracy in last 3 blocks of
# Phase 1. This mirrors the live screen in experiment/server/src/accuracy.js,
# which is unchanged: it counts only trials where the speaker sent a message
# and treats a missing selection as incorrect, which is what the response
# opportunity rule below reproduces. (`p1_listener` rather than
# `listener_trials`, which is the function that builds it.)
p1_listener <- listener_trials(trials, games, phase = 1)

# Get phase1Blocks per game
game_p1 <- games |> select(gameId, phase1Blocks)

listener_accuracy <- p1_listener |>
  left_join(game_p1, by = "gameId") |>
  filter(blockNum >= phase1Blocks - 3) |>
  group_by(gameId, playerId, originalGroup) |>
  summarise(
    accuracy = mean(correct, na.rm = TRUE),
    n_trials = n(),
    .groups = "drop"
  )

# Flag players below 2/3 accuracy
listener_accuracy <- listener_accuracy |>
  mutate(meets_threshold = accuracy >= 2/3)

# Check groups
group_accuracy <- listener_accuracy |>
  group_by(gameId, originalGroup) |>
  summarise(
    n_players = n(),
    n_meeting_threshold = sum(meets_threshold),
    prop_meeting = n_meeting_threshold / n_players,
    group_meets = prop_meeting >= 2/3,
    .groups = "drop"
  )

cat("\nGroup accuracy check (last 3 blocks Phase 1):\n")
group_accuracy |>
  knitr::kable()

excluded_groups <- group_accuracy |> filter(!group_meets)
if (nrow(excluded_groups) > 0) {
  cat("\nGroups flagged for exclusion:\n")
  print(excluded_groups)
} else {
  cat("\nNo groups flagged for exclusion.\n")
}
```

# Attrition

The preregistration commits to reporting attrition rates by condition and
testing whether dropout is differential across conditions (paper, "Dropout
handling"). A player counts as removed when the server took them out of the
game (idle for three rounds, low Phase 1 accuracy, a disbanded group, or too few
groups left). A game ran to completion when at least two groups reached the
end; a complete nine-player game kept every player. The Phase 2 analyses are
rerun on complete games in `02_primary_analysis.qmd`. The last column counts
players flagged for a Phase 1 description-length increase (the AI-use trigger).

```{r attrition}
pa <- player_attrition(players, games)
status <- game_status(players, games)

cat("Players by condition:\n")
print(knitr::kable(attrition_by_condition(pa), digits = 2))

cat("\nGames by condition:\n")
print(knitr::kable(games_by_condition(status), digits = 2))

if (any(pa$removed)) {
  cat("\nRemoved players:\n")
  print(knitr::kable(pa |> filter(removed) |> select(condition, gameId, originalGroup, reason, idleRounds) |>
                       mutate(gameId = substr(gameId, 1, 10)) |> arrange(condition)))
}

tests <- differential_dropout_tests(pa, status)
if (length(tests) == 0) {
  cat("\nNo dropout to test: no player was removed, or every player was.\n")
} else {
  if (!is.null(tests$players)) {
    cat("\nPlayers removed by condition (chi-square; players are clustered in games, so treat as descriptive):\n")
    print(tests$players)
  }
  if (!is.null(tests$games)) {
    cat("\nComplete nine-player games by condition (Fisher's exact test; the game is the unit):\n")
    print(tests$games)
  }
}
```

# Reshuffle health

In Phase 2 of the mixed conditions the server reassigns players to groups on
every trial so that each speaker has exactly one listener from their original
group. With nine players that always holds; after a removal the reshuffle
maximizes the number of trios that still meet the rule and the remaining players
form a pair, and a block's designated speaker may have to be replaced.
`reshuffle_health()` (analysis/R/reshuffle.R) reports what each game actually
produced: the Phase 2 roster at the first and last round, group-trials by size,
the share of trios that met the rule (1 whenever nine players remained), and the
number of speaker reassignments across both phases. Games whose roster shrank
are the ones the roster-size check in `02_primary_analysis.qmd` sets aside.

```{r reshuffle-health}
health <- reshuffle_health(trials, games, players)
print(knitr::kable(
  health |>
    mutate(gameId = substr(gameId, 1, 10)) |>
    select(
      condition, gameId, activeGroups, phase2_players_start, phase2_players_end,
      trios, trios_one_in_group, prop_trios_one_in_group, pairs, speaker_reassignments
    ),
  digits = 2
))
```

