---
title: "01: Outcome-Neutral Criteria"
format:
  html:
    toc: true
    code-fold: true
---

These models must pass before proceeding to full data collection.
All models use Phase 1 data only, pooled across conditions.

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(tidyboot)
library(here)

data_dir <- here("analysis", "processed")

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, "speaker_utterances.csv"), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)

has_adjacent <- file.exists(file.path(data_dir, "adjacent_similarities.csv"))
has_pairwise <- file.exists(file.path(data_dir, "pairwise_similarities.csv"))

adjacent_sim <- if (has_adjacent) read_csv(file.path(data_dir, "adjacent_similarities.csv"), show_col_types = FALSE) else tibble()
pairwise_sim <- if (has_pairwise) read_csv(file.path(data_dir, "pairwise_similarities.csv"), show_col_types = FALSE) else tibble()
```

```{r prep-data}
# Phase 1 speaker utterances only
p1_utts <- speaker_utts %>%
  filter(phaseNum == 1) %>%
  mutate(
    repNum_c = repNum - mean(repNum, na.rm = TRUE),
    log_uttLength = log(uttLength + 1),
    participant = playerId,
    group = originalGroup,
    tangram = target
  )

# Phase 1 listener trials: compute per-trial accuracy
p1_listener <- trials %>%
  filter(phaseNum == 1, role == "listener") %>%
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    group = originalGroup,
    tangram = target
  )

# Aggregate to per-round accuracy (proportion of listeners correct)
# Use blockNum + 1 as repNum (blocks are 0-indexed within phase)
p1_ref_accuracy <- p1_listener %>%
  group_by(gameId, roundId, blockNum, tangram, group) %>%
  summarise(
    ref_accuracy = mean(correct, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    repNum = blockNum + 1,
    repNum_c = repNum - mean(repNum, na.rm = TRUE)
  )

# Phase 1 adjacent similarities (phaseNum is in the CSV)
if (has_adjacent) {
  p1_adjacent <- adjacent_sim %>%
    filter(phaseNum == 1) %>%
    mutate(
      repNum_c = repNum - mean(repNum, na.rm = TRUE),
      participant = playerId,
      group = originalGroup,
      tangram = target
    )
}
```

# 1. Description Length Decreases Over Blocks

**Criterion:** Significant negative main effect of `repNum`.

```{r model-utt-length}
m_length <- lmer(
  log_uttLength ~ repNum_c + (repNum_c | participant) + (repNum_c | group) + (repNum_c | tangram),
  data = p1_utts,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

summary(m_length)
```

```{r plot-utt-length, fig.width=7, fig.height=4}
# Plot 1: Description length over blocks
length_by_block <- p1_utts %>%
  group_by(repNum) %>%
  tidyboot_mean(uttLength, nboot = 1000)

ggplot(length_by_block, aes(x = repNum, y = empirical_stat)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, fill = "steelblue") +
  geom_line(color = "steelblue", linewidth = 1) +
  geom_point(color = "steelblue", size = 2) +
  labs(
    x = "Repetition (block within Phase 1)",
    y = "Mean word count",
    title = "Description length decreases over repetitions"
  ) +
  theme_minimal()
```

# 2. Listener Accuracy Increases Over Blocks

**Criterion:** Significant positive main effect of `repNum`.

```{r model-accuracy}
m_accuracy <- lmer(
  ref_accuracy ~ repNum_c + (repNum_c | group) + (repNum_c | tangram),
  data = p1_ref_accuracy,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

summary(m_accuracy)
```

```{r plot-accuracy, fig.width=7, fig.height=4}
# Plot 2: Listener accuracy over blocks
acc_by_block <- p1_ref_accuracy %>%
  filter(!is.na(ref_accuracy), !is.na(repNum)) %>%
  group_by(repNum) %>%
  tidyboot_mean(ref_accuracy, nboot = 1000)

ggplot(acc_by_block, aes(x = repNum, y = empirical_stat)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, fill = "forestgreen") +
  geom_line(color = "forestgreen", linewidth = 1) +
  geom_point(color = "forestgreen", size = 2) +
  geom_hline(yintercept = 1/6, linetype = "dashed", color = "gray50") +
  annotate("text", x = max(acc_by_block$repNum), y = 1/6 + 0.02,
           label = "Chance (1/6)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Repetition (block within Phase 1)",
    y = "Proportion correct",
    title = "Listener accuracy increases over repetitions"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  theme_minimal()
```

# 3. Conventions Are Stable (Increasing Adjacent Similarity)

**Criterion:** Significant positive main effect of `repNum`.

```{r model-adjacent-sim}
if (has_adjacent) {
  p1_adjacent_valid <- p1_adjacent %>% filter(!is.na(simAdjacent))

  m_adjacent <- lmer(
    simAdjacent ~ repNum_c + (repNum_c | participant) + (repNum_c | group) + (repNum_c | tangram),
    data = p1_adjacent_valid,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
  )

  summary(m_adjacent)
} else {
  cat("Skipping: adjacent_similarities.csv not found. Run compute_embeddings.py first.\n")
}
```

```{r plot-adjacent-sim, fig.width=7, fig.height=4}
if (has_adjacent && exists("p1_adjacent_valid")) {
  # Plot 3: Adjacent similarity over blocks
  sim_by_block <- p1_adjacent_valid %>%
    group_by(repNum) %>%
    tidyboot_mean(simAdjacent, nboot = 1000)

  ggplot(sim_by_block, aes(x = repNum, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2, fill = "darkorange") +
    geom_line(color = "darkorange", linewidth = 1) +
    geom_point(color = "darkorange", size = 2) +
    labs(
      x = "Repetition (block within Phase 1)",
      y = "Cosine similarity (adjacent descriptions)",
      title = "Convention stability increases over repetitions"
    ) +
    theme_minimal()
}
```

# 4. Group-Specificity Exceeds Chance

**Criterion:** ≥80% of games show significant group-specificity (p < .05 via permutation test).

```{r group-specificity-functions}
fit_group_specificity <- function(pairwise_df) {
  # Fit: similarity ~ same_group + (1 | target) + (1 | participant_pair)
  # for each game separately
  game_ids <- unique(pairwise_df$gameId)

  results <- map_dfr(game_ids, function(gid) {
    game_data <- pairwise_df %>% filter(gameId == gid)

    if (nrow(game_data) < 5) return(NULL)
    if (n_distinct(game_data$sameGroup) < 2) return(NULL)

    tryCatch({
      model <- lmer(
        similarity ~ sameGroup + (1 | target) + (1 | participantPair),
        data = game_data,
        control = lmerControl(optimizer = "bobyqa")
      )
      coefs <- coef(summary(model))
      tibble(
        gameId = gid,
        coefficient = coefs["sameGroup", "Estimate"],
        std_error = coefs["sameGroup", "Std. Error"],
        t_value = coefs["sameGroup", "t value"]
      )
    }, error = function(e) NULL)
  })

  results
}

permutation_test <- function(pairwise_df, n_perm = 1000) {
  # For each game, permute group labels and recompute group-specificity
  game_ids <- unique(pairwise_df$gameId)

  results <- map_dfr(game_ids, function(gid) {
    game_data <- pairwise_df %>% filter(gameId == gid)

    if (nrow(game_data) < 5) return(NULL)
    if (n_distinct(game_data$sameGroup) < 2) return(NULL)

    # Observed coefficient
    obs_model <- tryCatch(
      lmer(similarity ~ sameGroup + (1 | target) + (1 | participantPair),
           data = game_data, control = lmerControl(optimizer = "bobyqa")),
      error = function(e) NULL
    )
    if (is.null(obs_model)) return(NULL)
    obs_coef <- coef(summary(obs_model))["sameGroup", "Estimate"]

    # Get unique speakers and their groups
    speakers <- game_data %>%
      select(speaker1, group1) %>%
      rename(speaker = speaker1, group = group1) %>%
      bind_rows(
        game_data %>% select(speaker2, group2) %>% rename(speaker = speaker2, group = group2)
      ) %>%
      distinct()

    # Permutation distribution
    perm_coefs <- map_dbl(seq_len(n_perm), function(i) {
      # Permute group assignments
      perm_groups <- speakers %>%
        mutate(perm_group = sample(group))

      # Remap same_group using permuted assignments
      perm_data <- game_data %>%
        left_join(perm_groups %>% select(speaker, perm_group),
                  by = c("speaker1" = "speaker")) %>%
        rename(perm_group1 = perm_group) %>%
        left_join(perm_groups %>% select(speaker, perm_group),
                  by = c("speaker2" = "speaker")) %>%
        rename(perm_group2 = perm_group) %>%
        mutate(sameGroup = as.numeric(perm_group1 == perm_group2))

      tryCatch({
        perm_model <- lmer(
          similarity ~ sameGroup + (1 | target) + (1 | participantPair),
          data = perm_data,
          control = lmerControl(optimizer = "bobyqa")
        )
        coef(summary(perm_model))["sameGroup", "Estimate"]
      }, error = function(e) NA_real_)
    })

    p_value <- mean(perm_coefs >= obs_coef, na.rm = TRUE)

    tibble(
      gameId = gid,
      obs_coefficient = obs_coef,
      perm_mean = mean(perm_coefs, na.rm = TRUE),
      perm_sd = sd(perm_coefs, na.rm = TRUE),
      p_value = p_value,
      perm_coefs = list(perm_coefs)
    )
  })

  results
}
```

```{r compute-group-specificity}
if (!has_pairwise) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_embeddings.py first.\n")
} else {
# Use Phase 1 final pairwise similarities
p1_pairwise <- pairwise_sim %>% filter(window == "phase1_final")

if (nrow(p1_pairwise) > 0) {
  gs_results <- fit_group_specificity(p1_pairwise)
  cat("Group-specificity coefficients (Phase 1 final):\n")
  print(gs_results)

  perm_results <- permutation_test(p1_pairwise, n_perm = 1000)
  cat("\nPermutation test results:\n")
  perm_results %>%
    select(gameId, obs_coefficient, perm_mean, p_value) %>%
    knitr::kable(digits = 4)

  prop_significant <- mean(perm_results$p_value < 0.05)
  cat(sprintf("\nProportion of games with significant group-specificity: %.1f%%\n",
              prop_significant * 100))
  cat(sprintf("Criterion (≥80%%): %s\n",
              ifelse(prop_significant >= 0.8, "PASS", "FAIL")))
} else {
  cat("No Phase 1 pairwise similarity data available.\n")
}
} # end has_pairwise
```

```{r plot-permutation, fig.width=10, fig.height=6}
# Plot 4: Permutation test histograms
if (exists("perm_results") && nrow(perm_results) > 0) {
  perm_plot_data <- perm_results %>%
    unnest(perm_coefs) %>%
    rename(perm_coef = perm_coefs)

  ggplot(perm_plot_data, aes(x = perm_coef)) +
    geom_histogram(bins = 40, fill = "gray70", color = "white") +
    geom_vline(aes(xintercept = obs_coefficient), color = "red", linewidth = 1) +
    facet_wrap(~gameId, scales = "free_y") +
    labs(
      x = "Group-specificity coefficient (permuted)",
      y = "Count",
      title = "Permutation test: observed vs. permuted group-specificity"
    ) +
    theme_minimal()
}
```

```{r plot-group-specificity-summary, fig.width=6, fig.height=4}
# Plot 5: Group-specificity summary dot plot
if (exists("gs_results") && nrow(gs_results) > 0) {
  # Join condition info
  gs_with_condition <- gs_results %>%
    left_join(games %>% select(gameId, condition), by = "gameId")

  ggplot(gs_with_condition, aes(x = condition, y = coefficient)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun.data = mean_cl_boot, color = "red", size = 0.8) +
    labs(
      x = "Condition",
      y = "Group-specificity coefficient",
      title = "Group-specificity by condition (Phase 1)"
    ) +
    theme_minimal()
}
```

# Summary of Outcome-Neutral Criteria

```{r summary-table}
criteria <- tibble(
  Criterion = c(
    "1. Description length decreases",
    "2. Listener accuracy increases",
    "3. Convention stability increases",
    "4. Group-specificity exceeds chance (≥80%)"
  ),
  Model = c(
    "log(uttLength) ~ repNum + REs",
    "refAccuracy ~ repNum + REs",
    "simAdjacent ~ repNum + REs",
    "Permutation test"
  ),
  `Expected Direction` = c("Negative", "Positive", "Positive", "p < .05")
)

knitr::kable(criteria, caption = "Outcome-neutral criteria checklist")
```
