---
title: "01: Convention-formation checks"
format:
  html:
    toc: true
    code-fold: true
---

These checks inform whether H1/H2 can be interpreted as convergence or
preservation of established group-specific conventions. Trend models pool
Phase 1 data from `refer_separated`, `refer_mixed`, and `social_mixed`, whose
Phase 2 manipulations have not yet been disclosed. `social_first` is excluded
from those models because its manipulation precedes Phase 1, but remains in
the condition-specific descriptive plots and game-level summaries.

No trend, permutation p-value, or proportion of significant games determines
whether primary analyses are run or reported. If initial differentiation is
not convincingly established, H1/H2 contrasts still estimate Phase 2 condition
differences, but do not establish erosion or preservation of pre-existing
conventions. Non-significance in that setting does not establish resistance
to mixing or failure of social incentives to preserve conventions. The checks
do not gate H3/H4, trigger recruitment, or change primary-analysis eligibility.
The separate eligibility and robustness-analysis rules remain unchanged.

Gray trajectories below show individual game means; bold trajectories show
the arithmetic mean across available games in each condition and block.
These descriptive plots have no inferential pass/fail threshold. Historical
four-condition pilot models are preserved separately in `SI_pilot.qmd`.

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_adjacent <- has_rows(adjacent_sim)
has_pairwise <- has_rows(pairwise_sim)

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/checks.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

```{r prep-data}
# Keep all conditions in descriptive reporting and only the common Phase 1
# procedure in trend models. Block is recentered after the restriction.
check_data <- convention_check_data(speaker_utts, trials, adjacent_sim, games)
p1_utts <- check_data$shared_phase1$utterances
p1_ref_accuracy <- check_data$shared_phase1$accuracy       # round-level, for plots
p1_listener <- check_data$shared_phase1$listener_trials    # listener-trial 0/1, for the model
p1_adjacent <- check_data$shared_phase1$adjacent
p1_utts_all <- check_data$all_conditions$utterances
p1_ref_accuracy_all <- check_data$all_conditions$accuracy
p1_adjacent_all <- check_data$all_conditions$adjacent

cat("Pooled trend-model conditions:", paste(H12_CONDITIONS, collapse = ", "), "\n")
```

# 1. Description length decreases over blocks

**Expected pattern:** A negative block effect in the shared Phase 1 procedure.
Report the estimate and uncertainty; significance is not required to proceed.

```{r model-utt-length}
# Maximal structure from the preregistration; fit_progressively() simplifies
# it only if it fails to converge and prints the steps it took. The game
# intercept is retained throughout (only slopes and correlations can go).
m_length <- fit_progressively(
  uttLength ~ blockNum_c + (1 | gameId) + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
  data = p1_utts
)

summary(m_length)
report_effect_sizes(m_length)
```

```{r plot-utt-length, fig.width=11, fig.height=7}
length_by_game <- p1_utts_all |>
  filter(is.finite(uttLength)) |>
  group_by(gameId, condition, blockNum) |>
  summarise(mean_length = mean(uttLength, na.rm = TRUE), .groups = "drop")

length_by_block <- length_by_game |>
  group_by(condition, blockNum) |>
  summarise(mean_length = mean(mean_length), .groups = "drop")

ggplot(length_by_block, aes(x = blockNum + 1, y = mean_length)) +
  geom_line(data = length_by_game,
            aes(group = gameId),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_line(color = "black", linewidth = 1) +
  geom_point(color = "black", size = 2) +
  facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
  labs(
    x = "Block",
    y = "Mean word count",
    title = "Phase 1 description length by condition",
    subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
  ) +
  scale_x_continuous(breaks = 1:6)
```

# 2. Listener accuracy increases over blocks

**Expected pattern:** A positive block effect in the shared Phase 1 procedure.
High and stable accuracy is also informative; an increase is not mandatory.

Referential accuracy is a binary listener-trial outcome under the
response-opportunity rule (`code_accuracy()`: a correct answer received by the
scoring cutoff is 1; an incorrect answer, timeout, or late arrival is 0;
trials without an opportunity leave the denominator), so the trend is fit as
a logistic mixed model on listener trials with the preregistered
random-effects structure plus the game intercept:
`correct ~ blockNum_c + (blockNum_c | group) + (blockNum_c | tangram) + (1 | gameId)`.
The effect size is the odds ratio per block.

```{r model-accuracy}
m_accuracy <- fit_progressively(
  correct ~ blockNum_c + (blockNum_c | group) + (blockNum_c | tangram) + (1 | gameId),
  data = p1_listener,
  family = binomial
)

summary(m_accuracy)
report_odds_ratios(m_accuracy)
cat(sprintf(
  "\nListener trials in the model: %d (%d games, %d groups)\n",
  nrow(p1_listener |> filter(!is.na(correct))),
  n_distinct(p1_listener$gameId), n_distinct(p1_listener$group)
))
```

```{r plot-accuracy, fig.width=11, fig.height=7}
acc_by_game <- p1_ref_accuracy_all |>
  filter(is.finite(ref_accuracy)) |>
  group_by(gameId, condition, blockNum) |>
  summarise(mean_acc = mean(ref_accuracy, na.rm = TRUE), .groups = "drop")

acc_by_block <- acc_by_game |>
  group_by(condition, blockNum) |>
  summarise(mean_acc = mean(mean_acc), .groups = "drop")

ggplot(acc_by_block, aes(x = blockNum + 1, y = mean_acc)) +
  geom_line(data = acc_by_game,
            aes(group = gameId),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_line(color = "black", linewidth = 1) +
  geom_point(color = "black", size = 2) +
  facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  annotate("text", x = 6, y = 1/16 + 0.02,
           label = "Blind-grid baseline (1/16)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block",
    y = "Proportion correct",
    title = "Phase 1 referential accuracy by condition",
    subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
  ) +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(limits = c(0, 1))
```

# 3. Conventions are stable (increasing adjacent similarity)

**Expected pattern:** A positive block effect in the shared Phase 1 procedure.
Interpret the trajectory alongside its level, not as a necessary significance test.

```{r model-adjacent-sim}
if (has_rows(p1_adjacent)) {
  p1_adjacent_valid <- p1_adjacent |> filter(!is.na(simAdjacent))

  m_adjacent <- fit_progressively(
    simAdjacent ~ blockNum_c + (1 | gameId) + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
    data = p1_adjacent_valid
  )

  summary(m_adjacent)
  report_effect_sizes(m_adjacent)
} else {
  cat("Skipping: no shared-Phase-1 adjacent similarity observations.\n")
}
```

```{r plot-adjacent-sim, fig.width=11, fig.height=7}
if (has_rows(p1_adjacent_all)) {
  sim_by_game <- p1_adjacent_all |>
    filter(is.finite(simAdjacent)) |>
    group_by(gameId, condition, blockNum) |>
    summarise(mean_sim = mean(simAdjacent, na.rm = TRUE), .groups = "drop")

  sim_by_block <- sim_by_game |>
    group_by(condition, blockNum) |>
    summarise(mean_sim = mean(mean_sim), .groups = "drop")

  ggplot(sim_by_block, aes(x = blockNum + 1, y = mean_sim)) +
    geom_line(data = sim_by_game,
              aes(group = gameId),
              color = "gray70", alpha = 0.5, linewidth = 0.4) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(color = "black", size = 2) +
    facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
    labs(
      x = "Block",
      y = "Cosine similarity (adjacent descriptions)",
      title = "Phase 1 adjacent-description similarity by condition",
      subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
    ) +
    scale_x_continuous(breaks = 1:6)
}
```

# 4. Endpoint group-specificity

Report game-level estimates and permutation p-values by condition. The shared
Phase 1 procedure also has a pooled descriptive summary. Neither an average
positive coefficient nor a proportion of significant games establishes that
most games developed substantially differentiated conventions. There is no
80% pass/fail threshold and no replacement binary support decision.

```{r compute-group-specificity, warning=FALSE}

if (!has_pairwise) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else {
# Use Phase 1 final pairwise similarities
p1_pairwise <- pairwise_sim |> filter(window == "phase1_final")

if (nrow(p1_pairwise) > 0) {
  # Uses cached RDS if available; pass force = TRUE to recompute
  gs_cached <- compute_group_specificity(p1_pairwise, derived_dir)
  gs_results <- gs_cached$gs_results
  perm_results <- gs_cached$perm_results

  if (!has_rows(gs_results)) {
    gs_results <- tibble(gameId = character(), coefficient = numeric(), std_error = numeric())
  }
  if (!has_rows(perm_results)) {
    perm_results <- tibble(gameId = character(), p_value = numeric())
  }

  # Retain games without estimable coefficients or permutation results as NA,
  # rather than silently treating them as successes or failures.
  gs_by_game <- games |>
    select(gameId, condition) |>
    left_join(gs_results |> select(gameId, coefficient, std_error), by = "gameId") |>
    left_join(perm_results |> select(gameId, p_value), by = "gameId") |>
    mutate(
      condition = factor(condition, levels = CONDITION_ORDER),
      shared_phase1 = condition %in% H12_CONDITIONS
    )
  cat("Group-specificity estimates and permutation p-values by game:\n")
  print(knitr::kable(gs_by_game, digits = 4))

  gs_by_condition <- gs_by_game |>
    group_by(condition) |>
    summarise(
      games = n(),
      estimable = sum(is.finite(coefficient)),
      mean_specificity = if (any(is.finite(coefficient))) mean(coefficient[is.finite(coefficient)]) else NA_real_,
      median_specificity = if (any(is.finite(coefficient))) median(coefficient[is.finite(coefficient)]) else NA_real_,
      .groups = "drop"
    )
  cat("\nDescriptive group-specificity summaries by condition:\n")
  print(knitr::kable(gs_by_condition, digits = 4))

  gs_shared_summary <- gs_by_game |>
    filter(shared_phase1) |>
    summarise(
      games = n(),
      estimable = sum(is.finite(coefficient)),
      mean_specificity = if (any(is.finite(coefficient))) mean(coefficient[is.finite(coefficient)]) else NA_real_,
      median_specificity = if (any(is.finite(coefficient))) median(coefficient[is.finite(coefficient)]) else NA_real_
    )
  cat("\nShared Phase 1 procedure: descriptive summary across games, not a pass/fail test:\n")
  print(knitr::kable(gs_shared_summary, digits = 4))
} else {
  cat("No Phase 1 pairwise similarity data available.\n")
}
} # end has_pairwise
```

```{r plot-permutation, fig.width=10, fig.height=6}
# Plot 4: Permutation test histograms
if (exists("perm_results") && nrow(perm_results) > 0) {
  perm_plot_data <- perm_results |>
    unnest(perm_coefs) |>
    rename(perm_coef = perm_coefs)

  ggplot(perm_plot_data, aes(x = perm_coef)) +
    geom_histogram(bins = 40, fill = "gray70", color = "white") +
    geom_vline(aes(xintercept = obs_coefficient), color = "red", linewidth = 1) +
    facet_wrap(~gameId, scales = "free_y") +
    labs(
      x = "Group-specificity coefficient (permuted)",
      y = "Count",
      title = "Permutation test: observed vs. permuted group-specificity"
    )
}
```

```{r plot-group-specificity-summary, fig.width=9, fig.height=5}
# Plot 5: Group-specificity summary dot plot
if (exists("gs_by_game") && any(is.finite(gs_by_game$coefficient))) {
  gs_with_condition <- gs_by_game |> filter(is.finite(coefficient))

  ggplot(gs_with_condition, aes(x = condition, y = coefficient)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun = mean, geom = "point", color = "black", size = 4) +
    scale_x_discrete(labels = CONDITION_LABELS) +
    labs(
      x = "Condition",
      y = "Group-specificity coefficient",
      title = "Group-specificity by condition (Phase 1)",
      caption = "Each small point is one game; large points are condition means.\nSocial-first is not part of the pooled starting-state check."
    )
}
```

# Summary of convention-formation checks

```{r summary-table}
checks <- tibble(
  Check = c(
    "1. Description length decreases",
    "2. Listener accuracy increases",
    "3. Convention stability increases",
    "4. Endpoint group-specificity"
  ),
  Model = c(
    "uttLength ~ blockNum + (1 | game) + slopes (lmer)",
    "correct ~ blockNum + (1 | game) + slopes (glmer, listener trials)",
    "simAdjacent ~ blockNum + (1 | game) + slopes (lmer)",
    "Permutation test"
  ),
  `Expected direction` = c("Negative", "Positive", "Positive", "Positive coefficient"),
  `Random effects fitted` = c(
    random_effects_structure(m_length),
    random_effects_structure(m_accuracy),
    if (exists("m_adjacent")) random_effects_structure(m_adjacent) else "not fit",
    "n/a"
  )
)

knitr::kable(checks, caption = "Convention-formation diagnostics, not analysis gates")
```

# Statistics for the manuscript

Every block effect and group-specificity summary reported above is written to
`writing/manuscript/stats/checks.tex` as a `\newcommand` macro, so the
manuscript never carries a hand-copied number. Names are prefixed `checks`.

```{r write-stats}
or_macros <- function(prefix, model, term) {
  if (is.null(model) || !inherits(model, c("glmerMod", "glm"))) return(list())
  tab <- odds_ratio_table(model) |> filter(term == .env$term)
  if (nrow(tab) == 0) return(list())
  setNames(
    list(fmt_stat(tab$odds_ratio), fmt_stat(tab$lower), fmt_stat(tab$upper)),
    paste0(prefix, c("OR", "ORLo", "ORHi"))
  )
}

macros <- c(
  macros,
  list(
    checksNGamesPooled = n_distinct(p1_utts$gameId),
    checksNUtterances = fmt_count(nrow(p1_utts)),
    checksNListenerTrials = fmt_count(sum(!is.na(p1_listener$correct)))
  ),
  coef_macros("checksLength", m_length, "blockNum_c"),
  coef_macros("checksAccuracy", m_accuracy, "blockNum_c"),
  or_macros("checksAccuracy", m_accuracy, "blockNum_c"),
  if (exists("m_adjacent")) coef_macros("checksAdjacent", m_adjacent, "blockNum_c") else list()
)

if (exists("gs_shared_summary") && nrow(gs_shared_summary) == 1) {
  macros <- c(macros, list(
    checksNGamesGS = gs_shared_summary$estimable,
    checksMeanGS = fmt_stat(gs_shared_summary$mean_specificity),
    checksMedianGS = fmt_stat(gs_shared_summary$median_specificity)
  ))
}
if (exists("gs_by_game") && "p_value" %in% names(gs_by_game)) {
  macros <- c(macros, list(
    checksNGamesSigGS = sum(gs_by_game$p_value < 0.05, na.rm = TRUE)
  ))
}

write_stats_tex(macros, stats_file("checks"), source = "analysis/01_outcome_neutral.qmd")
```
