---
title: "01: Convention-formation checks"
format:
  html:
    toc: true
    code-fold: true
---

These checks inform whether H1/H2 can be interpreted as convergence or
preservation of established group-specific conventions. Trend models pool
Phase 1 data from `refer_separated`, `refer_mixed`, and `social_mixed`, whose
Phase 2 manipulations have not yet been disclosed. `social_first` is excluded
from those models because its manipulation precedes Phase 1, but remains in
the condition-specific descriptive plots and game-level summaries.

No trend, permutation p-value, or proportion of significant games determines
whether primary analyses are run or reported. If initial differentiation is
not convincingly established, H1/H2 contrasts still estimate Phase 2 condition
differences, but do not establish erosion or preservation of pre-existing
conventions. Non-significance in that setting does not establish resistance
to mixing or failure of social incentives to preserve conventions. The checks
do not gate H3/H4, trigger recruitment, or change primary-analysis eligibility.
The separate eligibility and robustness-analysis rules remain unchanged.

Gray trajectories below show individual game means; bold trajectories show
the arithmetic mean across available games in each condition and block.
These descriptive plots have no inferential pass/fail threshold. Historical
four-condition pilot models are preserved separately in `SI_pilot.qmd`.

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_adjacent <- has_rows(adjacent_sim)
has_pairwise <- has_rows(pairwise_sim)

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/checks.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

```{r prep-data}
# Keep all conditions in descriptive reporting and only the common Phase 1
# procedure in trend models. Block is recentered after the restriction.
check_data <- convention_check_data(speaker_utts, trials, adjacent_sim, games)
p1_utts <- check_data$shared_phase1$utterances
p1_ref_accuracy <- check_data$shared_phase1$accuracy       # round-level, for plots
p1_listener <- check_data$shared_phase1$listener_trials    # listener-trial 0/1, for the model
p1_adjacent <- check_data$shared_phase1$adjacent
p1_utts_all <- check_data$all_conditions$utterances
p1_ref_accuracy_all <- check_data$all_conditions$accuracy
p1_adjacent_all <- check_data$all_conditions$adjacent

cat("Pooled trend-model conditions:", paste(H12_CONDITIONS, collapse = ", "), "\n")
```

# 1. Description length decreases over blocks

**Expected pattern:** A negative block effect in the shared Phase 1 procedure.
Report the estimate and uncertainty; significance is not required to proceed.

Word counts in this task are heavy-tailed, so length is log-transformed, as
in Hawkins et al. (2023); the block coefficient is a proportional change per
block. Untransformed means are plotted below.

No game term. In Phase 1 the three groups of a game never interact -- every
trial has `currentGroup == originalGroup` -- so the group is the conversing
unit and the game is only a shared recruitment session. These models test
block, which varies within every unit, so there is no between-game contrast
for a game term to serve as the reference for; simulated on this design it
leaves the block estimate and its standard error unchanged either way. The
one consequence is that any game-level variance is absorbed into the group
term, so the reported group SD is an upper bound. Boyce et al. (2024) do fit
`(block | gameId)` for the same outcome, but in their design the game *is*
the conversing unit, so our analogue of their term is the group. The Phase 2
length model keeps its by-game slope for the mirror-image reason: once mixing
starts the game becomes the interacting population.

```{r model-utt-length}
# Structure from the preregistration; fit_progressively() simplifies it only
# if it does not fit cleanly and prints the steps it took. The game intercept
# is retained throughout (only slopes and correlations can go).
p1_utts <- p1_utts |> mutate(logUttLength = log(uttLength))

m_length <- fit_progressively(
  logUttLength ~ blockNum_c + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram) + (1 | group:tangram),
  data = p1_utts |> filter(is.finite(logUttLength))
)

summary(m_length)
report_effect_sizes(m_length)
cat(sprintf(
  "\nBlock effect on the word scale: x%.3f per block (%.1f%% change).\n",
  exp(fixef(m_length)[["blockNum_c"]]),
  100 * (exp(fixef(m_length)[["blockNum_c"]]) - 1)
))
```

```{r plot-utt-length, fig.width=11, fig.height=7}
length_by_game <- p1_utts_all |>
  filter(is.finite(uttLength)) |>
  group_by(gameId, condition, blockNum) |>
  summarise(mean_length = mean(uttLength, na.rm = TRUE), .groups = "drop")

length_by_block <- length_by_game |>
  group_by(condition, blockNum) |>
  summarise(mean_length = mean(mean_length), .groups = "drop")

ggplot(length_by_block, aes(x = blockNum + 1, y = mean_length)) +
  geom_line(data = length_by_game,
            aes(group = gameId),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_line(color = "black", linewidth = 1) +
  geom_point(color = "black", size = 2) +
  facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
  labs(
    x = "Block",
    y = "Mean word count",
    title = "Phase 1 description length by condition",
    subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
  ) +
  scale_x_continuous(breaks = 1:6)
```

# 2. Listener accuracy increases over blocks

**Expected pattern:** A positive block effect in the shared Phase 1 procedure.
High and stable accuracy is also informative; an increase is not mandatory.

Referential accuracy is a binary listener-trial outcome under the
response-opportunity rule (`code_accuracy()`: a correct answer received by the
scoring cutoff is 1; an incorrect answer, timeout, or late arrival is 0;
trials without an opportunity leave the denominator), so the trend is fit as
a logistic mixed model on listener trials:
`correct ~ blockNum_c + (blockNum_c | group) + (1 | tangram) + (1 | group:tangram)`.
The effect size is the odds ratio per block.

This is Boyce et al.'s (2024) *preregistered* accuracy structure, mapped onto
this design: theirs was
`listener_accurate ~ block*... + (block*...|tangram) + (1|tangram*group) + (block|group)`,
and their code notes they fitted `(1 | gameId)` alone instead, "breaking this
pre-reg in the name of a model that doesn't take forever to fit". Their
`group` is the conversing unit, which here is the three-person group, not the
nine-person game. On the pilot, collapsing to their fitted structure moves the
block effect from OR 2.41 to 1.85, so the difference is not cosmetic.

It carries no participant term, which is also true of both their fitted and
their preregistered model. For accuracy the repeated-measures unit is the
group the listener is conversing in, not the listener: Hawkins et al. (2020,
2023) likewise put it on the pair or network.

`(1 | group:tangram)` is whether this group has a good label for this figure,
which is neither a group effect nor a tangram effect. It is the second-largest
component on the pilot (SD 0.71 on the log-odds scale, against 0.40 for
tangram), and 62 of 72 cells have the full 12 observations.

```{r model-accuracy}
m_accuracy <- fit_progressively(
  correct ~ blockNum_c + (blockNum_c | group) + (1 | tangram) + (1 | group:tangram),
  data = p1_listener,
  family = binomial
)

summary(m_accuracy)
report_odds_ratios(m_accuracy)
cat(sprintf(
  "\nListener trials in the model: %d (%d games, %d groups)\n",
  nrow(p1_listener |> filter(!is.na(correct))),
  n_distinct(p1_listener$gameId), n_distinct(p1_listener$group)
))
```

```{r plot-accuracy, fig.width=11, fig.height=7}
acc_by_game <- p1_ref_accuracy_all |>
  filter(is.finite(ref_accuracy)) |>
  group_by(gameId, condition, blockNum) |>
  summarise(mean_acc = mean(ref_accuracy, na.rm = TRUE), .groups = "drop")

acc_by_block <- acc_by_game |>
  group_by(condition, blockNum) |>
  summarise(mean_acc = mean(mean_acc), .groups = "drop")

ggplot(acc_by_block, aes(x = blockNum + 1, y = mean_acc)) +
  geom_line(data = acc_by_game,
            aes(group = gameId),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_line(color = "black", linewidth = 1) +
  geom_point(color = "black", size = 2) +
  facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  annotate("text", x = 6, y = 1/16 + 0.02,
           label = "Blind-grid baseline (1/16)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block",
    y = "Proportion correct",
    title = "Phase 1 referential accuracy by condition",
    subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
  ) +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(limits = c(0, 1))
```

# 3. Conventions are stable (increasing adjacent similarity)

**Expected pattern:** A positive block effect in the shared Phase 1 procedure.
Interpret the trajectory alongside its level, not as a necessary significance test.

No participant term here, unlike the length model. `simAdjacent` compares a
description with the one that preceded it for the same group and tangram, and
because speaker roles rotate those two come from different people on 97% of
occasions in the pilot -- so the `playerId` on the row is just whoever
happened to speak second. Attributing the similarity to that person is the
same arbitrary choice the pairwise model avoids with its multiple-membership
speaker term. The quantity belongs to the group's trajectory, which
`(blockNum_c | group)` and `(1 | group:tangram)` carry. Hawkins et al. (2020)
do include a speaker effect in their version because theirs compares
successive utterances by the *same* speaker.

```{r model-adjacent-sim}
if (has_rows(p1_adjacent)) {
  p1_adjacent_valid <- p1_adjacent |> filter(!is.na(simAdjacent))

  m_adjacent <- fit_progressively(
    simAdjacent ~ blockNum_c + (blockNum_c | group) + (1 | tangram) + (1 | group:tangram),
    data = p1_adjacent_valid
  )

  summary(m_adjacent)
  report_effect_sizes(m_adjacent)
} else {
  cat("Skipping: no shared-Phase-1 adjacent similarity observations.\n")
}
```

```{r plot-adjacent-sim, fig.width=11, fig.height=7}
if (has_rows(p1_adjacent_all)) {
  sim_by_game <- p1_adjacent_all |>
    filter(is.finite(simAdjacent)) |>
    group_by(gameId, condition, blockNum) |>
    summarise(mean_sim = mean(simAdjacent, na.rm = TRUE), .groups = "drop")

  sim_by_block <- sim_by_game |>
    group_by(condition, blockNum) |>
    summarise(mean_sim = mean(mean_sim), .groups = "drop")

  ggplot(sim_by_block, aes(x = blockNum + 1, y = mean_sim)) +
    geom_line(data = sim_by_game,
              aes(group = gameId),
              color = "gray70", alpha = 0.5, linewidth = 0.4) +
    geom_line(color = "black", linewidth = 1) +
    geom_point(color = "black", size = 2) +
    facet_wrap(~condition, ncol = 2, labeller = as_labeller(CONDITION_LABELS)) +
    labs(
      x = "Block",
      y = "Cosine similarity (adjacent descriptions)",
      title = "Phase 1 adjacent-description similarity by condition",
      subtitle = "Social-first is descriptive here, not part of the shared-Phase-1 trend model."
    ) +
    scale_x_continuous(breaks = 1:6)
}
```

# 4. Endpoint group-specificity

Report game-level estimates and permutation p-values by condition. The shared
Phase 1 procedure also has a pooled descriptive summary. Neither an average
positive coefficient nor a proportion of significant games establishes that
most games developed substantially differentiated conventions. There is no
80% pass/fail threshold and no replacement binary support decision.

```{r compute-group-specificity, warning=FALSE}

if (!has_pairwise) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else {
# Use Phase 1 final pairwise similarities
p1_pairwise <- pairwise_sim |> filter(window == "phase1_final")

if (nrow(p1_pairwise) > 0) {
  # Uses cached RDS if available; pass force = TRUE to recompute
  gs_cached <- compute_group_specificity(p1_pairwise, derived_dir)
  gs_results <- gs_cached$gs_results
  perm_results <- gs_cached$perm_results

  if (!has_rows(gs_results)) {
    gs_results <- tibble(gameId = character(), coefficient = numeric(), std_error = numeric())
  }
  if (!has_rows(perm_results)) {
    perm_results <- tibble(gameId = character(), p_value = numeric())
  }

  # Retain games without estimable coefficients or permutation results as NA,
  # rather than silently treating them as successes or failures.
  gs_by_game <- games |>
    select(gameId, condition) |>
    left_join(gs_results |> select(gameId, coefficient, std_error), by = "gameId") |>
    left_join(perm_results |> select(gameId, p_value), by = "gameId") |>
    mutate(
      condition = factor(condition, levels = CONDITION_ORDER),
      shared_phase1 = condition %in% H12_CONDITIONS
    )
  cat("Group-specificity estimates and permutation p-values by game:\n")
  print(knitr::kable(gs_by_game, digits = 4))

  gs_by_condition <- gs_by_game |>
    group_by(condition) |>
    summarise(
      games = n(),
      estimable = sum(is.finite(coefficient)),
      mean_specificity = if (any(is.finite(coefficient))) mean(coefficient[is.finite(coefficient)]) else NA_real_,
      median_specificity = if (any(is.finite(coefficient))) median(coefficient[is.finite(coefficient)]) else NA_real_,
      .groups = "drop"
    )
  cat("\nDescriptive group-specificity summaries by condition:\n")
  print(knitr::kable(gs_by_condition, digits = 4))

  gs_shared_summary <- gs_by_game |>
    filter(shared_phase1) |>
    summarise(
      games = n(),
      estimable = sum(is.finite(coefficient)),
      mean_specificity = if (any(is.finite(coefficient))) mean(coefficient[is.finite(coefficient)]) else NA_real_,
      median_specificity = if (any(is.finite(coefficient))) median(coefficient[is.finite(coefficient)]) else NA_real_
    )
  cat("\nShared Phase 1 procedure: descriptive summary across games, not a pass/fail test:\n")
  print(knitr::kable(gs_shared_summary, digits = 4))
} else {
  cat("No Phase 1 pairwise similarity data available.\n")
}
} # end has_pairwise
```

```{r plot-permutation, fig.width=10, fig.height=6}
# Plot 4: Permutation test histograms
if (exists("perm_results") && nrow(perm_results) > 0) {
  perm_plot_data <- perm_results |>
    unnest(perm_coefs) |>
    rename(perm_coef = perm_coefs)

  ggplot(perm_plot_data, aes(x = perm_coef)) +
    geom_histogram(bins = 40, fill = "gray70", color = "white") +
    geom_vline(aes(xintercept = obs_coefficient), color = "red", linewidth = 1) +
    facet_wrap(~gameId, scales = "free_y") +
    labs(
      x = "Group-specificity coefficient (permuted)",
      y = "Count",
      title = "Permutation test: observed vs. permuted group-specificity"
    )
}
```

```{r plot-group-specificity-summary, fig.width=9, fig.height=5}
# Plot 5: Group-specificity summary dot plot
if (exists("gs_by_game") && any(is.finite(gs_by_game$coefficient))) {
  gs_with_condition <- gs_by_game |> filter(is.finite(coefficient))

  ggplot(gs_with_condition, aes(x = condition, y = coefficient)) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun = mean, geom = "point", color = "black", size = 4) +
    scale_x_discrete(labels = CONDITION_LABELS) +
    labs(
      x = "Condition",
      y = "Group-specificity coefficient",
      title = "Group-specificity by condition (Phase 1)",
      caption = "Each small point is one game; large points are condition means.\nSocial-first is not part of the pooled starting-state check."
    )
}
```

# Summary of convention-formation checks

```{r summary-table}
checks <- tibble(
  Check = c(
    "1. Description length decreases",
    "2. Listener accuracy increases",
    "3. Convention stability increases",
    "4. Endpoint group-specificity"
  ),
  Model = c(
    "log(uttLength) ~ blockNum + (1 | game) + slopes (lmer)",
    "correct ~ blockNum + (1 | game) + slopes (glmer, listener trials)",
    "simAdjacent ~ blockNum + (1 | game) + slopes (lmer)",
    "Permutation test"
  ),
  `Expected direction` = c("Negative", "Positive", "Positive", "Positive coefficient"),
  `Random effects fitted` = c(
    random_effects_structure(m_length),
    random_effects_structure(m_accuracy),
    if (exists("m_adjacent")) random_effects_structure(m_adjacent) else "not fit",
    "n/a"
  )
)

knitr::kable(checks, caption = "Convention-formation diagnostics, not analysis gates")
```

# Statistics for the manuscript

Every block effect and group-specificity summary reported above is written to
`writing/manuscript/stats/checks.tex` as a `\newcommand` macro, so the
manuscript never carries a hand-copied number. Names are prefixed `checks`.

```{r write-stats}
or_macros <- function(prefix, model, term) {
  if (is.null(model) || !inherits(model, c("glmerMod", "glm"))) return(list())
  tab <- odds_ratio_table(model) |> filter(term == .env$term)
  if (nrow(tab) == 0) return(list())
  setNames(
    list(fmt_stat(tab$odds_ratio), fmt_stat(tab$lower), fmt_stat(tab$upper)),
    paste0(prefix, c("OR", "ORLo", "ORHi"))
  )
}

macros <- c(
  macros,
  list(
    checksNGamesPooled = n_distinct(p1_utts$gameId),
    checksNUtterances = fmt_count(nrow(p1_utts)),
    checksNListenerTrials = fmt_count(sum(!is.na(p1_listener$correct)))
  ),
  coef_macros("checksLength", m_length, "blockNum_c"),
  coef_macros("checksAccuracy", m_accuracy, "blockNum_c"),
  or_macros("checksAccuracy", m_accuracy, "blockNum_c"),
  if (exists("m_adjacent")) coef_macros("checksAdjacent", m_adjacent, "blockNum_c") else list()
)

if (exists("gs_shared_summary") && nrow(gs_shared_summary) == 1) {
  macros <- c(macros, list(
    checksNGamesGS = gs_shared_summary$estimable,
    checksMeanGS = fmt_stat(gs_shared_summary$mean_specificity),
    checksMedianGS = fmt_stat(gs_shared_summary$median_specificity)
  ))
}
if (exists("gs_by_game") && "p_value" %in% names(gs_by_game)) {
  macros <- c(macros, list(
    checksNGamesSigGS = sum(gs_by_game$p_value < 0.05, na.rm = TRUE)
  ))
}

write_stats_tex(macros, stats_file("checks"), source = "analysis/01_outcome_neutral.qmd")
```
