---
title: "02: Primary Analysis — H1 & H2"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyverse)
library(lme4)
library(lmerTest)
library(emmeans)
library(here)

data_dir <- here("analysis", "data")

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
pairwise_sim <- read_csv(file.path(data_dir, "pairwise_similarities.csv"), show_col_types = FALSE)
```

# Compute Group-Specificity per Game per Phase

For each game, fit:
`similarity ~ sameGroup + (1 | target) + (1 | participantPair)`

Extract the `sameGroup` coefficient as the group-specificity score.

```{r fit-group-specificity}
fit_gs_by_game_window <- function(pw_df) {
  game_ids <- unique(pw_df$gameId)

  map_dfr(game_ids, function(gid) {
    game_data <- pw_df %>% filter(gameId == gid)

    if (nrow(game_data) < 5) return(NULL)
    if (n_distinct(game_data$sameGroup) < 2) return(NULL)

    tryCatch({
      model <- lmer(
        similarity ~ sameGroup + (1 | target) + (1 | participantPair),
        data = game_data,
        control = lmerControl(optimizer = "bobyqa")
      )
      coefs <- coef(summary(model))
      tibble(
        gameId = gid,
        coefficient = coefs["sameGroup", "Estimate"],
        std_error = coefs["sameGroup", "Std. Error"],
        t_value = coefs["sameGroup", "t value"]
      )
    }, error = function(e) NULL)
  })
}

# Phase 1 final
gs_p1 <- pairwise_sim %>%
  filter(window == "phase1_final") %>%
  fit_gs_by_game_window() %>%
  rename(gs_phase1 = coefficient, se_phase1 = std_error) %>%
  select(gameId, gs_phase1, se_phase1)

# Phase 2 final
gs_p2 <- pairwise_sim %>%
  filter(window == "phase2_final") %>%
  fit_gs_by_game_window() %>%
  rename(gs_phase2 = coefficient, se_phase2 = std_error) %>%
  select(gameId, gs_phase2, se_phase2)

# Combine
gs_combined <- gs_p1 %>%
  inner_join(gs_p2, by = "gameId") %>%
  left_join(games %>% select(gameId, condition), by = "gameId")

cat("Group-specificity estimates:\n")
gs_combined %>%
  knitr::kable(digits = 4, caption = "Group-specificity by game and phase")
```

# Weighted Least Squares Regression

```
group_specificity_phase2 ~ condition + group_specificity_phase1
```
Weights: `1 / SE_phase2^2` (inverse variance of Phase 2 estimates).

```{r wls-model}
gs_combined <- gs_combined %>%
  mutate(
    weight = 1 / se_phase2^2,
    condition = factor(condition, levels = c("refer_mixed", "refer_separated", "social_mixed"))
  )

n_conditions <- n_distinct(gs_combined$condition)

if (n_conditions >= 2) {
  wls_model <- lm(
    gs_phase2 ~ condition + gs_phase1,
    data = gs_combined,
    weights = weight
  )
  summary(wls_model)
} else {
  cat(sprintf("Only %d condition(s) in data (%s). Need ≥2 for WLS model.\n",
              n_conditions, paste(unique(gs_combined$condition), collapse = ", ")))
  wls_model <- NULL
}
```

# Planned Contrasts

Two-tailed tests at α = 0.05.

- **H1:** `refer_separated > refer_mixed`
- **H2:** `social_mixed > refer_mixed`

```{r planned-contrasts}
if (!is.null(wls_model) && n_conditions >= 3) {
  em <- emmeans(wls_model, "condition")
  contrasts <- contrast(em, list(
    H1_sep_vs_mixed = c(-1, 1, 0),   # refer_separated - refer_mixed
    H2_social_vs_mixed = c(-1, 0, 1)  # social_mixed - refer_mixed
  ))
  summary(contrasts, infer = c(TRUE, TRUE))
} else {
  cat("Skipping contrasts: need all 3 conditions.\n")
  contrasts <- NULL
}
```

```{r contrast-details}
if (!is.null(contrasts)) {
  cat("\nDetailed contrast results:\n")
  contrast_summary <- summary(contrasts, infer = c(TRUE, TRUE))
  for (i in seq_len(nrow(contrast_summary))) {
    row <- contrast_summary[i, ]
    cat(sprintf(
      "\n%s:\n  Estimate = %.4f, 95%% CI [%.4f, %.4f], p = %.4f\n",
      row$contrast, row$estimate, row$lower.CL, row$upper.CL, row$p.value
    ))
  }
}
```

# Plots

```{r plot-gs-paired, fig.width=8, fig.height=5}
# Plot 6: Group-specificity Phase 1 vs Phase 2 by condition
gs_long <- gs_combined %>%
  pivot_longer(
    cols = c(gs_phase1, gs_phase2),
    names_to = "phase",
    values_to = "group_specificity"
  ) %>%
  mutate(phase = ifelse(phase == "gs_phase1", "Phase 1", "Phase 2"))

ggplot(gs_long, aes(x = phase, y = group_specificity, group = gameId)) +
  geom_line(alpha = 0.5, color = "gray50") +
  geom_point(aes(color = condition), size = 3) +
  facet_wrap(~condition) +
  labs(
    x = "Phase",
    y = "Group-specificity coefficient",
    title = "Group-specificity across phases by condition"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r plot-gs-phase2, fig.width=7, fig.height=5}
# Plot 7: Phase 2 group-specificity by condition (primary result)
ggplot(gs_combined, aes(x = condition, y = gs_phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  labs(
    x = "Condition",
    y = "Phase 2 group-specificity",
    title = "Group-specificity in Phase 2 by condition"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r plot-wls-forest, fig.width=7, fig.height=4}
# Plot 8: WLS coefficient forest plot
if (!is.null(wls_model)) {
  wls_coefs <- broom::tidy(wls_model, conf.int = TRUE) %>%
    filter(term != "(Intercept)")

  ggplot(wls_coefs, aes(x = estimate, y = term)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
    geom_point(size = 3) +
    labs(
      x = "Coefficient estimate",
      y = "",
      title = "WLS model coefficients"
    ) +
    theme_minimal()
} else {
  cat("WLS model not fitted (insufficient conditions).\n")
}
```

```{r plot-within-between, fig.width=10, fig.height=5}
# Plot 9: Within- vs between-group similarity by condition
pw_summary <- pairwise_sim %>%
  left_join(games %>% select(gameId, condition), by = "gameId") %>%
  mutate(
    groupType = ifelse(sameGroup == 1, "Within-group", "Between-group"),
    phase = ifelse(window == "phase1_final", "Phase 1 final", "Phase 2 final")
  )

pw_means <- pw_summary %>%
  group_by(condition, phase, groupType) %>%
  summarise(
    mean_sim = mean(similarity, na.rm = TRUE),
    se_sim = sd(similarity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(pw_means, aes(x = condition, y = mean_sim, fill = groupType)) +
  geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_sim - se_sim, ymax = mean_sim + se_sim),
    position = position_dodge(0.8), width = 0.2
  ) +
  facet_wrap(~phase) +
  labs(
    x = "Condition",
    y = "Mean cosine similarity",
    fill = "Group type",
    title = "Within- vs between-group similarity by condition and phase"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
```

# Optional: Bayes Factors (if null results)

Uncomment and run if contrasts are non-significant.

```{r bayes-factors, eval=FALSE}
# library(brms)
# library(bridgesampling)
#
# # Refit with brms using Cauchy(0, sqrt(2)/2) prior on condition effects
# prior <- c(
#   set_prior("cauchy(0, 0.707)", class = "b")
# )
#
# brm_model <- brm(
#   gs_phase2 | weights(weight) ~ condition + gs_phase1,
#   data = gs_combined,
#   prior = prior,
#   iter = 10000,
#   warmup = 2000,
#   chains = 4,
#   save_pars = save_pars(all = TRUE)
# )
#
# # Compute Bayes factors via bridge sampling
# brm_null <- brm(
#   gs_phase2 | weights(weight) ~ gs_phase1,
#   data = gs_combined,
#   iter = 10000,
#   warmup = 2000,
#   chains = 4,
#   save_pars = save_pars(all = TRUE)
# )
#
# bf <- bayes_factor(brm_model, brm_null)
# cat("Bayes factor (full vs null):", bf$bf, "\n")
```
