---
title: "02: Primary analyses -- H1 through H4"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyboot)
library(patchwork)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_pairwise <- has_rows(pairwise_sim)
has_block_pairwise <- has_rows(block_pairwise_sim)
if (!has_pairwise) {
  cat("WARNING: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
  cat("Most analyses in this notebook will be skipped.\n")
}

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/primary.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

# Inference and reporting policy

Each prespecified prediction is evaluated separately using a two-tailed,
unadjusted test at α = 0.05. Support requires both p < .05 and the predicted
effect direction; a significant effect in the opposite direction is reported
as contrary to the prediction. All planned contrasts, including non-significant
results, are reported with estimates, pointwise 95% confidence intervals, and
exact unadjusted p-values. This is a per-test policy, not family-wise error
control across the collection of tests.

H3a, H3b, and each of the three H3c description-property predictions have
separate support decisions. There is no omnibus H3 support decision, and one
significant H3c result does not establish the other description-property
predictions. Secondary and robustness analyses do not substitute for the
prespecified primary tests.

The convention-formation checks in `01_outcome_neutral.qmd` pool the three
conditions with the shared Phase 1 procedure for their trend models. They
inform interpretation of H1/H2 as convergence or preservation of established
group differences, but do not determine whether the contrasts are reported.
If initial differentiation is not convincingly established, we will report
Phase 2 condition differences without claiming erosion or preservation of
pre-existing conventions; non-significant contrasts in that setting do not
establish resistance to mixing or failure of social incentives to preserve
conventions. These checks do not gate H3/H4 or primary-analysis eligibility.

# Compute group-specificity per game per phase

For each game and endpoint window, fit:
`similarity ~ sameGroup + (1 | target) + (1 | speaker) + (1 | speakerPair)`

Extract the `sameGroup` coefficient as the group-specificity score.

The speaker term is a multiple-membership random intercept. Each observation
is a pair of speakers describing the same tangram, so every speaker has one
random effect shared across both pair positions, with a common variance and
equal weights of 1/2. The `speakerPair` intercept is the unordered pair of
speakers: it captures resemblance that belongs to two particular people rather
than to either of them, which is where a shared convention resides, and it is
retained even when its variance is estimated at the boundary. This replaces
the earlier `(1 | speaker1) + (1 | speaker2)`, which split one person's effect
across two variance components and made the estimate depend on an arbitrary
ordering within each pair. Fitting is by REML through lme4's modular interface
(`analysis/R/multimembership.R`); the design-matched simulations that check
convergence, ordering invariance, and interval calibration are in
`analysis/simulate_mm_speaker.R`. The pilot SI keeps the earlier structure,
as the paper records.

Each game's standard error is reported as within-game uncertainty and feeds
the inverse-variance weights of the robustness reruns below; it does not
determine a game's weight in the primary regressions, which give every
eligible game equal weight.

## Endpoint coverage

An endpoint estimate requires analyzable descriptions from at least two
speakers in each of at least two original groups within the window, and both
within-group and between-group pairs. H1/H2 use games with both endpoints;
H3a uses the Phase 1 endpoint only. A missing endpoint is never replaced from
another window. Every game-window is listed with its eligibility and, where
there is no estimate, the reason.

```{r fit-group-specificity}
if (!has_pairwise) {
  cat("Skipping group-specificity analysis: no pairwise similarity data.\n")
} else {
  # game_specificity_table() (analysis/R/group_specificity.R) is the one place
  # the per-game estimates are fit; it returns every game-window with an
  # eligibility flag and a note. gs_wide() gives one row per game with the
  # Phase 1 and Phase 2 estimates (NA where an endpoint is not estimable) and
  # the inverse-variance weights for the robustness reruns.
  gs_table <- game_specificity_table(pairwise_sim, games)
  gs_combined <- gs_wide(gs_table)

  cat("Endpoint coverage by window and condition:\n")
  print(knitr::kable(endpoint_coverage_summary(gs_table)))

  not_estimable <- gs_table |> filter(!estimable)
  if (nrow(not_estimable) > 0) {
    cat("\nGame-windows without an estimate:\n")
    print(knitr::kable(
      not_estimable |>
        mutate(gameId = substr(gameId, 1, 10)) |>
        select(condition, gameId, window, n_pairs, n_speakers, n_groups_two_speakers, both_levels, note)
    ))
  } else {
    cat("\nEvery game-window is eligible and estimable.\n")
  }

  cat(sprintf(
    "\nGames with both endpoints (H1/H2): %d of %d; with the Phase 1 endpoint (H3a): %d.\n",
    sum(gs_combined$both_endpoints), nrow(gs_combined), sum(gs_combined$eligible_phase1)
  ))
  cat("\nGroup-specificity estimates:\n")
  print(knitr::kable(gs_combined |> select(gameId, condition, gs_phase1, se_phase1, gs_phase2, se_phase2),
                     digits = 4, caption = "Group-specificity by game and phase (NA: endpoint not estimable)"))
}
```

# Ordinary least squares regression

```
group_specificity_phase2 ~ condition + group_specificity_phase1
```
Every eligible game receives equal weight (ordinary least squares). A game is
eligible when both endpoint windows are estimable under the coverage rule
above. Inverse-variance weighting (`1 / SE_phase2^2`) is the robustness check
reported right after the planned contrasts, not an alternative primary
support criterion.

Restricted to the three conditions that share the same Phase 1 procedure
(`refer_separated`, `refer_mixed`, `social_mixed`). In `social_first` the
manipulation precedes Phase 1, so its Phase 1 group-specificity is an outcome
(H3a) rather than a pre-manipulation baseline, and pooling it into the common
covariate slope could bias the H1 and H2 contrasts. `social_first` is kept in
the descriptive plots below.

```{r ols-model}
if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity data.\n")
} else {
  # fit_h12() (analysis/R/contrasts.R) restricts to H12_CONDITIONS, drops
  # games without both endpoints, and returns the data, the lm, and the
  # planned contrasts. weights = NULL is ordinary least squares.
  h12 <- fit_h12(gs_combined)
  gs_h12 <- h12$data
  ols_model <- h12$model
  contrasts <- h12$contrasts
  if (is.null(ols_model)) {
    cat(sprintf("Skipping the H1/H2 model: %s.\n", h12$note))
  } else {
    cat(sprintf("Games in the H1/H2 model: %d\n", nrow(gs_h12)))
    print(knitr::kable(gs_h12 |> count(condition, name = "games")))
    print(summary(ols_model))
    report_effect_sizes(ols_model)
  }
}
```

# Planned contrasts

Two-tailed tests at α = 0.05.

- **H1:** `refer_separated > refer_mixed`
- **H2:** `social_mixed > refer_mixed`

```{r planned-contrasts}
if (exists("contrasts") && !is.null(contrasts)) {
  print(summary(contrasts, infer = c(TRUE, TRUE)))
} else {
  contrasts <- NULL
  cat("Skipping contrasts: need all three H1/H2 conditions.\n")
}
```

Game-level Cohen's *d* for each contrast (unadjusted difference in Phase 2
group-specificity between the two conditions, pooled SD; the quantity the
power analysis was based on). Not estimable with one game per condition.

```{r effect-sizes-h1-h2}
if (!is.null(contrasts)) {
  rbind(
    cohens_d_games(gs_h12, "gs_phase2", "condition", "refer_separated", "refer_mixed"),
    cohens_d_games(gs_h12, "gs_phase2", "condition", "social_mixed", "refer_mixed")
  )
}
```

## Robustness: inverse-variance weights

The preregistered robustness check repeats the same contrasts with weighted
least squares, each game weighted by `1 / SE_phase2^2` (the inverse variance
of its Phase 2 group-specificity estimate). It is reported beside the primary
OLS contrasts and does not replace them.

```{r robustness-wls-h1-h2}
if (exists("h12") && !is.null(h12$model)) {
  h12_wls <- fit_h12(gs_combined, weights = "weight")
  print_weighted_check(h12, h12_wls)
} else {
  cat("Primary H1/H2 model not fit; nothing to compare.\n")
}
```

Bayes factors for non-significant contrasts (preregistered: brms refit,
standardized estimates, Cauchy(0, sqrt(2)/2) prior on the effect size,
two-sided; see `bayes_factors.R`). The full model is the primary
three-condition OLS model (`gs_phase2 ~ condition + gs_phase1`, equal
weights) refit in brms; the null for each contrast equates the two conditions
it compares and keeps everything else, so BF10 concerns that planned contrast
only.

```{r bayes-factors-h1-h2}
if (!is.null(contrasts)) {
  p <- as.data.frame(summary(contrasts, infer = c(TRUE, TRUE)))$p.value
  bf_h1 <- maybe_bayes_factor(p[1], "H1 refer_separated vs refer_mixed", function() {
    bf_game_level_contrast(gs_h12, "gs_phase2", levels = H12_CONDITIONS,
                           contrast = c("refer_separated", "refer_mixed"),
                           covariates = "gs_phase1", name = "H1")
  })
  bf_h2 <- maybe_bayes_factor(p[2], "H2 social_mixed vs refer_mixed", function() {
    bf_game_level_contrast(gs_h12, "gs_phase2", levels = H12_CONDITIONS,
                           contrast = c("social_mixed", "refer_mixed"),
                           covariates = "gs_phase1", name = "H2")
  })
}
```

```{r contrast-details}
if (!is.null(contrasts)) {
  cat("\nDetailed contrast results:\n")
  contrast_summary <- summary(contrasts, infer = c(TRUE, TRUE))
  for (i in seq_len(nrow(contrast_summary))) {
    row <- contrast_summary[i, ]
    cat(sprintf(
      "\n%s:\n  Estimate = %.4f, 95%% CI [%.4f, %.4f], p = %.4f\n",
      row$contrast, row$estimate, row$lower.CL, row$upper.CL, row$p.value
    ))
  }
}
```

# Plots

```{r plot-gs-paired, fig.width=8, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 6: Group-specificity Phase 1 vs Phase 2 by condition
gs_long <- gs_combined |>
  pivot_longer(
    cols = c(gs_phase1, gs_phase2),
    names_to = "phase",
    values_to = "group_specificity"
  ) |>
  filter(!is.na(group_specificity)) |>
  mutate(phase = ifelse(phase == "gs_phase1", "Phase 1", "Phase 2"))

ggplot(gs_long, aes(x = phase, y = group_specificity, group = gameId)) +
  geom_line(alpha = 0.5, color = "gray50") +
  geom_point(aes(color = condition), size = 3) +
  facet_wrap(~condition) +
  scale_color_condition() +
  labs(
    x = "Phase",
    y = "Group-specificity coefficient",
    title = "Group-specificity across phases by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-gs-phase2, fig.width=7, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 7: Phase 2 group-specificity by condition (primary result)
ggplot(gs_combined |> filter(!is.na(gs_phase2)), aes(x = condition, y = gs_phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Phase 2 group-specificity",
    title = "Group-specificity in Phase 2 by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-ols-forest, fig.width=7, fig.height=4}
# Plot 8: OLS coefficient forest plot
if (exists("ols_model") && !is.null(ols_model)) {
  ols_coefs <- broom::tidy(ols_model, conf.int = TRUE) |>
    filter(term != "(Intercept)")

  ggplot(ols_coefs, aes(x = estimate, y = term)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
    geom_point(size = 3) +
    labs(
      x = "Coefficient estimate",
      y = "",
      title = "H1/H2 OLS model coefficients"
    )
} else {
  cat("OLS model not fitted (insufficient conditions).\n")
}
```

```{r plot-within-between, fig.width=10, fig.height=5}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping plot: no pairwise similarity data.\n")
} else {
# Plot 9: Within- vs between-group similarity by condition. All three
# windows are shown separately; phase2_early is descriptive and is not part
# of the H1/H2 comparison, which uses the two final windows.
pw_summary <- pairwise_sim |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  label_window() |>
  mutate(groupType = ifelse(sameGroup == 1, "Within-group", "Between-group"))

pw_means <- pw_summary |>
  group_by(condition, window_label, groupType) |>
  summarise(
    mean_sim = mean(similarity, na.rm = TRUE),
    se_sim = sd(similarity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(pw_means, aes(x = condition, y = mean_sim, fill = groupType)) +
  geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_sim - se_sim, ymax = mean_sim + se_sim),
    position = position_dodge(0.8), width = 0.2
  ) +
  facet_wrap(~window_label) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Condition",
    y = "Mean cosine similarity",
    fill = "Group type",
    title = "Within- vs between-group similarity by condition and phase"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```


# H3: Effect of anticipated social goal on Phase 1 conventions

Restricted to `social_mixed` vs `social_first`. Tests whether anticipating
a social-signaling task shapes convention formation *before* Phase 2.
Conclusions are drawn separately for group-specificity, early within-group
alignment, and each description property, rather than for H3 as a whole.

## H3a: Phase 1 group-specificity

Ordinary least squares: `group_specificity_phase1 ~ condition`, restricted to
the two social conditions, with equal weight for each game whose Phase 1
endpoint is estimable (the Phase 2 endpoint is not required). Contrast:
`social_first > social_mixed`. The inverse-variance rerun (weights
`1 / SE_phase1^2`) follows as a robustness check.

```{r h3a-gs}
if (!exists("gs_combined")) {
  cat("Skipping H3a: no group-specificity data.\n")
} else {
  h3a <- fit_h3a(gs_combined)
  gs_social <- h3a$data
  ols_h3a <- h3a$model
  contrast_h3a <- h3a$contrasts

  if (!is.null(ols_h3a) && !is.null(contrast_h3a) && nrow(gs_social) >= 4) {
    cat(sprintf("Games in the H3a model: %d\n", nrow(gs_social)))
    print(knitr::kable(gs_social |> count(condition, name = "games")))
    print(summary(ols_h3a))
    report_effect_sizes(ols_h3a)
    print(summary(contrast_h3a, infer = c(TRUE, TRUE)))

    print(cohens_d_games(gs_social, "gs_phase1", "condition", "social_first", "social_mixed"))

    p_h3a <- as.data.frame(summary(contrast_h3a, infer = c(TRUE, TRUE)))$p.value
    # The primary two-condition OLS model refit in brms, unweighted; the null
    # drops condition
    bf_h3a <- maybe_bayes_factor(p_h3a, "H3a social_first vs social_mixed", function() {
      bf_game_level_contrast(gs_social, "gs_phase1", levels = SOCIAL_CONDITIONS,
                             contrast = c("social_first", "social_mixed"), name = "H3a")
    })

    cat("\nRobustness: inverse-variance weights (1 / SE_phase1^2):\n")
    print_weighted_check(h3a, fit_h3a(gs_combined, weights = "weight_p1"))
  } else {
    contrast_h3a <- NULL
    cat("Insufficient data for H3a (need both social conditions with at least two games each):",
        if (!is.null(h3a$note)) h3a$note else "", "\n")
  }
}
```

```{r plot-h3a, fig.width=6, fig.height=5}
if (exists("gs_social") && nrow(gs_social) > 0) {
  ggplot(gs_social, aes(x = condition, y = gs_phase1, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Phase 1 group-specificity",
      title = "H3a: Phase 1 group-specificity by social condition"
    ) +
    theme(legend.position = "none")
}
```

## H3b: Early within-group alignment

The primary contrast is `social_first > social_mixed` at Phase 1 block 3,
when all three original group members have had one scheduled speaker turn.
Blocks are counted from 1 here, as in the paper; the exported `blockNum` is
0-indexed, so `within_group_block_pairs()` converts it once to a 1-indexed
`block` column and the model uses blocks 2-6 on that scale (0-indexed
blockNum 1-5), with the milestone at block 3 (0-indexed blockNum 2).
Within-group pairwise similarities are averaged within group, tangram, and
block. Available pairs contribute equally; means with no valid pairs are
omitted, without substituting another block for a missing block-3 observation.

The model uses categorical block and its interaction with condition, with
random effects for game, group nested within game, tangram, and repeated
group--tangram observations. The starting formula is `H3B_FORMULA` in
`analysis/R/convergence.R`; random effects follow the shared simplification
procedure. Game-level replication is required for the condition comparison.

H3b is supported by a positive block-3 contrast with a two-tailed p-value below
.05, with its raw similarity difference and 95% confidence interval reported.
This tests greater alignment by a fixed early milestone, not the exact time
a convention is established. Neither an omnibus interaction nor a linear
slope difference is a prerequisite or an alternative support criterion.

```{r h3b-convergence}
if (!has_block_pairwise || nrow(block_pairwise_sim) == 0) {
  cat("Skipping H3b: block_pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else {
  h3b <- fit_h3b(within_group_block_means(
    block_pairwise_sim, games, phase = 1, conditions = SOCIAL_CONDITIONS
  ))
  p1_within_social <- h3b$data
  m_h3b <- h3b$model
  if (!is.null(m_h3b)) {
    print(summary(m_h3b))
    print(h3b$summary)
    cat("Primary block-3 support criterion met:", h3b$supported, "\n")
    # Bayes factors are preregistered for H1, H2, H3a, H4a, and H4b only; H3b
    # reports the frequentist contrast, and a non-significant contrast is not
    # evidence for the null.
  } else {
    cat("H3b inference skipped:", h3b$note, "\n")
  }
}
```

### Secondary description of the linear trend

This model describes differences in the average linear increase in similarity.
It does not determine primary H3b support, including when its interaction is
significant but the block-3 contrast is not.

```{r h3b-secondary-trend}
if (exists("h3b") && !is.null(h3b$model)) {
  # block_c is the 1-indexed block centered at the block-3 milestone
  m_h3b_trend <- fit_progressively(H3B_TREND_FORMULA, data = h3b$data)
  print(summary(m_h3b_trend))
  slopes_h3b <- emtrends(m_h3b_trend, "condition", var = "block_c",
                         lmer.df = "satterthwaite", lmerTest.limit = Inf)
  print(summary(contrast(slopes_h3b, list(
    secondary_slope_socialfirst_minus_socialmixed = c(-1, 1)
  )), infer = c(TRUE, TRUE)))
}
```

```{r plot-h3b, fig.width=8, fig.height=5}
if (exists("p1_within_social") && nrow(p1_within_social) > 0) {
  # Bootstrap game means, not the overlapping pairs or groups within a game.
  # `block` is 1-indexed (paper numbering); the milestone is block 3.
  conv_games <- p1_within_social |>
    mutate(block = as.integer(as.character(block))) |>
    group_by(condition, gameId, group, block) |>
    summarise(similarity = mean(similarity), .groups = "drop") |>
    group_by(condition, gameId, block) |>
    summarise(similarity = mean(similarity), .groups = "drop")
  conv_by_block <- conv_games |>
    group_by(condition, block) |>
    group_modify(function(d, keys) {
      if (nrow(d) < 2L) {
        return(tibble(empirical_stat = mean(d$similarity),
                      ci_lower = NA_real_, ci_upper = NA_real_))
      }
      tidyboot_mean(d, similarity, nboot = 1000) |>
        select(empirical_stat, ci_lower, ci_upper)
    })

  ggplot(conv_by_block, aes(x = block, y = empirical_stat, color = condition)) +
    geom_vline(xintercept = H3B_BLOCK, linetype = "dotted", color = "gray50") +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    scale_x_continuous(breaks = 1:6, limits = c(1, 6)) +
    labs(
      x = "Phase 1 block (1-indexed)",
      y = "Within-group cosine similarity",
      color = "Condition",
      title = "H3b: Within-group alignment by social condition",
      subtitle = "The primary comparison is at block 3 (dotted line).",
      caption = "Intervals resample games; a single pilot game per condition cannot estimate between-game uncertainty."
    )
}
```

## H3c: Description properties

Compare concreteness, lexical uniqueness, and word frequency of final Phase 1
descriptions between `social_mixed` and `social_first`.

These are three separate predictions: lower concreteness, higher lexical
uniqueness, and lower word frequency in `social_first`. Each is supported only
if its own condition contrast has a two-tailed, unadjusted p < .05 and the
predicted direction. All three results are reported; a lower-frequency effect
alone does not support the concreteness or lexical-uniqueness predictions.

Each measure is modeled on each speaker's final Phase 1 description of each
tangram with the preregistered structure
`measure ~ condition + (1 | game) + (1 | group) + (1 | participant) + (condition | tangram)`,
where `group` is the game-unique original group. The game and group
intercepts are retained through any simplification.

All three measures are precomputed in Python (`compute_derived.py`) with
consistent tokenization: each utterance is lowercased, split into alphabetic
sequences, filtered for NLTK English stopwords, and only tokens with
length > 1 are retained as *content words*.

- **Concreteness**: the proportion of content words matching concrete term
  stems (prefix matching) from Boyce et al. 2024. Four categories: geometric
  (e.g., *triangle*, *square*, *diamond*), body parts (e.g., *arm*, *head*,
  *leg*), positional (e.g., *left*, *right*, *top*), and posture (e.g.,
  *kick*, *kneel*, *stand*). Prefix matching means *squar* matches both
  *square* and *squares*.
- **Lexical uniqueness**: the proportion of content words in a description
  that do not appear in any other original group's descriptions of the same
  tangram, pooled across games. The comparison vocabulary is restricted to the
  same phase, so labels that other groups borrow in Phase 2 cannot
  retroactively make a Phase 1 word non-unique.
- **Word frequency**: the mean Zipf frequency of content words, computed via
  the `wordfreq` Python package (which draws on SUBTLEX and other corpora).
  Common words score higher (e.g., "the" ≈ 7.7); rare words score lower.

```{r h3c-description-properties}
if (!has_rows(desc_props) || !has_rows(lex_uniq)) {
  cat("Missing H3c data files. Run compute_description_properties() and compute_lexical_uniqueness() first.\n")
} else {
  # Each speaker's final Phase 1 description of each tangram, social conditions
  p1_final_social <- final_phase1_properties(desc_props, lex_uniq, games, SOCIAL_CONDITIONS)

  if (nrow(p1_final_social) > 0 && n_distinct(p1_final_social$condition) >= 2) {
    # Concreteness
    p1_concrete <- p1_final_social |> filter(!is.na(concreteness))
    if (nrow(p1_concrete) > 0) {
      cat("=== Concreteness ===\n")
      m_concrete <- fit_progressively(
        concreteness ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (condition | target),
        data = p1_concrete
      )
      print(summary(m_concrete))
      report_effect_sizes(m_concrete)
    }

    # Lexical uniqueness
    p1_unique <- p1_final_social |> filter(!is.na(uniqueness))
    if (nrow(p1_unique) > 0) {
      cat("\n=== Lexical Uniqueness ===\n")
      m_uniqueness <- fit_progressively(
        uniqueness ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (condition | target),
        data = p1_unique
      )
      print(summary(m_uniqueness))
      report_effect_sizes(m_uniqueness)
    }

    # Word frequency
    p1_freq <- p1_final_social |> filter(!is.na(mean_zipf_freq))
    if (nrow(p1_freq) > 0) {
      cat("\n=== Word Frequency ===\n")
      m_freq <- fit_progressively(
        mean_zipf_freq ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (condition | target),
        data = p1_freq
      )
      print(summary(m_freq))
      report_effect_sizes(m_freq)
    }
  } else {
    cat("Insufficient data for H3c (need both social conditions with Phase 1 utterances).\n")
  }
}
```

```{r plot-h3c, fig.width=14, fig.height=4.5}
if (exists("p1_final_social") && "concreteness" %in% names(p1_final_social)) {
  p1_h3c <- p1_final_social |>
    filter(!is.na(concreteness), !is.na(uniqueness))

  p_concrete <- ggplot(p1_h3c, aes(x = condition, y = concreteness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Concrete terms") +
    theme(legend.position = "none")

  p_unique <- ggplot(p1_h3c, aes(x = condition, y = uniqueness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Lexical uniqueness") +
    theme(legend.position = "none")

  p_freq <- ggplot(p1_h3c |> filter(!is.na(mean_zipf_freq)),
                   aes(x = condition, y = mean_zipf_freq, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Mean Zipf frequency", title = "Word frequency") +
    theme(legend.position = "none")

  fig_h3c <- patchwork::wrap_plots(p_concrete, p_unique, p_freq) +
    plot_annotation(
      title = "H3c: Description properties in final Phase 1 utterances",
      tag_levels = "a"
    ) &
    TAG_THEME

  print(fig_h3c)
}
```

# H4: Downstream trade-off (social-first vs social-mixed)

Restricted to `social_mixed` vs `social_first` in Phase 2. Tests whether
anticipated social goals produce a trade-off: higher social accuracy but
lower referential accuracy. Both H4a and H4b are tested regardless of the H3
results. These condition contrasts do not establish mediation through the
Phase 1 convention measures.

## Response opportunities and nonresponse

Both accuracy outcomes are proportions of eligible listener response
opportunities: an active listener assigned to the task on a played trial with
a speaker message available during the response period. A correct answer
received by the server's scoring cutoff counts as 1; an incorrect answer, an
ordinary timeout, and a late arrival all count as 0. Trials with no response
opportunity leave the denominator rather than counting as failures, so these
outcomes measure successful listener performance when a speaker message was
available, not accuracy conditional on submitting an answer.

The summaries below separate nonresponse from incorrect answers. They are
descriptive and are not alternative support criteria for H4a or H4b.

```{r response-opportunities}
ref_p2 <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS)
social_p2 <- social_guess_trials(social_guesses, games, SOCIAL_CONDITIONS)

response_opportunity_summary(ref_p2) |>
  knitr::kable(digits = 3, caption = "Referential accuracy (H4b outcome)")

if (has_rows(social_p2)) {
  response_opportunity_summary(social_p2) |>
    knitr::kable(digits = 3, caption = "Social-identification accuracy (H4a outcome)")
}
```

```{r technical-exclusions}
# Documented technical failures are recorded in
# data/<dataset>/technical_exclusions.csv before the primary contrasts are
# examined, and become missing rather than unsuccessful. Missing or late
# responses alone do not qualify.
excl <- technical_exclusions()
if (nrow(excl) == 0) {
  cat("No technical failures recorded.\n")
} else {
  excl |>
    count(outcome, reason, name = "observations") |>
    knitr::kable(caption = "Recorded technical failures")
}
```

## Social-identification components and opportunities

Overall social-identification accuracy over eligible response opportunities is
the primary H4a outcome. The table below breaks it into the parts the
preregistration commits to reporting, by condition: how many in-group and
out-group response opportunities there actually were, the proportion answered
by the server's scoring cutoff, accuracy among those on-time answers, the
in-group hit rate (answering "same group" to a speaker from the guesser's own
original group), the out-group false-alarm rate (answering "same group" to a
speaker from another group), and balanced accuracy, the mean of the hit rate
and the out-group correct-rejection rate. Excluded observations are counted in
`excluded` and left out of every rate. These components separate the signal
from the response bias and show how the composition of the opportunities
changes after dropout; they are descriptive and are not alternative support
criteria for H4a. The short version is in `00_data_overview.qmd`.

```{r h4a-components}
if (!has_rows(social_p2)) {
  cat("No social-identification opportunities in the social conditions.\n")
} else {
  social_components <- social_identification_components(social_p2, trials)
  print(knitr::kable(
    social_components,
    digits = 3,
    caption = "Signal-detection components of social identification, by condition"
  ))
  if (any(social_components$speaker_unknown > 0)) {
    cat(sprintf(
      "\n%d scored opportunit(ies) have no recorded speaker group and count in neither the in-group nor the out-group column.\n",
      sum(social_components$speaker_unknown)
    ))
  }
}
```

## H4a: Social-identification accuracy

`glmer(social_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: social-identification accuracy depends on
both the listener's ability to detect cues and the speaker's tendency to produce
identifiable language.

Contrast: `social_first > social_mixed`.

```{r h4a-social-accuracy}
if (nrow(social_guesses) > 0) {
  # speakerId comes from preprocessing, attributed by (roundId, currentGroup):
  # a roundId is shared across all groups in a game, so joining on
  # (gameId, blockNum, target) would match three speakers (one per group)
  # and, because blockNum resets per phase, Phase 1 speakers too
  social_h4 <- social_guess_trials(social_guesses, games, SOCIAL_CONDITIONS)

  if (n_distinct(social_h4$condition) >= 2 && nrow(social_h4) > 0) {
    m_h4a <- fit_progressively(
      correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
      data = social_h4,
      family = binomial
    )
    print(summary(m_h4a))
    report_effect_sizes(m_h4a)

    if (inherits(m_h4a, "glmerMod")) {
      em_h4a <- emmeans(m_h4a, "condition", type = "response")
      contrast_h4a <- contrast(em_h4a, list(
        H4a_socialfirst_gt_socialmixed = c(-1, 1)
      ))
      print(summary(contrast_h4a, infer = c(TRUE, TRUE)))

      p_h4a <- as.data.frame(summary(contrast_h4a, infer = c(TRUE, TRUE)))$p.value
      # Same formula that converged above; the null drops the fixed effect of condition
      bf_h4a <- maybe_bayes_factor(p_h4a, "H4a social_first vs social_mixed", function() {
        bf_glmm_condition(social_h4, formula(m_h4a), levels = SOCIAL_CONDITIONS, name = "H4a")
      })
    }
  } else {
    cat("Insufficient data for H4a (need both social conditions).\n")
  }
} else {
  cat("No social guess data for H4a.\n")
}
```

```{r plot-h4a, fig.width=6, fig.height=5}
if (exists("social_h4") && nrow(social_h4) > 0) {
  h4a_summary <- social_h4 |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4a_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct social guesses",
      title = "H4a: Social-identification accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

## H4b: Referential accuracy

`glmer(ref_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: referential accuracy depends on both the
listener's interpretation and the quality of the speaker's description.

Contrast: `social_mixed > social_first` (lower ref accuracy expected in social_first).

```{r h4b-ref-accuracy}
# One speaker per (game, round, group): the group key is required because a
# roundId is shared across all groups in a game
p2_listener_social <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS) |>
  attach_speaker(trials)

if (n_distinct(p2_listener_social$condition) >= 2 && nrow(p2_listener_social) > 0) {
  m_h4b <- fit_progressively(
    correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
    data = p2_listener_social,
    family = binomial
  )
  print(summary(m_h4b))
  report_effect_sizes(m_h4b)

  if (inherits(m_h4b, "glmerMod")) {
    em_h4b <- emmeans(m_h4b, "condition", type = "response")
    contrast_h4b <- contrast(em_h4b, list(
      H4b_socialmixed_gt_socialfirst = c(1, -1)
    ))
    print(summary(contrast_h4b, infer = c(TRUE, TRUE)))

    p_h4b <- as.data.frame(summary(contrast_h4b, infer = c(TRUE, TRUE)))$p.value
    bf_h4b <- maybe_bayes_factor(p_h4b, "H4b social_mixed vs social_first", function() {
      bf_glmm_condition(p2_listener_social, formula(m_h4b), levels = SOCIAL_CONDITIONS, name = "H4b")
    })
  }
} else {
  cat("Insufficient data for H4b (need both social conditions with Phase 2 listener trials).\n")
}
```

```{r plot-h4b, fig.width=6, fig.height=5}
if (exists("p2_listener_social") && nrow(p2_listener_social) > 0) {
  h4b_summary <- p2_listener_social |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4b_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct",
      title = "H4b: Referential accuracy by social condition (Phase 2)"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```


# Robustness: Phase 2 results on a subset of games

The preregistration commits to sensitivity analyses for the Phase 2 analyses
(H1, H2, and H4) on subsets of games: complete nine-player games and games
that kept all three original groups. `robustness_rerun()`
(analysis/R/contrasts.R) refits the H1/H2 equal-weight OLS regression and the
H4 models on a subset of games and prints the planned contrasts next to the
primary ones. If the restriction excludes no games, the results are identical
to the primary analyses by construction and nothing is refit. The paper
explicitly does not restrict analyses to games with statistically significant
Phase 1 group specificity, so no such rerun appears here.

```{r robustness-helper}
# The H4 models to rerun, with their planned contrast weights
h4_specs <- list(
  # `late` names the lateness flag for each outcome. Lateness is established
  # separately per outcome: lateClick is a tangram selection that arrived
  # after the deadline, lateSocialGuess the social-guessing analogue.
  list(name = "H4a", data = if (exists("social_h4")) social_h4 else NULL,
       model = if (exists("m_h4a")) m_h4a else NULL,
       late = "lateSocialGuess",
       weights = list(H4a_socialfirst_gt_socialmixed = c(-1, 1))),
  list(name = "H4b", data = if (exists("p2_listener_social")) p2_listener_social else NULL,
       model = if (exists("m_h4b")) m_h4b else NULL,
       late = "lateClick",
       weights = list(H4b_socialmixed_gt_socialfirst = c(1, -1)))
)
```

## Description length as a covariate

The paper commits to this check if description length differs across
conditions in a way that tracks group specificity: each game's similarity model
gets the pair's length difference and mean length as covariates, and the H1/H2
regression is rerun on the adjusted estimates. The diagnostic and the rerun are
both shown so the decision is visible.

```{r robustness-length-covariate}
if (!has_pairwise || !all(c("length1", "length2") %in% names(pairwise_sim))) {
  cat("Skipping: pairwise_similarities.csv has no length1/length2 columns (regenerate with compute_derived.py).\n")
} else if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity estimates.\n")
} else {
  # Diagnostic: does mean Phase 2 description length differ by condition and track gs_phase2?
  length_by_game <- speaker_utts |>
    filter(phaseNum == 2) |>
    group_by(gameId) |>
    summarise(mean_length_p2 = mean(uttLength, na.rm = TRUE), .groups = "drop")
  diag <- gs_combined |> filter(both_endpoints) |> left_join(length_by_game, by = "gameId")
  cat("Mean Phase 2 description length by condition (games with both endpoints):\n")
  print(knitr::kable(diag |> group_by(condition) |>
                       summarise(games = n(), mean_length = mean(mean_length_p2, na.rm = TRUE), .groups = "drop"), digits = 2))
  if (nrow(diag) >= 4) {
    r <- cor(diag$mean_length_p2, diag$gs_phase2, use = "complete.obs")
    cat(sprintf("\nGame-level correlation between mean Phase 2 length and Phase 2 group specificity: r = %.2f\n", r))
  }

  # Group specificity with the pair's length difference and mean length as covariates
  pw_len <- pairwise_sim |>
    mutate(lengthDiff = abs(length1 - length2), meanLength = (length1 + length2) / 2)
  gs_adj <- gs_wide(game_specificity_table(pw_len, games, covariates = c("lengthDiff", "meanLength")))
  cat("\nGroup specificity with and without the length covariates:\n")
  print(knitr::kable(gs_combined |> select(gameId, condition, gs_phase1, gs_phase2) |>
                       inner_join(gs_adj |> select(gameId, gs_phase1_adj = gs_phase1, gs_phase2_adj = gs_phase2), by = "gameId") |>
                       mutate(gameId = substr(gameId, 1, 8)), digits = 3))

  h12_adj <- fit_h12(gs_adj)
  if (!is.null(h12_adj$contrasts) && all(table(h12_adj$data$condition) >= 2)) {
    cat("\nH1/H2 with length-adjusted group specificity:\n")
    print(knitr::kable(bind_rows(contrast_table(contrasts, "primary"),
                                 contrast_table(h12_adj$contrasts, "length-adjusted")), digits = 3))
  } else {
    cat("\nH1/H2 rerun needs at least two games per condition; skipped.\n")
  }
}
```

## Complete nine-player games

The complete-games restriction from the paper's dropout-handling section
(`complete_games()` in `analysis/R/attrition.R`).

```{r robustness-complete-games}
if (exists("gs_combined")) {
  complete <- complete_games(players, games)
  cat(sprintf("%d of %d games kept all nine players.\n", length(complete), sum(!is.na(games$condition))))
  robustness_rerun(complete, "complete nine-player games",
                   gs = gs_combined, primary_contrasts = contrasts, h4 = h4_specs)
} else {
  cat("Skipping: no group-specificity estimates.\n")
}
```

## Phase 2 roster size

Exposure in the mixed conditions depends on how many original groups reach
Phase 2. With three groups each player can meet six others and shares a group
with each of them 12 times in expectation over the 36 trials; when a group was
removed at the accuracy screen, the six remaining players reshuffle across two
groups and every out-group pair meets twice as often. The preregistration lets
such games contribute under the outcome-specific inclusion rules, so the Phase 2
results are checked here with the number of active original groups as a
covariate and, as a sensitivity split, on the games that kept all three groups
throughout Phase 2. The covariate is `activeGroupsMin`, the minimum number of
active original groups during Phase 2 (a group that is removed partway through
Phase 2 counts as lost for the whole check); when the pipeline has not written
that column, `activeGroups`, the number of original groups still active at the
end of the game, is used with a note. The per-game roster is in
`00_data_overview.qmd`.

```{r robustness-roster-size}
if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity estimates.\n")
} else {
  if ("activeGroupsMin" %in% names(games)) {
    roster <- games |>
      filter(!is.na(condition)) |>
      transmute(gameId, activeGroupsMin = as.integer(activeGroupsMin))
    cat("Covariate: activeGroupsMin, the minimum number of active original groups during Phase 2.\n")
  } else {
    roster <- games |>
      filter(!is.na(condition)) |>
      transmute(gameId, activeGroupsMin = as.integer(activeGroups))
    cat("NOTE: games.csv has no activeGroupsMin column (regenerate with the current pipeline);",
        "falling back to activeGroups, the number of original groups active at the end of the game.\n")
  }
  three_groups <- roster |> filter(activeGroupsMin == 3) |> pull(gameId)
  cat(sprintf("%d of %d games kept all three original groups throughout Phase 2.\n",
              length(three_groups), nrow(roster)))
  robustness_rerun(three_groups, "games with three active groups throughout Phase 2",
                   gs = gs_combined, primary_contrasts = contrasts, h4 = h4_specs)

  gs_rs <- gs_combined |>
    left_join(roster, by = "gameId") |>
    mutate(activeGroupsMin = factor(activeGroupsMin))
  if (n_distinct(gs_rs$activeGroupsMin) < 2) {
    cat("\nEvery game has the same minimum number of active groups; the covariate is not estimable.\n")
  } else {
    cat("\nGames per condition and minimum number of active groups:\n")
    print(knitr::kable(gs_rs |> count(condition, activeGroupsMin)))
    # With few games the covariate can be nested in condition (every two-group
    # game in one condition), which emmeans refuses to contrast; report that
    # rather than stop the render.
    not_estimable <- function(e) list(model = NULL, contrasts = NULL, note = conditionMessage(e))
    h12_rs <- tryCatch(fit_h12(gs_rs, covariate = "gs_phase1 + activeGroupsMin"), error = not_estimable)
    if (is.null(h12_rs$contrasts) || any(is.na(coef(h12_rs$model)))) {
      cat("\nH1/H2 with the minimum number of active groups as a covariate is not estimable in this dataset",
          if (!is.null(h12_rs$note)) paste0(" (", h12_rs$note, ")"), ".\n", sep = "")
    } else {
      cat("\nH1/H2 with the minimum number of active groups as a covariate:\n")
      print(knitr::kable(bind_rows(contrast_table(contrasts, "primary"),
                                   contrast_table(h12_rs$contrasts, "with active groups")), digits = 3))
    }

    for (spec in h4_specs) {
      if (is.null(spec$data) || is.null(spec$model) || !"gameId" %in% names(spec$data)) next
      d <- spec$data |>
        select(-any_of(c("activeGroups", "activeGroupsMin"))) |>
        left_join(roster, by = "gameId") |>
        mutate(activeGroupsMin = factor(activeGroupsMin))
      if (n_distinct(d$activeGroupsMin) < 2) next
      m_rs <- fit_progressively(update(H4_FORMULA, . ~ . + activeGroupsMin), data = d, family = binomial)
      if (!inherits(m_rs, "glmerMod")) next
      cat(sprintf("\n%s with the number of active groups as a covariate (%s):\n", spec$name, random_effects_structure(m_rs)))
      rs_contrast <- tryCatch(
        contrast_table(contrast(emmeans(m_rs, "condition", type = "response"), spec$weights), "with active groups"),
        error = function(e) { cat("  not estimable in this dataset:", conditionMessage(e), "\n"); NULL }
      )
      if (is.null(rs_contrast)) next
      print(knitr::kable(bind_rows(
        if (inherits(spec$model, "glmerMod")) contrast_table(contrast(emmeans(spec$model, "condition", type = "response"), spec$weights), "primary"),
        rs_contrast
      ), digits = 3))
    }
  }
}
```

## Tangram set as a covariate

Tangram sets are balanced within condition, and the paper states that tangram
set enters the game-level condition regressions only in a sensitivity
analysis, not as a primary covariate. This section is that sensitivity
analysis: the H1/H2 regression and the H4 models are refit with tangram set as
a covariate.

```{r robustness-tangram-set}
if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity estimates.\n")
} else if (n_distinct(games$tangramSet[!is.na(games$condition)]) < 2) {
  cat("Only one tangram set in this dataset; the covariate is not estimable (the pilot used set 0 throughout).\n")
} else {
  gs_ts <- gs_combined |>
    left_join(games |> select(gameId, tangramSet), by = "gameId") |>
    mutate(tangramSet = factor(tangramSet))
  cat("Games per condition and tangram set:\n")
  print(knitr::kable(gs_ts |> count(condition, tangramSet)))

  h12_ts <- fit_h12(gs_ts, covariate = "gs_phase1 + tangramSet")
  if (!is.null(h12_ts$contrasts)) {
    cat("\nH1/H2 with tangram set as a covariate:\n")
    print(knitr::kable(bind_rows(contrast_table(contrasts, "primary"),
                                 contrast_table(h12_ts$contrasts, "with tangram set")), digits = 3))
  } else {
    cat("\nH1/H2 rerun not possible:", h12_ts$note, "\n")
  }

  for (spec in h4_specs) {
    if (is.null(spec$data) || is.null(spec$model) || !"tangramSet" %in% names(spec$data)) next
    d <- spec$data |> mutate(tangramSet = factor(tangramSet))
    if (n_distinct(d$tangramSet) < 2) next
    m_ts <- fit_progressively(update(H4_FORMULA, . ~ . + tangramSet), data = d, family = binomial)
    if (!inherits(m_ts, "glmerMod")) next
    cat(sprintf("\n%s with tangram set as a covariate (%s):\n", spec$name, random_effects_structure(m_ts)))
    print(knitr::kable(bind_rows(
      if (inherits(spec$model, "glmerMod")) contrast_table(contrast(emmeans(spec$model, "condition", type = "response"), spec$weights), "primary"),
      contrast_table(contrast(emmeans(m_ts, "condition", type = "response"), spec$weights), "with tangram set")
    ), digits = 3))
  }
}
```

## Late arrivals

A response that reached the server after its scoring cutoff earns no points
and counts as unsuccessful in the primary coding. Lateness is established
separately for each outcome: `lateClick` flags a tangram selection, and
`lateSocialGuess` the social guess, so a trial can be late for one outcome
and on time for the other. Missing or late responses alone are not treated
as technical failures.

If any late arrivals are present, their frequency is reported by condition
and the primary accuracy comparisons are repeated with them excluded rather
than counted as unsuccessful.

```{r robustness-late-arrivals}
for (spec in h4_specs) {
  if (is.null(spec$data) || !spec$late %in% names(spec$data)) next
  late_flag <- spec$data[[spec$late]] %in% TRUE

  cat(sprintf("\n%s late arrivals (%s):\n", spec$name, spec$late))
  print(knitr::kable(
    late_arrival_summary(spec$data, late = .data[[spec$late]]), digits = 4
  ))

  if (!any(late_flag)) {
    cat("No late arrivals; the sensitivity analysis is not needed.\n")
    next
  }
  if (is.null(spec$model) || !inherits(spec$model, "glmerMod")) {
    cat("Primary model unavailable; cannot repeat the comparison.\n")
    next
  }

  d_on_time <- spec$data |> filter(!late_flag)
  m_on_time <- fit_progressively(formula(spec$model), data = d_on_time,
                                 family = binomial)
  if (!inherits(m_on_time, "glmerMod")) {
    cat("Model did not converge with late arrivals excluded.\n")
    next
  }
  cat(sprintf("%s with late arrivals excluded (%s):\n",
              spec$name, random_effects_structure(m_on_time)))
  print(knitr::kable(bind_rows(
    contrast_table(contrast(emmeans(spec$model, "condition", type = "response"),
                            spec$weights), "primary"),
    contrast_table(contrast(emmeans(m_on_time, "condition", type = "response"),
                            spec$weights), "late arrivals excluded")
  ), digits = 3))
}
```

# Robustness: content-word Jaccard similarity

The preregistered robustness measure for every similarity-based hypothesis is
the Jaccard overlap of the two descriptions' content-word sets, which replaces
the sentence-embedding cosine with a purely lexical measure.
`compute_derived.py` writes the same pairs as the SBERT tables, in the same
order, with `similarity` recomputed as that overlap
(`pairwise_similarities_jaccard.csv`, `block_pairwise_similarities_jaccard.csv`).
A pair in which either description has no content words has no Jaccard value;
those pairs are omitted, as the preregistration specifies, so their coverage
is reported first and every estimate below rests on the pairs that remain.
H1, H2, H3a, and the H3b block-3 contrast are refit on these tables and shown
beside the primary SBERT results.

```{r robustness-jaccard-coverage}
if (!has_rows(pairwise_sim_jaccard)) {
  cat(sprintf(
    "Skipping: %s not found in analysis/derived/%s/ (regenerate with compute_derived.py).\n",
    pairwise_file("jaccard"), DATASET
  ))
} else {
  cat("Endpoint pairs with and without a Jaccard similarity:\n")
  print(knitr::kable(similarity_coverage(pairwise_sim_jaccard, games), digits = 3))
}
if (has_rows(block_pairwise_sim_jaccard)) {
  cat("\nWithin-group block pairs with and without a Jaccard similarity:\n")
  print(knitr::kable(
    similarity_coverage(block_pairwise_sim_jaccard, games, by = c("phaseNum", "condition")),
    digits = 3
  ))
}
```

```{r robustness-jaccard-h12-h3a}
if (!has_rows(pairwise_sim_jaccard)) {
  cat("Skipping: no Jaccard pairwise similarities.\n")
} else {
  # Drop the pairs without a Jaccard value, then run the same equal-weight
  # endpoint pipeline as the primary analysis
  gs_table_jaccard <- game_specificity_table(
    pairwise_sim_jaccard |> filter(is.finite(similarity)),
    games,
    measure = "jaccard"
  )
  gs_jaccard <- gs_wide(gs_table_jaccard)

  cat("Endpoint coverage on the Jaccard table:\n")
  print(knitr::kable(endpoint_coverage_summary(gs_table_jaccard)))

  h12_jaccard <- fit_h12(gs_jaccard)
  h12_jaccard_reported <- !is.null(h12_jaccard$contrasts) &&
    all(table(h12_jaccard$data$condition) >= 2)
  if (!h12_jaccard_reported) {
    cat("\nH1/H2 on Jaccard similarity needs at least two games per condition; skipped",
        if (!is.null(h12_jaccard$note)) paste0(" (", h12_jaccard$note, ")"), ".\n", sep = "")
  } else {
    cat(sprintf("\nH1/H2 on Jaccard similarity (%d games):\n", nrow(h12_jaccard$data)))
    print(knitr::kable(bind_rows(
      contrast_table(if (exists("contrasts")) contrasts else NULL, "primary (SBERT)"),
      contrast_table(h12_jaccard$contrasts, "Jaccard")
    ), digits = 3))
  }

  h3a_jaccard <- fit_h3a(gs_jaccard)
  h3a_jaccard_reported <- !is.null(h3a_jaccard$contrasts) && nrow(h3a_jaccard$data) >= 4
  if (!h3a_jaccard_reported) {
    cat("\nH3a on Jaccard similarity needs both social conditions with at least two games each; skipped",
        if (!is.null(h3a_jaccard$note)) paste0(" (", h3a_jaccard$note, ")"), ".\n", sep = "")
  } else {
    cat(sprintf("\nH3a on Jaccard similarity (%d games):\n", nrow(h3a_jaccard$data)))
    print(knitr::kable(bind_rows(
      contrast_table(if (exists("contrast_h3a")) contrast_h3a else NULL, "primary (SBERT)"),
      contrast_table(h3a_jaccard$contrasts, "Jaccard")
    ), digits = 3))
  }
}
```

```{r robustness-jaccard-h3b}
if (!has_rows(block_pairwise_sim_jaccard)) {
  cat(sprintf(
    "Skipping H3b on Jaccard similarity: %s not found.\n",
    block_pairwise_file("jaccard")
  ))
} else {
  h3b_jaccard <- fit_h3b(within_group_block_means(
    block_pairwise_sim_jaccard, games, phase = 1, conditions = SOCIAL_CONDITIONS
  ))
  if (is.null(h3b_jaccard$model)) {
    cat("H3b on Jaccard similarity skipped:", h3b_jaccard$note, "\n")
  } else {
    cat(sprintf("H3b block-%d contrast on Jaccard similarity (%s):\n",
                H3B_BLOCK, random_effects_structure(h3b_jaccard$model)))
    print(knitr::kable(bind_rows(
      contrast_table(if (exists("h3b")) h3b$contrast else NULL, "primary (SBERT)"),
      contrast_table(h3b_jaccard$contrast, "Jaccard")
    ), digits = 3))
    cat("Block-3 support criterion met on Jaccard similarity:", h3b_jaccard$supported, "\n")
  }
}
```

# Statistics for the manuscript

Every planned-contrast estimate, confidence interval, p-value, and Bayes
factor printed above is written to `writing/manuscript/stats/primary.tex` as a
`\newcommand` macro, with names prefixed `primary`, so the manuscript never
carries a hand-copied number. A hypothesis the dataset cannot support
contributes no macro, which makes an unavailable number fail loudly in LaTeX
rather than read as a result.

```{r write-stats}
if (exists("gs_combined") && has_rows(gs_combined)) {
  macros <- c(macros, list(
    primaryNGamesBothEndpoints = sum(gs_combined$both_endpoints),
    primaryNGamesPhaseOneEndpoint = sum(gs_combined$eligible_phase1 %in% TRUE),
    primaryNGameWindows = nrow(gs_table)
  ))
}

# H1 and H2: the primary equal-weight OLS contrasts, their Bayes factors, and
# the inverse-variance robustness rerun
if (exists("contrasts") && !is.null(contrasts)) {
  macros <- c(
    macros,
    list(primaryNGamesHOneTwo = nrow(gs_h12)),
    contrast_macros("primaryHOne", contrasts, row = 1),
    contrast_macros("primaryHTwo", contrasts, row = 2)
  )
  if (exists("bf_h1")) macros <- c(macros, bf_macros("primaryHOne", bf_h1))
  if (exists("bf_h2")) macros <- c(macros, bf_macros("primaryHTwo", bf_h2))
}
if (exists("h12_wls") && !is.null(h12_wls$contrasts)) {
  macros <- c(
    macros,
    contrast_macros("primaryHOneWls", h12_wls$contrasts, row = 1),
    contrast_macros("primaryHTwoWls", h12_wls$contrasts, row = 2)
  )
}

# H3a
if (exists("contrast_h3a") && !is.null(contrast_h3a)) {
  macros <- c(
    macros,
    list(primaryNGamesHThreeA = nrow(gs_social)),
    contrast_macros("primaryHThreeA", contrast_h3a, row = 1)
  )
  if (exists("bf_h3a")) macros <- c(macros, bf_macros("primaryHThreeA", bf_h3a))
}

# H3b: the block-3 contrast
if (exists("h3b") && !is.null(h3b$summary) && nrow(h3b$summary) == 1) {
  sm <- h3b$summary
  lo <- if ("lower.CL" %in% names(sm)) sm$lower.CL else sm$asymp.LCL
  hi <- if ("upper.CL" %in% names(sm)) sm$upper.CL else sm$asymp.UCL
  macros <- c(macros, list(
    primaryHThreeBBlock = H3B_BLOCK,
    primaryHThreeBEst = fmt_stat(sm$estimate),
    primaryHThreeBLo = fmt_stat(lo),
    primaryHThreeBHi = fmt_stat(hi),
    primaryHThreeBP = fmt_pval(sm$p.value)
  ))
}

# H3c: the three description properties (reference level is social_mixed, so
# the coefficient is social_first minus social_mixed)
if (exists("m_concrete")) macros <- c(macros, coef_macros("primaryHThreeCConcrete", m_concrete, "conditionsocial_first"))
if (exists("m_uniqueness")) macros <- c(macros, coef_macros("primaryHThreeCUnique", m_uniqueness, "conditionsocial_first"))
if (exists("m_freq")) macros <- c(macros, coef_macros("primaryHThreeCFreq", m_freq, "conditionsocial_first"))

# H4a and H4b: odds-ratio contrasts on the response scale, with their Bayes
# factors
if (exists("contrast_h4a")) {
  macros <- c(macros, contrast_macros("primaryHFourA", contrast_h4a, row = 1))
  if (exists("bf_h4a")) macros <- c(macros, bf_macros("primaryHFourA", bf_h4a))
}
if (exists("contrast_h4b")) {
  macros <- c(macros, contrast_macros("primaryHFourB", contrast_h4b, row = 1))
  if (exists("bf_h4b")) macros <- c(macros, bf_macros("primaryHFourB", bf_h4b))
}

# The signal-detection components of H4a, pooled over the social conditions
if (exists("social_components") && has_rows(social_components)) {
  tot <- function(col) sum(social_components[[col]], na.rm = TRUE)
  macros <- c(macros, list(
    primaryNSocialOpportunities = fmt_count(tot("opportunities")),
    primaryNSocialInGroup = fmt_count(tot("in_group_opportunities")),
    primaryNSocialOutGroup = fmt_count(tot("out_group_opportunities")),
    primarySocialOnTimeRate = fmt_stat(tot("on_time") / tot("scored")),
    primarySocialHitRate = fmt_stat(sum(social_components$hit_rate * social_components$in_group_answered, na.rm = TRUE) /
                                      tot("in_group_answered")),
    primarySocialFalseAlarmRate = fmt_stat(sum(social_components$false_alarm_rate * social_components$out_group_answered, na.rm = TRUE) /
                                             tot("out_group_answered"))
  ))
}

# The Jaccard robustness reruns
if (exists("h12_jaccard_reported") && isTRUE(h12_jaccard_reported)) {
  macros <- c(
    macros,
    contrast_macros("primaryHOneJaccard", h12_jaccard$contrasts, row = 1),
    contrast_macros("primaryHTwoJaccard", h12_jaccard$contrasts, row = 2)
  )
}
if (exists("h3a_jaccard_reported") && isTRUE(h3a_jaccard_reported)) {
  macros <- c(macros, contrast_macros("primaryHThreeAJaccard", h3a_jaccard$contrasts, row = 1))
}
if (exists("h3b_jaccard") && !is.null(h3b_jaccard$contrast)) {
  macros <- c(macros, contrast_macros("primaryHThreeBJaccard", h3b_jaccard$contrast, row = 1))
}

write_stats_tex(macros, stats_file("primary"), source = "analysis/02_primary_analysis.qmd")
```
