---
title: "02: Primary Analysis — H1 & H2"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyboot)
library(patchwork)
source(here::here("analysis", "config.R"))

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, utterances_file), show_col_types = FALSE)
social_guesses <- read_csv(file.path(data_dir, "social_guesses.csv"), show_col_types = FALSE)

has_pairwise <- file.exists(file.path(derived_dir, "pairwise_similarities.csv"))
if (!has_pairwise) {
  cat("WARNING: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
  cat("Most analyses in this notebook will be skipped.\n")
  pairwise_sim <- tibble()
} else {
  pairwise_sim <- read_csv(file.path(derived_dir, "pairwise_similarities.csv"), show_col_types = FALSE)
}

has_block_pairwise <- file.exists(file.path(derived_dir, "block_pairwise_similarities.csv"))
block_pairwise_sim <- if (has_block_pairwise) {
  read_csv(file.path(derived_dir, "block_pairwise_similarities.csv"), show_col_types = FALSE)
} else {
  tibble()
}
```

# Compute Group-Specificity per Game per Phase

For each game, fit:
`similarity ~ sameGroup + (1 | target) + (1 | speaker1) + (1 | speaker2)`

Extract the `sameGroup` coefficient as the group-specificity score.

```{r fit-group-specificity}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping group-specificity analysis: no pairwise similarity data.\n")
} else {
fit_gs_by_game_window <- function(pw_df) {
  game_ids <- unique(pw_df$gameId)

  map_dfr(game_ids, function(gid) {
    game_data <- pw_df |> filter(gameId == gid)

    if (nrow(game_data) < 5) return(NULL)
    if (n_distinct(game_data$sameGroup) < 2) return(NULL)

    tryCatch({
      model <- lmer(
        similarity ~ sameGroup + (1 | target) + (1 | speaker1) + (1 | speaker2),
        data = game_data,
        control = lmerControl(optimizer = "bobyqa")
      )
      coefs <- coef(summary(model))
      tibble(
        gameId = gid,
        coefficient = coefs["sameGroup", "Estimate"],
        std_error = coefs["sameGroup", "Std. Error"],
        t_value = coefs["sameGroup", "t value"]
      )
    }, error = function(e) NULL)
  })
}

# Phase 1 final
gs_p1 <- pairwise_sim |>
  filter(window == "phase1_final") |>
  fit_gs_by_game_window() |>
  rename(gs_phase1 = coefficient, se_phase1 = std_error) |>
  select(gameId, gs_phase1, se_phase1)

# Phase 2 final
gs_p2 <- pairwise_sim |>
  filter(window == "phase2_final") |>
  fit_gs_by_game_window() |>
  rename(gs_phase2 = coefficient, se_phase2 = std_error) |>
  select(gameId, gs_phase2, se_phase2)

# Combine
gs_combined <- gs_p1 |>
  inner_join(gs_p2, by = "gameId") |>
  left_join(games |> select(gameId, condition), by = "gameId")

cat("Group-specificity estimates:\n")
gs_combined |>
  knitr::kable(digits = 4, caption = "Group-specificity by game and phase")
} # end has_pairwise
```

# Weighted Least Squares Regression

```
group_specificity_phase2 ~ condition + group_specificity_phase1
```
Weights: `1 / SE_phase2^2` (inverse variance of Phase 2 estimates).

```{r wls-model}
if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity data.\n")
} else {
gs_combined <- gs_combined |>
  mutate(
    weight = 1 / se_phase2^2,
    condition = factor(condition, levels = CONDITION_ORDER)
  )

n_conditions <- n_distinct(gs_combined$condition)

if (n_conditions >= 2) {
  wls_model <- lm(
    gs_phase2 ~ condition + gs_phase1,
    data = gs_combined,
    weights = weight
  )
  summary(wls_model)
  report_effect_sizes(wls_model)
} else {
  cat(sprintf("Only %d condition(s) in data (%s). Need ≥2 for WLS model.\n",
              n_conditions, paste(unique(gs_combined$condition), collapse = ", ")))
  wls_model <- NULL
}
} # end exists("gs_combined")
```

# Planned Contrasts

Two-tailed tests at α = 0.05.

- **H1:** `refer_separated > refer_mixed`
- **H2:** `social_mixed > refer_mixed`

```{r planned-contrasts}
if (!is.null(wls_model) && n_conditions >= 4) {
  em <- emmeans(wls_model, "condition")
  # Order: refer_separated, refer_mixed, social_mixed, social_first
  contrasts <- contrast(em, list(
    H1_sep_vs_mixed = c(1, -1, 0, 0),    # refer_separated - refer_mixed
    H2_social_vs_mixed = c(0, -1, 1, 0)  # social_mixed - refer_mixed
  ))
  summary(contrasts, infer = c(TRUE, TRUE))
} else {
  cat("Skipping contrasts: need all 3 conditions.\n")
  contrasts <- NULL
}
```

```{r contrast-details}
if (!is.null(contrasts)) {
  cat("\nDetailed contrast results:\n")
  contrast_summary <- summary(contrasts, infer = c(TRUE, TRUE))
  for (i in seq_len(nrow(contrast_summary))) {
    row <- contrast_summary[i, ]
    cat(sprintf(
      "\n%s:\n  Estimate = %.4f, 95%% CI [%.4f, %.4f], p = %.4f\n",
      row$contrast, row$estimate, row$lower.CL, row$upper.CL, row$p.value
    ))
  }
}
```

# Plots

```{r plot-gs-paired, fig.width=8, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 6: Group-specificity Phase 1 vs Phase 2 by condition
gs_long <- gs_combined |>
  pivot_longer(
    cols = c(gs_phase1, gs_phase2),
    names_to = "phase",
    values_to = "group_specificity"
  ) |>
  mutate(phase = ifelse(phase == "gs_phase1", "Phase 1", "Phase 2"))

ggplot(gs_long, aes(x = phase, y = group_specificity, group = gameId)) +
  geom_line(alpha = 0.5, color = "gray50") +
  geom_point(aes(color = condition), size = 3) +
  facet_wrap(~condition) +
  scale_color_condition() +
  labs(
    x = "Phase",
    y = "Group-specificity coefficient",
    title = "Group-specificity across phases by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-gs-phase2, fig.width=7, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 7: Phase 2 group-specificity by condition (primary result)
ggplot(gs_combined, aes(x = condition, y = gs_phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Phase 2 group-specificity",
    title = "Group-specificity in Phase 2 by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-wls-forest, fig.width=7, fig.height=4}
# Plot 8: WLS coefficient forest plot
if (!is.null(wls_model)) {
  wls_coefs <- broom::tidy(wls_model, conf.int = TRUE) |>
    filter(term != "(Intercept)")

  ggplot(wls_coefs, aes(x = estimate, y = term)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
    geom_point(size = 3) +
    labs(
      x = "Coefficient estimate",
      y = "",
      title = "WLS model coefficients"
    )
} else {
  cat("WLS model not fitted (insufficient conditions).\n")
}
```

```{r plot-within-between, fig.width=10, fig.height=5}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping plot: no pairwise similarity data.\n")
} else {
# Plot 9: Within- vs between-group similarity by condition
pw_summary <- pairwise_sim |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  mutate(
    groupType = ifelse(sameGroup == 1, "Within-group", "Between-group"),
    phase = ifelse(window == "phase1_final", "Phase 1 final", "Phase 2 final")
  )

pw_means <- pw_summary |>
  group_by(condition, phase, groupType) |>
  summarise(
    mean_sim = mean(similarity, na.rm = TRUE),
    se_sim = sd(similarity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(pw_means, aes(x = condition, y = mean_sim, fill = groupType)) +
  geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_sim - se_sim, ymax = mean_sim + se_sim),
    position = position_dodge(0.8), width = 0.2
  ) +
  facet_wrap(~phase) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Condition",
    y = "Mean cosine similarity",
    fill = "Group type",
    title = "Within- vs between-group similarity by condition and phase"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

# Optional: Bayes Factors (if null results)

Uncomment and run if contrasts are non-significant.

```{r bayes-factors, eval=FALSE}
# library(brms)
# library(bridgesampling)
#
# # Refit with brms using Cauchy(0, sqrt(2)/2) prior on condition effects
# prior <- c(
#   set_prior("cauchy(0, 0.707)", class = "b")
# )
#
# brm_model <- brm(
#   gs_phase2 | weights(weight) ~ condition + gs_phase1,
#   data = gs_combined,
#   prior = prior,
#   iter = 10000,
#   warmup = 2000,
#   chains = 4,
#   save_pars = save_pars(all = TRUE)
# )
#
# # Compute Bayes factors via bridge sampling
# brm_null <- brm(
#   gs_phase2 | weights(weight) ~ gs_phase1,
#   data = gs_combined,
#   iter = 10000,
#   warmup = 2000,
#   chains = 4,
#   save_pars = save_pars(all = TRUE)
# )
#
# bf <- bayes_factor(brm_model, brm_null)
# cat("Bayes factor (full vs null):", bf$bf, "\n")
```

# H3: Effect of Anticipated Social Goal on Phase 1 Conventions

Restricted to `social_mixed` vs `social_first`. Tests whether anticipating
a social-signaling task shapes convention formation *before* Phase 2.

## H3a: Phase 1 Group-Specificity

WLS regression: `group_specificity_phase1 ~ condition`, weighted by inverse
variance of Phase 1 GS estimates. Contrast: `social_first > social_mixed`.

```{r h3a-gs}
if (!exists("gs_combined")) {
  cat("Skipping H3a: no group-specificity data.\n")
} else {
  gs_social <- gs_combined |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    mutate(
      weight_p1 = 1 / se_phase1^2,
      condition = factor(condition, levels = c("social_mixed", "social_first"))
    )

  if (n_distinct(gs_social$condition) >= 2 && nrow(gs_social) >= 4) {
    wls_h3a <- lm(
      gs_phase1 ~ condition,
      data = gs_social,
      weights = weight_p1
    )
    summary(wls_h3a)
    report_effect_sizes(wls_h3a)

    em_h3a <- emmeans(wls_h3a, "condition")
    contrast_h3a <- contrast(em_h3a, list(
      H3a_socialfirst_gt_socialmixed = c(-1, 1)  # social_first - social_mixed
    ))
    summary(contrast_h3a, infer = c(TRUE, TRUE))
  } else {
    cat("Insufficient data for H3a (need both social conditions).\n")
  }
}
```

```{r plot-h3a, fig.width=6, fig.height=5}
if (exists("gs_social") && nrow(gs_social) > 0) {
  ggplot(gs_social, aes(x = condition, y = gs_phase1, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Phase 1 group-specificity",
      title = "H3a: Phase 1 group-specificity by social condition"
    ) +
    theme(legend.position = "none")
}
```

## H3b: Within-Group Convergence Rate

`within_group_similarity ~ condition * block + (block | group) + (block | tangram)`,
restricted to Phase 1, within-group pairs, social conditions only.
Group (not game) is the natural unit of replication for within-group similarity,
and condition is between-game so cannot have a random slope within game or group.

```{r h3b-convergence}
if (!has_block_pairwise || nrow(block_pairwise_sim) == 0) {
  cat("Skipping H3b: block_pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else {
  # Filter to Phase 1, within-group pairs, social conditions
  p1_within_social <- block_pairwise_sim |>
    filter(phaseNum == 1, sameGroup == 1) |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    mutate(
      blockNum_c = blockNum - mean(blockNum, na.rm = TRUE),
      condition = factor(condition, levels = c("social_mixed", "social_first"))
    )

  # Derive group ID for within-group pairs (group1 == group2 since sameGroup == 1)
  p1_within_social <- p1_within_social |>
    mutate(group = paste0(gameId, "_", group1))

  if (nrow(p1_within_social) > 0 && n_distinct(p1_within_social$condition) >= 2) {
    # Group is the natural unit of replication for within-group similarity.
    # Condition is between-game, so it cannot have a random slope within group.
    m_h3b <- fit_progressively(
      list(
        maximal    = similarity ~ condition * blockNum_c + (blockNum_c | group) + (blockNum_c | target),
        intercepts = similarity ~ condition * blockNum_c + (1 | group) + (1 | target)
      ),
      data = p1_within_social
    )
    summary(m_h3b)
    report_effect_sizes(m_h3b)
  } else {
    cat("Insufficient data for H3b convergence analysis.\n")
  }
}
```

```{r plot-h3b, fig.width=8, fig.height=5}
if (exists("p1_within_social") && nrow(p1_within_social) > 0) {
  conv_by_block <- p1_within_social |>
    filter(!is.na(similarity)) |>
    group_by(condition, blockNum) |>
    tidyboot_mean(similarity, nboot = 1000)

  ggplot(conv_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number (Phase 1)",
      y = "Within-group cosine similarity",
      color = "Condition",
      title = "H3b: Within-group convergence rate by social condition"
    )
}
```

## H3c: Description Properties

Compare concreteness, lexical uniqueness, and word frequency of final Phase 1
descriptions between `social_mixed` and `social_first`.

All three measures are precomputed in Python (`compute_derived.py`) with
consistent tokenization: each utterance is lowercased, split into alphabetic
sequences, filtered for standard English stopwords, and only tokens with
length > 1 are retained as *content words*.

- **Concreteness**: the proportion of content words matching concrete term
  stems (prefix matching) from Boyce et al. 2024. Four categories: geometric
  (e.g., *triangle*, *square*, *diamond*), body parts (e.g., *arm*, *head*,
  *leg*), positional (e.g., *left*, *right*, *top*), and posture (e.g.,
  *kick*, *kneel*, *stand*). Prefix matching means *squar* matches both
  *square* and *squares*.
- **Lexical uniqueness**: the proportion of content words in a description
  that do not appear in any other group's descriptions for the same tangram
  across games.
- **Word frequency**: the mean Zipf frequency of content words, computed via
  the `wordfreq` Python package (which draws on SUBTLEX and other corpora).
  Common words score higher (e.g., "the" ≈ 7.7); rare words score lower.

```{r h3c-description-properties}
has_desc_props <- file.exists(file.path(derived_dir, "description_properties.csv"))
has_lex_uniq <- file.exists(file.path(derived_dir, "lexical_uniqueness.csv"))

if (!has_desc_props || !has_lex_uniq) {
  cat("Missing H3c data files. Run compute_description_properties() and compute_lexical_uniqueness() first.\n")
} else {
  desc_props <- read_csv(file.path(derived_dir, "description_properties.csv"), show_col_types = FALSE)
  lex_uniq <- read_csv(file.path(derived_dir, "lexical_uniqueness.csv"), show_col_types = FALSE)

  # Merge, filter to final Phase 1 block, social conditions only
  p1_final_social <- desc_props |>
    left_join(
      lex_uniq |> select(gameId, playerId, target, blockNum, phaseNum, uniqueness),
      by = c("gameId", "playerId", "target", "blockNum", "phaseNum")
    ) |>
    filter(phaseNum == 1) |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    group_by(gameId, playerId, target) |>
    filter(blockNum == max(blockNum)) |>
    ungroup() |>
    mutate(condition = factor(condition, levels = c("social_mixed", "social_first")))

  if (nrow(p1_final_social) > 0 && n_distinct(p1_final_social$condition) >= 2) {
    # Concreteness
    p1_concrete <- p1_final_social |> filter(!is.na(concreteness))
    if (nrow(p1_concrete) > 0) {
      cat("=== Concreteness ===\n")
      m_concrete <- fit_progressively(
        list(
          maximal    = concreteness ~ condition + (1 | playerId) + (condition | target),
          intercepts = concreteness ~ condition + (1 | playerId) + (1 | target)
        ),
        data = p1_concrete
      )
      summary(m_concrete)
      report_effect_sizes(m_concrete)
    }

    # Lexical uniqueness
    p1_unique <- p1_final_social |> filter(!is.na(uniqueness))
    if (nrow(p1_unique) > 0) {
      cat("\n=== Lexical Uniqueness ===\n")
      m_uniqueness <- fit_progressively(
        list(
          maximal    = uniqueness ~ condition + (1 | playerId) + (condition | target),
          intercepts = uniqueness ~ condition + (1 | playerId) + (1 | target)
        ),
        data = p1_unique
      )
      summary(m_uniqueness)
      report_effect_sizes(m_uniqueness)
    }

    # Word frequency
    p1_freq <- p1_final_social |> filter(!is.na(mean_zipf_freq))
    if (nrow(p1_freq) > 0) {
      cat("\n=== Word Frequency ===\n")
      m_freq <- fit_progressively(
        list(
          maximal    = mean_zipf_freq ~ condition + (1 | playerId) + (condition | target),
          intercepts = mean_zipf_freq ~ condition + (1 | playerId) + (1 | target)
        ),
        data = p1_freq
      )
      summary(m_freq)
      report_effect_sizes(m_freq)
    }
  } else {
    cat("Insufficient data for H3c (need both social conditions with Phase 1 utterances).\n")
  }
}
```

```{r plot-h3c, fig.width=14, fig.height=4.5}
if (exists("p1_final_social") && "concreteness" %in% names(p1_final_social)) {
  p1_h3c <- p1_final_social |>
    filter(!is.na(concreteness), !is.na(uniqueness))

  p_concrete <- ggplot(p1_h3c, aes(x = condition, y = concreteness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Concrete terms") +
    theme(legend.position = "none")

  p_unique <- ggplot(p1_h3c, aes(x = condition, y = uniqueness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Lexical uniqueness") +
    theme(legend.position = "none")

  p_freq <- ggplot(p1_h3c |> filter(!is.na(mean_zipf_freq)),
                   aes(x = condition, y = mean_zipf_freq, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Mean Zipf frequency", title = "Word frequency") +
    theme(legend.position = "none")

  fig_h3c <- patchwork::wrap_plots(p_concrete, p_unique, p_freq) +
    plot_annotation(
      title = "H3c: Description properties in final Phase 1 utterances",
      tag_levels = "a"
    ) &
    TAG_THEME

  print(fig_h3c)
}
```

# H4: Downstream Trade-Off (Social-First vs Social-Mixed)

Restricted to `social_mixed` vs `social_first` in Phase 2. Tests whether
anticipated social goals produce a trade-off: higher social accuracy but
lower referential accuracy. Contingent on H3a.

## H4a: Social-Identification Accuracy

`glmer(social_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: social-identification accuracy depends on
both the listener's ability to detect cues and the speaker's tendency to produce
identifiable language.

Contrast: `social_first > social_mixed`.

```{r h4a-social-accuracy}
if (nrow(social_guesses) > 0) {
  # Join speaker identity from trials (one speaker per game × block × target)
  speaker_lookup <- trials |>
    filter(role == "speaker") |>
    select(gameId, blockNum, target, playerId) |>
    rename(speakerId = playerId) |>
    distinct()

  social_h4 <- social_guesses |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    left_join(speaker_lookup, by = c("gameId", "blockNum", "target")) |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    mutate(
      correct = as.numeric(socialGuessCorrect == TRUE | socialGuessCorrect == "true"),
      condition = factor(condition, levels = c("social_mixed", "social_first")),
      blockNum_c = blockNum - mean(blockNum, na.rm = TRUE)
    )

  if (n_distinct(social_h4$condition) >= 2 && nrow(social_h4) > 0) {
    m_h4a <- fit_progressively(
      list(
        maximal    = correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
        drop_cond  = correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (1 | target),
        intercepts = correct ~ condition + blockNum_c + (1 | gameId) + (1 | playerId) + (1 | speakerId) + (1 | target)
      ),
      data = social_h4,
      family = binomial
    )
    summary(m_h4a)
    report_effect_sizes(m_h4a)

    if (inherits(m_h4a, "glmerMod")) {
      em_h4a <- emmeans(m_h4a, "condition", type = "response")
      contrast_h4a <- contrast(em_h4a, list(
        H4a_socialfirst_gt_socialmixed = c(-1, 1)
      ))
      summary(contrast_h4a, infer = c(TRUE, TRUE))
    }
  } else {
    cat("Insufficient data for H4a (need both social conditions).\n")
  }
} else {
  cat("No social guess data for H4a.\n")
}
```

```{r plot-h4a, fig.width=6, fig.height=5}
if (exists("social_h4") && nrow(social_h4) > 0) {
  h4a_summary <- social_h4 |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4a_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct social guesses",
      title = "H4a: Social-identification accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

## H4b: Referential Accuracy

`glmer(ref_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: referential accuracy depends on both the
listener's interpretation and the quality of the speaker's description.

Contrast: `social_mixed > social_first` (lower ref accuracy expected in social_first).

```{r h4b-ref-accuracy}
# Join speaker identity onto listener trials
speaker_lookup_h4b <- trials |>
  filter(role == "speaker") |>
  select(gameId, blockNum, target, playerId) |>
  rename(speakerId = playerId) |>
  distinct()

p2_listener_social <- trials |>
  filter(phaseNum == 2, role == "listener") |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  left_join(speaker_lookup_h4b, by = c("gameId", "blockNum", "target")) |>
  filter(condition %in% c("social_mixed", "social_first")) |>
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    condition = factor(condition, levels = c("social_mixed", "social_first")),
    blockNum_c = blockNum - mean(blockNum, na.rm = TRUE)
  )

if (n_distinct(p2_listener_social$condition) >= 2 && nrow(p2_listener_social) > 0) {
  m_h4b <- fit_progressively(
    list(
      maximal    = correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
      drop_cond  = correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (1 | target),
      intercepts = correct ~ condition + blockNum_c + (1 | gameId) + (1 | playerId) + (1 | speakerId) + (1 | target)
    ),
    data = p2_listener_social,
    family = binomial
  )
  summary(m_h4b)
  report_effect_sizes(m_h4b)

  if (inherits(m_h4b, "glmerMod")) {
    em_h4b <- emmeans(m_h4b, "condition", type = "response")
    contrast_h4b <- contrast(em_h4b, list(
      H4b_socialmixed_gt_socialfirst = c(1, -1)
    ))
    summary(contrast_h4b, infer = c(TRUE, TRUE))
  }
} else {
  cat("Insufficient data for H4b (need both social conditions with Phase 2 listener trials).\n")
}
```

```{r plot-h4b, fig.width=6, fig.height=5}
if (exists("p2_listener_social") && nrow(p2_listener_social) > 0) {
  h4b_summary <- p2_listener_social |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4b_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct",
      title = "H4b: Referential accuracy by social condition (Phase 2)"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```
