---
title: "02: Primary analysis — H1 & H2"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(emmeans)
library(tidyboot)
library(patchwork)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_pairwise <- has_rows(pairwise_sim)
has_block_pairwise <- has_rows(block_pairwise_sim)
if (!has_pairwise) {
  cat("WARNING: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
  cat("Most analyses in this notebook will be skipped.\n")
}
```

# Compute group-specificity per game per phase

For each game, fit:
`similarity ~ sameGroup + (1 | target) + (1 | speaker1) + (1 | speaker2)`

Extract the `sameGroup` coefficient as the group-specificity score.

```{r fit-group-specificity}
if (!has_pairwise) {
  cat("Skipping group-specificity analysis: no pairwise similarity data.\n")
} else {
  # game_specificity_table() (analysis/R/group_specificity.R) is the one place
  # the per-game estimates are fit; gs_wide() gives one row per game with the
  # Phase 1 and Phase 2 estimates and the inverse-variance weights.
  gs_table <- game_specificity_table(pairwise_sim, games)
  gs_combined <- gs_wide(gs_table)

  cat("Group-specificity estimates:\n")
  print(knitr::kable(gs_combined |> select(gameId, condition, gs_phase1, se_phase1, gs_phase2, se_phase2),
                     digits = 4, caption = "Group-specificity by game and phase"))
}
```

# Weighted least squares regression

```
group_specificity_phase2 ~ condition + group_specificity_phase1
```
Weights: `1 / SE_phase2^2` (inverse variance of Phase 2 estimates).

Restricted to the three conditions that share the same Phase 1 procedure
(`refer_separated`, `refer_mixed`, `social_mixed`). In `social_first` the
manipulation precedes Phase 1, so its Phase 1 group-specificity is an outcome
(H3a) rather than a pre-manipulation baseline, and pooling it into the common
covariate slope could bias the H1 and H2 contrasts. `social_first` is kept in
the descriptive plots below.

```{r wls-model}
if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity data.\n")
} else {
  # fit_h12_wls() (analysis/R/contrasts.R) restricts to H12_CONDITIONS and
  # returns the data, the lm, and the planned contrasts
  h12 <- fit_h12_wls(gs_combined)
  gs_h12 <- h12$data
  wls_model <- h12$model
  contrasts <- h12$contrasts
  if (is.null(wls_model)) {
    cat(sprintf("Skipping the H1/H2 model: %s.\n", h12$note))
  } else {
    print(summary(wls_model))
    report_effect_sizes(wls_model)
  }
}
```

# Planned contrasts

Two-tailed tests at α = 0.05.

- **H1:** `refer_separated > refer_mixed`
- **H2:** `social_mixed > refer_mixed`

```{r planned-contrasts}
if (exists("contrasts") && !is.null(contrasts)) {
  print(summary(contrasts, infer = c(TRUE, TRUE)))
} else {
  contrasts <- NULL
  cat("Skipping contrasts: need all three H1/H2 conditions.\n")
}
```

Game-level Cohen's *d* for each contrast (unadjusted difference in Phase 2
group-specificity between the two conditions, pooled SD; the quantity the
power analysis was based on). Not estimable with one game per condition.

```{r effect-sizes-h1-h2}
if (!is.null(contrasts)) {
  rbind(
    cohens_d_games(gs_h12, "gs_phase2", "condition", "refer_separated", "refer_mixed"),
    cohens_d_games(gs_h12, "gs_phase2", "condition", "social_mixed", "refer_mixed")
  )
}
```

Bayes factors for non-significant contrasts (preregistered: brms refit,
standardized estimates, Cauchy(0, sqrt(2)/2) prior on the effect size; see
`bayes_factors.R`). Each contrast is tested on its two conditions, keeping the
Phase 1 covariate and the inverse-variance weights.

```{r bayes-factors-h1-h2}
if (!is.null(contrasts)) {
  p <- as.data.frame(summary(contrasts, infer = c(TRUE, TRUE)))$p.value
  maybe_bayes_factor(p[1], "H1 refer_separated vs refer_mixed", function() {
    bf_wls_contrast(gs_h12, "gs_phase2", levels = c("refer_mixed", "refer_separated"),
                    covariates = "gs_phase1", weights = "weight", name = "H1")
  })
  maybe_bayes_factor(p[2], "H2 social_mixed vs refer_mixed", function() {
    bf_wls_contrast(gs_h12, "gs_phase2", levels = c("refer_mixed", "social_mixed"),
                    covariates = "gs_phase1", weights = "weight", name = "H2")
  })
}
```

```{r contrast-details}
if (!is.null(contrasts)) {
  cat("\nDetailed contrast results:\n")
  contrast_summary <- summary(contrasts, infer = c(TRUE, TRUE))
  for (i in seq_len(nrow(contrast_summary))) {
    row <- contrast_summary[i, ]
    cat(sprintf(
      "\n%s:\n  Estimate = %.4f, 95%% CI [%.4f, %.4f], p = %.4f\n",
      row$contrast, row$estimate, row$lower.CL, row$upper.CL, row$p.value
    ))
  }
}
```

# Plots

```{r plot-gs-paired, fig.width=8, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 6: Group-specificity Phase 1 vs Phase 2 by condition
gs_long <- gs_combined |>
  pivot_longer(
    cols = c(gs_phase1, gs_phase2),
    names_to = "phase",
    values_to = "group_specificity"
  ) |>
  mutate(phase = ifelse(phase == "gs_phase1", "Phase 1", "Phase 2"))

ggplot(gs_long, aes(x = phase, y = group_specificity, group = gameId)) +
  geom_line(alpha = 0.5, color = "gray50") +
  geom_point(aes(color = condition), size = 3) +
  facet_wrap(~condition) +
  scale_color_condition() +
  labs(
    x = "Phase",
    y = "Group-specificity coefficient",
    title = "Group-specificity across phases by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-gs-phase2, fig.width=7, fig.height=5}
if (!exists("gs_combined")) {
  cat("Skipping plot: no group-specificity data.\n")
} else {
# Plot 7: Phase 2 group-specificity by condition (primary result)
ggplot(gs_combined, aes(x = condition, y = gs_phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Phase 2 group-specificity",
    title = "Group-specificity in Phase 2 by condition"
  ) +
  theme(legend.position = "none")
}
```

```{r plot-wls-forest, fig.width=7, fig.height=4}
# Plot 8: WLS coefficient forest plot
if (!is.null(wls_model)) {
  wls_coefs <- broom::tidy(wls_model, conf.int = TRUE) |>
    filter(term != "(Intercept)")

  ggplot(wls_coefs, aes(x = estimate, y = term)) +
    geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
    geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.2) +
    geom_point(size = 3) +
    labs(
      x = "Coefficient estimate",
      y = "",
      title = "WLS model coefficients"
    )
} else {
  cat("WLS model not fitted (insufficient conditions).\n")
}
```

```{r plot-within-between, fig.width=10, fig.height=5}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping plot: no pairwise similarity data.\n")
} else {
# Plot 9: Within- vs between-group similarity by condition
pw_summary <- pairwise_sim |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  mutate(
    groupType = ifelse(sameGroup == 1, "Within-group", "Between-group"),
    phase = ifelse(window == "phase1_final", "Phase 1 final", "Phase 2 final")
  )

pw_means <- pw_summary |>
  group_by(condition, phase, groupType) |>
  summarise(
    mean_sim = mean(similarity, na.rm = TRUE),
    se_sim = sd(similarity, na.rm = TRUE) / sqrt(n()),
    .groups = "drop"
  )

ggplot(pw_means, aes(x = condition, y = mean_sim, fill = groupType)) +
  geom_col(position = position_dodge(0.8), width = 0.7, alpha = 0.8) +
  geom_errorbar(
    aes(ymin = mean_sim - se_sim, ymax = mean_sim + se_sim),
    position = position_dodge(0.8), width = 0.2
  ) +
  facet_wrap(~phase) +
  scale_fill_brewer(palette = "Pastel1") +
  labs(
    x = "Condition",
    y = "Mean cosine similarity",
    fill = "Group type",
    title = "Within- vs between-group similarity by condition and phase"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```


# H3: Effect of anticipated social goal on Phase 1 conventions

Restricted to `social_mixed` vs `social_first`. Tests whether anticipating
a social-signaling task shapes convention formation *before* Phase 2.

## H3a: Phase 1 group-specificity

WLS regression: `group_specificity_phase1 ~ condition`, weighted by inverse
variance of Phase 1 GS estimates. Contrast: `social_first > social_mixed`.

```{r h3a-gs}
if (!exists("gs_combined")) {
  cat("Skipping H3a: no group-specificity data.\n")
} else {
  gs_social <- gs_combined |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    mutate(
      weight_p1 = 1 / se_phase1^2,
      condition = factor(condition, levels = c("social_mixed", "social_first"))
    )

  if (n_distinct(gs_social$condition) >= 2 && nrow(gs_social) >= 4) {
    wls_h3a <- lm(
      gs_phase1 ~ condition,
      data = gs_social,
      weights = weight_p1
    )
    print(summary(wls_h3a))
    report_effect_sizes(wls_h3a)

    em_h3a <- emmeans(wls_h3a, "condition")
    contrast_h3a <- contrast(em_h3a, list(
      H3a_socialfirst_gt_socialmixed = c(-1, 1)  # social_first - social_mixed
    ))
    print(summary(contrast_h3a, infer = c(TRUE, TRUE)))

    print(cohens_d_games(gs_social, "gs_phase1", "condition", "social_first", "social_mixed"))

    p_h3a <- as.data.frame(summary(contrast_h3a, infer = c(TRUE, TRUE)))$p.value
    maybe_bayes_factor(p_h3a, "H3a social_first vs social_mixed", function() {
      bf_wls_contrast(gs_social, "gs_phase1", levels = SOCIAL_CONDITIONS, weights = "weight_p1", name = "H3a")
    })
  } else {
    cat("Insufficient data for H3a (need both social conditions).\n")
  }
}
```

```{r plot-h3a, fig.width=6, fig.height=5}
if (exists("gs_social") && nrow(gs_social) > 0) {
  ggplot(gs_social, aes(x = condition, y = gs_phase1, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.1, size = 3, alpha = 0.7) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    geom_hline(yintercept = 0, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Phase 1 group-specificity",
      title = "H3a: Phase 1 group-specificity by social condition"
    ) +
    theme(legend.position = "none")
}
```

## H3b: Within-group convergence rate

`within_group_similarity ~ condition * block + (block | group) + (block | tangram)`,
restricted to Phase 1, within-group pairs, social conditions only.
Group (not game) is the natural unit of replication for within-group similarity,
and condition is between-game so cannot have a random slope within game or group.

```{r h3b-convergence}
if (!has_block_pairwise || nrow(block_pairwise_sim) == 0) {
  cat("Skipping H3b: block_pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else {
  # Phase 1, within-group pairs, social conditions, game-qualified group id
  p1_within_social <- within_group_block_pairs(block_pairwise_sim, games, phase = 1, conditions = SOCIAL_CONDITIONS)

  if (nrow(p1_within_social) > 0 && n_distinct(p1_within_social$condition) >= 2) {
    # Group is the natural unit of replication for within-group similarity.
    # Condition is between-game, so it cannot have a random slope within group.
    m_h3b <- fit_progressively(
      similarity ~ condition * blockNum_c + (blockNum_c | group) + (blockNum_c | target),
      data = p1_within_social
    )
    print(summary(m_h3b))
    report_effect_sizes(m_h3b)
  } else {
    cat("Insufficient data for H3b convergence analysis.\n")
  }
}
```

```{r plot-h3b, fig.width=8, fig.height=5}
if (exists("p1_within_social") && nrow(p1_within_social) > 0) {
  conv_by_block <- p1_within_social |>
    filter(!is.na(similarity)) |>
    group_by(condition, blockNum) |>
    tidyboot_mean(similarity, nboot = 1000)

  ggplot(conv_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number (Phase 1)",
      y = "Within-group cosine similarity",
      color = "Condition",
      title = "H3b: Within-group convergence rate by social condition"
    )
}
```

## H3c: Description properties

Compare concreteness, lexical uniqueness, and word frequency of final Phase 1
descriptions between `social_mixed` and `social_first`.

All three measures are precomputed in Python (`compute_derived.py`) with
consistent tokenization: each utterance is lowercased, split into alphabetic
sequences, filtered for NLTK English stopwords, and only tokens with
length > 1 are retained as *content words*.

- **Concreteness**: the proportion of content words matching concrete term
  stems (prefix matching) from Boyce et al. 2024. Four categories: geometric
  (e.g., *triangle*, *square*, *diamond*), body parts (e.g., *arm*, *head*,
  *leg*), positional (e.g., *left*, *right*, *top*), and posture (e.g.,
  *kick*, *kneel*, *stand*). Prefix matching means *squar* matches both
  *square* and *squares*.
- **Lexical uniqueness**: the proportion of content words in a description
  that do not appear in any other group's descriptions for the same tangram
  across games.
- **Word frequency**: the mean Zipf frequency of content words, computed via
  the `wordfreq` Python package (which draws on SUBTLEX and other corpora).
  Common words score higher (e.g., "the" ≈ 7.7); rare words score lower.

```{r h3c-description-properties}
if (!has_rows(desc_props) || !has_rows(lex_uniq)) {
  cat("Missing H3c data files. Run compute_description_properties() and compute_lexical_uniqueness() first.\n")
} else {
  # Each speaker's final Phase 1 description of each tangram, social conditions
  p1_final_social <- final_phase1_properties(desc_props, lex_uniq, games, SOCIAL_CONDITIONS)

  if (nrow(p1_final_social) > 0 && n_distinct(p1_final_social$condition) >= 2) {
    # Concreteness
    p1_concrete <- p1_final_social |> filter(!is.na(concreteness))
    if (nrow(p1_concrete) > 0) {
      cat("=== Concreteness ===\n")
      m_concrete <- fit_progressively(
        concreteness ~ condition + (1 | playerId) + (condition | target),
        data = p1_concrete
      )
      print(summary(m_concrete))
      report_effect_sizes(m_concrete)
    }

    # Lexical uniqueness
    p1_unique <- p1_final_social |> filter(!is.na(uniqueness))
    if (nrow(p1_unique) > 0) {
      cat("\n=== Lexical Uniqueness ===\n")
      m_uniqueness <- fit_progressively(
        uniqueness ~ condition + (1 | playerId) + (condition | target),
        data = p1_unique
      )
      print(summary(m_uniqueness))
      report_effect_sizes(m_uniqueness)
    }

    # Word frequency
    p1_freq <- p1_final_social |> filter(!is.na(mean_zipf_freq))
    if (nrow(p1_freq) > 0) {
      cat("\n=== Word Frequency ===\n")
      m_freq <- fit_progressively(
        mean_zipf_freq ~ condition + (1 | playerId) + (condition | target),
        data = p1_freq
      )
      print(summary(m_freq))
      report_effect_sizes(m_freq)
    }
  } else {
    cat("Insufficient data for H3c (need both social conditions with Phase 1 utterances).\n")
  }
}
```

```{r plot-h3c, fig.width=14, fig.height=4.5}
if (exists("p1_final_social") && "concreteness" %in% names(p1_final_social)) {
  p1_h3c <- p1_final_social |>
    filter(!is.na(concreteness), !is.na(uniqueness))

  p_concrete <- ggplot(p1_h3c, aes(x = condition, y = concreteness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Concrete terms") +
    theme(legend.position = "none")

  p_unique <- ggplot(p1_h3c, aes(x = condition, y = uniqueness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Proportion", title = "Lexical uniqueness") +
    theme(legend.position = "none")

  p_freq <- ggplot(p1_h3c |> filter(!is.na(mean_zipf_freq)),
                   aes(x = condition, y = mean_zipf_freq, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.2, alpha = 0.3) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = NULL, y = "Mean Zipf frequency", title = "Word frequency") +
    theme(legend.position = "none")

  fig_h3c <- patchwork::wrap_plots(p_concrete, p_unique, p_freq) +
    plot_annotation(
      title = "H3c: Description properties in final Phase 1 utterances",
      tag_levels = "a"
    ) &
    TAG_THEME

  print(fig_h3c)
}
```

# H4: Downstream trade-off (social-first vs social-mixed)

Restricted to `social_mixed` vs `social_first` in Phase 2. Tests whether
anticipated social goals produce a trade-off: higher social accuracy but
lower referential accuracy. Contingent on H3a.

## H4a: Social-identification accuracy

`glmer(social_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: social-identification accuracy depends on
both the listener's ability to detect cues and the speaker's tendency to produce
identifiable language.

Contrast: `social_first > social_mixed`.

```{r h4a-social-accuracy}
if (nrow(social_guesses) > 0) {
  # speakerId comes from preprocessing, attributed by (roundId, currentGroup):
  # a roundId is shared across all groups in a game, so joining on
  # (gameId, blockNum, target) would match three speakers (one per group)
  # and, because blockNum resets per phase, Phase 1 speakers too
  social_h4 <- social_guess_trials(social_guesses, games, SOCIAL_CONDITIONS)

  if (n_distinct(social_h4$condition) >= 2 && nrow(social_h4) > 0) {
    m_h4a <- fit_progressively(
      correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
      data = social_h4,
      family = binomial
    )
    print(summary(m_h4a))
    report_effect_sizes(m_h4a)

    if (inherits(m_h4a, "glmerMod")) {
      em_h4a <- emmeans(m_h4a, "condition", type = "response")
      contrast_h4a <- contrast(em_h4a, list(
        H4a_socialfirst_gt_socialmixed = c(-1, 1)
      ))
      print(summary(contrast_h4a, infer = c(TRUE, TRUE)))

      p_h4a <- as.data.frame(summary(contrast_h4a, infer = c(TRUE, TRUE)))$p.value
      # Same formula that converged above; the null drops the fixed effect of condition
      maybe_bayes_factor(p_h4a, "H4a social_first vs social_mixed", function() {
        bf_glmm_condition(social_h4, formula(m_h4a), levels = SOCIAL_CONDITIONS, name = "H4a")
      })
    }
  } else {
    cat("Insufficient data for H4a (need both social conditions).\n")
  }
} else {
  cat("No social guess data for H4a.\n")
}
```

```{r plot-h4a, fig.width=6, fig.height=5}
if (exists("social_h4") && nrow(social_h4) > 0) {
  h4a_summary <- social_h4 |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4a_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct social guesses",
      title = "H4a: Social-identification accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

## H4b: Referential accuracy

`glmer(ref_correct ~ condition + block_num + (block_num | game) + (block_num | listener) + (block_num | speaker) + (condition | tangram), family = binomial)`

Crossed listener and speaker random effects: referential accuracy depends on both the
listener's interpretation and the quality of the speaker's description.

Contrast: `social_mixed > social_first` (lower ref accuracy expected in social_first).

```{r h4b-ref-accuracy}
# One speaker per (game, round, group): the group key is required because a
# roundId is shared across all groups in a game
p2_listener_social <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS) |>
  attach_speaker(trials)

if (n_distinct(p2_listener_social$condition) >= 2 && nrow(p2_listener_social) > 0) {
  m_h4b <- fit_progressively(
    correct ~ condition + blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) + (condition | target),
    data = p2_listener_social,
    family = binomial
  )
  print(summary(m_h4b))
  report_effect_sizes(m_h4b)

  if (inherits(m_h4b, "glmerMod")) {
    em_h4b <- emmeans(m_h4b, "condition", type = "response")
    contrast_h4b <- contrast(em_h4b, list(
      H4b_socialmixed_gt_socialfirst = c(1, -1)
    ))
    print(summary(contrast_h4b, infer = c(TRUE, TRUE)))

    p_h4b <- as.data.frame(summary(contrast_h4b, infer = c(TRUE, TRUE)))$p.value
    maybe_bayes_factor(p_h4b, "H4b social_mixed vs social_first", function() {
      bf_glmm_condition(p2_listener_social, formula(m_h4b), levels = SOCIAL_CONDITIONS, name = "H4b")
    })
  }
} else {
  cat("Insufficient data for H4b (need both social conditions with Phase 2 listener trials).\n")
}
```

```{r plot-h4b, fig.width=6, fig.height=5}
if (exists("p2_listener_social") && nrow(p2_listener_social) > 0) {
  h4b_summary <- p2_listener_social |>
    filter(!is.na(correct)) |>
    group_by(condition) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(h4b_summary, aes(x = condition, y = empirical_stat, fill = condition)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    scale_fill_condition() +
    labs(
      x = "Condition",
      y = "Proportion correct",
      title = "H4b: Referential accuracy by social condition (Phase 2)"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```


# Robustness: Phase 2 results on a subset of games

The preregistration commits to two robustness checks for the Phase 2 analyses
(H1, H2, and H4): restricting to complete nine-player games, and restricting to
games whose Phase 1 group specificity exceeded chance in the permutation test.
`robustness_rerun()` (analysis/R/contrasts.R) refits the H1/H2 weighted
regression and the H4 models on a subset of games and prints the planned
contrasts next to the primary ones. If the restriction excludes no games, the
results are identical to the primary analyses by construction and nothing is
refit.

```{r robustness-helper}
# The H4 models to rerun, with their planned contrast weights
h4_specs <- list(
  list(name = "H4a", data = if (exists("social_h4")) social_h4 else NULL,
       model = if (exists("m_h4a")) m_h4a else NULL,
       weights = list(H4a_socialfirst_gt_socialmixed = c(-1, 1))),
  list(name = "H4b", data = if (exists("p2_listener_social")) p2_listener_social else NULL,
       model = if (exists("m_h4b")) m_h4b else NULL,
       weights = list(H4b_socialmixed_gt_socialfirst = c(1, -1)))
)
```

## Games with significant Phase 1 group specificity

The permutation test is the one run for the outcome-neutral criterion
(`01_outcome_neutral.qmd`); its cached results are reused when the pairwise
similarities are unchanged.

```{r robustness-significant-gs}
if (has_pairwise && exists("gs_combined")) {
  perm_p1 <- compute_group_specificity(pairwise_sim |> filter(window == "phase1_final"), derived_dir)$perm_results
  significant_games <- perm_p1 |> filter(p_value < 0.05) |> pull(gameId)
  cat(sprintf("%d of %d games have significant Phase 1 group specificity (permutation p < .05).\n",
              length(significant_games), nrow(perm_p1)))
  robustness_rerun(significant_games, "games with significant Phase 1 group specificity",
                   gs = gs_combined, primary_contrasts = contrasts, h4 = h4_specs)
} else {
  cat("Skipping: no pairwise similarities.\n")
}
```

## Description length as a covariate

The paper commits to this check if description length differs across
conditions in a way that tracks group specificity: each game's similarity model
gets the pair's length difference and mean length as covariates, and the H1/H2
regression is rerun on the adjusted estimates. The diagnostic and the rerun are
both shown so the decision is visible.

```{r robustness-length-covariate}
if (!has_pairwise || !all(c("length1", "length2") %in% names(pairwise_sim))) {
  cat("Skipping: pairwise_similarities.csv has no length1/length2 columns (regenerate with compute_derived.py).\n")
} else if (!exists("gs_combined")) {
  cat("Skipping: no group-specificity estimates.\n")
} else {
  # Diagnostic: does mean Phase 2 description length differ by condition and track gs_phase2?
  length_by_game <- speaker_utts |>
    filter(phaseNum == 2) |>
    group_by(gameId) |>
    summarise(mean_length_p2 = mean(uttLength, na.rm = TRUE), .groups = "drop")
  diag <- gs_combined |> left_join(length_by_game, by = "gameId")
  cat("Mean Phase 2 description length by condition:\n")
  print(knitr::kable(diag |> group_by(condition) |>
                       summarise(games = n(), mean_length = mean(mean_length_p2), .groups = "drop"), digits = 2))
  if (nrow(diag) >= 4) {
    r <- cor(diag$mean_length_p2, diag$gs_phase2, use = "complete.obs")
    cat(sprintf("\nGame-level correlation between mean Phase 2 length and Phase 2 group specificity: r = %.2f\n", r))
  }

  # Group specificity with the pair's length difference and mean length as covariates
  pw_len <- pairwise_sim |>
    mutate(lengthDiff = abs(length1 - length2), meanLength = (length1 + length2) / 2)
  gs_adj <- gs_wide(game_specificity_table(pw_len, games, covariates = c("lengthDiff", "meanLength")))
  cat("\nGroup specificity with and without the length covariates:\n")
  print(knitr::kable(gs_combined |> select(gameId, condition, gs_phase1, gs_phase2) |>
                       inner_join(gs_adj |> select(gameId, gs_phase1_adj = gs_phase1, gs_phase2_adj = gs_phase2), by = "gameId") |>
                       mutate(gameId = substr(gameId, 1, 8)), digits = 3))

  h12_adj <- fit_h12_wls(gs_adj)
  if (!is.null(h12_adj$contrasts) && all(table(h12_adj$data$condition) >= 2)) {
    cat("\nH1/H2 with length-adjusted group specificity:\n")
    print(knitr::kable(bind_rows(contrast_table(contrasts, "primary"),
                                 contrast_table(h12_adj$contrasts, "length-adjusted")), digits = 3))
  } else {
    cat("\nH1/H2 rerun needs at least two games per condition; skipped.\n")
  }
}
```

## Complete nine-player games

The complete-games restriction from the paper's dropout-handling section
(`complete_games()` in `analysis/R/attrition.R`).

```{r robustness-complete-games}
if (exists("gs_combined")) {
  complete <- complete_games(players, games)
  cat(sprintf("%d of %d games kept all nine players.\n", length(complete), sum(!is.na(games$condition))))
  robustness_rerun(complete, "complete nine-player games",
                   gs = gs_combined, primary_contrasts = contrasts, h4 = h4_specs)
} else {
  cat("Skipping: no group-specificity estimates.\n")
}
```
