---
title: "03: Secondary analyses"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
library(emmeans)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_phase_change <- has_rows(phase_change)
has_pairwise <- has_rows(pairwise_sim)

# Real games only: test games have no condition
n_conditions <- n_distinct(games$condition, na.rm = TRUE)

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/secondary.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

# 1. Semantic change (Phase 1 → Phase 2)

Cosine similarity between final Phase 1 and final Phase 2 descriptions per participant × tangram.
Higher similarity = less semantic change.

**Predictions:** `refer_mixed` < `refer_separated`, `social_mixed` > `refer_mixed`

```{r model-semantic-change}
if (!has_phase_change || nrow(phase_change) == 0) {
  cat("Skipping: phase_change_similarities.csv not found. Run compute_derived.py first.\n")
} else {
phase_change <- phase_change |>
  mutate(condition = factor(condition, levels = CONDITION_ORDER)) |>
  add_group_id()

if (n_conditions >= 2) {
  # Paper formula: sim_phase1_phase2 ~ condition + (1 | game) + (1 | group)
  #   + (1 | participant) + (condition | tangram); the game and group
  #   intercepts are retained through any simplification
  m_semantic <- fit_progressively(
    simPhase1Phase2 ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (condition | target),
    data = phase_change
  )
  print(summary(m_semantic))
  report_effect_sizes(m_semantic)

  # Preregistered predictions: refer_separated > refer_mixed and
  # social_mixed > refer_mixed (higher similarity = less semantic change)
  if (all(c("refer_separated", "refer_mixed", "social_mixed") %in% phase_change$condition)) {
    em_semantic <- emmeans(m_semantic, "condition")
    lv <- levels(droplevels(phase_change$condition))
    contrasts_semantic <- contrast(em_semantic, list(
      sep_gt_mixed    = pairwise_weights(lv, "refer_separated", "refer_mixed"),
      social_gt_mixed = pairwise_weights(lv, "social_mixed", "refer_mixed")
    ))
    cat("\nPlanned contrasts (positive = predicted direction):\n")
    print(summary(contrasts_semantic, infer = c(TRUE, TRUE)))
  } else {
    cat("\nPlanned contrasts need refer_separated, refer_mixed, and social_mixed.\n")
  }

  # Baseline drift: refer_separated should be near ceiling
  drift <- phase_change |> filter(condition == "refer_separated", !is.na(simPhase1Phase2))
  if (nrow(drift) > 1) {
    drift_ci <- tidyboot_mean(drift, simPhase1Phase2, nboot = 1000)
    cat(sprintf("\nrefer_separated Phase 1 to Phase 2 similarity: M = %.3f, 95%% CI [%.3f, %.3f] (ceiling = 1)\n",
                drift_ci$empirical_stat, drift_ci$ci_lower, drift_ci$ci_upper))
  }
} else {
  # With only 1 condition, fit without condition predictor
  m_semantic <- lmer(
    simPhase1Phase2 ~ 1 + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target),
    data = phase_change,
    control = lmerControl(optimizer = "bobyqa")
  )
  cat(sprintf("Only %d condition(s). Intercept-only model:\n", n_conditions))
  print(summary(m_semantic))
}
} # end has_phase_change
```

```{r plot-semantic-change, fig.width=7, fig.height=5}
if (has_phase_change && nrow(phase_change) > 0) {
# Plot 10: Semantic change by condition
ggplot(phase_change, aes(x = condition, y = simPhase1Phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Cosine similarity (Phase 1 final → Phase 2 final)",
    title = "Semantic change across phases by condition",
    subtitle = "Higher = less change from Phase 1 to Phase 2"
  ) +
  theme(legend.position = "none")
}
```

# 2. Referential accuracy in Phase 2

**Predictions:** Above chance (1/16), main effect of `blockNum`, ordered: separated > mixed > social_mixed.

```{r prep-accuracy}
# Phase 2 listener trials with correct, condition, and centered block
p2_listener <- listener_trials(trials, games, phase = 2)
```

```{r model-accuracy}
n_games <- n_distinct(p2_listener$gameId)

if (n_conditions >= 2 && n_games >= 2) {
  # Paper formula: condition * blockNum slopes on tangram, blockNum on game/player
  m_ref_accuracy <- fit_progressively(
    correct ~ condition * blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (condition * blockNum_c | target),
    data = p2_listener,
    family = binomial
  )
} else if (n_games >= 2) {
  cat("Only 1 condition; fitting without condition predictor.\n")
  m_ref_accuracy <- fit_progressively(
    correct ~ blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | target),
    data = p2_listener,
    family = binomial
  )
} else {
  m_ref_accuracy <- glm(correct ~ blockNum_c, data = p2_listener, family = binomial)
  cat("Only 1 game; using fixed-effects logistic model.\n")
}

summary(m_ref_accuracy)
report_effect_sizes(m_ref_accuracy)
```

The three preregistered predictions: accuracy above chance (1/16) in every
condition, a positive main effect of block averaged over conditions, and the
ordering refer_separated > refer_mixed > social_mixed > social_first averaged
over blocks.

```{r accuracy-tests}
if (inherits(m_ref_accuracy, "glmerMod") && "condition" %in% all.vars(formula(m_ref_accuracy))) {
  em_acc <- emmeans(m_ref_accuracy, "condition")   # at the mean block (blockNum_c = 0)

  cat("Above chance? One-sided test of each condition's accuracy against 1/16 (logit scale):\n")
  print(test(em_acc, null = qlogis(1 / 16), side = ">"))
  cat("\nEstimated accuracy by condition (response scale):\n")
  print(summary(em_acc, type = "response"))

  cat("\nBlock slope averaged over conditions (log-odds per block), then per condition:\n")
  print(test(emtrends(m_ref_accuracy, ~ 1, var = "blockNum_c")))
  print(test(emtrends(m_ref_accuracy, ~ condition, var = "blockNum_c")))

  lv <- levels(droplevels(p2_listener$condition))
  ordering <- list()
  if (all(c("refer_separated", "refer_mixed") %in% lv)) ordering$sep_gt_mixed <- pairwise_weights(lv, "refer_separated", "refer_mixed")
  if (all(c("refer_mixed", "social_mixed") %in% lv)) ordering$mixed_gt_social <- pairwise_weights(lv, "refer_mixed", "social_mixed")
  if (all(c("social_mixed", "social_first") %in% lv)) ordering$social_gt_socialfirst <- pairwise_weights(lv, "social_mixed", "social_first")
  if (length(ordering)) {
    ordering_contrasts <- contrast(em_acc, ordering)
    cat("\nPredicted ordering, consecutive contrasts (positive = predicted direction, log-odds):\n")
    print(summary(ordering_contrasts, infer = c(TRUE, TRUE)))
  }
} else {
  cat("Preregistered accuracy tests need the condition-by-block mixed model.\n")
}
```

```{r plot-accuracy, fig.width=8, fig.height=5}
# Plot 11: Referential accuracy over Phase 2 blocks by condition
acc_by_block <- p2_listener |>
  filter(!is.na(correct)) |>
  group_by(condition, blockNum) |>
  tidyboot_mean(correct, nboot = 1000)

ggplot(acc_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  annotate("text", x = max(acc_by_block$blockNum), y = 1/16 + 0.02,
           label = "Chance (1/16)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block number",
    y = "Proportion correct",
    color = "Condition",
    title = "Referential accuracy in Phase 2"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_condition() +
  scale_fill_condition()
```

# 3. In-group vs out-group listener accuracy (social conditions)

Within `social_mixed` and `social_first`: do listeners understand in-group
speakers better, and is any H4b cost concentrated among out-group listeners?
The preregistration extends the H4b model with in-group status and its
interaction with condition, retaining the block term and the H4b
random-effects structure:

```
ref_correct ~ in_group_speaker * condition + block_num
  + (block_num | game) + (block_num | listener) + (block_num | speaker)
  + (condition | tangram)
```

The social-first minus social-mixed contrast is reported separately for
in-group and for out-group listeners (as odds ratios on the response scale,
with confidence intervals), regardless of the primary H4b result, together
with the interaction (the ratio of those two odds ratios). Significance in
one subgroup but not the other is not itself evidence for an interaction.

```{r model-ingroup}
# Phase 2 listener trials in the social conditions, each with its speaker and
# in-group status (attach_speaker() joins on the listener's current group, since
# a roundId is shared by all groups in a game). The condition factor keeps only
# the two social levels so the random slope has no empty-level columns.
social_listener <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS) |>
  attach_speaker(trials) |>
  filter(!is.na(inGroupSpeaker))

if (nrow(social_listener) > 0 && n_distinct(social_listener$condition) >= 2) {
  m_ingroup <- fit_progressively(
    correct ~ inGroupSpeaker * condition + blockNum_c +
      (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | speakerId) +
      (condition | target),
    data = social_listener,
    family = binomial
  )
  print(summary(m_ingroup))
  report_odds_ratios(m_ingroup)

  if (inherits(m_ingroup, "glmerMod")) {
    # social_first minus social_mixed within each listener type, as odds
    # ratios (emmeans back-transforms the log-odds contrast to the response
    # scale), and the interaction as the ratio of the two odds ratios.
    em_cond_by_group <- emmeans(m_ingroup, ~ condition | inGroupSpeaker, type = "response")
    ct_within <- contrast(em_cond_by_group, list(
      socialfirst_vs_socialmixed = pairwise_weights(SOCIAL_CONDITIONS, "social_first", "social_mixed")
    ))
    cat("\nSocial-first vs social-mixed within out-group and within in-group listeners (odds ratio):\n")
    print(summary(ct_within, infer = c(TRUE, TRUE)))
    ingroup_contrasts <- contrast_table(ct_within, "in-group status")

    cat("\nInteraction: ratio of the in-group to the out-group odds ratio (does social-first amplify the in-group advantage?):\n")
    em_full <- emmeans(m_ingroup, ~ condition * inGroupSpeaker, type = "response")
    ct_interaction <- contrast(em_full, interaction = c("revpairwise", "revpairwise"))
    print(summary(ct_interaction, infer = c(TRUE, TRUE)))

    # The in-group advantage itself, within each condition and overall
    em_group_by_cond <- emmeans(m_ingroup, ~ inGroupSpeaker | condition, type = "response")
    cat("\nIn-group vs out-group listeners within each condition (odds ratio):\n")
    print(summary(contrast(em_group_by_cond, "revpairwise"), infer = c(TRUE, TRUE)))
    cat("\nAveraged over conditions:\n")
    print(summary(contrast(emmeans(m_ingroup, ~ inGroupSpeaker, type = "response"), "revpairwise"),
                  infer = c(TRUE, TRUE)))
    cat("\nEstimated accuracy by condition and listener type (response scale, at the mean block):\n")
    print(summary(em_full))
  }
} else {
  cat("Need Phase 2 listener trials from both social conditions.\n")
}
```

```{r plot-ingroup-by-condition, fig.width=8, fig.height=5}
if (exists("social_listener") && nrow(social_listener) > 0 &&
    "inGroupSpeaker" %in% names(social_listener) &&
    n_distinct(social_listener$condition) >= 2) {
  ingroup_by_cond <- social_listener |>
    filter(!is.na(correct), !is.na(inGroupSpeaker)) |>
    mutate(speaker_type = ifelse(inGroupSpeaker == "in_group", "In-group", "Out-group")) |>
    group_by(condition, speaker_type) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(ingroup_by_cond, aes(x = speaker_type, y = empirical_stat, fill = speaker_type)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    facet_wrap(~condition) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    labs(
      x = "Speaker type",
      y = "Mean listener accuracy",
      title = "In-group vs out-group accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

# 4. Group-specificity vs social accuracy

Do games with higher Phase 2 social-identification accuracy also have more
group-specific expressions, after accounting for condition and Phase 1
group-specificity? The preregistered model is the ordinary least-squares
regression

```
group_specificity_phase2 ~ social_accuracy + condition + group_specificity_phase1
```

restricted to the two social conditions, with equal weight for each eligible
game (both endpoints estimable). The inverse-variance rerun (weights
`1 / SE_phase2^2`) is a robustness check. A positive social-accuracy
coefficient would indicate an association conditional on condition and
baseline group-specificity, not that distinctive language causes better
social identification.

```{r model-gs-social}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else if (nrow(social_guesses) > 0) {
  # Mean social accuracy per game over eligible response opportunities
  social_acc_by_game <- social_guess_trials(social_guesses, games) |>
    group_by(gameId) |>
    summarise(social_accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

  # The same per-game estimates as the primary analysis
  gs_social <- gs_wide(game_specificity_table(pairwise_sim, games)) |>
    inner_join(social_acc_by_game, by = "gameId")

  fit_gs_social <- function(weights = NULL) {
    fit_game_level(
      gs_social,
      outcome = "gs_phase2",
      predictors = c("social_accuracy", "gs_phase1"),
      conditions = SOCIAL_CONDITIONS,
      weights = weights
    )
  }
  gs_social_fit <- fit_gs_social()
  m_gs_social <- gs_social_fit$model
  if (!is.null(m_gs_social) && nrow(gs_social_fit$data) >= 4) {
    cat(sprintf("Games in the model: %d\n", nrow(gs_social_fit$data)))
    print(knitr::kable(gs_social_fit$data |> count(condition, name = "games")))
    print(summary(m_gs_social))
    report_effect_sizes(m_gs_social)
    sa_coef <- broom::tidy(m_gs_social, conf.int = TRUE) |>
      filter(term == "social_accuracy")
    cat(sprintf(
      "\nSocial accuracy coefficient (OLS): %.3f, 95%% CI [%.3f, %.3f], p = %s\n",
      sa_coef$estimate, sa_coef$conf.low, sa_coef$conf.high, fmt_pval(sa_coef$p.value)
    ))

    gs_social_wls <- fit_gs_social(weights = "weight")
    if (!is.null(gs_social_wls$model)) {
      sa_coef_w <- broom::tidy(gs_social_wls$model, conf.int = TRUE) |>
        filter(term == "social_accuracy")
      cat(sprintf(
        "Robustness, inverse-variance weights (1 / SE_phase2^2): %.3f, 95%% CI [%.3f, %.3f], p = %s\n",
        sa_coef_w$estimate, sa_coef_w$conf.low, sa_coef_w$conf.high, fmt_pval(sa_coef_w$p.value)
      ))
    }
  } else {
    cat("Not enough games in the two social conditions for the regression",
        if (!is.null(gs_social_fit$note)) paste0("(", gs_social_fit$note, ")"), "\n")
  }
} else {
  cat("No social guess data available.\n")
}
```

```{r plot-gs-social, fig.width=7, fig.height=5}
# Plot 13: Social accuracy vs group-specificity scatter
if (exists("gs_social_fit") && nrow(gs_social_fit$data) >= 3) {
  ggplot(gs_social_fit$data, aes(x = social_accuracy, y = gs_phase2, color = condition)) +
    geom_point(size = 3) +
    geom_smooth(aes(group = 1), method = "lm", se = TRUE, alpha = 0.2, color = "gray40") +
    scale_color_condition() +
    labs(
      x = "Mean social-identification accuracy (Phase 2)",
      y = "Phase 2 group-specificity",
      color = "Condition",
      title = "Social accuracy vs group-specificity (social conditions)"
    )
} else {
  cat("Not enough data points for scatter plot.\n")
}
```

# 5. Social-signaling accuracy over repetitions

Does social identification improve over blocks?

```{r model-social-accuracy}
if (nrow(social_guesses) > 0) {
  social_guesses_model <- social_guess_trials(social_guesses, games)

  n_social_games <- n_distinct(social_guesses_model$gameId)
  if (n_social_games >= 2) {
    m_social_acc <- fit_progressively(
      correct ~ blockNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
      data = social_guesses_model,
      family = binomial
    )
  } else {
    m_social_acc <- glm(correct ~ blockNum_c, data = social_guesses_model, family = binomial)
    cat("Only 1 game; using fixed-effects logistic model.\n")
  }

  print(summary(m_social_acc))
  report_effect_sizes(m_social_acc)

  cat(sprintf("\nOverall social accuracy: %.1f%%\n",
              mean(social_guesses_model$correct, na.rm = TRUE) * 100))
  cat(sprintf("Chance level: 50%%\n"))
} else {
  cat("No social guess data.\n")
}
```

```{r plot-social-accuracy, fig.width=7, fig.height=4}
# Plot 14: Social-signaling accuracy over blocks (pooled), over eligible
# response opportunities (code_accuracy() via social_guess_trials())
if (nrow(social_guesses) > 0) {
  social_by_block <- social_guess_trials(social_guesses, games) |>
    filter(!is.na(correct)) |>
    group_by(blockNum) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(social_by_block, aes(x = blockNum, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                alpha = 0.2, fill = "black") +
    geom_line(color = "black", linewidth = 1) +
    geom_point(color = "black", size = 2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    annotate("text", x = max(social_by_block$blockNum), y = 0.52,
             label = "Chance (50%)", hjust = 1, color = "gray50", size = 3) +
    labs(
      x = "Block number",
      y = "Proportion correct social guesses",
      title = "Social-signaling accuracy over Phase 2 blocks"
    ) +
    scale_y_continuous(limits = c(0, 1))
}
```

```{r plot-social-accuracy-by-condition, fig.width=8, fig.height=5}
# Social accuracy trajectories by condition (paper: compare across social
# conditions), over eligible response opportunities
if (nrow(social_guesses) > 0) {
  social_with_cond <- social_guess_trials(social_guesses, games, SOCIAL_CONDITIONS)

  if (n_distinct(social_with_cond$condition) >= 2) {
    social_by_block_cond <- social_with_cond |>
      filter(!is.na(correct)) |>
      group_by(condition, blockNum) |>
      tidyboot_mean(correct, nboot = 1000)

    ggplot(social_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                     color = condition)) +
      geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                  alpha = 0.15, color = NA) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
      scale_color_condition() +
      scale_fill_condition() +
      labs(
        x = "Block number",
        y = "Proportion correct social guesses",
        color = "Condition",
        title = "Social-signaling accuracy trajectories by condition"
      ) +
      scale_y_continuous(limits = c(0, 1))
  }
}
```

# 6. Description length in Phase 2

How does description length evolve across Phase 2 blocks by condition?

Paper formula:
`utt_length ~ condition * block_num + (1 | game) + (1 | group) + (1 | participant) + (1 | tangram)`,
with `group` the game-unique original group.

```{r model-utt-length-phase2}
p2_utts <- phase2_utterances(speaker_utts, games)

n_cond <- n_distinct(p2_utts$condition)
if (n_cond >= 2) {
  m_p2_length <- fit_progressively(
    uttLength ~ condition * blockNum_c + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target),
    data = p2_utts
  )
} else {
  m_p2_length <- fit_progressively(
    uttLength ~ blockNum_c + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target),
    data = p2_utts
  )
  cat("Only 1 condition; fitting without condition predictor.\n")
}

summary(m_p2_length)
report_effect_sizes(m_p2_length)

if (n_cond >= 2 && inherits(m_p2_length, "merMod")) {
  # The paper's predictions are per condition: continued shortening or a
  # plateau in refer_separated, a possible initial lengthening in refer_mixed,
  # short descriptions in the social conditions.
  cat("\nBlock slope (words per block) by condition:\n")
  p2_length_slopes <- as.data.frame(test(emtrends(m_p2_length, ~ condition, var = "blockNum_c")))
  print(p2_length_slopes)
  cat("\nMean Phase 2 description length by condition (at the mean block):\n")
  print(summary(emmeans(m_p2_length, "condition")))
}
```

```{r plot-utt-length-phase2, fig.width=8, fig.height=5}
if (nrow(p2_utts) > 0) {
  length_by_block_cond <- p2_utts |>
    group_by(condition, blockNum) |>
    tidyboot_mean(uttLength, nboot = 1000)

  ggplot(length_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                   color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number",
      y = "Mean word count",
      color = "Condition",
      title = "Description length over Phase 2 blocks by condition"
    )
}
```

# 7. Utterance properties across the four conditions

The paper's "Utterance properties" paragraph: the H3c measures (concreteness,
lexical uniqueness, word frequency; see `compute_derived.py`) applied to all
four conditions, to explore whether group-specific language maintained
passively (refer_separated) differs from language maintained under a
social-signaling goal (social_mixed). Each measure is modeled on each speaker's
final description of each tangram within a phase, with the H3c formula
`measure ~ condition + (1 | game) + (1 | group) + (1 | participant) + (condition | tangram)`
(`group` is the game-unique original group). The named contrast is
refer_separated minus social_mixed; all pairwise comparisons follow
(Tukey-adjusted). Exploratory.

```{r utterance-properties-all-conditions}
PROPERTY_LABELS <- c(concreteness = "Concreteness", uniqueness = "Lexical uniqueness",
                     mean_zipf_freq = "Word frequency (Zipf)")

if (!has_rows(desc_props) || !has_rows(lex_uniq)) {
  cat("Skipping: description_properties.csv or lexical_uniqueness.csv not found.\n")
} else {
  property_summaries <- list()
  for (ph in c(1, 2)) {
    props <- final_phase_properties(desc_props, lex_uniq, games, phase = ph)
    if (nrow(props) == 0 || n_distinct(props$condition) < 2) {
      cat(sprintf("Phase %d: not enough conditions with data.\n", ph))
      next
    }
    for (measure in names(PROPERTY_LABELS)) {
      d <- props |> filter(!is.na(.data[[measure]]))
      if (nrow(d) == 0) next
      cat(sprintf("\n=== Phase %d: %s ===\n", ph, PROPERTY_LABELS[[measure]]))
      f <- as.formula(sprintf(
        "%s ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (condition | target)", measure
      ))
      m <- fit_progressively(f, data = d)
      em <- emmeans(m, "condition")
      property_summaries[[length(property_summaries) + 1]] <- as.data.frame(summary(em)) |>
        mutate(phase = paste("Phase", ph), measure = PROPERTY_LABELS[[measure]])
      lv <- levels(droplevels(d$condition))
      if (all(c("refer_separated", "social_mixed") %in% lv)) {
        cat("Passive vs active maintenance (refer_separated minus social_mixed):\n")
        print(summary(contrast(em, list(passive_minus_active = pairwise_weights(lv, "refer_separated", "social_mixed"))),
                      infer = c(TRUE, TRUE)))
      }
      cat("All pairwise comparisons (Tukey):\n")
      print(summary(pairs(em)))
    }
  }
  if (length(property_summaries)) {
    property_plot <- bind_rows(property_summaries)
    print(
      ggplot(property_plot, aes(x = condition, y = emmean, color = condition)) +
        geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
        facet_grid(measure ~ phase, scales = "free_y") +
        scale_color_condition() +
        labs(x = "Condition", y = "Estimated marginal mean (95% CI)") +
        theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
    )
  }
}
```

# Statistics for the manuscript

Every contrast, coefficient, confidence interval, and p-value reported above
goes into `writing/manuscript/stats/secondary.tex` as a `\newcommand` macro,
with names prefixed `secondary`. A comparison the dataset cannot support
contributes no macro, so an unavailable number fails loudly in LaTeX rather
than reading as a result. The secondary analyses report frequentist results
only; Bayes factors are preregistered for the primary hypotheses.

```{r write-stats}
# emmeans labels an interaction contrast with one column per factor rather
# than a single `contrast` column, so pull the estimate, interval, and p-value
# by position and accept whichever names the back-transform produced.
emm_macros <- function(prefix, obj, row = 1L, digits = 3) {
  if (is.null(obj)) return(list())
  tab <- as.data.frame(summary(obj, infer = c(TRUE, TRUE)))
  if (nrow(tab) < row) return(list())
  pick <- function(candidates) {
    hit <- intersect(candidates, names(tab))
    if (length(hit) == 0) return(NA_real_)
    tab[[hit[1]]][row]
  }
  setNames(
    list(
      fmt_stat(pick(c("estimate", "odds.ratio", "ratio")), digits),
      fmt_stat(pick(c("lower.CL", "asymp.LCL")), digits),
      fmt_stat(pick(c("upper.CL", "asymp.UCL")), digits),
      fmt_pval(pick("p.value"))
    ),
    paste0(prefix, c("Est", "Lo", "Hi", "P"))
  )
}

if (exists("contrasts_semantic")) {
  macros <- c(
    macros,
    contrast_macros("secondarySemanticSepVsMixed", contrasts_semantic, row = 1),
    contrast_macros("secondarySemanticSocialVsMixed", contrasts_semantic, row = 2)
  )
}
if (exists("drift_ci")) {
  macros <- c(macros, list(
    secondaryDriftMean = fmt_stat(drift_ci$empirical_stat),
    secondaryDriftLo = fmt_stat(drift_ci$ci_lower),
    secondaryDriftHi = fmt_stat(drift_ci$ci_upper)
  ))
}

if (exists("ordering_contrasts")) {
  # contrast_table() gives the `lower`/`upper` names contrast_macros() expects
  ord <- contrast_table(ordering_contrasts, "")
  prefixes <- c(sep_gt_mixed = "secondaryAccSepVsMixed",
                mixed_gt_social = "secondaryAccMixedVsSocial",
                social_gt_socialfirst = "secondaryAccSocialVsSocialFirst")
  for (i in seq_len(nrow(ord))) {
    prefix <- prefixes[as.character(ord$contrast[i])]
    if (is.na(prefix)) next
    macros <- c(macros, contrast_macros(unname(prefix), ord, row = i))
  }
}
if (exists("m_ref_accuracy")) {
  macros <- c(macros, coef_macros("secondaryAccBlock", m_ref_accuracy, "blockNum_c"))
}

if (exists("ct_within")) {
  within_tab <- as.data.frame(summary(ct_within, infer = c(TRUE, TRUE)))
  # One row per in-group status; name the macros by the status rather than the row
  for (i in seq_len(nrow(within_tab))) {
    status <- as.character(within_tab$inGroupSpeaker[i])
    prefix <- if (identical(status, "in_group")) {
      "secondaryIngroupWithinIn"
    } else if (identical(status, "out_group")) {
      "secondaryIngroupWithinOut"
    } else {
      NULL
    }
    if (is.null(prefix)) next
    macros <- c(macros, emm_macros(prefix, ct_within, row = i))
  }
}
if (exists("ct_interaction")) {
  macros <- c(macros, emm_macros("secondaryIngroupInteraction", ct_interaction, row = 1))
}

if (exists("sa_coef") && nrow(sa_coef) == 1) {
  macros <- c(macros, list(
    secondaryGsSocialEst = fmt_stat(sa_coef$estimate),
    secondaryGsSocialLo = fmt_stat(sa_coef$conf.low),
    secondaryGsSocialHi = fmt_stat(sa_coef$conf.high),
    secondaryGsSocialP = fmt_pval(sa_coef$p.value),
    secondaryGsSocialNGames = nrow(gs_social_fit$data)
  ))
}
if (exists("sa_coef_w") && nrow(sa_coef_w) == 1) {
  macros <- c(macros, list(
    secondaryGsSocialWlsEst = fmt_stat(sa_coef_w$estimate),
    secondaryGsSocialWlsLo = fmt_stat(sa_coef_w$conf.low),
    secondaryGsSocialWlsHi = fmt_stat(sa_coef_w$conf.high),
    secondaryGsSocialWlsP = fmt_pval(sa_coef_w$p.value)
  ))
}

if (exists("m_social_acc")) {
  macros <- c(macros, coef_macros("secondarySocialAccBlock", m_social_acc, "blockNum_c"))
}
if (exists("social_guesses_model") && has_rows(social_guesses_model)) {
  macros <- c(macros, list(
    secondarySocialAccMean = fmt_stat(mean(social_guesses_model$correct, na.rm = TRUE)),
    secondarySocialAccN = fmt_count(sum(!is.na(social_guesses_model$correct)))
  ))
}

if (exists("m_p2_length")) {
  macros <- c(macros, coef_macros("secondaryLengthBlock", m_p2_length, "blockNum_c"))
}

write_stats_tex(macros, stats_file("secondary"), source = "analysis/03_secondary_analysis.qmd")
```
