---
title: "03: Secondary Analyses"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
source(here::here("analysis", "config.R"))

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, "speaker_utterances.csv"), show_col_types = FALSE)
social_guesses <- read_csv(file.path(data_dir, "social_guesses.csv"), show_col_types = FALSE)

has_phase_change <- file.exists(file.path(data_dir, "phase_change_similarities.csv"))
has_pairwise <- file.exists(file.path(data_dir, "pairwise_similarities.csv"))

phase_change <- if (has_phase_change) read_csv(file.path(data_dir, "phase_change_similarities.csv"), show_col_types = FALSE) else tibble()
pairwise_sim <- if (has_pairwise) read_csv(file.path(data_dir, "pairwise_similarities.csv"), show_col_types = FALSE) else tibble()

n_conditions <- n_distinct(games$condition)
```

# 1. Semantic Change (Phase 1 → Phase 2)

Cosine similarity between final Phase 1 and final Phase 2 descriptions per participant × tangram.
Higher similarity = less semantic change.

**Predictions:** `refer_mixed` < `refer_separated`, `social_mixed` > `refer_mixed`

```{r model-semantic-change}
if (!has_phase_change || nrow(phase_change) == 0) {
  cat("Skipping: phase_change_similarities.csv not found. Run compute_embeddings.py first.\n")
} else {
phase_change <- phase_change %>%
  mutate(condition = factor(condition, levels = CONDITION_ORDER))

if (n_conditions >= 2) {
  m_semantic <- lmer(
    simPhase1Phase2 ~ condition + (1 | playerId) + (1 | target),
    data = phase_change,
    control = lmerControl(optimizer = "bobyqa")
  )
  summary(m_semantic)
} else {
  # With only 1 condition, fit without condition predictor
  m_semantic <- lmer(
    simPhase1Phase2 ~ 1 + (1 | playerId) + (1 | target),
    data = phase_change,
    control = lmerControl(optimizer = "bobyqa")
  )
  cat(sprintf("Only %d condition(s). Intercept-only model:\n", n_conditions))
  summary(m_semantic)
}
} # end has_phase_change
```

```{r plot-semantic-change, fig.width=7, fig.height=5}
if (has_phase_change && nrow(phase_change) > 0) {
# Plot 10: Semantic change by condition
ggplot(phase_change, aes(x = condition, y = simPhase1Phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Cosine similarity (Phase 1 final → Phase 2 final)",
    title = "Semantic change across phases by condition",
    subtitle = "Higher = less change from Phase 1 to Phase 2"
  ) +
  theme(legend.position = "none")
}
```

# 2. Referential Accuracy in Phase 2

**Predictions:** Above chance (1/6), main effect of `repNum`, ordered: separated > mixed > social_mixed.

```{r prep-accuracy}
# Phase 2 listener accuracy — trial-level binary outcome
p2_listener <- trials %>%
  filter(phaseNum == 2, role == "listener") %>%
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    repNum = blockNum + 1
  ) %>%
  left_join(games %>% select(gameId, condition), by = "gameId") %>%
  mutate(
    repNum_c = repNum - mean(repNum, na.rm = TRUE),
    condition = factor(condition, levels = CONDITION_ORDER)
  )
```

```{r model-accuracy}
n_games <- n_distinct(p2_listener$gameId)

if (n_conditions >= 2 && n_games >= 2) {
  m_ref_accuracy <- tryCatch(
    glmer(
      correct ~ condition * repNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
      data = p2_listener,
      family = binomial,
      control = glmerControl(optimizer = "bobyqa")
    ),
    error = function(e) {
      cat("glmer failed:", e$message, "\nFalling back to lmer on proportions.\n")
      lmer(correct ~ condition * repNum_c + (1 | gameId),
           data = p2_listener, control = lmerControl(optimizer = "bobyqa"))
    }
  )
} else if (n_games >= 2) {
  m_ref_accuracy <- tryCatch(
    glmer(
      correct ~ repNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
      data = p2_listener,
      family = binomial,
      control = glmerControl(optimizer = "bobyqa")
    ),
    error = function(e) {
      cat("glmer failed:", e$message, "\nFalling back to lmer.\n")
      lmer(correct ~ repNum_c + (1 | gameId),
           data = p2_listener, control = lmerControl(optimizer = "bobyqa"))
    }
  )
  cat("Only 1 condition; fitting without condition predictor.\n")
} else {
  m_ref_accuracy <- glm(correct ~ repNum_c, data = p2_listener, family = binomial)
  cat("Only 1 game; using fixed-effects logistic model.\n")
}

summary(m_ref_accuracy)
```

```{r plot-accuracy, fig.width=8, fig.height=5}
# Plot 11: Referential accuracy over Phase 2 blocks by condition
acc_by_block <- p2_listener %>%
  filter(!is.na(correct)) %>%
  group_by(condition, blockNum) %>%
  tidyboot_mean(correct, nboot = 1000)

ggplot(acc_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1/6, linetype = "dashed", color = "gray50") +
  annotate("text", x = max(acc_by_block$blockNum), y = 1/6 + 0.02,
           label = "Chance", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block number",
    y = "Proportion correct",
    color = "Condition",
    title = "Referential accuracy in Phase 2"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_condition() +
  scale_fill_condition()
```

# 3. In-Group vs Out-Group Listener Accuracy (Social Conditions)

Within `social_mixed` and `social_first`: do listeners understand in-group speakers better?

```{r model-ingroup}
social_listener <- trials %>%
  filter(phaseNum == 2, role == "listener") %>%
  left_join(games %>% select(gameId, condition), by = "gameId") %>%
  filter(condition %in% c("social_mixed", "social_first")) %>%
  mutate(correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"))

if (nrow(social_listener) > 0) {
  # Determine in-group status: listener's original group matches speaker's original group
  speaker_groups <- trials %>%
    filter(role == "speaker") %>%
    select(gameId, roundId, playerId, originalGroup) %>%
    rename(speakerId = playerId, speakerGroup = originalGroup) %>%
    distinct()

  social_listener <- social_listener %>%
    left_join(speaker_groups, by = c("gameId", "roundId")) %>%
    mutate(inGroupSpeaker = as.numeric(originalGroup == speakerGroup))

  # Paper formula: ref_correct ~ in_group_speaker + (1 | speaker) + (1 | listener) + (1 | tangram)
  # Note: paper omits (1 | game) here; participant IDs are unique across games.
  m_ingroup <- tryCatch(
    glmer(
      correct ~ inGroupSpeaker + (1 | playerId) + (1 | speakerId) + (1 | target),
      data = social_listener,
      family = binomial,
      control = glmerControl(optimizer = "bobyqa")
    ),
    error = function(e) {
      cat("glmer failed (likely sparse data):", e$message, "\n")
      cat("Falling back to fixed-effects logistic model.\n")
      glm(correct ~ inGroupSpeaker, data = social_listener, family = binomial)
    }
  )

  summary(m_ingroup)
} else {
  cat("No social condition listener data available.\n")
}
```

```{r plot-ingroup, fig.width=6, fig.height=4}
# Plot 12: In-group vs out-group accuracy (social conditions)
if (exists("social_listener") && nrow(social_listener) > 0 &&
    "inGroupSpeaker" %in% names(social_listener)) {
  ingroup_acc <- social_listener %>%
    filter(!is.na(correct), !is.na(inGroupSpeaker)) %>%
    mutate(speaker_type = ifelse(inGroupSpeaker == 1, "In-group", "Out-group")) %>%
    group_by(speaker_type) %>%
    tidyboot_mean(correct, nboot = 1000)

  ggplot(ingroup_acc, aes(x = speaker_type, y = empirical_stat, fill = speaker_type)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    labs(
      x = "Speaker type",
      y = "Mean listener accuracy",
      title = "In-group vs out-group accuracy (social conditions)"
    ) +
    theme(legend.position = "none")
}
```

```{r plot-ingroup-by-condition, fig.width=8, fig.height=5}
if (exists("social_listener") && nrow(social_listener) > 0 &&
    "inGroupSpeaker" %in% names(social_listener) &&
    n_distinct(social_listener$condition) >= 2) {
  ingroup_by_cond <- social_listener %>%
    filter(!is.na(correct), !is.na(inGroupSpeaker)) %>%
    mutate(speaker_type = ifelse(inGroupSpeaker == 1, "In-group", "Out-group")) %>%
    group_by(condition, speaker_type) %>%
    tidyboot_mean(correct, nboot = 1000)

  ggplot(ingroup_by_cond, aes(x = speaker_type, y = empirical_stat, fill = speaker_type)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    facet_wrap(~condition) +
    geom_hline(yintercept = 1/6, linetype = "dashed", color = "gray50") +
    labs(
      x = "Speaker type",
      y = "Mean listener accuracy",
      title = "In-group vs out-group accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

# 4. Group-Specificity vs Social Accuracy Correlation

Does maintaining group-specific language correlate with social identification accuracy?

```{r model-gs-social}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_embeddings.py first.\n")
} else if (nrow(social_guesses) > 0) {
  # Compute mean social accuracy per game
  social_acc_by_game <- social_guesses %>%
    mutate(correct = as.numeric(
      socialGuessCorrect == TRUE | socialGuessCorrect == "true"
    )) %>%
    group_by(gameId) %>%
    summarise(social_accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

  # Get Phase 2 group-specificity per game
  gs_p2 <- pairwise_sim %>%
    filter(window == "phase2_final") %>%
    group_by(gameId) %>%
    group_modify(~ {
      if (nrow(.x) < 5 || n_distinct(.x$sameGroup) < 2) return(tibble())
      tryCatch({
        model <- lmer(
          similarity ~ sameGroup + (1 | target) + (1 | speaker1) + (1 | speaker2),
          data = .x, control = lmerControl(optimizer = "bobyqa")
        )
        tibble(gs_phase2 = coef(summary(model))["sameGroup", "Estimate"])
      }, error = function(e) tibble())
    }) %>%
    ungroup()

  # Get Phase 1 group-specificity
  gs_p1 <- pairwise_sim %>%
    filter(window == "phase1_final") %>%
    group_by(gameId) %>%
    group_modify(~ {
      if (nrow(.x) < 5 || n_distinct(.x$sameGroup) < 2) return(tibble())
      tryCatch({
        model <- lmer(
          similarity ~ sameGroup + (1 | target) + (1 | speaker1) + (1 | speaker2),
          data = .x, control = lmerControl(optimizer = "bobyqa")
        )
        tibble(gs_phase1 = coef(summary(model))["sameGroup", "Estimate"])
      }, error = function(e) tibble())
    }) %>%
    ungroup()

  gs_social <- gs_p2 %>%
    inner_join(gs_p1, by = "gameId") %>%
    inner_join(social_acc_by_game, by = "gameId")

  if (nrow(gs_social) >= 3) {
    m_gs_social <- lm(gs_phase2 ~ social_accuracy + gs_phase1, data = gs_social)
    summary(m_gs_social)
  } else {
    cat(sprintf("Not enough games (%d) for regression (need ≥3).\n", nrow(gs_social)))
  }
} else {
  cat("No social guess data available.\n")
}
```

```{r plot-gs-social, fig.width=6, fig.height=5}
# Plot 13: Social accuracy vs group-specificity scatter
if (exists("gs_social") && nrow(gs_social) >= 3) {
  ggplot(gs_social, aes(x = social_accuracy, y = gs_phase2)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
    labs(
      x = "Mean social accuracy",
      y = "Phase 2 group-specificity",
      title = "Social accuracy vs group-specificity (social_mixed)"
    ) +
    NULL
} else {
  cat("Not enough data points for scatter plot.\n")
}
```

# 5. Social-Signaling Accuracy Over Repetitions

Does social identification improve over blocks?

```{r model-social-accuracy}
if (nrow(social_guesses) > 0) {
  social_guesses_model <- social_guesses %>%
    mutate(
      correct = as.numeric(
        socialGuessCorrect == TRUE | socialGuessCorrect == "true"
      ),
      blockNum_c = blockNum - mean(blockNum, na.rm = TRUE)
    )

  n_social_games <- n_distinct(social_guesses_model$gameId)
  if (n_social_games >= 2) {
    # Paper: social_correct ~ rep_num + (1 | game) + (1 | participant) + (1 | tangram)
    m_social_acc <- tryCatch(
      glmer(
        correct ~ blockNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
        data = social_guesses_model,
        family = binomial,
        control = glmerControl(optimizer = "bobyqa")
      ),
      error = function(e) {
        cat("glmer failed:", e$message, "\nFalling back to lmer.\n")
        lmer(correct ~ blockNum_c + (1 | gameId),
             data = social_guesses_model, control = lmerControl(optimizer = "bobyqa"))
      }
    )
  } else {
    m_social_acc <- glm(correct ~ blockNum_c, data = social_guesses_model, family = binomial)
    cat("Only 1 game; using fixed-effects logistic model.\n")
  }

  summary(m_social_acc)

  cat(sprintf("\nOverall social accuracy: %.1f%%\n",
              mean(social_guesses_model$correct, na.rm = TRUE) * 100))
  cat(sprintf("Chance level: 50%%\n"))
} else {
  cat("No social guess data.\n")
}
```

```{r plot-social-accuracy, fig.width=7, fig.height=4}
# Plot 14: Social-signaling accuracy over blocks (pooled)
if (nrow(social_guesses) > 0) {
  social_by_block <- social_guesses %>%
    mutate(correct = as.numeric(
      socialGuessCorrect == TRUE | socialGuessCorrect == "true"
    )) %>%
    filter(!is.na(correct)) %>%
    group_by(blockNum) %>%
    tidyboot_mean(correct, nboot = 1000)

  ggplot(social_by_block, aes(x = blockNum, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                alpha = 0.2, fill = ACCENT$social) +
    geom_line(color = ACCENT$social, linewidth = 1) +
    geom_point(color = ACCENT$social, size = 2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    annotate("text", x = max(social_by_block$blockNum), y = 0.52,
             label = "Chance (50%)", hjust = 1, color = "gray50", size = 3) +
    labs(
      x = "Block number",
      y = "Proportion correct social guesses",
      title = "Social-signaling accuracy over Phase 2 blocks"
    ) +
    scale_y_continuous(limits = c(0, 1))
}
```

```{r plot-social-accuracy-by-condition, fig.width=8, fig.height=5}
# Social accuracy trajectories by condition (paper: compare across social conditions)
if (nrow(social_guesses) > 0) {
  social_with_cond <- social_guesses %>%
    left_join(games %>% select(gameId, condition), by = "gameId") %>%
    filter(condition %in% c("social_mixed", "social_first")) %>%
    mutate(correct = as.numeric(
      socialGuessCorrect == TRUE | socialGuessCorrect == "true"
    ))

  if (n_distinct(social_with_cond$condition) >= 2) {
    social_by_block_cond <- social_with_cond %>%
      filter(!is.na(correct)) %>%
      group_by(condition, blockNum) %>%
      tidyboot_mean(correct, nboot = 1000)

    ggplot(social_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                     color = condition)) +
      geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                  alpha = 0.15, color = NA) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
      scale_color_condition() +
      scale_fill_condition() +
      labs(
        x = "Block number",
        y = "Proportion correct social guesses",
        color = "Condition",
        title = "Social-signaling accuracy trajectories by condition"
      ) +
      scale_y_continuous(limits = c(0, 1))
  }
}
```

# 6. Description Length in Phase 2

How does description length evolve across Phase 2 blocks by condition?

```{r plot-utt-length-phase2, fig.width=8, fig.height=5}
p2_utts <- speaker_utts %>%
  filter(phaseNum == 2) %>%
  left_join(games %>% select(gameId, condition), by = "gameId") %>%
  mutate(condition = factor(condition, levels = CONDITION_ORDER))

if (nrow(p2_utts) > 0) {
  length_by_block_cond <- p2_utts %>%
    group_by(condition, blockNum) %>%
    tidyboot_mean(uttLength, nboot = 1000)

  ggplot(length_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                   color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number",
      y = "Mean word count",
      color = "Condition",
      title = "Description length over Phase 2 blocks by condition"
    )
}
```
