---
title: "03: Secondary analyses"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
library(emmeans)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_phase_change <- has_rows(phase_change)
has_pairwise <- has_rows(pairwise_sim)

# Real games only: test games have no condition
n_conditions <- n_distinct(games$condition, na.rm = TRUE)

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/secondary.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

# 1. Semantic change (Phase 1 → Phase 2)

Cosine similarity between final Phase 1 and final Phase 2 descriptions per participant × tangram.
Higher similarity = less semantic change.

**Predictions:** `refer_mixed` < `refer_separated`, `social_mixed` > `refer_mixed`

```{r model-semantic-change}
if (!has_phase_change || nrow(phase_change) == 0) {
  cat("Skipping: phase_change_similarities.csv not found. Run compute_derived.py first.\n")
} else {
phase_change <- phase_change |>
  mutate(condition = factor(condition, levels = CONDITION_ORDER)) |>
  add_group_id()

if (n_conditions >= 2) {
  # Paper formula: sim_phase1_phase2 ~ condition + (1 | game) + (1 | group)
  #   + (1 | participant) + (1 | tangram); the game and group
  #   intercepts are retained through any simplification
  m_semantic <- fit_progressively(
    simPhase1Phase2 ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target),
    data = phase_change
  )
  print(summary(m_semantic))
  report_effect_sizes(m_semantic)

  # Preregistered predictions: refer_separated > refer_mixed and
  # social_mixed > refer_mixed (higher similarity = less semantic change)
  if (all(c("refer_separated", "refer_mixed", "social_mixed") %in% phase_change$condition)) {
    em_semantic <- emmeans(m_semantic, "condition")
    lv <- levels(droplevels(phase_change$condition))
    contrasts_semantic <- contrast(em_semantic, list(
      sep_gt_mixed    = pairwise_weights(lv, "refer_separated", "refer_mixed"),
      social_gt_mixed = pairwise_weights(lv, "social_mixed", "refer_mixed")
    ))
    cat("\nPlanned contrasts (positive = predicted direction):\n")
    print(summary(contrasts_semantic, infer = c(TRUE, TRUE)))
  } else {
    cat("\nPlanned contrasts need refer_separated, refer_mixed, and social_mixed.\n")
  }

  # Baseline drift: refer_separated should be near ceiling
  drift <- phase_change |> filter(condition == "refer_separated", !is.na(simPhase1Phase2))
  if (nrow(drift) > 1) {
    drift_ci <- tidyboot_mean(drift, simPhase1Phase2, nboot = 1000)
    cat(sprintf("\nrefer_separated Phase 1 to Phase 2 similarity: M = %.3f, 95%% CI [%.3f, %.3f] (ceiling = 1)\n",
                drift_ci$empirical_stat, drift_ci$ci_lower, drift_ci$ci_upper))
  }
} else {
  # With only 1 condition, fit without condition predictor
  m_semantic <- lmer(
    simPhase1Phase2 ~ 1 + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target),
    data = phase_change,
    control = lmerControl(optimizer = "bobyqa")
  )
  cat(sprintf("Only %d condition(s). Intercept-only model:\n", n_conditions))
  print(summary(m_semantic))
}
} # end has_phase_change
```

```{r plot-semantic-change, fig.width=7, fig.height=5}
if (has_phase_change && nrow(phase_change) > 0) {
# Plot 10: Semantic change by condition
ggplot(phase_change, aes(x = condition, y = simPhase1Phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Cosine similarity (Phase 1 final → Phase 2 final)",
    title = "Semantic change across phases by condition",
    subtitle = "Higher = less change from Phase 1 to Phase 2"
  ) +
  theme(legend.position = "none")
}
```

# 2. Referential accuracy in Phase 2

**Predictions:** Above chance (1/16), main effect of `blockNum`, ordered: separated > mixed > social_mixed.

The item cell here is `gameId:target`, not `group:target` as in Phase 1 and in
the Phase 2 length model. Referential accuracy is a property of the *match*
between speaker and listener, and in the mixed conditions the listener is
usually not from the speaker's original group (in the pilot: 70/140 of
`refer_mixed` trials, 144/214 of `social_mixed`, 131/180 of `social_first`),
so what matters is whether the game as a whole has converged on that figure.
Fitting both on the pilot bears this out: `gameId:target` takes SD 0.61 and
`group:target` goes to 0.00 beside it. The length model goes the other way --
there `group:target` is 0.39 and `gameId:target` collapses -- because a
description's length reflects the speaker's own carried-over convention
rather than a match with anyone.

```{r prep-accuracy}
# Phase 2 listener trials with correct, condition, and centered block
p2_listener <- listener_trials(trials, games, phase = 2)
```

```{r model-accuracy}
n_games <- n_distinct(p2_listener$gameId)

if (n_conditions >= 2 && n_games >= 2) {
  # Paper formula: condition * blockNum slopes on tangram, blockNum on game/player
  m_ref_accuracy <- fit_progressively(
    correct ~ condition * blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (1 | target) + (1 | gameId:target),
    data = p2_listener,
    family = binomial
  )
} else if (n_games >= 2) {
  cat("Only 1 condition; fitting without condition predictor.\n")
  m_ref_accuracy <- fit_progressively(
    correct ~ blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (1 | target) + (1 | gameId:target),
    data = p2_listener,
    family = binomial
  )
} else {
  m_ref_accuracy <- glm(correct ~ blockNum_c, data = p2_listener, family = binomial)
  cat("Only 1 game; using fixed-effects logistic model.\n")
}

summary(m_ref_accuracy)
report_effect_sizes(m_ref_accuracy)
```

The three preregistered predictions: accuracy above chance (1/16) in every
condition, a positive main effect of block averaged over conditions, and the
ordering refer_separated > refer_mixed > social_mixed > social_first averaged
over blocks.

```{r accuracy-tests}
if (inherits(m_ref_accuracy, "glmerMod") && "condition" %in% all.vars(formula(m_ref_accuracy))) {
  em_acc <- emmeans(m_ref_accuracy, "condition")   # at the mean block (blockNum_c = 0)

  cat("Above chance? One-sided test of each condition's accuracy against 1/16 (logit scale):\n")
  print(test(em_acc, null = qlogis(1 / 16), side = ">"))
  cat("\nEstimated accuracy by condition (response scale):\n")
  print(summary(em_acc, type = "response"))

  cat("\nBlock slope averaged over conditions (log-odds per block), then per condition:\n")
  print(test(emtrends(m_ref_accuracy, ~ 1, var = "blockNum_c")))
  print(test(emtrends(m_ref_accuracy, ~ condition, var = "blockNum_c")))

  lv <- levels(droplevels(p2_listener$condition))
  ordering <- list()
  if (all(c("refer_separated", "refer_mixed") %in% lv)) ordering$sep_gt_mixed <- pairwise_weights(lv, "refer_separated", "refer_mixed")
  if (all(c("refer_mixed", "social_mixed") %in% lv)) ordering$mixed_gt_social <- pairwise_weights(lv, "refer_mixed", "social_mixed")
  if (all(c("social_mixed", "social_first") %in% lv)) ordering$social_gt_socialfirst <- pairwise_weights(lv, "social_mixed", "social_first")
  if (length(ordering)) {
    ordering_contrasts <- contrast(em_acc, ordering)
    cat("\nPredicted ordering, consecutive contrasts (positive = predicted direction, log-odds):\n")
    print(summary(ordering_contrasts, infer = c(TRUE, TRUE)))
  }
} else {
  cat("Preregistered accuracy tests need the condition-by-block mixed model.\n")
}
```

```{r plot-accuracy, fig.width=8, fig.height=5}
# Plot 11: Referential accuracy over Phase 2 blocks by condition
acc_by_block <- p2_listener |>
  filter(!is.na(correct)) |>
  group_by(condition, blockNum) |>
  tidyboot_mean(correct, nboot = 1000)

ggplot(acc_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  annotate("text", x = max(acc_by_block$blockNum), y = 1/16 + 0.02,
           label = "Chance (1/16)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block number",
    y = "Proportion correct",
    color = "Condition",
    title = "Referential accuracy in Phase 2"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_condition() +
  scale_fill_condition()
```

# 3. In-group vs out-group listener accuracy (social conditions)

Within `social_mixed` and `social_first`: do listeners understand in-group
speakers better, and is any H4b cost concentrated among out-group listeners?
The preregistration extends the H4b model with in-group status and its
interaction with condition, retaining the block term and the H4b
random-effects structure:

```
ref_correct ~ in_group_speaker * condition + block_num
  + (1 | game) + (1 | listener) + (1 | speaker) + (1 | tangram)
```

The two questions this section answers are estimated by two different models,
because they live at different levels.

**The in-group advantage itself, and whether condition amplifies it**, are
within-listener comparisons: the reshuffle rule gives each trio exactly one
in-group listener, so a listener is an in-group listener on some trials and an
out-group listener on others. By the rule used elsewhere in these notebooks
that calls for by-game and by-listener random slopes on `in_group_speaker`,
and a trial-level model without them is anticonservative -- simulated on this
design, a nominal .05 test runs near .26 for the in-group effect and .28 for
its interaction with condition. But those slopes cannot be estimated from
about a dozen Bernoulli trials per listener per cell; across simulated slope
SDs from 0.2 to 0.8 `fit_progressively()` dropped them every time. So these
two questions are answered by the paired form of the contrast instead: one
in-group advantage score per listener, modelled at the listener level
(`ingroup_advantage()` in `analysis/R/prepare.R`). The same simulations put
this at .083 and .058. Hawkins et al. (2020) use the same device for their
within- versus across-speaker comparison.

```
advantage ~ condition + (1 | game)
```

The intercept is the in-group advantage pooled over conditions; the condition
coefficient is the interaction, i.e. whether anticipating the social goal
amplifies it.

There is no trial-level model here. One would need by-listener and by-game
slopes on `inGroupSpeaker` to test those two quantities honestly, and they
cannot be estimated from about a dozen Bernoulli trials per listener per
cell -- across simulated slope SDs from 0.2 to 0.8 the ladder dropped them
every time, leaving an intercepts-only model that runs a nominal .05 test
near .26. In-group and out-group accuracy rates are reported descriptively
below instead.

```{r model-ingroup}
# Phase 2 listener trials in the social conditions, each with its speaker and
# in-group status (attach_speaker() joins on the listener's current group, since
# a roundId is shared by all groups in a game). The condition factor keeps only
# the two social levels so the random slope has no empty-level columns.
social_listener <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS) |>
  attach_speaker(trials) |>
  filter(!is.na(inGroupSpeaker))

if (nrow(social_listener) > 0 && n_distinct(social_listener$condition) >= 2) {
  # The in-group advantage and its interaction with condition, in paired
  # form: one score per listener, which collapses the within-listener
  # dimension into the outcome and leaves an estimable between-listener
  # model. See ingroup_advantage() in analysis/R/prepare.R.
  adv <- ingroup_advantage(social_listener)
  cat(sprintf(
    "In-group advantage scores: %d listeners in %d games (median %d in-group and %d out-group trials each)\n",
    nrow(adv), n_distinct(adv$gameId), median(adv$n_in), median(adv$n_out)
  ))
  if (nrow(adv) >= 4 && n_distinct(adv$condition) >= 2) {
    m_adv <- fit_progressively(advantage ~ condition + (1 | gameId), data = adv)
    print(summary(m_adv))
    ca <- coef(summary(m_adv))
    cat(sprintf(
      "\nIn-group advantage pooled over conditions: %+.3f (SE %.3f, p = %s)\n",
      ca["(Intercept)", "Estimate"], ca["(Intercept)", "Std. Error"],
      fmt_pval(ca["(Intercept)", "Pr(>|t|)"])
    ))
    cat("Does the anticipated social goal amplify it? (condition coefficient)\n")
    print(emmeans(m_adv, "condition"))
    print(summary(contrast(emmeans(m_adv, "condition"), list(
      socialfirst_minus_socialmixed = pairwise_weights(
        SOCIAL_CONDITIONS, "social_first", "social_mixed"
      )
    )), infer = c(TRUE, TRUE)))
    # Thin cells are noisier; rerun on listeners with at least five each.
    adv5 <- ingroup_advantage(social_listener, min_trials = 5L)
    if (nrow(adv5) >= 4 && n_distinct(adv5$condition) >= 2) {
      m_adv5 <- fit_progressively(advantage ~ condition + (1 | gameId), data = adv5,
                                  verbose = FALSE)
      cat(sprintf("\nSensitivity, cells of >= 5 trials (%d listeners): intercept %+.3f, p = %s\n",
                  nrow(adv5), coef(summary(m_adv5))["(Intercept)", "Estimate"],
                  fmt_pval(coef(summary(m_adv5))["(Intercept)", "Pr(>|t|)"])))
    }
  } else {
    cat("Too few listeners with both in-group and out-group trials for the advantage model.\n")
  }

  # Descriptive in-group and out-group accuracy rates, which the paired model
  # above does not print. Raw per-listener means aggregated by condition, so
  # they carry no distributional assumption; the test is the advantage model.
  cat("\nDescriptive accuracy by condition and listener type (per-listener means):\n")
  print(knitr::kable(
    adv |>
      tidyr::pivot_longer(c(acc_in, acc_out), names_to = "listener_type",
                          values_to = "accuracy") |>
      mutate(listener_type = ifelse(listener_type == "acc_in", "in-group", "out-group")) |>
      group_by(condition, listener_type) |>
      summarise(listeners = dplyr::n(), accuracy = mean(accuracy), .groups = "drop"),
    digits = 3
  ))
} else {
  cat("Need Phase 2 listener trials from both social conditions.\n")
}
```

```{r plot-ingroup-by-condition, fig.width=8, fig.height=5}
if (exists("social_listener") && nrow(social_listener) > 0 &&
    "inGroupSpeaker" %in% names(social_listener) &&
    n_distinct(social_listener$condition) >= 2) {
  ingroup_by_cond <- social_listener |>
    filter(!is.na(correct), !is.na(inGroupSpeaker)) |>
    mutate(speaker_type = ifelse(inGroupSpeaker == "in_group", "In-group", "Out-group")) |>
    group_by(condition, speaker_type) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(ingroup_by_cond, aes(x = speaker_type, y = empirical_stat, fill = speaker_type)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    facet_wrap(~condition) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    labs(
      x = "Speaker type",
      y = "Mean listener accuracy",
      title = "In-group vs out-group accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

# 4. Group-specificity vs social accuracy

Do games with higher Phase 2 social-identification accuracy also have more
group-specific expressions, after accounting for condition and Phase 1
group-specificity? The preregistered model is the ordinary least-squares
regression

```
group_specificity_phase2 ~ social_accuracy + condition + group_specificity_phase1
```

restricted to the two social conditions, with equal weight for each eligible
game (both endpoints estimable). The inverse-variance rerun (weights
`1 / SE_phase2^2`) is a robustness check. A positive social-accuracy
coefficient would indicate an association conditional on condition and
baseline group-specificity, not that distinctive language causes better
social identification.

```{r model-gs-social}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else if (nrow(social_guesses) > 0) {
  # Mean social accuracy per game over eligible response opportunities
  social_acc_by_game <- social_guess_trials(social_guesses, games) |>
    group_by(gameId) |>
    summarise(social_accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

  # The same per-game estimates as the primary analysis
  gs_social <- gs_wide(game_specificity_table(pairwise_sim, games)) |>
    inner_join(social_acc_by_game, by = "gameId")

  fit_gs_social <- function(weights = NULL) {
    fit_game_level(
      gs_social,
      outcome = "gs_phase2",
      predictors = c("social_accuracy", "gs_phase1"),
      conditions = SOCIAL_CONDITIONS,
      weights = weights,
      # No condition term: the analysis plan specifies this as an association
      # across the two social conditions, not a comparison between them.
      include_condition = FALSE
    )
  }
  gs_social_fit <- fit_gs_social()
  m_gs_social <- gs_social_fit$model
  if (!is.null(m_gs_social) && nrow(gs_social_fit$data) >= 4) {
    cat(sprintf("Games in the model: %d\n", nrow(gs_social_fit$data)))
    print(knitr::kable(gs_social_fit$data |> count(condition, name = "games")))
    print(summary(m_gs_social))
    report_effect_sizes(m_gs_social)
    sa_coef <- broom::tidy(m_gs_social, conf.int = TRUE) |>
      filter(term == "social_accuracy")
    cat(sprintf(
      "\nSocial accuracy coefficient (OLS): %.3f, 95%% CI [%.3f, %.3f], p = %s\n",
      sa_coef$estimate, sa_coef$conf.low, sa_coef$conf.high, fmt_pval(sa_coef$p.value)
    ))

    gs_social_wls <- fit_gs_social(weights = "weight")
    if (!is.null(gs_social_wls$model)) {
      sa_coef_w <- broom::tidy(gs_social_wls$model, conf.int = TRUE) |>
        filter(term == "social_accuracy")
      cat(sprintf(
        "Robustness, inverse-variance weights (1 / SE_phase2^2): %.3f, 95%% CI [%.3f, %.3f], p = %s\n",
        sa_coef_w$estimate, sa_coef_w$conf.low, sa_coef_w$conf.high, fmt_pval(sa_coef_w$p.value)
      ))
    }
  } else {
    cat("Not enough games in the two social conditions for the regression",
        if (!is.null(gs_social_fit$note)) paste0("(", gs_social_fit$note, ")"), "\n")
  }
} else {
  cat("No social guess data available.\n")
}
```

```{r plot-gs-social, fig.width=7, fig.height=5}
# Plot 13: Social accuracy vs group-specificity scatter
if (exists("gs_social_fit") && nrow(gs_social_fit$data) >= 3) {
  ggplot(gs_social_fit$data, aes(x = social_accuracy, y = gs_phase2, color = condition)) +
    geom_point(size = 3) +
    geom_smooth(aes(group = 1), method = "lm", se = TRUE, alpha = 0.2, color = "gray40") +
    scale_color_condition() +
    labs(
      x = "Mean social-identification accuracy (Phase 2)",
      y = "Phase 2 group-specificity",
      color = "Condition",
      title = "Social accuracy vs group-specificity (social conditions)"
    )
} else {
  cat("Not enough data points for scatter plot.\n")
}
```

# 5. Social-signaling accuracy over repetitions

Does social identification improve over blocks?

```{r model-social-accuracy}
if (nrow(social_guesses) > 0) {
  social_guesses_model <- social_guess_trials(social_guesses, games)

  n_social_games <- n_distinct(social_guesses_model$gameId)
  if (n_social_games >= 2) {
    m_social_acc <- fit_progressively(
      correct ~ blockNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
      data = social_guesses_model,
      family = binomial
    )
  } else {
    m_social_acc <- glm(correct ~ blockNum_c, data = social_guesses_model, family = binomial)
    cat("Only 1 game; using fixed-effects logistic model.\n")
  }

  print(summary(m_social_acc))
  report_effect_sizes(m_social_acc)

  cat(sprintf("\nOverall social accuracy: %.1f%%\n",
              mean(social_guesses_model$correct, na.rm = TRUE) * 100))
  cat(sprintf("Chance level: 50%%\n"))
} else {
  cat("No social guess data.\n")
}
```

```{r plot-social-accuracy, fig.width=7, fig.height=4}
# Plot 14: Social-signaling accuracy over blocks (pooled), over eligible
# response opportunities (code_accuracy() via social_guess_trials())
if (nrow(social_guesses) > 0) {
  social_by_block <- social_guess_trials(social_guesses, games) |>
    filter(!is.na(correct)) |>
    group_by(blockNum) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(social_by_block, aes(x = blockNum, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                alpha = 0.2, fill = "black") +
    geom_line(color = "black", linewidth = 1) +
    geom_point(color = "black", size = 2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    annotate("text", x = max(social_by_block$blockNum), y = 0.52,
             label = "Chance (50%)", hjust = 1, color = "gray50", size = 3) +
    labs(
      x = "Block number",
      y = "Proportion correct social guesses",
      title = "Social-signaling accuracy over Phase 2 blocks"
    ) +
    scale_y_continuous(limits = c(0, 1))
}
```

```{r plot-social-accuracy-by-condition, fig.width=8, fig.height=5}
# Social accuracy trajectories by condition (paper: compare across social
# conditions), over eligible response opportunities
if (nrow(social_guesses) > 0) {
  social_with_cond <- social_guess_trials(social_guesses, games, SOCIAL_CONDITIONS)

  if (n_distinct(social_with_cond$condition) >= 2) {
    social_by_block_cond <- social_with_cond |>
      filter(!is.na(correct)) |>
      group_by(condition, blockNum) |>
      tidyboot_mean(correct, nboot = 1000)

    ggplot(social_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                     color = condition)) +
      geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                  alpha = 0.15, color = NA) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
      scale_color_condition() +
      scale_fill_condition() +
      labs(
        x = "Block number",
        y = "Proportion correct social guesses",
        color = "Condition",
        title = "Social-signaling accuracy trajectories by condition"
      ) +
      scale_y_continuous(limits = c(0, 1))
  }
}
```

# 6. Description length in Phase 2

How does description length evolve across Phase 2 blocks by condition?

Paper formula:
`log(utt_length) ~ condition * block_num + (1 | game) + (1 | group) + (1 | participant) + (1 | tangram)`,
with `group` the game-unique original group. Length is log-transformed for the
same reason as in the convention-formation check: word counts here are
heavy-tailed (Hawkins et al. 2023). Slopes and marginal means are
back-transformed to the word scale for reporting.

```{r model-utt-length-phase2}
p2_utts <- phase2_utterances(speaker_utts, games) |>
  mutate(logUttLength = log(uttLength)) |>
  filter(is.finite(logUttLength))

n_cond <- n_distinct(p2_utts$condition)
if (n_cond >= 2) {
  m_p2_length <- fit_progressively(
    logUttLength ~ condition * blockNum_c + (blockNum_c | gameId) + (1 | group) + (1 | playerId) + (blockNum_c | target) + (1 | group:target),
    data = p2_utts
  )
} else {
  m_p2_length <- fit_progressively(
    logUttLength ~ blockNum_c + (blockNum_c | gameId) + (1 | group) + (1 | playerId) + (blockNum_c | target) + (1 | group:target),
    data = p2_utts
  )
  cat("Only 1 condition; fitting without condition predictor.\n")
}

summary(m_p2_length)
report_effect_sizes(m_p2_length)

if (n_cond >= 2 && inherits(m_p2_length, "merMod")) {
  # The paper's predictions are per condition: continued shortening or a
  # plateau in refer_separated, a possible initial lengthening in refer_mixed,
  # short descriptions in the social conditions.
  cat("\nBlock slope by condition (log scale, and as a multiplier per block):\n")
  p2_length_slopes <- as.data.frame(test(emtrends(m_p2_length, ~ condition, var = "blockNum_c")))
  p2_length_slopes$per_block_multiplier <- exp(p2_length_slopes$blockNum_c.trend)
  print(p2_length_slopes)
  cat("\nMean Phase 2 description length by condition, in words (at the mean block):\n")
  print(summary(emmeans(m_p2_length, "condition"), type = "response"))
}
```

```{r plot-utt-length-phase2, fig.width=8, fig.height=5}
if (nrow(p2_utts) > 0) {
  length_by_block_cond <- p2_utts |>
    group_by(condition, blockNum) |>
    tidyboot_mean(uttLength, nboot = 1000)

  ggplot(length_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                   color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number",
      y = "Mean word count",
      color = "Condition",
      title = "Description length over Phase 2 blocks by condition"
    )
}
```

# 7. Utterance properties across the four conditions

The paper's "Utterance properties" paragraph: the H3c measures (concreteness,
lexical uniqueness, word frequency; see `compute_derived.py`) applied to all
four conditions, to explore whether group-specific language maintained
passively (refer_separated) differs from language maintained under a
social-signaling goal (social_mixed). Each measure is modeled on each speaker's
final description of each tangram within a phase, using the same structures
as H3c: the two proportion measures as binomial counts `cbind(k, n - k)` with
a per-description intercept, and word frequency as a linear model, each with
`(1 | game) + (1 | group) + (1 | participant) + (1 | tangram)` (`group` is
the game-unique original group). The named contrast is refer_separated minus
social_mixed; all pairwise comparisons follow (Tukey-adjusted). Exploratory.

```{r utterance-properties-all-conditions}
PROPERTY_LABELS <- c(concreteness = "Concreteness", uniqueness = "Lexical uniqueness",
                     mean_zipf_freq = "Word frequency (Zipf)")
# The count column behind each proportion; NULL means the measure is linear.
PROPERTY_NUMERATORS <- list(concreteness = "n_concrete",
                            uniqueness = "n_unique_words",
                            mean_zipf_freq = NULL)

if (!has_rows(desc_props) || !has_rows(lex_uniq)) {
  cat("Skipping: description_properties.csv or lexical_uniqueness.csv not found.\n")
} else {
  property_summaries <- list()
  for (ph in c(1, 2)) {
    props <- final_phase_properties(desc_props, lex_uniq, games, phase = ph)
    if (nrow(props) == 0 || n_distinct(props$condition) < 2) {
      cat(sprintf("Phase %d: not enough conditions with data.\n", ph))
      next
    }
    for (measure in names(PROPERTY_LABELS)) {
      # The two proportions are binomial in their content-word counts; word
      # frequency is a mean of a continuous quantity and stays linear.
      numerator <- PROPERTY_NUMERATORS[[measure]]
      if (is.null(numerator)) {
        d <- props |> filter(!is.na(.data[[measure]]))
        f <- as.formula(sprintf(
          "%s ~ condition + (1 | gameId) + (1 | group) + (1 | playerId) + (1 | target)", measure
        ))
        fam <- NULL
      } else {
        d <- word_level_measure(props, numerator, "is_target")
        f <- is_target ~ condition + (1 | gameId) + (1 | group) +
          (1 | playerId) + (1 | target) + (1 | descriptionId)
        fam <- binomial
      }
      if (nrow(d) == 0) next
      cat(sprintf("\n=== Phase %d: %s ===\n", ph, PROPERTY_LABELS[[measure]]))
      m <- fit_progressively(f, data = d, family = fam)
      em <- if (is.null(fam)) {
        emmeans(m, "condition")
      } else {
        emmeans(m, "condition", type = "response")
      }
      # A response-scale binomial grid calls the estimate `prob`; a linear one
      # calls it `emmean`. Normalize so the two stack into one plot.
      em_df <- as.data.frame(summary(em))
      est_col <- intersect(c("emmean", "prob", "response"), names(em_df))[1]
      property_summaries[[length(property_summaries) + 1]] <- em_df |>
        rename(estimate = all_of(est_col)) |>
        mutate(phase = paste("Phase", ph), measure = PROPERTY_LABELS[[measure]])
      lv <- levels(droplevels(d$condition))
      if (all(c("refer_separated", "social_mixed") %in% lv)) {
        cat("Passive vs active maintenance (refer_separated minus social_mixed):\n")
        print(summary(contrast(em, list(passive_minus_active = pairwise_weights(lv, "refer_separated", "social_mixed"))),
                      infer = c(TRUE, TRUE)))
      }
      cat("All pairwise comparisons (Tukey):\n")
      print(summary(pairs(em)))
    }
  }
  if (length(property_summaries)) {
    property_plot <- bind_rows(property_summaries)
    print(
      ggplot(property_plot, aes(x = condition, y = estimate, color = condition)) +
        geom_pointrange(aes(ymin = lower.CL, ymax = upper.CL)) +
        facet_grid(measure ~ phase, scales = "free_y") +
        scale_color_condition() +
        labs(x = "Condition", y = "Estimated marginal mean (95% CI)") +
        theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
    )
  }
}
```

# Statistics for the manuscript

Every contrast, coefficient, confidence interval, and p-value reported above
goes into `writing/manuscript/stats/secondary.tex` as a `\newcommand` macro,
with names prefixed `secondary`. A comparison the dataset cannot support
contributes no macro, so an unavailable number fails loudly in LaTeX rather
than reading as a result. The secondary analyses report frequentist results
only; Bayes factors are preregistered for the primary hypotheses.

```{r write-stats}
# emmeans labels an interaction contrast with one column per factor rather
# than a single `contrast` column, so pull the estimate, interval, and p-value
# by position and accept whichever names the back-transform produced.
emm_macros <- function(prefix, obj, row = 1L, digits = 3) {
  if (is.null(obj)) return(list())
  tab <- as.data.frame(summary(obj, infer = c(TRUE, TRUE)))
  if (nrow(tab) < row) return(list())
  pick <- function(candidates) {
    hit <- intersect(candidates, names(tab))
    if (length(hit) == 0) return(NA_real_)
    tab[[hit[1]]][row]
  }
  setNames(
    list(
      fmt_stat(pick(c("estimate", "odds.ratio", "ratio")), digits),
      fmt_stat(pick(c("lower.CL", "asymp.LCL")), digits),
      fmt_stat(pick(c("upper.CL", "asymp.UCL")), digits),
      fmt_pval(pick("p.value"))
    ),
    paste0(prefix, c("Est", "Lo", "Hi", "P"))
  )
}

if (exists("contrasts_semantic")) {
  macros <- c(
    macros,
    contrast_macros("secondarySemanticSepVsMixed", contrasts_semantic, row = 1),
    contrast_macros("secondarySemanticSocialVsMixed", contrasts_semantic, row = 2)
  )
}
if (exists("drift_ci")) {
  macros <- c(macros, list(
    secondaryDriftMean = fmt_stat(drift_ci$empirical_stat),
    secondaryDriftLo = fmt_stat(drift_ci$ci_lower),
    secondaryDriftHi = fmt_stat(drift_ci$ci_upper)
  ))
}

if (exists("ordering_contrasts")) {
  # contrast_table() gives the `lower`/`upper` names contrast_macros() expects
  ord <- contrast_table(ordering_contrasts, "")
  prefixes <- c(sep_gt_mixed = "secondaryAccSepVsMixed",
                mixed_gt_social = "secondaryAccMixedVsSocial",
                social_gt_socialfirst = "secondaryAccSocialVsSocialFirst")
  for (i in seq_len(nrow(ord))) {
    prefix <- prefixes[as.character(ord$contrast[i])]
    if (is.na(prefix)) next
    macros <- c(macros, contrast_macros(unname(prefix), ord, row = i))
  }
}
if (exists("m_ref_accuracy")) {
  macros <- c(macros, coef_macros("secondaryAccBlock", m_ref_accuracy, "blockNum_c"))
}

if (exists("ct_within")) {
  within_tab <- as.data.frame(summary(ct_within, infer = c(TRUE, TRUE)))
  # One row per in-group status; name the macros by the status rather than the row
  for (i in seq_len(nrow(within_tab))) {
    status <- as.character(within_tab$inGroupSpeaker[i])
    prefix <- if (identical(status, "in_group")) {
      "secondaryIngroupWithinIn"
    } else if (identical(status, "out_group")) {
      "secondaryIngroupWithinOut"
    } else {
      NULL
    }
    if (is.null(prefix)) next
    macros <- c(macros, emm_macros(prefix, ct_within, row = i))
  }
}
if (exists("ct_interaction")) {
  macros <- c(macros, emm_macros("secondaryIngroupInteraction", ct_interaction, row = 1))
}

if (exists("sa_coef") && nrow(sa_coef) == 1) {
  macros <- c(macros, list(
    secondaryGsSocialEst = fmt_stat(sa_coef$estimate),
    secondaryGsSocialLo = fmt_stat(sa_coef$conf.low),
    secondaryGsSocialHi = fmt_stat(sa_coef$conf.high),
    secondaryGsSocialP = fmt_pval(sa_coef$p.value),
    secondaryGsSocialNGames = nrow(gs_social_fit$data)
  ))
}
if (exists("sa_coef_w") && nrow(sa_coef_w) == 1) {
  macros <- c(macros, list(
    secondaryGsSocialWlsEst = fmt_stat(sa_coef_w$estimate),
    secondaryGsSocialWlsLo = fmt_stat(sa_coef_w$conf.low),
    secondaryGsSocialWlsHi = fmt_stat(sa_coef_w$conf.high),
    secondaryGsSocialWlsP = fmt_pval(sa_coef_w$p.value)
  ))
}

if (exists("m_social_acc")) {
  macros <- c(macros, coef_macros("secondarySocialAccBlock", m_social_acc, "blockNum_c"))
}
if (exists("social_guesses_model") && has_rows(social_guesses_model)) {
  macros <- c(macros, list(
    secondarySocialAccMean = fmt_stat(mean(social_guesses_model$correct, na.rm = TRUE)),
    secondarySocialAccN = fmt_count(sum(!is.na(social_guesses_model$correct)))
  ))
}

if (exists("m_p2_length")) {
  macros <- c(macros, coef_macros("secondaryLengthBlock", m_p2_length, "blockNum_c"))
}

write_stats_tex(macros, stats_file("secondary"), source = "analysis/03_secondary_analysis.qmd")
```
