---
title: "03: Secondary analyses"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
library(emmeans)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_phase_change <- has_rows(phase_change)
has_pairwise <- has_rows(pairwise_sim)

n_conditions <- n_distinct(games$condition)
```

# 1. Semantic change (Phase 1 → Phase 2)

Cosine similarity between final Phase 1 and final Phase 2 descriptions per participant × tangram.
Higher similarity = less semantic change.

**Predictions:** `refer_mixed` < `refer_separated`, `social_mixed` > `refer_mixed`

```{r model-semantic-change}
if (!has_phase_change || nrow(phase_change) == 0) {
  cat("Skipping: phase_change_similarities.csv not found. Run compute_derived.py first.\n")
} else {
phase_change <- phase_change |>
  mutate(condition = factor(condition, levels = CONDITION_ORDER))

if (n_conditions >= 2) {
  m_semantic <- fit_progressively(
    simPhase1Phase2 ~ condition + (1 | playerId) + (condition | target),
    data = phase_change
  )
  print(summary(m_semantic))
  report_effect_sizes(m_semantic)

  # Preregistered predictions: refer_separated > refer_mixed and
  # social_mixed > refer_mixed (higher similarity = less semantic change)
  if (all(c("refer_separated", "refer_mixed", "social_mixed") %in% phase_change$condition)) {
    em_semantic <- emmeans(m_semantic, "condition")
    lv <- levels(droplevels(phase_change$condition))
    contrasts_semantic <- contrast(em_semantic, list(
      sep_gt_mixed    = pairwise_weights(lv, "refer_separated", "refer_mixed"),
      social_gt_mixed = pairwise_weights(lv, "social_mixed", "refer_mixed")
    ))
    cat("\nPlanned contrasts (positive = predicted direction):\n")
    print(summary(contrasts_semantic, infer = c(TRUE, TRUE)))
  } else {
    cat("\nPlanned contrasts need refer_separated, refer_mixed, and social_mixed.\n")
  }

  # Baseline drift: refer_separated should be near ceiling
  drift <- phase_change |> filter(condition == "refer_separated", !is.na(simPhase1Phase2))
  if (nrow(drift) > 1) {
    drift_ci <- tidyboot_mean(drift, simPhase1Phase2, nboot = 1000)
    cat(sprintf("\nrefer_separated Phase 1 to Phase 2 similarity: M = %.3f, 95%% CI [%.3f, %.3f] (ceiling = 1)\n",
                drift_ci$empirical_stat, drift_ci$ci_lower, drift_ci$ci_upper))
  }
} else {
  # With only 1 condition, fit without condition predictor
  m_semantic <- lmer(
    simPhase1Phase2 ~ 1 + (1 | playerId) + (1 | target),
    data = phase_change,
    control = lmerControl(optimizer = "bobyqa")
  )
  cat(sprintf("Only %d condition(s). Intercept-only model:\n", n_conditions))
  print(summary(m_semantic))
}
} # end has_phase_change
```

```{r plot-semantic-change, fig.width=7, fig.height=5}
if (has_phase_change && nrow(phase_change) > 0) {
# Plot 10: Semantic change by condition
ggplot(phase_change, aes(x = condition, y = simPhase1Phase2, fill = condition)) +
  geom_violin(alpha = 0.3) +
  geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
  stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
  scale_fill_condition() +
  labs(
    x = "Condition",
    y = "Cosine similarity (Phase 1 final → Phase 2 final)",
    title = "Semantic change across phases by condition",
    subtitle = "Higher = less change from Phase 1 to Phase 2"
  ) +
  theme(legend.position = "none")
}
```

# 2. Referential accuracy in Phase 2

**Predictions:** Above chance (1/16), main effect of `blockNum`, ordered: separated > mixed > social_mixed.

```{r prep-accuracy}
# Phase 2 listener trials with correct, condition, and centered block
p2_listener <- listener_trials(trials, games, phase = 2)
```

```{r model-accuracy}
n_games <- n_distinct(p2_listener$gameId)

if (n_conditions >= 2 && n_games >= 2) {
  # Paper formula: condition * blockNum slopes on tangram, blockNum on game/player
  m_ref_accuracy <- fit_progressively(
    correct ~ condition * blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (condition * blockNum_c | target),
    data = p2_listener,
    family = binomial
  )
} else if (n_games >= 2) {
  cat("Only 1 condition; fitting without condition predictor.\n")
  m_ref_accuracy <- fit_progressively(
    correct ~ blockNum_c + (blockNum_c | gameId) + (blockNum_c | playerId) + (blockNum_c | target),
    data = p2_listener,
    family = binomial
  )
} else {
  m_ref_accuracy <- glm(correct ~ blockNum_c, data = p2_listener, family = binomial)
  cat("Only 1 game; using fixed-effects logistic model.\n")
}

summary(m_ref_accuracy)
report_effect_sizes(m_ref_accuracy)
```

The three preregistered predictions: accuracy above chance (1/16) in every
condition, a positive main effect of block averaged over conditions, and the
ordering refer_separated > refer_mixed > social_mixed > social_first averaged
over blocks.

```{r accuracy-tests}
if (inherits(m_ref_accuracy, "glmerMod") && "condition" %in% all.vars(formula(m_ref_accuracy))) {
  em_acc <- emmeans(m_ref_accuracy, "condition")   # at the mean block (blockNum_c = 0)

  cat("Above chance? One-sided test of each condition's accuracy against 1/16 (logit scale):\n")
  print(test(em_acc, null = qlogis(1 / 16), side = ">"))
  cat("\nEstimated accuracy by condition (response scale):\n")
  print(summary(em_acc, type = "response"))

  cat("\nBlock slope averaged over conditions (log-odds per block), then per condition:\n")
  print(test(emtrends(m_ref_accuracy, ~ 1, var = "blockNum_c")))
  print(test(emtrends(m_ref_accuracy, ~ condition, var = "blockNum_c")))

  lv <- levels(droplevels(p2_listener$condition))
  ordering <- list()
  if (all(c("refer_separated", "refer_mixed") %in% lv)) ordering$sep_gt_mixed <- pairwise_weights(lv, "refer_separated", "refer_mixed")
  if (all(c("refer_mixed", "social_mixed") %in% lv)) ordering$mixed_gt_social <- pairwise_weights(lv, "refer_mixed", "social_mixed")
  if (all(c("social_mixed", "social_first") %in% lv)) ordering$social_gt_socialfirst <- pairwise_weights(lv, "social_mixed", "social_first")
  if (length(ordering)) {
    cat("\nPredicted ordering, consecutive contrasts (positive = predicted direction, log-odds):\n")
    print(summary(contrast(em_acc, ordering), infer = c(TRUE, TRUE)))
  }
} else {
  cat("Preregistered accuracy tests need the condition-by-block mixed model.\n")
}
```

```{r plot-accuracy, fig.width=8, fig.height=5}
# Plot 11: Referential accuracy over Phase 2 blocks by condition
acc_by_block <- p2_listener |>
  filter(!is.na(correct)) |>
  group_by(condition, blockNum) |>
  tidyboot_mean(correct, nboot = 1000)

ggplot(acc_by_block, aes(x = blockNum, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.15, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  annotate("text", x = max(acc_by_block$blockNum), y = 1/16 + 0.02,
           label = "Chance (1/16)", hjust = 1, color = "gray50", size = 3) +
  labs(
    x = "Block number",
    y = "Proportion correct",
    color = "Condition",
    title = "Referential accuracy in Phase 2"
  ) +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_condition() +
  scale_fill_condition()
```

# 3. In-group vs out-group listener accuracy (social conditions)

Within `social_mixed` and `social_first`: do listeners understand in-group speakers better?

```{r model-ingroup}
# Phase 2 listener trials in the social conditions, each with its speaker and
# in-group status (attach_speaker() joins on the listener's current group, since
# a roundId is shared by all groups in a game). The condition factor keeps only
# the two social levels so the random slope has no empty-level columns.
social_listener <- listener_trials(trials, games, phase = 2, conditions = SOCIAL_CONDITIONS) |>
  attach_speaker(trials) |>
  filter(!is.na(inGroupSpeaker))

if (nrow(social_listener) > 0) {
  # Paper formula: ref_correct ~ in_group_speaker * condition + (1 | speaker) + (1 | listener) + (condition | tangram)

  m_ingroup <- fit_progressively(
    correct ~ inGroupSpeaker * condition + (1 | playerId) + (1 | speakerId) + (condition | target),
    data = social_listener,
    family = binomial
  )
  print(summary(m_ingroup))
  report_effect_sizes(m_ingroup)

  if (inherits(m_ingroup, "glmerMod")) {
    # Prediction: in-group listeners are more accurate; the interaction asks
    # whether social_first amplifies that advantage.
    em_ingroup <- emmeans(m_ingroup, ~ inGroupSpeaker | condition)
    cat("\nIn-group minus out-group (log-odds) within each condition:\n")
    print(summary(contrast(em_ingroup, "revpairwise"), infer = c(TRUE, TRUE)))
    cat("\nAveraged over conditions:\n")
    print(summary(contrast(emmeans(m_ingroup, ~ inGroupSpeaker), "revpairwise"), infer = c(TRUE, TRUE)))
    cat("\nEstimated accuracy (response scale):\n")
    print(summary(em_ingroup, type = "response"))
  }
} else {
  cat("No social condition listener data available.\n")
}
```

```{r plot-ingroup-by-condition, fig.width=8, fig.height=5}
if (exists("social_listener") && nrow(social_listener) > 0 &&
    "inGroupSpeaker" %in% names(social_listener) &&
    n_distinct(social_listener$condition) >= 2) {
  ingroup_by_cond <- social_listener |>
    filter(!is.na(correct), !is.na(inGroupSpeaker)) |>
    mutate(speaker_type = ifelse(inGroupSpeaker == "in_group", "In-group", "Out-group")) |>
    group_by(condition, speaker_type) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(ingroup_by_cond, aes(x = speaker_type, y = empirical_stat, fill = speaker_type)) +
    geom_col(alpha = 0.7, width = 0.6) +
    geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2) +
    facet_wrap(~condition) +
    geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
    labs(
      x = "Speaker type",
      y = "Mean listener accuracy",
      title = "In-group vs out-group accuracy by condition"
    ) +
    scale_y_continuous(limits = c(0, 1)) +
    theme(legend.position = "none")
}
```

# 4. Group-specificity vs social accuracy correlation

Does maintaining group-specific language correlate with social identification accuracy?

```{r model-gs-social}
if (!has_pairwise || nrow(pairwise_sim) == 0) {
  cat("Skipping: pairwise_similarities.csv not found. Run compute_derived.py first.\n")
} else if (nrow(social_guesses) > 0) {
  # Mean social accuracy per game
  social_acc_by_game <- social_guess_trials(social_guesses, games) |>
    group_by(gameId) |>
    summarise(social_accuracy = mean(correct, na.rm = TRUE), .groups = "drop")

  # The same per-game estimates as the primary analysis (weight = 1 / SE^2 of Phase 2)
  gs_social <- gs_wide(game_specificity_table(pairwise_sim, games)) |>
    inner_join(social_acc_by_game, by = "gameId")

  if (nrow(gs_social) >= 3) {
    m_gs_social <- lm(gs_phase2 ~ social_accuracy + gs_phase1, data = gs_social, weights = weight)
    print(summary(m_gs_social))
    report_effect_sizes(m_gs_social)
  } else {
    cat(sprintf("Not enough games (%d) for regression (need ≥3).\n", nrow(gs_social)))
  }
} else {
  cat("No social guess data available.\n")
}
```

```{r plot-gs-social, fig.width=6, fig.height=5}
# Plot 13: Social accuracy vs group-specificity scatter
if (exists("gs_social") && nrow(gs_social) >= 3) {
  ggplot(gs_social, aes(x = social_accuracy, y = gs_phase2)) +
    geom_point(size = 3) +
    geom_smooth(method = "lm", se = TRUE, alpha = 0.2) +
    labs(
      x = "Mean social accuracy",
      y = "Phase 2 group-specificity",
      title = "Social accuracy vs group-specificity (social_mixed)"
    ) +
    NULL
} else {
  cat("Not enough data points for scatter plot.\n")
}
```

# 5. Social-signaling accuracy over repetitions

Does social identification improve over blocks?

```{r model-social-accuracy}
if (nrow(social_guesses) > 0) {
  social_guesses_model <- social_guess_trials(social_guesses, games)

  n_social_games <- n_distinct(social_guesses_model$gameId)
  if (n_social_games >= 2) {
    m_social_acc <- fit_progressively(
      correct ~ blockNum_c + (1 | gameId) + (1 | playerId) + (1 | target),
      data = social_guesses_model,
      family = binomial
    )
  } else {
    m_social_acc <- glm(correct ~ blockNum_c, data = social_guesses_model, family = binomial)
    cat("Only 1 game; using fixed-effects logistic model.\n")
  }

  print(summary(m_social_acc))
  report_effect_sizes(m_social_acc)

  cat(sprintf("\nOverall social accuracy: %.1f%%\n",
              mean(social_guesses_model$correct, na.rm = TRUE) * 100))
  cat(sprintf("Chance level: 50%%\n"))
} else {
  cat("No social guess data.\n")
}
```

```{r plot-social-accuracy, fig.width=7, fig.height=4}
# Plot 14: Social-signaling accuracy over blocks (pooled)
if (nrow(social_guesses) > 0) {
  social_by_block <- social_guesses |>
    mutate(correct = as.numeric(
      socialGuessCorrect == TRUE | socialGuessCorrect == "true"
    )) |>
    filter(!is.na(correct)) |>
    group_by(blockNum) |>
    tidyboot_mean(correct, nboot = 1000)

  ggplot(social_by_block, aes(x = blockNum, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper),
                alpha = 0.2, fill = "black") +
    geom_line(color = "black", linewidth = 1) +
    geom_point(color = "black", size = 2) +
    geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
    annotate("text", x = max(social_by_block$blockNum), y = 0.52,
             label = "Chance (50%)", hjust = 1, color = "gray50", size = 3) +
    labs(
      x = "Block number",
      y = "Proportion correct social guesses",
      title = "Social-signaling accuracy over Phase 2 blocks"
    ) +
    scale_y_continuous(limits = c(0, 1))
}
```

```{r plot-social-accuracy-by-condition, fig.width=8, fig.height=5}
# Social accuracy trajectories by condition (paper: compare across social conditions)
if (nrow(social_guesses) > 0) {
  social_with_cond <- social_guesses |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    mutate(correct = as.numeric(
      socialGuessCorrect == TRUE | socialGuessCorrect == "true"
    ))

  if (n_distinct(social_with_cond$condition) >= 2) {
    social_by_block_cond <- social_with_cond |>
      filter(!is.na(correct)) |>
      group_by(condition, blockNum) |>
      tidyboot_mean(correct, nboot = 1000)

    ggplot(social_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                     color = condition)) +
      geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                  alpha = 0.15, color = NA) +
      geom_line(linewidth = 1) +
      geom_point(size = 2) +
      geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
      scale_color_condition() +
      scale_fill_condition() +
      labs(
        x = "Block number",
        y = "Proportion correct social guesses",
        color = "Condition",
        title = "Social-signaling accuracy trajectories by condition"
      ) +
      scale_y_continuous(limits = c(0, 1))
  }
}
```

# 6. Description length in Phase 2

How does description length evolve across Phase 2 blocks by condition?

Paper formula: `utt_length ~ condition * block_num + (1 | participant) + (1 | tangram)`

```{r model-utt-length-phase2}
p2_utts <- phase2_utterances(speaker_utts, games)

n_cond <- n_distinct(p2_utts$condition)
if (n_cond >= 2) {
  # Paper formula: utt_length ~ condition * block_num + (1 | participant) + (1 | tangram)
  m_p2_length <- fit_progressively(
    uttLength ~ condition * blockNum_c + (1 | playerId) + (1 | target),
    data = p2_utts
  )
} else {
  m_p2_length <- fit_progressively(
    uttLength ~ blockNum_c + (1 | playerId) + (1 | target),
    data = p2_utts
  )
  cat("Only 1 condition; fitting without condition predictor.\n")
}

summary(m_p2_length)
report_effect_sizes(m_p2_length)

if (n_cond >= 2 && inherits(m_p2_length, "merMod")) {
  # The paper's predictions are per condition: continued shortening or a
  # plateau in refer_separated, a possible initial lengthening in refer_mixed,
  # short descriptions in the social conditions.
  cat("\nBlock slope (words per block) by condition:\n")
  print(test(emtrends(m_p2_length, ~ condition, var = "blockNum_c")))
  cat("\nMean Phase 2 description length by condition (at the mean block):\n")
  print(summary(emmeans(m_p2_length, "condition")))
}
```

```{r plot-utt-length-phase2, fig.width=8, fig.height=5}
if (nrow(p2_utts) > 0) {
  length_by_block_cond <- p2_utts |>
    group_by(condition, blockNum) |>
    tidyboot_mean(uttLength, nboot = 1000)

  ggplot(length_by_block_cond, aes(x = blockNum, y = empirical_stat,
                                   color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.15, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    scale_color_condition() +
    scale_fill_condition() +
    labs(
      x = "Block number",
      y = "Mean word count",
      color = "Condition",
      title = "Description length over Phase 2 blocks by condition"
    )
}
```
