---
title: "04: Exploratory analyses"
format:
  html:
    toc: true
    code-fold: true
---

```{r setup, message=FALSE, warning=FALSE}
library(tidyboot)
library(patchwork)
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

has_adjacent <- has_rows(adjacent_sim)

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/exploratory.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

```{r prep-data}
# Add condition and a block number that runs across both phases (blockNum
# resets per phase; add_global_block() uses each game's own Phase 1 length)
utts_with_condition <- speaker_utts |> add_condition(games) |> add_global_block(games)

# Listener trials of both phases under the response-opportunity rule
# (code_accuracy() via listener_trials(): timeouts and late arrivals are 0,
# trials without an opportunity leave the denominator)
listeners_with_condition <- bind_rows(
  listener_trials(trials, games, phase = 1),
  listener_trials(trials, games, phase = 2)
) |>
  filter(!is.na(correct)) |>
  add_global_block(games)

if (has_adjacent) {
  adj_with_condition <- adjacent_sim |> add_condition(games) |> add_global_block(games)
}
```

# 1. Temporal dynamics: full game

```{r plot-full-temporal, fig.width=10, fig.height=10}
phase_boundary <- games$phase1Blocks[1] - 0.5

# Plot 15: Description length over ALL blocks by condition
length_all <- utts_with_condition |>
  filter(!is.na(uttLength)) |>
  group_by(condition, globalBlock) |>
  tidyboot_mean(uttLength, nboot = 1000)

p_length <- ggplot(length_all, aes(x = globalBlock, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_vline(xintercept = phase_boundary, linetype = "dashed", alpha = 0.5) +
  annotate("text", x = phase_boundary, y = Inf,
           label = "Phase 1 | Phase 2", vjust = 1.5, size = 3, alpha = 0.5) +
  labs(x = "Block", y = "Mean word count", title = "(a) Description length") +
  scale_color_condition() +
  scale_fill_condition() +
  theme(legend.position = "bottom")

# Plot 16a: Referential accuracy over all blocks
acc_all <- listeners_with_condition |>
  filter(!is.na(correct)) |>
  group_by(condition, globalBlock) |>
  tidyboot_mean(correct, nboot = 1000)

p_acc <- ggplot(acc_all, aes(x = globalBlock, y = empirical_stat, color = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
              alpha = 0.12, color = NA) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 1/16, linetype = "dotted", color = "gray50") +
  geom_vline(xintercept = phase_boundary, linetype = "dashed", alpha = 0.5) +
  labs(x = "Block", y = "Proportion correct", title = "(b) Referential accuracy") +
  scale_y_continuous(limits = c(0, 1)) +
  scale_color_condition() +
  scale_fill_condition() +
  theme(legend.position = "bottom")

# Plot 16b: Adjacent similarity over all blocks
if (exists("adj_with_condition")) {
  adj_all <- adj_with_condition |>
    filter(!is.na(simAdjacent)) |>
    group_by(condition, globalBlock) |>
    tidyboot_mean(simAdjacent, nboot = 1000)

  p_sim <- ggplot(adj_all, aes(x = globalBlock, y = empirical_stat, color = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper, fill = condition),
                alpha = 0.12, color = NA) +
    geom_line(linewidth = 1) +
    geom_point(size = 2) +
    geom_vline(xintercept = phase_boundary, linetype = "dashed", alpha = 0.5) +
    labs(x = "Block", y = "Cosine similarity (adjacent)", title = "(c) Convention stability") +
    scale_color_condition() +
    scale_fill_condition() +
    theme(legend.position = "bottom")
}

# Plot 16: Combined temporal dynamics panel
if (exists("p_sim")) {
  p_length / p_acc / p_sim +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
} else {
  p_length / p_acc +
    plot_layout(guides = "collect") &
    theme(legend.position = "bottom")
}
```

# 2. UMAP visualization

```{r plot-umap, fig.width=12, fig.height=8}
# Plot 17: UMAP embedding space
umap_file <- file.path(derived_dir, "umap_projections.csv")
has_umap <- file.exists(umap_file)

if (has_umap) {
  umap_data <- read_csv(umap_file, show_col_types = FALSE)

  # Add condition from games if not present
  if (!"condition" %in% names(umap_data)) {
    umap_data <- umap_data |>
      left_join(games |> select(gameId, condition), by = "gameId")
  }

  umap_data <- umap_data |>
    mutate(
      condition = factor(condition, levels = CONDITION_ORDER),
      phase_label = ifelse(phaseNum == 1, "Phase 1", "Phase 2")
    )

  ggplot(umap_data, aes(x = umap_x, y = umap_y, color = originalGroup)) +
    geom_point(alpha = 0.5, size = 1.5) +
    facet_grid(phase_label ~ condition) +
    labs(
      x = "UMAP 1",
      y = "UMAP 2",
      color = "Original group",
      title = "UMAP projections of speaker descriptions"
    ) +
    scale_color_group() +
    theme(legend.position = "bottom")
} else {
  cat("No UMAP projections file found. Run compute_derived.py first.\n")
}
```

```{r plot-umap-example, fig.width=14, fig.height=8}
# Plot 18: Example game trajectory
if (has_umap && exists("umap_data")) {
  example_game <- umap_data |>
    count(gameId) |>
    filter(n >= 20) |>
    slice(1) |>
    pull(gameId)

  if (length(example_game) > 0) {
    example_data <- umap_data |>
      filter(gameId == example_game) |>
      mutate(globalBlock = ifelse(phaseNum == 1, blockNum,
                                   blockNum + games$phase1Blocks[1]))

    ggplot(example_data, aes(x = umap_x, y = umap_y, color = originalGroup)) +
      geom_point(size = 2, alpha = 0.7) +
      facet_wrap(~globalBlock, ncol = 4) +
      labs(
        x = "UMAP 1",
        y = "UMAP 2",
        color = "Original group",
        title = paste("Example game trajectory:", example_game)
      ) +
      scale_color_group() +
      theme(legend.position = "bottom")
  }
}
```


# Statistics for the manuscript

This notebook is descriptive: it reports trajectories and embedding
projections rather than hypothesis tests, so the macros it writes to
`writing/manuscript/stats/exploratory.tex` are the observation counts and the
endpoint means the figures rest on. Names are prefixed `exploratory`.

```{r write-stats}
# The mean of a measure in its own first and last observed block, with the two
# block numbers, so a measure that only starts once a pair of descriptions
# exists (adjacent similarity) is not reported as missing in block 0.
endpoint_macros <- function(prefix, df, value, digits = 3) {
  d <- df |> filter(is.finite(.data[[value]]), !is.na(globalBlock))
  if (nrow(d) == 0) return(list())
  first <- min(d$globalBlock)
  last <- max(d$globalBlock)
  setNames(
    list(
      fmt_count(nrow(d)),
      first,
      last,
      fmt_stat(mean(d[[value]][d$globalBlock == first]), digits),
      fmt_stat(mean(d[[value]][d$globalBlock == last]), digits)
    ),
    paste0(prefix, c("N", "FirstBlock", "LastBlock", "First", "Last"))
  )
}

macros <- c(
  macros,
  endpoint_macros("exploratoryLength", utts_with_condition, "uttLength", digits = 2),
  endpoint_macros("exploratoryAccuracy", listeners_with_condition, "correct"),
  if (exists("adj_with_condition")) {
    endpoint_macros("exploratorySimAdjacent", adj_with_condition, "simAdjacent")
  } else {
    list()
  }
)
if (exists("has_umap") && has_umap) {
  macros <- c(macros, list(
    exploratoryNUmapPoints = fmt_count(nrow(umap_data))
  ))
}

write_stats_tex(macros, stats_file("exploratory"), source = "analysis/04_exploratory.qmd")
```
