---
title: "05: Exit survey responses"
format:
  html:
    toc: true
    code-fold: true
---

The post-test survey (paper, "Post-test survey and group identification
check") asks whether the instructions were understood, how much participants
identified with and felt close to their Phase 1 group (7-point items), whether
they noticed their group developing its own way of describing the pictures,
their strategy, and, on a second page, whether they felt they were playing with
other humans, plus age, gender, education, pay fairness, and open feedback.
Players removed at the Phase 1 accuracy check complete the survey too.

The pilot predates the revamped survey: it has the strategy question (as
`strength`) and the demographics but none of the group or felt-human items, so
those sections report "not collected" on the pilot data.

```{r setup, message=FALSE, warning=FALSE}
source(here::here("analysis", "config.R"))
list2env(load_tables(), envir = globalenv())

survey <- exit_survey_responses(players, games)
cat("Respondents (any survey field answered):", nrow(survey), "of",
    nrow(players |> semi_join(games |> filter(!is.na(condition)), by = "gameId")), "players in real games\n")

# Statistics this notebook reports, written as LaTeX macros at the end
# (writing/manuscript/stats/survey.tex)
macros <- list()
```

```{r dataset-banner, results='asis', echo=FALSE}
dataset_banner(games = games)
```

# Who responded

```{r response-summary}
if (nrow(survey) == 0) {
  cat("No survey responses in this dataset.\n")
} else {
  print(knitr::kable(
    survey |>
      group_by(condition, respondent, .drop = FALSE) |>
      summarise(
        respondents = n(),
        page2 = if (has_field(survey, "age")) sum(!is.na(age)) else NA_integer_,
        .groups = "drop"
      ) |>
      filter(respondents > 0),
    caption = "Respondents by condition and whether they completed the game (page2: also answered the demographics page)"
  ))
}
```

# Group identification and closeness

The two 7-point items (1 = not at all, 7 = very much). Descriptive by
condition with a Kruskal-Wallis test across conditions; the paper treats these
as a manipulation check, not a hypothesis test.

```{r likert-items, fig.width=8, fig.height=4}
likert_items <- c(groupIdentification = "Sense of being part of the Phase 1 group",
                  groupCloseness = "Closeness to the Phase 1 group")
present <- names(likert_items)[vapply(names(likert_items), function(f) has_field(survey, f), logical(1))]
if (length(present) == 0) {
  cat("Not collected in this dataset (the group items were added after the pilot).\n")
} else {
  for (item in present) {
    cat(sprintf("\n=== %s ===\n", likert_items[[item]]))
    print(knitr::kable(likert_by_condition(survey, item), digits = 2))
    d <- survey |> filter(!is.na(.data[[item]]))
    if (n_distinct(d$condition) >= 2) print(kruskal.test(d[[item]] ~ d$condition))
  }
  long <- survey |>
    select(condition, respondent, all_of(present)) |>
    pivot_longer(all_of(present), names_to = "item", values_to = "rating") |>
    filter(!is.na(rating)) |>
    mutate(item = likert_items[item])
  ggplot(long, aes(x = condition, y = rating, color = condition)) +
    geom_jitter(width = 0.15, height = 0.1, alpha = 0.4, size = 1.5) +
    stat_summary(fun.data = mean_cl_normal, color = "black", size = 0.6) +
    facet_wrap(~item) +
    scale_color_condition() +
    scale_y_continuous(breaks = 1:7, limits = c(0.8, 7.2)) +
    labs(x = "Condition", y = "Rating (1-7)") +
    theme(legend.position = "none", axis.text.x = element_text(angle = 30, hjust = 1))
}
```

# Awareness of group-specific language

```{r group-language}
if (!has_field(survey, "groupLanguage")) {
  cat("Not collected in this dataset.\n")
} else {
  gl <- survey |>
    filter(!is.na(groupLanguage)) |>
    group_by(condition, .drop = FALSE) |>
    summarise(n = n(), yes = sum(groupLanguage == "yes"), prop_yes = yes / n, .groups = "drop") |>
    filter(n > 0)
  print(knitr::kable(gl, digits = 2, caption = "Noticed the group developing its own way of describing the pictures"))
  d <- survey |> filter(!is.na(groupLanguage))
  if (n_distinct(d$condition) >= 2 && n_distinct(d$groupLanguage) >= 2) {
    print(suppressWarnings(chisq.test(table(droplevels(d$condition), d$groupLanguage))))
  }
}
```

# Felt they were playing with humans

Games in which any respondent answered "no" are flagged for detailed
inspection of their chat logs (paper, "AI detection").

```{r felt-human}
if (!has_field(survey, "feltHuman")) {
  cat("Not collected in this dataset (the item was added after the pilot).\n")
} else {
  fh <- survey |>
    filter(!is.na(feltHuman)) |>
    group_by(condition, .drop = FALSE) |>
    summarise(n = n(), said_no = sum(feltHuman == "no"), prop_no = said_no / n, .groups = "drop") |>
    filter(n > 0)
  print(knitr::kable(fh, digits = 2, caption = "Respondents who did not feel they were playing with humans"))
  flags <- felt_human_flags(survey)
  if (nrow(flags) == 0) {
    cat("\nNo game flagged: every respondent felt they were playing with humans.\n")
  } else {
    cat(sprintf("\n%d game(s) flagged for inspection:\n", nrow(flags)))
    print(knitr::kable(flags |> mutate(gameId = substr(gameId, 1, 10))))
    group_flags <- felt_human_flags(survey, level = "group")
    cat(sprintf("\n%d original group(s) flagged for inspection of their chat logs:\n", nrow(group_flags)))
    print(knitr::kable(group_flags |> mutate(gameId = substr(gameId, 1, 10)) |>
                         select(condition, gameId, originalGroup, respondents, said_no, players_saying_no)))
  }
}
```

# Comprehension and pay fairness

```{r comprehension-fairness}
if (has_field(survey, "understood")) {
  print(knitr::kable(survey |> filter(!is.na(understood)) |> count(condition, understood),
                     caption = "Understood the instructions"))
}
if (has_field(survey, "fair")) {
  print(knitr::kable(survey |> filter(!is.na(fair), fair != "") |>
                       select(condition, originalGroup, respondent, fair) |> arrange(condition, originalGroup),
                     caption = "Was the pay fair (free text)"))
}
```

# Demographics

```{r demographics}
if (has_field(survey, "age")) {
  print(knitr::kable(survey |> filter(!is.na(age)) |> group_by(condition, .drop = FALSE) |>
                       summarise(n = n(), mean = mean(age), median = median(age), min = min(age), max = max(age), .groups = "drop") |>
                       filter(n > 0), digits = 1, caption = "Age by condition"))
}
if (has_field(survey, "gender")) {
  print(knitr::kable(survey |> filter(!is.na(gender)) |> count(condition, gender), caption = "Gender by condition"))
}
if (has_field(survey, "education")) {
  print(knitr::kable(survey |> filter(!is.na(education)) |> count(condition, education), caption = "Education by condition"))
}
if (!any(vapply(c("age", "gender", "education"), function(f) has_field(survey, f), logical(1)))) {
  cat("No demographics in this dataset.\n")
}
```

# Strategies and open feedback

```{r free-text}
if (has_field(survey, "strategy")) {
  print(knitr::kable(survey |> filter(!is.na(strategy), strategy != "") |>
                       select(condition, originalGroup, respondent, strategy) |> arrange(condition, originalGroup),
                     caption = "Strategy (free text)"))
}
if (has_field(survey, "feedback")) {
  print(knitr::kable(survey |> filter(!is.na(feedback), feedback != "") |>
                       select(condition, originalGroup, respondent, feedback) |> arrange(condition, originalGroup),
                     caption = "Open feedback"))
}
```

# Statistics for the manuscript

The survey numbers the manuscript quotes are written to
`writing/manuscript/stats/survey.tex` as `\newcommand` macros, with names
prefixed `survey`. An item the dataset never collected contributes no macro
rather than a zero, so a missing number fails loudly in LaTeX instead of
reading as a result.

```{r write-stats}
item_macros <- function(prefix, field) {
  if (!has_field(survey, field)) return(list())
  v <- survey[[field]][!is.na(survey[[field]])]
  if (length(v) == 0) return(list())
  setNames(
    list(length(v), fmt_stat(mean(v), 2), fmt_stat(sd(v), 2)),
    paste0(prefix, c("N", "Mean", "SD"))
  )
}

yes_macros <- function(prefix, field) {
  if (!has_field(survey, field)) return(list())
  v <- survey[[field]][!is.na(survey[[field]])]
  if (length(v) == 0) return(list())
  setNames(
    list(length(v), sum(v == "yes"), fmt_stat(mean(v == "yes"))),
    paste0(prefix, c("N", "Yes", "PropYes"))
  )
}

macros <- c(
  macros,
  list(
    surveyNRespondents = nrow(survey),
    surveyNPlayers = nrow(players |> semi_join(games |> filter(!is.na(condition)), by = "gameId")),
    surveyNRemovedRespondents = sum(survey$respondent == "removed", na.rm = TRUE)
  ),
  item_macros("surveyIdentification", "groupIdentification"),
  item_macros("surveyCloseness", "groupCloseness"),
  yes_macros("surveyGroupLanguage", "groupLanguage"),
  item_macros("surveyAge", "age")
)

if (has_field(survey, "feltHuman")) {
  fh_all <- survey$feltHuman[!is.na(survey$feltHuman)]
  flags <- felt_human_flags(survey)
  group_flags <- felt_human_flags(survey, level = "group")
  macros <- c(macros, list(
    surveyFeltHumanN = length(fh_all),
    surveyFeltHumanNo = sum(fh_all == "no"),
    surveyFeltHumanPropNo = fmt_stat(mean(fh_all == "no")),
    surveyNGamesFlagged = nrow(flags),
    surveyNGroupsFlagged = nrow(group_flags)
  ))
}
if (has_field(survey, "understood")) {
  u <- survey$understood[!is.na(survey$understood)]
  macros <- c(macros, list(
    surveyUnderstoodN = length(u),
    surveyUnderstoodYes = sum(u == "yes")
  ))
}

write_stats_tex(macros, stats_file("survey"), source = "analysis/05_exit_survey.qmd")
```
