---
title: "SI: Exploratory visualizations"
format:
  html:
    toc: true
    code-fold: true
---

Exploratory pilot visualizations that complement the main SI analyses.

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
library(patchwork)
source(here::here("analysis", "config.R"))

# Override paths: pilot data
data_dir <- here("data", "pilots")
figures_dir <- here("analysis", "pilot_outputs", "exploratory")
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Re-resolve utterances file for the pilot data dir
utterances_file <- if (file.exists(file.path(data_dir, "speaker_utterances_filtered.csv"))) {
  "speaker_utterances_filtered.csv"
} else {
  "speaker_utterances.csv"
}

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, utterances_file), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)
```

# 1. UMAP semantic space

2D UMAP projections of speaker descriptions (SBERT embeddings), showing how
each tangram's description evolves across the 6 blocks of Phase 2. Each
trajectory is one speaker's descriptions of a tangram across blocks, colored by
original group. Converging trajectories indicate groups adopting similar
descriptions; parallel or diverging trajectories indicate maintained
distinctiveness.

```{r umap-data}
umap <- read_csv(file.path(data_dir, "umap_projections.csv"), show_col_types = FALSE) |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  mutate(
    condition = factor(condition, levels = CONDITION_ORDER),
    originalGroup = factor(originalGroup, levels = GROUP_ORDER)
  )

# Phase 2 utterances: one point per speaker × tangram × block
p2_trajectories <- umap |>
  filter(phaseNum == 2) |>
  mutate(block_label = blockNum + 1 + PHASE2_OFFSET)
```

```{r umap-plot, fig.width=12, fig.height=4.5}
fig_umap <- ggplot(p2_trajectories,
                    aes(x = umap_x, y = umap_y, color = originalGroup,
                        group = interaction(playerId, target))) +
  geom_path(alpha = 0.5, linewidth = 0.5) +
  geom_point(aes(size = block_label), alpha = 0.7) +
  scale_size_continuous(range = c(1, 3.5), name = "Block") +
  facet_wrap(~condition, nrow = 1, scales = "free",
             labeller = labeller(condition = function(x) format_condition(x))) +
  scale_color_group() +
  labs(x = "UMAP 1", y = "UMAP 2", color = "Group") +
  guides(color = guide_legend(order = 1), size = guide_legend(order = 2)) +
  theme(legend.position = "bottom",
        axis.text = element_blank(),
        axis.ticks = element_blank())

fig_umap

save_fig(fig_umap, "SI_umap_semantic_space.pdf", width = 12, height = 4.5)
```

To quantify convergence, we measure the mean between-group Euclidean distance
between group centroids (per tangram) across Phase 2 blocks. Decreasing
distance indicates groups are converging in semantic space.

```{r umap-distances}
# Group centroids per tangram per block, then between-group distances
centroids <- p2_trajectories |>
  group_by(condition, originalGroup, target, blockNum, block_label) |>
  summarise(umap_x = mean(umap_x), umap_y = mean(umap_y), .groups = "drop")

centroid_dists <- centroids |>
  group_by(condition, target, blockNum, block_label) |>
  group_modify(function(grp, key) {
    n <- nrow(grp)
    if (n < 2) return(tibble())
    pairs <- combn(seq_len(n), 2)
    tibble(
      d = sqrt((grp$umap_x[pairs[1, ]] - grp$umap_x[pairs[2, ]])^2 +
               (grp$umap_y[pairs[1, ]] - grp$umap_y[pairs[2, ]])^2)
    )
  }) |>
  ungroup()

centroid_summary <- centroid_dists |>
  group_by(condition, block_label) |>
  tidyboot_mean(d, nboot = 1000) |>
  ungroup()

ggplot(centroid_summary, aes(x = block_label, y = empirical_stat,
                              color = condition, fill = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_condition(name = "Condition") +
  scale_fill_condition(name = "Condition") +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  labs(x = "Block", y = "Mean between-group centroid distance (UMAP)",
       subtitle = "Group convergence in semantic space across Phase 2") +
  theme(legend.position = "bottom")
```


# 2. Group similarity matrices

How similar are descriptions between and within groups across Phase 2? Each
heatmap shows the 3×3 matrix of mean pairwise cosine similarity between
groups (averaged across tangrams and speaker pairs) for one block and
condition. Diagonal cells = within-group similarity; off-diagonal =
between-group. In mixed conditions, off-diagonal cells should brighten over
time as groups converge. Group C was lost in the `refer_mixed` game (grey
cells).

```{r sim-matrix-data}
block_pw <- read_csv(file.path(data_dir, "block_pairwise_similarities.csv"),
                     show_col_types = FALSE) |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  continuous_block() |>
  mutate(
    block1 = block + 1,
    condition = factor(condition, levels = CONDITION_ORDER)
  )

# Canonicalize group pairs and compute cell means
cell_means <- block_pw |>
  filter(phaseNum == 2) |>
  mutate(
    grp_min = pmin(group1, group2),
    grp_max = pmax(group1, group2)
  ) |>
  group_by(condition, grp_min, grp_max, block1) |>
  summarise(mean_sim = mean(similarity, na.rm = TRUE), .groups = "drop")

# Mirror to fill full symmetric 3×3 matrix
lower_tri <- cell_means |> rename(grp_row = grp_min, grp_col = grp_max)
upper_tri <- cell_means |>
  filter(grp_min != grp_max) |>
  rename(grp_row = grp_max, grp_col = grp_min)

matrix_data <- bind_rows(lower_tri, upper_tri) |>
  mutate(
    block_label = paste0("Block ", block1),
    block_label = factor(block_label, levels = paste0("Block ", 7:12)),
    grp_row = factor(grp_row, levels = rev(GROUP_ORDER)),
    grp_col = factor(grp_col, levels = GROUP_ORDER),
    condition = factor(condition, levels = CONDITION_ORDER)
  ) |>
  complete(condition, grp_row, grp_col, block_label)
```

```{r sim-matrix-plot, fig.width=14, fig.height=8}
fig_sim_matrix <- ggplot(matrix_data,
                          aes(x = grp_col, y = grp_row, fill = mean_sim)) +
  geom_tile(color = "white", linewidth = 0.8) +
  geom_text(aes(label = sprintf("%.2f", mean_sim)), size = 2.8, na.rm = TRUE) +
  scale_fill_viridis_c(
    option = "inferno",
    na.value = "grey90",
    name = "Mean cosine\nsimilarity"
  ) +
  facet_grid(
    condition ~ block_label,
    labeller = labeller(condition = function(x) format_condition(x))
  ) +
  labs(x = "Group", y = "Group") +
  coord_equal() +
  theme(
    axis.text.x = element_text(color = GROUP_COLORS[GROUP_ORDER]),
    axis.text.y = element_text(color = GROUP_COLORS[rev(GROUP_ORDER)]),
    strip.text = element_text(size = 14),
    strip.text.y = element_text(angle = 0),
    panel.spacing = unit(0.3, "lines"),
    legend.position = "right"
  )

fig_sim_matrix

save_fig(fig_sim_matrix, "SI_group_similarity_matrices.pdf", width = 14, height = 8)
```


# 3. Cross-group lexical borrowing

When groups mix in Phase 2, do speakers start adopting terms from other groups?
We measure borrowing rate: the proportion of another group's Phase 1 signature
terms that appear in a speaker's Phase 2 descriptions. Higher borrowing in
mixed conditions would indicate convergence at the lexical level.

```{r borrowing-data}
borrowing <- read_csv(file.path(data_dir, "cross_group_borrowing.csv"),
                      show_col_types = FALSE) |>
  mutate(condition = factor(condition, levels = CONDITION_ORDER))
```

```{r borrowing-by-condition, fig.width=8, fig.height=4.5}
# Average borrowing rate per speaker (across all source groups and tangrams)
borrowing_by_speaker <- borrowing |>
  group_by(condition, gameId, playerId, originalGroup) |>
  summarise(
    mean_borrowing = mean(borrowingRate, na.rm = TRUE),
    total_borrowed = sum(nBorrowed),
    total_source_terms = sum(nSourceTerms),
    .groups = "drop"
  )

# Bootstrapped summary by condition
borrowing_summary <- borrowing_by_speaker |>
  group_by(condition) |>
  tidyboot_mean(mean_borrowing, nboot = 1000) |>
  ungroup()

p_borrow <- ggplot(borrowing_summary, aes(x = condition, y = empirical_stat,
                                           fill = condition)) +
  geom_col(width = 0.6, alpha = 0.8) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15) +
  geom_jitter(data = borrowing_by_speaker,
              aes(x = condition, y = mean_borrowing, color = condition),
              width = 0.15, size = 2, alpha = 0.5, inherit.aes = FALSE) +
  scale_fill_condition(guide = "none") +
  scale_color_condition(guide = "none") +
  scale_x_discrete(labels = function(x) format_condition(x)) +
  labs(x = NULL, y = "Mean borrowing rate",
       subtitle = "Proportion of other groups' Phase 1 terms adopted in Phase 2") +
  coord_cartesian(ylim = c(0, NA))

p_borrow
```

```{r borrowing-over-blocks, fig.width=9, fig.height=4}
# Borrowing rate trajectory across Phase 2 blocks
borrowing_by_block <- borrowing |>
  group_by(condition, blockNum) |>
  tidyboot_mean(borrowingRate, nboot = 1000) |>
  ungroup() |>
  mutate(block1 = blockNum + 1 + PHASE2_OFFSET)  # continuous block numbering

p_borrow_traj <- ggplot(borrowing_by_block,
                         aes(x = block1, y = empirical_stat,
                             color = condition, fill = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_condition(name = "Condition") +
  scale_fill_condition(name = "Condition") +
  scale_x_continuous(breaks = scales::breaks_width(1)) +
  labs(x = "Block", y = "Borrowing rate",
       subtitle = "Cross-group term adoption across Phase 2 blocks") +
  theme(legend.position = "bottom")

p_borrow_traj
```

```{r borrowing-combined, fig.width=12, fig.height=4.5}
fig_borrowing <- p_borrow + p_borrow_traj +
  plot_layout(widths = c(1, 1.3))

fig_borrowing

save_fig(fig_borrowing, "SI_lexical_borrowing.pdf", width = 12, height = 4.5)
```


# 4. Within-group convergence rate (H3b)

Do groups that anticipate social goals (`social_first`) converge faster within
groups during Phase 1 than `social_mixed`? We plot within-group pairwise
similarity across Phase 1 blocks for these two conditions.

```{r convergence-data}
# block_pw loaded in section 2

# Phase 1 within-group pairs for social conditions
p1_within <- block_pw |>
  filter(phaseNum == 1, sameGroup == 1,
         condition %in% c("social_mixed", "social_first"))
```

```{r convergence-plot, fig.width=8, fig.height=5}
# Bootstrapped means
convergence_summary <- p1_within |>
  group_by(condition, block1) |>
  tidyboot_mean(similarity, nboot = 1000) |>
  ungroup()

# Per-tangram means for background points
convergence_by_tangram <- p1_within |>
  group_by(condition, block1, target) |>
  summarise(mean_sim = mean(similarity, na.rm = TRUE), .groups = "drop")

p_convergence <- ggplot(convergence_summary,
                         aes(x = block1, y = empirical_stat,
                             color = condition, fill = condition)) +
  geom_point(data = convergence_by_tangram,
             aes(x = block1, y = mean_sim, color = condition),
             alpha = 0.2, size = 1.5, shape = 16,
             position = position_jitterdodge(jitter.width = 0.1,
                                             dodge.width = 0.4)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15, color = NA,
              position = position_dodge(width = 0.3)) +
  geom_line(linewidth = 0.8, position = position_dodge(width = 0.3)) +
  geom_point(size = 2.5, position = position_dodge(width = 0.3)) +
  scale_color_condition(name = "Condition") +
  scale_fill_condition(name = "Condition") +
  scale_x_continuous(breaks = 1:6) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Phase 1 block", y = "Within-group pairwise similarity",
       subtitle = "Within-group convergence: social-first vs social mixed") +
  theme(legend.position = "bottom")

p_convergence

save_fig(p_convergence, "SI_within_group_convergence.pdf", width = 8, height = 5)
```

For comparison, here are all four conditions overlaid:

```{r convergence-all-conditions, fig.width=9, fig.height=5}
p1_within_all <- block_pw |>
  filter(phaseNum == 1, sameGroup == 1)

convergence_all <- p1_within_all |>
  group_by(condition, block1) |>
  tidyboot_mean(similarity, nboot = 1000) |>
  ungroup()

p_conv_all <- ggplot(convergence_all,
                      aes(x = block1, y = empirical_stat,
                          color = condition, fill = condition)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 2.5) +
  scale_color_condition(name = "Condition") +
  scale_fill_condition(name = "Condition") +
  scale_x_continuous(breaks = 1:6) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(x = "Phase 1 block", y = "Within-group pairwise similarity",
       subtitle = "Within-group convergence across all conditions") +
  theme(legend.position = "bottom")

p_conv_all
```


# 5. Tangram-level variation

Some tangrams may be easier to develop conventions for than others. We look at
three indicators at the end of Phase 1: description length, listener accuracy,
and within-group pairwise similarity — broken down by tangram.

```{r tangram-data}
# Final Phase 1 block utterances
utts_p1_final <- speaker_utts |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  filter(phaseNum == 1) |>
  filter(blockNum == max(blockNum))

# Final Phase 1 block accuracy
trials_with_condition <- trials |>
  left_join(games |> select(gameId, condition), by = "gameId")

acc_p1_final <- trials_with_condition |>
  filter(phaseNum == 1, role == "listener", !is.na(clickedCorrect)) |>
  filter(blockNum == max(blockNum))

# Final Phase 1 block within-group similarity
sim_p1_final <- block_pw |>
  filter(phaseNum == 1, sameGroup == 1) |>
  filter(blockNum == max(blockNum))
```

```{r tangram-variation, fig.width=12, fig.height=4.5}
# (a) Description length by tangram
length_by_tangram <- utts_p1_final |>
  group_by(target) |>
  tidyboot_mean(uttLength, nboot = 1000) |>
  ungroup() |>
  mutate(target = fct_reorder(target, empirical_stat))

p_tang_len <- ggplot(length_by_tangram, aes(x = target, y = empirical_stat)) +
  geom_col(fill = "gray60", width = 0.6) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15) +
  geom_jitter(data = utts_p1_final, aes(x = target, y = uttLength),
              width = 0.15, alpha = 0.3, size = 1.2, color = "gray30") +
  labs(x = NULL, y = "Words", subtitle = "Description length") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

# (b) Accuracy by tangram
acc_by_tangram <- acc_p1_final |>
  mutate(correct = as.numeric(clickedCorrect)) |>
  group_by(target) |>
  tidyboot_mean(correct, nboot = 1000) |>
  ungroup() |>
  mutate(target = fct_reorder(target, empirical_stat))

p_tang_acc <- ggplot(acc_by_tangram, aes(x = target, y = empirical_stat)) +
  geom_col(fill = "gray60", width = 0.6) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  labs(x = NULL, y = "Proportion correct", subtitle = "Listener accuracy") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

# (c) Within-group similarity by tangram
sim_by_tangram <- sim_p1_final |>
  group_by(target) |>
  tidyboot_mean(similarity, nboot = 1000) |>
  ungroup() |>
  mutate(target = fct_reorder(target, empirical_stat))

p_tang_sim <- ggplot(sim_by_tangram, aes(x = target, y = empirical_stat)) +
  geom_col(fill = "gray60", width = 0.6) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.15) +
  labs(x = NULL, y = "Cosine similarity", subtitle = "Within-group similarity") +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))

fig_tangram <- p_tang_len + p_tang_acc + p_tang_sim +
  plot_annotation(title = "Tangram-level variation (Phase 1 final block)")

fig_tangram

save_fig(fig_tangram, "SI_tangram_variation.pdf", width = 14, height = 5)
```

We can also check whether tangram difficulty interacts with condition by
looking at group-specificity per tangram:

```{r tangram-specificity, fig.width=9, fig.height=5}
# Group specificity per tangram (Phase 1 final block)
within_final <- block_pw |>
  filter(phaseNum == 1, blockNum == max(blockNum[phaseNum == 1]),
         sameGroup == 1) |>
  group_by(condition, target) |>
  summarise(sim_within = mean(similarity, na.rm = TRUE), .groups = "drop")

between_final <- block_pw |>
  filter(phaseNum == 1, blockNum == max(blockNum[phaseNum == 1]),
         sameGroup == 0) |>
  group_by(condition, target) |>
  summarise(sim_between = mean(similarity, na.rm = TRUE), .groups = "drop")

specificity_tangram <- inner_join(within_final, between_final,
                                   by = c("condition", "target")) |>
  mutate(specificity = sim_within - sim_between)

ggplot(specificity_tangram, aes(x = target, y = specificity,
                                 fill = condition)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6, alpha = 0.8) +
  geom_hline(yintercept = 0, linetype = "dotted", color = "gray50") +
  scale_fill_condition(name = "Condition") +
  labs(x = "Tangram", y = "Group specificity\n(within − between similarity)",
       subtitle = "Phase 1 final block: group specificity by tangram and condition") +
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, hjust = 1))
```
