---
title: "SI: Pilot Data"
format:
  html:
    toc: true
    code-fold: true
---

This notebook generates the numbers and figures for the Supplementary Information
pilot data section. Run it to extract statistics for the LaTeX writeup.

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
source(here::here("analysis", "config.R"))

# Override paths: pilot notebook reads from pilot data, saves to pilot figures
data_dir <- here("data", "pilots")
figures_dir <- here("analysis", "pilot_outputs")
dir.create(figures_dir, showWarnings = FALSE, recursive = TRUE)

# Re-resolve utterances file for the pilot data dir
utterances_file <- if (file.exists(file.path(data_dir, "speaker_utterances_filtered.csv"))) {
  "speaker_utterances_filtered.csv"
} else {
  "speaker_utterances.csv"
}

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
players <- read_csv(file.path(data_dir, "players.csv"), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)
messages <- read_csv(file.path(data_dir, "messages.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, utterances_file), show_col_types = FALSE)
social_guesses <- read_csv(file.path(data_dir, "social_guesses.csv"), show_col_types = FALSE)

has_adjacent <- file.exists(file.path(data_dir, "adjacent_similarities.csv"))
has_pairwise <- file.exists(file.path(data_dir, "pairwise_similarities.csv"))

adjacent_sim <- if (has_adjacent) read_csv(file.path(data_dir, "adjacent_similarities.csv"), show_col_types = FALSE) else tibble()
pairwise_sim <- if (has_pairwise) read_csv(file.path(data_dir, "pairwise_similarities.csv"), show_col_types = FALSE) else tibble()
```

# 1. Sample Description

```{r sample-description}
n_games <- nrow(games)
n_players <- nrow(players)
n_per_game <- n_players / n_games

cat(sprintf("Pilot games: %d (1 per condition)\n", n_games))
cat(sprintf("Total players: %d (%d per game)\n", n_players, n_per_game))
cat("\nConditions:\n")
games |>
  select(gameId, condition, activeGroups, phase1Blocks, phase2Blocks) |>
  knitr::kable()
```

```{r completion-rates}
# Active groups (3 = all groups survived, <3 = dropout)
games_complete <- games |>
  mutate(all_groups_active = activeGroups == 3)

cat(sprintf("\nGames with all 3 groups active through completion: %d / %d\n",
            sum(games_complete$all_groups_active), n_games))

# Per-game active groups
games |>
  select(condition, activeGroups) |>
  knitr::kable(col.names = c("Condition", "Active groups at end"))
```

```{r session-duration}
# Compute session duration from message timestamps
duration_by_game <- messages |>
  group_by(gameId) |>
  summarise(
    first_msg = min(timestamp, na.rm = TRUE),
    last_msg = max(timestamp, na.rm = TRUE),
    duration_min = (last_msg - first_msg) / 60000,
    .groups = "drop"
  ) |>
  left_join(games |> select(gameId, condition), by = "gameId")

cat("Session duration (minutes, from first to last message):\n")
duration_by_game |>
  select(condition, duration_min) |>
  mutate(duration_min = round(duration_min, 1)) |>
  knitr::kable(col.names = c("Condition", "Duration (min)"))

cat(sprintf("\nMedian session duration: %.1f minutes\n", median(duration_by_game$duration_min)))
cat(sprintf("Range: %.1f - %.1f minutes\n",
            min(duration_by_game$duration_min), max(duration_by_game$duration_min)))
```

```{r idle-players}
# Check for idle players
idle_players <- players |>
  filter(idleRounds > 0) |>
  select(gameId, playerId, originalGroup, idleRounds, isActive) |>
  left_join(games |> select(gameId, condition), by = "gameId")

if (nrow(idle_players) > 0) {
  cat(sprintf("Players with idle rounds: %d\n", nrow(idle_players)))
  idle_players |>
    select(condition, originalGroup, idleRounds, isActive) |>
    knitr::kable()
} else {
  cat("No idle players detected.\n")
}
```

# 2. Outcome-Neutral Criteria

## Prep data

```{r prep-data}
# Phase 1 speaker utterances
p1_utts <- speaker_utts |>
  filter(phaseNum == 1) |>
  mutate(
    blockNum_c = blockNum - mean(blockNum, na.rm = TRUE),
    log_uttLength = log(uttLength + 1),
    participant = playerId,
    group = originalGroup,
    tangram = target
  )

# Phase 1 listener accuracy
p1_listener <- trials |>
  filter(phaseNum == 1, role == "listener") |>
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    group = originalGroup,
    tangram = target
  )

p1_ref_accuracy <- p1_listener |>
  group_by(gameId, roundId, blockNum, tangram, group) |>
  summarise(
    ref_accuracy = mean(correct, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(
    blockNum_c = blockNum - mean(blockNum, na.rm = TRUE)
  )

# Phase 1 adjacent similarities
if (has_adjacent) {
  p1_adjacent <- adjacent_sim |>
    filter(phaseNum == 1) |>
    mutate(
      blockNum_c = blockNum - mean(blockNum, na.rm = TRUE),
      participant = playerId,
      group = originalGroup,
      tangram = target
    )
}
```

## 2a. Description Length Decreases

```{r model-length}
m_length <- lmer(
  log_uttLength ~ blockNum_c + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
  data = p1_utts,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

s <- summary(m_length)
coefs <- coef(s)
cat("=== Description Length Model ===\n")
print(s)
cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
            coefs["blockNum_c", "Estimate"],
            coefs["blockNum_c", "Std. Error"],
            coefs["blockNum_c", "t value"]))
```

```{r descriptive-length}
# Descriptive stats for text (by blockNum, not repNum)
length_first <- p1_utts |> filter(blockNum == min(blockNum)) |> pull(uttLength)
length_last <- p1_utts |> filter(blockNum == max(blockNum)) |> pull(uttLength)

cat(sprintf("First block: M = %.1f words (SD = %.1f)\n",
            mean(length_first), sd(length_first)))
cat(sprintf("Last block: M = %.1f words (SD = %.1f)\n",
            mean(length_last), sd(length_last)))
```

## 2b. Listener Accuracy Increases

```{r model-accuracy}
m_accuracy <- lmer(
  ref_accuracy ~ blockNum_c + (blockNum_c | group) + (blockNum_c | tangram),
  data = p1_ref_accuracy,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

s <- summary(m_accuracy)
coefs <- coef(s)
cat("=== Listener Accuracy Model ===\n")
print(s)
cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
            coefs["blockNum_c", "Estimate"],
            coefs["blockNum_c", "Std. Error"],
            coefs["blockNum_c", "t value"]))
```

```{r descriptive-accuracy}
acc_first <- p1_ref_accuracy |> filter(blockNum == min(blockNum)) |> pull(ref_accuracy)
acc_last <- p1_ref_accuracy |> filter(blockNum == max(blockNum)) |> pull(ref_accuracy)

cat(sprintf("First block accuracy: M = %.2f (SD = %.2f)\n",
            mean(acc_first, na.rm = TRUE), sd(acc_first, na.rm = TRUE)))
cat(sprintf("Last block accuracy: M = %.2f (SD = %.2f)\n",
            mean(acc_last, na.rm = TRUE), sd(acc_last, na.rm = TRUE)))
cat(sprintf("Chance level: %.4f (1/16)\n", 1/16))
```

## 2c. Convention Stability Increases

```{r model-adjacent}
if (has_adjacent) {
  p1_adjacent_valid <- p1_adjacent |> filter(!is.na(simAdjacent), blockNum > 0)

  m_adjacent <- lmer(
    simAdjacent ~ blockNum_c + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
    data = p1_adjacent_valid,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
  )

  s <- summary(m_adjacent)
  coefs <- coef(s)
  cat("=== Adjacent Similarity Model ===\n")
  print(s)
  cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
              coefs["blockNum_c", "Estimate"],
              coefs["blockNum_c", "Std. Error"],
              coefs["blockNum_c", "t value"]))
} else {
  cat("Skipping: adjacent_similarities.csv not found.\n")
}
```

## 2d. Group-Specificity Exceeds Chance

```{r compute-group-specificity, warning=FALSE}
source(here::here("analysis", "group_specificity.R"))

if (!has_pairwise) {
  cat("Skipping: pairwise_similarities.csv not found.\n")
} else {
  p1_pairwise <- pairwise_sim |> filter(window == "phase1_final")

  if (nrow(p1_pairwise) > 0) {
    # Uses cached RDS if available; pass force = TRUE to recompute
    gs_cached <- compute_group_specificity(p1_pairwise, data_dir)
    gs_results <- gs_cached$gs_results
    perm_results <- gs_cached$perm_results

    cat("Group-specificity coefficients (Phase 1 final):\n")
    gs_results |>
      left_join(games |> select(gameId, condition), by = "gameId") |>
      select(condition, coefficient, std_error, t_value) |>
      mutate(across(where(is.numeric), ~round(., 3))) |>
      knitr::kable()

    cat("\nPermutation test results:\n")
    perm_results |>
      left_join(games |> select(gameId, condition), by = "gameId") |>
      select(condition, obs_coefficient, perm_mean, p_value) |>
      mutate(across(where(is.numeric), ~round(., 4))) |>
      knitr::kable()

    prop_significant <- mean(perm_results$p_value < 0.05)
    cat(sprintf("\nGames with significant group-specificity: %d / %d (%.0f%%)\n",
                sum(perm_results$p_value < 0.05), nrow(perm_results),
                prop_significant * 100))
    cat(sprintf("Criterion (≥80%%): %s\n",
                ifelse(prop_significant >= 0.8, "PASS", "FAIL")))
  }
}
```

# 3. Combined Figure: Convention Formation (SI Figure 1)

Three-panel figure showing description length, listener accuracy, and convention
stability over Phase 1 blocks. Thin gray lines show per-group trajectories
(4 games × 3 groups = 12 lines), with the bold line and ribbon showing the
overall mean ± 95% bootstrapped CI.

**Note on CIs:** Bootstrapped CIs here resample individual observations
(utterances or trials), not games or groups. With only 4 pilot games, these
CIs reflect observation-level variability and overstate precision relative to
the true game-level uncertainty. They are intended to show directional trends,
not for inference. The per-group lines provide a more honest picture of
variability across independent units.

```{r si-fig-convention, fig.width=14, fig.height=4.5}
library(patchwork)

# Per-group means for spaghetti lines
length_by_group <- p1_utts |>
  group_by(gameId, group, blockNum) |>
  summarise(mean_length = mean(uttLength, na.rm = TRUE), .groups = "drop")

# Panel A: Description length (by blockNum, not repNum)
length_by_block <- p1_utts |>
  group_by(blockNum) |>
  tidyboot_mean(uttLength, nboot = 1000)

p_length <- ggplot(length_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
  geom_line(data = length_by_group,
            aes(x = blockNum + 1, y = mean_length, group = interaction(gameId, group)),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(x = "Block", y = "Words per description", title = "Description length") +
  scale_x_continuous(breaks = 1:6, labels = as.integer)

# Per-group accuracy
acc_by_group <- p1_ref_accuracy |>
  filter(!is.na(ref_accuracy)) |>
  group_by(gameId, group, blockNum) |>
  summarise(mean_acc = mean(ref_accuracy, na.rm = TRUE), .groups = "drop")

# Panel B: Listener accuracy
acc_by_block <- p1_ref_accuracy |>
  filter(!is.na(ref_accuracy), !is.na(blockNum)) |>
  group_by(blockNum) |>
  tidyboot_mean(ref_accuracy, nboot = 1000)

p_accuracy <- ggplot(acc_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
  geom_line(data = acc_by_group,
            aes(x = blockNum + 1, y = mean_acc, group = interaction(gameId, group)),
            color = "gray70", alpha = 0.5, linewidth = 0.4) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  geom_hline(yintercept = 2/3, linetype = "dashed", color = "red", alpha = 0.7) +
  labs(x = "Block", y = "Proportion correct", title = "Listener accuracy") +
  scale_x_continuous(breaks = 1:6, labels = as.integer) +
  scale_y_continuous(limits = c(0, 1))

# Panel C: Adjacent similarity
if (has_adjacent && exists("p1_adjacent")) {
  p1_adjacent_valid <- p1_adjacent |> filter(!is.na(simAdjacent), blockNum > 0)

  sim_by_group <- p1_adjacent_valid |>
    group_by(gameId, group, blockNum) |>
    summarise(mean_sim = mean(simAdjacent, na.rm = TRUE), .groups = "drop")

  sim_by_block <- p1_adjacent_valid |>
    group_by(blockNum) |>
    tidyboot_mean(simAdjacent, nboot = 1000)

  p_sim <- ggplot(sim_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
    geom_line(data = sim_by_group,
              aes(x = blockNum + 1, y = mean_sim, group = interaction(gameId, group)),
              color = "gray70", alpha = 0.5, linewidth = 0.4) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(x = "Block", y = "Cosine similarity", title = "Convention stability") +
    scale_x_continuous(breaks = 1:6, labels = as.integer, limits = c(1, 6)) +
    scale_y_continuous(limits = c(0, 1))

  fig1 <- p_length + p_accuracy + p_sim +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 20))
} else {
  fig1 <- p_length + p_accuracy +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 20))
}

fig1

save_fig(fig1, "SI_convention_formation.pdf", width = 14, height = 4.5)
```

# 4. Permutation Test Figure (SI Figure 2)

For each pilot game, we test whether group-specificity exceeds chance using a
permutation test. We fit a linear mixed-effects model predicting pairwise
cosine similarity from a binary same-group indicator (with random intercepts
for tangram and both speakers). The observed coefficient captures how much more
similar within-group descriptions are compared to between-group descriptions.

To build a null distribution, we permute the group labels of speakers 1,000
times and refit the model each time. The histograms show these null
distributions; the red vertical line marks the observed coefficient. The
p-value is the proportion of permuted coefficients that equal or exceed the
observed value. We require ≥80% of games to show significant group-specificity
(p < .05) to satisfy the outcome-neutral criterion.

```{r si-fig-permutation, fig.width=12, fig.height=3}
if (exists("perm_results") && nrow(perm_results) > 0) {
  perm_plot_data <- perm_results |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    unnest(perm_coefs) |>
    rename(perm_coef = perm_coefs) |>
    mutate(condition_label = format_condition(condition))

  obs_labels <- perm_results |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    mutate(
      condition_label = format_condition(condition),
      sig_label = sprintf("p = %.3f", p_value),
    )

  fig2 <- ggplot(perm_plot_data, aes(x = perm_coef)) +
    geom_histogram(bins = 40, fill = "gray75", color = "white") +
    geom_vline(data = obs_labels,
               aes(xintercept = obs_coefficient),
               color = "#ce3045", linewidth = 1) +
    geom_text(data = obs_labels,
              aes(x = obs_coefficient, y = Inf, label = sig_label),
              vjust = 1.5, hjust = 1.1, color = "#ce3045", size = 4) +
    facet_wrap(~condition_label, nrow = 1, scales = "free_y") +
    labs(
      x = "Group-specificity coefficient (permuted)",
      y = "Count",
    )

  print(fig2)
  save_fig(fig2, "SI_permutation_test.pdf", width = 12, height = 3)
}
```

# 5. Group-Specificity Trajectory (SI Figure 3)

Within- vs between-group similarity over all blocks, showing how group-specificity
emerges in Phase 1 and evolves in Phase 2.

**How pairwise similarity is computed:** For each block and tangram, we take
each speaker's most recent description of that tangram up to and including the
current block (within the same phase). We then compute pairwise cosine
similarity (using SBERT embeddings) between all speaker pairs for that tangram.
This is a cumulative approach, not a sliding window: at block 0, only speakers
who described the tangram in block 0 are included; by block 5, each speaker's
representation is their latest description from any block 0–5. Pairs are
labeled as "within group" or "between group" based on original group
assignment. Group-specificity is the difference (within − between).

**Figure 3a (within vs between):** Faded points show per-tangram mean similarity
for each block (1 game per condition facet, 6 tangrams per block per pair type).
Lines and ribbons show the overall mean ± 95% bootstrapped CI.

**Figure 3b (specificity trajectory):** Faded points show per-tangram
group-specificity (within − between similarity) for each block and condition.
Lines and ribbons show the mean ± 95% bootstrapped CI across tangrams.

**Note on CIs:** Same caveat as above — CIs resample pairwise similarity
values (or per-tangram specificity scores), not games. With 1 game per
condition, these reflect within-game variability across tangrams and speaker
pairs, not between-game uncertainty. The individual points provide the
underlying spread.

```{r load-block-pairwise}
has_block_pw <- file.exists(file.path(data_dir, "block_pairwise_similarities.csv"))
if (has_block_pw) {
  block_pw <- read_csv(file.path(data_dir, "block_pairwise_similarities.csv"),
                       show_col_types = FALSE) |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    continuous_block() |>
    mutate(
      block1 = block + 1,
      pair_type = ifelse(sameGroup == 1, "Within group", "Between group"),
      condition = factor(condition, levels = CONDITION_ORDER)
    )
} else {
  cat("Skipping: block_pairwise_similarities.csv not found.\n")
}
```

```{r si-fig-within-between, fig.width=10, fig.height=3.7}
if (has_block_pw) {
  # Bootstrapped means by block, pair type, and condition
  wb_summary <- block_pw |>
    group_by(condition, block1, pair_type) |>
    tidyboot_mean(similarity, nboot = 1000) |>
    ungroup()

  # Per-tangram means for jittered points
  wb_by_tangram <- block_pw |>
    group_by(condition, block1, pair_type, target) |>
    summarise(mean_sim = mean(similarity, na.rm = TRUE), .groups = "drop")

  fig3a <- ggplot(wb_summary, aes(x = block1, y = empirical_stat,
                                   color = pair_type, fill = pair_type)) +
    annotate("rect", xmin = PHASE_BOUNDARY, xmax = Inf,
             ymin = -Inf, ymax = Inf, fill = "gray90", alpha = 0.5) +
    geom_vline(xintercept = PHASE_BOUNDARY, color = "gray70", linetype = "dotted") +
    geom_point(data = wb_by_tangram,
               aes(x = block1, y = mean_sim, color = pair_type),
               alpha = 0.15, size = 0.8, position = position_jitter(width = 0.2)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2,
                color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    facet_wrap(~condition, nrow = 1,
               labeller = labeller(condition = function(x) format_condition(x))) +
    scale_fill_brewer(name = "Pair type", palette = "Accent") +
    scale_color_brewer(name = "Pair type", palette = "Accent") +
    scale_x_continuous(breaks = scales::breaks_width(1)) +
    labs(x = "Block", y = "Pairwise cosine similarity") +
    coord_cartesian(ylim = c(0, 1)) +
    theme(legend.position = "bottom",
          legend.box.margin = margin(t = -10, b = -10))

}
```

```{r si-fig-specificity-trajectory, fig.width=9, fig.height=4}
if (has_block_pw) {
  # Compute group-specificity = within - between per block, condition, tangram
  within_means <- block_pw |>
    filter(sameGroup == 1) |>
    group_by(condition, block1, target) |>
    summarise(sim_within = mean(similarity, na.rm = TRUE), .groups = "drop")

  between_means <- block_pw |>
    filter(sameGroup == 0) |>
    group_by(condition, block1, target) |>
    summarise(sim_between = mean(similarity, na.rm = TRUE), .groups = "drop")

  specificity_by_tangram <- inner_join(within_means, between_means,
                                       by = c("condition", "block1", "target")) |>
    mutate(specificity = sim_within - sim_between)

  # Bootstrapped means of the tangram-level specificity
  spec_summary <- specificity_by_tangram |>
    group_by(condition, block1) |>
    tidyboot_mean(specificity, nboot = 1000) |>
    ungroup() |>
    mutate(condition = factor(condition, levels = CONDITION_ORDER))

  # Per-tangram specificity points (jittered, faded)
  spec_tangram_plot <- specificity_by_tangram |>
    mutate(condition = factor(condition, levels = CONDITION_ORDER))

  fig3b <- ggplot(spec_summary, aes(x = block1, y = empirical_stat,
                                     color = condition, fill = condition)) +
    annotate("rect", xmin = PHASE_BOUNDARY, xmax = Inf,
             ymin = -Inf, ymax = Inf, fill = "gray90", alpha = 0.5) +
    geom_vline(xintercept = PHASE_BOUNDARY, color = "gray70", linetype = "dotted") +
    geom_hline(yintercept = 0, color = "gray", linetype = "dotted", alpha = 0.5) +
    geom_point(data = spec_tangram_plot,
               aes(x = block1, y = specificity, color = condition),
               alpha = 0.15, size = 1, position = position_jitter(width = 0.15)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12,
                color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    scale_color_condition(labels = function(x) format_condition(x), name = "Condition") +
    scale_fill_condition(labels = function(x) format_condition(x), name = "Condition") +
    scale_x_continuous(breaks = scales::breaks_width(1), limits = c(1, NA)) +
    coord_cartesian(ylim = c(-0.05, 0.85)) +
    labs(x = "Block",
         y = "Group specificity") +
    theme(legend.position = "none")
}
```

# 6. Phase 1 vs Phase 2 Comparisons (SI Figure 4)

These plots correspond directly to the paper's primary analyses, which compare
the final three blocks of Phase 1 (baseline) to the final three blocks of
Phase 2 (post-manipulation). Group-specificity is computed from pairwise
cosine similarity between speaker descriptions within the same window, using
the model described in the outcome-neutral criteria (similarity ~ same_group +
random effects for tangram and speakers).

Group-specificity coefficient (the same_group effect from the pairwise
similarity model) for Phase 1 final vs Phase 2 final, by condition. Lines
connect the same game across phases. This is the primary dependent variable
for H1 and H2.

NOTE: I couldnt figure out how to use patchwork right for the dimensions here so ive edited the output aspect stuff on illustrator

```{r si-fig-phase-comparison, fig.width=6.5, fig.height=3.5}
if (has_pairwise) {
  # Fit group-specificity model per game per window
  gs_by_window <- map_dfr(c("phase1_final", "phase2_final"), function(w) {
    pw_w <- pairwise_sim |> filter(window == w)
    gs <- fit_group_specificity(pw_w)
    gs$window <- w
    gs
  }) |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    mutate(
      phase = ifelse(window == "phase1_final", "Phase 1\n(final 3 blocks)",
                     "Phase 2\n(final 3 blocks)"),
      phase = factor(phase, levels = c("Phase 1\n(final 3 blocks)",
                                        "Phase 2\n(final 3 blocks)")),
      condition = factor(condition, levels = CONDITION_ORDER)
    )

  fig4 <- ggplot(gs_by_window, aes(x = phase, y = coefficient,
                                    color = condition)) +
    geom_line(aes(group = interaction(gameId, condition)), linewidth = 0.6, alpha = 0.6) +
    geom_point(size = 3) +
    scale_color_condition(labels = function(x) format_condition(x), name = "Condition") +
    coord_cartesian(ylim = c(-0.05, 0.85)) +
    labs(x = "Phase", y = "Group-specificity") +
    theme(legend.position = "none")

}
```

```{r si-fig-group-specificity-combined, fig.width=12, fig.height=7.7}
if (exists("fig3a") && exists("fig3b") && exists("fig4")) {
  # Extract condition legend from fig3b
  p_for_legend <- fig3b +
    theme(legend.position = "bottom",
          legend.text = element_text(size = 14),
          legend.title = element_text(size = 14),
          legend.key.size = unit(1, "cm"),
          legend.box.margin = margin(t = -10))
  legend_grob <- ggplotGrob(p_for_legend)$grobs[[
    which(ggplotGrob(p_for_legend)$layout$name == "guide-box-bottom")
  ]]

  design <- "
    AAAAAAAAAA
    AAAAAAAAAA
    BBBBB#CCCC
    BBBBB#CCCC
    BBBBB#CCCC
    DDDDDDDDDD
  "

  fig_specificity_combined <- fig3a + fig3b + fig4 +
    wrap_elements(legend_grob) +
    plot_layout(design = design,
                heights = c(0.65, 0.65, 0.65, 0.5, 0.5, 0.16)) +
    plot_annotation(tag_levels = list(c("a", "b", "c", "")),
                    theme = theme(plot.margin = margin(t = 8))) &
    theme(plot.tag = element_text(face = "bold"),
          plot.tag.position = c(0, 1.07))

  print(fig_specificity_combined)
  save_fig(fig_specificity_combined, "SI_group_specificity.pdf",
           width = 12, height = 7.7)
}
```

# 7. Secondary Analysis Previews

Pilot previews of the secondary analyses described in the paper. With only
1 game per condition these are purely descriptive — included to show the
data pipeline works end-to-end and to give a qualitative sense of the patterns.

**Panel (a) Referential accuracy:** Listener accuracy over both phases by
condition. Gray shading marks Phase 2. Thin lines show per-group trajectories.
Dashed line marks chance (1/16).

**Panel (b) Description length:** Speaker word count over both phases by
condition. Thin lines show per-group trajectories.

**Panel (c) Social-signaling accuracy:** Social guess accuracy (social
conditions only). Data exists only in Phase 2; x-axis matches the other
panels. Dashed line marks chance (50%).

**Panel (d) Convention stability:** Cosine similarity between successive
descriptions of the same tangram by the same speaker, across both phases.
Higher values = more stable conventions. Thin lines show per-group
trajectories.

```{r si-fig-secondary-combined, fig.width=12, fig.height=8}
# --- Data prep ---

# (a) Referential accuracy
all_listener <- trials |>
  filter(role == "listener") |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  continuous_block() |>
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    condition = factor(condition, levels = CONDITION_ORDER),
    group = originalGroup,
    block1 = block + 1
  )

acc_group_all <- all_listener |>
  group_by(condition, gameId, group, block1) |>
  summarise(mean_acc = mean(correct, na.rm = TRUE), .groups = "drop")

# Split to avoid tidyboot failing on zero-variance groups
acc_data <- all_listener |> filter(!is.na(correct))
acc_variance <- acc_data |>
  group_by(condition, block1) |>
  summarise(sd = sd(correct), m = mean(correct), .groups = "drop")

acc_boot <- acc_data |>
  semi_join(acc_variance |> filter(sd > 0), by = c("condition", "block1")) |>
  group_by(condition, block1) |>
  tidyboot_mean(correct, nboot = 1000) |>
  ungroup()

acc_novar <- acc_variance |>
  filter(sd == 0) |>
  transmute(condition, block1,
            empirical_stat = m, ci_lower = m, ci_upper = m)

acc_block_all <- bind_rows(acc_boot, acc_novar)

# (b) Description length
all_utts <- speaker_utts |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  continuous_block() |>
  mutate(
    condition = factor(condition, levels = CONDITION_ORDER),
    group = originalGroup,
    block1 = block + 1
  )

len_group_all <- all_utts |>
  group_by(condition, gameId, group, block1) |>
  summarise(mean_length = mean(uttLength, na.rm = TRUE), .groups = "drop")

len_block_all <- all_utts |>
  group_by(condition, block1) |>
  tidyboot_mean(uttLength, nboot = 1000) |>
  ungroup()

# (c) Social-signaling accuracy — use continuous block (Phase 2 only)
social_guesses_cond <- social_guesses |>
  filter(!is.na(socialGuessCorrect)) |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  mutate(
    correct = as.numeric(socialGuessCorrect == TRUE),
    condition = factor(condition, levels = CONDITION_ORDER),
    group = originalGroup,
    block1 = blockNum + PHASE2_OFFSET + 1  # align to continuous x-axis
  )

sg_group <- social_guesses_cond |>
  group_by(condition, gameId, group, block1) |>
  summarise(mean_acc = mean(correct, na.rm = TRUE), .groups = "drop")

sg_block <- social_guesses_cond |>
  group_by(condition, block1) |>
  tidyboot_mean(correct, nboot = 1000) |>
  ungroup()

# (d) Convention stability (adjacent similarity across both phases)
all_adjacent <- adjacent_sim |>
  filter(!is.na(simAdjacent)) |>
  left_join(games |> select(gameId, condition), by = "gameId") |>
  continuous_block() |>
  mutate(
    condition = factor(condition, levels = CONDITION_ORDER),
    group = originalGroup,
    block1 = block + 1
  )

# --- Shared phase boundary layers ---
phase2_bg <- list(
  annotate("rect", xmin = PHASE_BOUNDARY, xmax = Inf,
           ymin = -Inf, ymax = Inf, fill = "gray90", alpha = 0.5),
  geom_vline(xintercept = PHASE_BOUNDARY, color = "gray70", linetype = "dotted"),
  scale_x_continuous(breaks = 1:12, limits = c(1, 12), labels = as.integer),
  scale_color_condition(labels = function(x) format_condition(x), name = NULL,
                        limits = CONDITION_ORDER, drop = FALSE),
  scale_fill_condition(labels = function(x) format_condition(x), name = NULL,
                       limits = CONDITION_ORDER, drop = FALSE)
)

# --- Panel (a): Referential accuracy ---
p_acc <- ggplot(acc_block_all, aes(x = block1, y = empirical_stat,
                                    color = condition, fill = condition)) +
  phase2_bg +
  geom_line(data = acc_group_all,
            aes(x = block1, y = mean_acc,
                group = interaction(gameId, group, condition)),
            alpha = 0.25, linewidth = 0.3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 1/16, linetype = "dashed", color = "gray50") +
  labs(x = "Block", y = "Proportion correct",
       title = "Referential accuracy") +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "none")

# --- Panel (b): Description length ---
p_len <- ggplot(len_block_all, aes(x = block1, y = empirical_stat,
                                    color = condition, fill = condition)) +
  phase2_bg +
  geom_line(data = len_group_all,
            aes(x = block1, y = mean_length,
                group = interaction(gameId, group, condition)),
            alpha = 0.25, linewidth = 0.3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(x = "Block", y = "Words per description",
       title = "Description length") +
  theme(legend.position = "none")

# --- Panel (c): Social-signaling accuracy ---
p_sg <- ggplot(sg_block, aes(x = block1, y = empirical_stat,
                              color = condition, fill = condition)) +
  phase2_bg +
  geom_line(data = sg_group,
            aes(x = block1, y = mean_acc,
                group = interaction(gameId, group, condition)),
            alpha = 0.3, linewidth = 0.3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  geom_hline(yintercept = 0.5, linetype = "dashed", color = "gray50") +
  labs(x = "Block", y = "Proportion correct",
       title = "Social-signaling accuracy") +
  scale_y_continuous(limits = c(0, 1)) +
  theme(legend.position = "none")

# --- Panel (d): Convention stability ---
adj_group_all <- all_adjacent |>
  group_by(condition, gameId, group, block1) |>
  summarise(mean_sim = mean(simAdjacent, na.rm = TRUE), .groups = "drop")

adj_block_all <- all_adjacent |>
  group_by(condition, block1) |>
  tidyboot_mean(simAdjacent, nboot = 1000) |>
  ungroup()

p_stab <- ggplot(adj_block_all, aes(x = block1, y = empirical_stat,
                                     color = condition, fill = condition)) +
  phase2_bg +
  geom_line(data = adj_group_all,
            aes(x = block1, y = mean_sim,
                group = interaction(gameId, group, condition)),
            alpha = 0.25, linewidth = 0.3) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12, color = NA) +
  geom_line(linewidth = 0.8) +
  geom_point(size = 1.8) +
  labs(x = "Block", y = "Cosine similarity\n(adjacent descriptions)",
       title = "Convention stability") +
  theme(legend.position = "none")

# --- Combine with shared legend at bottom ---
# Extract legend from panel a (has all 4 conditions) as a standalone grob
p_for_legend <- p_acc +
  scale_color_condition(labels = function(x) format_condition(x), name = "Condition") +
  guides(fill = "none") +
  theme(legend.position = "bottom",
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.key.size = unit(1.2, "cm"))
legend_grob <- ggplotGrob(p_for_legend)$grobs[[
  which(ggplotGrob(p_for_legend)$layout$name == "guide-box-bottom")
]]

fig_secondary <- (p_acc + p_len) / (p_sg + p_stab) / wrap_elements(legend_grob) +
  plot_layout(heights = c(1, 1, 0.08)) +
  plot_annotation(tag_levels = list(c("a", "b", "c", "d", ""))) &
  theme(plot.tag = element_text(face = "bold", size = 20))

print(fig_secondary)
save_fig(fig_secondary, "SI_secondary_analyses.pdf", width = 12, height = 8)
```

# 8. H3c: Description Properties Preview

Descriptive preview of the H3c measures comparing `social_mixed` vs
`social_first` in Phase 1. H3c predicts that anticipating a social-identification
task leads to descriptions that are less concrete, more lexically unique, and
use lower-frequency vocabulary.

All three measures are precomputed in Python (`compute_embeddings.py`) with
consistent tokenization: each utterance is lowercased, split into alphabetic
sequences, filtered for standard English stopwords, and only tokens with
length > 1 are retained as *content words*.

- **Concreteness**: the proportion of content words matching concrete term
  stems (prefix matching) from Boyce et al. 2024. Four categories: geometric
  (e.g., *triangle*, *square*, *diamond*), body parts (e.g., *arm*, *head*,
  *leg*), positional (e.g., *left*, *right*, *top*), and posture (e.g.,
  *kick*, *kneel*, *stand*).
- **Lexical uniqueness**: the proportion of content words in a description
  that do not appear in any other group's descriptions for the same tangram
  in the same game.
- **Word frequency**: the mean Zipf frequency of content words, computed via
  the `wordfreq` Python package (which draws on SUBTLEX and other corpora).
  Common words score higher (e.g., "the" ≈ 7.7); rare words score lower.

With 1 game per social condition, these are purely descriptive — included
to verify the H3c analysis pipeline and give a qualitative sense of the patterns.

```{r h3c-pilot-data}
# All three H3c measures are precomputed in Python with consistent tokenization
# (same stopwords and content-word extraction). See compute_embeddings.py.
has_desc_props <- file.exists(file.path(data_dir, "description_properties.csv"))
has_lex_uniq <- file.exists(file.path(data_dir, "lexical_uniqueness.csv"))

if (!has_desc_props || !has_lex_uniq) {
  cat("Missing H3c data files. Run compute_description_properties() and compute_lexical_uniqueness() first.\n")
} else {
  desc_props <- read_csv(file.path(data_dir, "description_properties.csv"), show_col_types = FALSE)
  lex_uniq <- read_csv(file.path(data_dir, "lexical_uniqueness.csv"), show_col_types = FALSE)

  # Merge properties + uniqueness, filter to final Phase 1 block, social conditions
  p1_final_social <- desc_props |>
    left_join(
      lex_uniq |> select(gameId, playerId, target, blockNum, phaseNum, uniqueness),
      by = c("gameId", "playerId", "target", "blockNum", "phaseNum")
    ) |>
    filter(phaseNum == 1) |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    filter(condition %in% c("social_mixed", "social_first")) |>
    group_by(gameId, playerId, target) |>
    filter(blockNum == max(blockNum)) |>
    ungroup() |>
    mutate(condition = factor(condition, levels = c("social_mixed", "social_first")))
}
```

```{r h3c-plot, fig.width=14, fig.height=4.5}
if (exists("p1_final_social") && nrow(p1_final_social) > 0) {
  p1_h3c <- p1_final_social |>
    filter(!is.na(concreteness), !is.na(uniqueness)) |>
    select(condition, concreteness, uniqueness, mean_zipf_freq)

  # Panel A: Concrete terms
  p_concrete <- ggplot(p1_h3c, aes(x = condition, y = concreteness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = "Condition", y = "Proportion", title = "Concrete terms") +
    theme(legend.position = "none")

  # Panel B: Lexical uniqueness
  p_unique <- ggplot(p1_h3c, aes(x = condition, y = uniqueness, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = "Condition", y = "Proportion", title = "Lexical uniqueness") +
    theme(legend.position = "none")

  # Panel C: Word frequency
  p_freq <- ggplot(p1_h3c |> filter(!is.na(mean_zipf_freq)),
                   aes(x = condition, y = mean_zipf_freq, fill = condition)) +
    geom_violin(alpha = 0.3) +
    geom_jitter(width = 0.15, size = 1.5, alpha = 0.4) +
    stat_summary(fun.data = mean_cl_boot, color = "black", size = 0.8) +
    scale_fill_condition() +
    scale_x_discrete(labels = function(x) format_condition(x)) +
    labs(x = "Condition", y = "Mean Zipf frequency", title = "Word frequency") +
    theme(legend.position = "none")

  fig_h3c <- p_concrete + p_unique + p_freq +
    plot_annotation(
      tag_levels = "a"
    ) &
    theme(plot.tag = element_text(face = "bold", size = 20))

  print(fig_h3c)
  save_fig(fig_h3c, "SI_h3c_description_properties.pdf", width = 14, height = 4.5)
}
```

```{r h3c-descriptive}
if (exists("p1_final_social") && nrow(p1_final_social) > 0) {
  cat("=== H3c descriptive statistics (final Phase 1 block, social conditions) ===\n\n")
  p1_final_social |>
    group_by(condition) |>
    summarise(
      n = n(),
      concreteness_M = round(mean(concreteness, na.rm = TRUE), 3),
      concreteness_SD = round(sd(concreteness, na.rm = TRUE), 3),
      uniqueness_M = round(mean(uniqueness, na.rm = TRUE), 3),
      uniqueness_SD = round(sd(uniqueness, na.rm = TRUE), 3),
      zipf_freq_M = round(mean(mean_zipf_freq, na.rm = TRUE), 3),
      zipf_freq_SD = round(sd(mean_zipf_freq, na.rm = TRUE), 3),
      .groups = "drop"
    ) |>
    knitr::kable()
}
```

# 9. Summary Table

```{r summary-table}
summary_rows <- list()

# Length
s_len <- coef(summary(m_length))
summary_rows[[1]] <- tibble(
  Criterion = "Description length decreases",
  `β` = round(s_len["blockNum_c", "Estimate"], 3),
  SE = round(s_len["blockNum_c", "Std. Error"], 3),
  `t` = round(s_len["blockNum_c", "t value"], 2),
  Result = ifelse(s_len["blockNum_c", "Estimate"] < 0, "PASS", "FAIL")
)

# Accuracy
s_acc <- coef(summary(m_accuracy))
summary_rows[[2]] <- tibble(
  Criterion = "Listener accuracy increases",
  `β` = round(s_acc["blockNum_c", "Estimate"], 3),
  SE = round(s_acc["blockNum_c", "Std. Error"], 3),
  `t` = round(s_acc["blockNum_c", "t value"], 2),
  Result = ifelse(s_acc["blockNum_c", "Estimate"] > 0, "PASS", "FAIL")
)

# Adjacent similarity
if (has_adjacent && exists("m_adjacent")) {
  s_adj <- coef(summary(m_adjacent))
  summary_rows[[3]] <- tibble(
    Criterion = "Convention stability increases",
    `β` = round(s_adj["blockNum_c", "Estimate"], 3),
    SE = round(s_adj["blockNum_c", "Std. Error"], 3),
    `t` = round(s_adj["blockNum_c", "t value"], 2),
    Result = ifelse(s_adj["blockNum_c", "Estimate"] > 0, "PASS", "FAIL")
  )
}

# Group-specificity
if (exists("perm_results")) {
  prop_sig <- mean(perm_results$p_value < 0.05)
  summary_rows[[length(summary_rows) + 1]] <- tibble(
    Criterion = "Group-specificity (≥80% significant)",
    `β` = NA,
    SE = NA,
    `t` = NA,
    Result = sprintf("%d/%d games sig. — %s",
                     sum(perm_results$p_value < 0.05),
                     nrow(perm_results),
                     ifelse(prop_sig >= 0.8, "PASS", "FAIL"))
  )
}

bind_rows(summary_rows) |>
  knitr::kable(caption = "Outcome-neutral criteria summary")
```

# 9. Descriptive Statistics for LaTeX

Collect all numbers needed for the SI text in one place.

```{r latex-numbers}
cat("====== Numbers for LaTeX SI section ======\n\n")

# Sample
cat(sprintf("Games: %d, Players: %d\n", n_games, n_players))
cat(sprintf("Conditions: %s\n", paste(games$condition, collapse = ", ")))

# Duration
cat(sprintf("\nMedian session duration: %.1f min (range: %.1f–%.1f)\n",
            median(duration_by_game$duration_min),
            min(duration_by_game$duration_min),
            max(duration_by_game$duration_min)))

# Completion
n_complete <- sum(games$activeGroups == 3)
cat(sprintf("Games with all groups active: %d/%d\n", n_complete, n_games))

# Description length
s_len <- coef(summary(m_length))
cat(sprintf("\nDescription length: β = %.3f, SE = %.3f, t = %.2f\n",
            s_len["blockNum_c", "Estimate"],
            s_len["blockNum_c", "Std. Error"],
            s_len["blockNum_c", "t value"]))
cat(sprintf("  Block 1: M = %.1f words; Block 6: M = %.1f words\n",
            mean(length_first), mean(length_last)))

# Accuracy
s_acc <- coef(summary(m_accuracy))
cat(sprintf("\nAccuracy: β = %.3f, SE = %.3f, t = %.2f\n",
            s_acc["blockNum_c", "Estimate"],
            s_acc["blockNum_c", "Std. Error"],
            s_acc["blockNum_c", "t value"]))
cat(sprintf("  Block 1: M = %.2f; Block 6: M = %.2f\n",
            mean(acc_first, na.rm = TRUE),
            mean(acc_last, na.rm = TRUE)))

# Adjacent similarity
if (has_adjacent && exists("m_adjacent")) {
  s_adj <- coef(summary(m_adjacent))
  cat(sprintf("\nAdjacent similarity: β = %.3f, SE = %.3f, t = %.2f\n",
              s_adj["blockNum_c", "Estimate"],
              s_adj["blockNum_c", "Std. Error"],
              s_adj["blockNum_c", "t value"]))
}

# Group-specificity
if (exists("perm_results")) {
  cat(sprintf("\nGroup-specificity: %d/%d games significant (p < .05)\n",
              sum(perm_results$p_value < 0.05), nrow(perm_results)))
  perm_results |>
    left_join(games |> select(gameId, condition), by = "gameId") |>
    select(condition, obs_coefficient, p_value) |>
    mutate(across(where(is.numeric), ~round(., 4))) |>
    print()
}

# Idle / dropout
n_idle <- nrow(players |> filter(idleRounds > 0))
n_inactive <- nrow(players |> filter(isActive == FALSE | isActive == "false"))
cat(sprintf("\nPlayers with idle rounds: %d/%d\n", n_idle, n_players))
cat(sprintf("Players marked inactive: %d/%d\n", n_inactive, n_players))

# Total trials and messages
cat(sprintf("\nTotal trials: %d\n", nrow(trials)))
cat(sprintf("Total messages: %d\n", nrow(messages)))
cat(sprintf("Speaker utterances: %d\n", nrow(speaker_utts)))
```

# 10. Write Stats to LaTeX

Write all statistics as `\newcommand` definitions for `\input` in the paper.

```{r write-stats-tex}
stats_dir <- here::here("paper", "stats")
dir.create(stats_dir, showWarnings = FALSE, recursive = TRUE)

cmds <- c()
cmd <- function(name, value) paste0("\\newcommand{\\", name, "}{", value, "}")

# Sample
cmds <- c(cmds,
  cmd("pilotNGames", n_games),
  cmd("pilotNPlayers", n_players)
)

# Duration
cmds <- c(cmds,
  cmd("pilotDuration", sprintf("%.1f", median(duration_by_game$duration_min))),
  cmd("pilotDurationMin", sprintf("%.1f", min(duration_by_game$duration_min))),
  cmd("pilotDurationMax", sprintf("%.1f", max(duration_by_game$duration_min)))
)

# Completion
cmds <- c(cmds,
  cmd("pilotNComplete", sum(games$activeGroups == 3))
)

# Description length model
s_len <- coef(summary(m_length))
cmds <- c(cmds,
  cmd("pilotBetaLen", sprintf("%.3f", s_len["blockNum_c", "Estimate"])),
  cmd("pilotSELen", sprintf("%.3f", s_len["blockNum_c", "Std. Error"])),
  cmd("pilotTLen", sprintf("%.2f", s_len["blockNum_c", "t value"])),
  cmd("pilotMOneLen", sprintf("%.1f", mean(length_first))),
  cmd("pilotMSixLen", sprintf("%.1f", mean(length_last)))
)

# Accuracy model
s_acc <- coef(summary(m_accuracy))
cmds <- c(cmds,
  cmd("pilotBetaAcc", sprintf("%.3f", s_acc["blockNum_c", "Estimate"])),
  cmd("pilotSEAcc", sprintf("%.3f", s_acc["blockNum_c", "Std. Error"])),
  cmd("pilotTAcc", sprintf("%.2f", s_acc["blockNum_c", "t value"])),
  cmd("pilotMOneAcc", sprintf("%.2f", mean(acc_first, na.rm = TRUE))),
  cmd("pilotMSixAcc", sprintf("%.2f", mean(acc_last, na.rm = TRUE)))
)

# Adjacent similarity model
if (has_adjacent && exists("m_adjacent")) {
  s_adj <- coef(summary(m_adjacent))
  cmds <- c(cmds,
    cmd("pilotBetaSim", sprintf("%.3f", s_adj["blockNum_c", "Estimate"])),
    cmd("pilotSESim", sprintf("%.3f", s_adj["blockNum_c", "Std. Error"])),
    cmd("pilotTSim", sprintf("%.2f", s_adj["blockNum_c", "t value"]))
  )
}

# Group-specificity permutation
if (exists("perm_results")) {
  cmds <- c(cmds,
    cmd("pilotNSig", sum(perm_results$p_value < 0.05))
  )
}

# Idle / dropout
cmds <- c(cmds,
  cmd("pilotNIdle", nrow(players |> filter(idleRounds > 0))),
  cmd("pilotNInactive", nrow(players |> filter(isActive == FALSE | isActive == "false")))
)

header <- "% AUTO-GENERATED by analysis/SI_pilot.qmd — do not edit by hand."
writeLines(c(header, cmds), file.path(stats_dir, "pilot.tex"))
cat("Wrote", length(cmds), "commands to paper/stats/pilot.tex\n")
```
