---
title: "SI: Pilot Data"
format:
  html:
    toc: true
    code-fold: true
---

This notebook generates the numbers and figures for the Supplementary Information
pilot data section. Run it to extract statistics for the LaTeX writeup.

```{r setup, message=FALSE, warning=FALSE}
library(lme4)
library(lmerTest)
library(tidyboot)
source(here::here("analysis", "config.R"))

games <- read_csv(file.path(data_dir, "games.csv"), show_col_types = FALSE)
players <- read_csv(file.path(data_dir, "players.csv"), show_col_types = FALSE)
trials <- read_csv(file.path(data_dir, "trials.csv"), show_col_types = FALSE)
messages <- read_csv(file.path(data_dir, "messages.csv"), show_col_types = FALSE)
speaker_utts <- read_csv(file.path(data_dir, "speaker_utterances.csv"), show_col_types = FALSE)
social_guesses <- read_csv(file.path(data_dir, "social_guesses.csv"), show_col_types = FALSE)

has_adjacent <- file.exists(file.path(data_dir, "adjacent_similarities.csv"))
has_pairwise <- file.exists(file.path(data_dir, "pairwise_similarities.csv"))

adjacent_sim <- if (has_adjacent) read_csv(file.path(data_dir, "adjacent_similarities.csv"), show_col_types = FALSE) else tibble()
pairwise_sim <- if (has_pairwise) read_csv(file.path(data_dir, "pairwise_similarities.csv"), show_col_types = FALSE) else tibble()
```

# 1. Sample Description

```{r sample-description}
n_games <- nrow(games)
n_players <- nrow(players)
n_per_game <- n_players / n_games

cat(sprintf("Pilot games: %d (1 per condition)\n", n_games))
cat(sprintf("Total players: %d (%d per game)\n", n_players, n_per_game))
cat("\nConditions:\n")
games %>%
  select(gameId, condition, activeGroups, phase1Blocks, phase2Blocks) %>%
  knitr::kable()
```

```{r completion-rates}
# Active groups (3 = all groups survived, <3 = dropout)
games_complete <- games %>%
  mutate(all_groups_active = activeGroups == 3)

cat(sprintf("\nGames with all 3 groups active through completion: %d / %d\n",
            sum(games_complete$all_groups_active), n_games))

# Per-game active groups
games %>%
  select(condition, activeGroups) %>%
  knitr::kable(col.names = c("Condition", "Active groups at end"))
```

```{r session-duration}
# Compute session duration from message timestamps
duration_by_game <- messages %>%
  group_by(gameId) %>%
  summarise(
    first_msg = min(timestamp, na.rm = TRUE),
    last_msg = max(timestamp, na.rm = TRUE),
    duration_min = (last_msg - first_msg) / 60000,
    .groups = "drop"
  ) %>%
  left_join(games %>% select(gameId, condition), by = "gameId")

cat("Session duration (minutes, from first to last message):\n")
duration_by_game %>%
  select(condition, duration_min) %>%
  mutate(duration_min = round(duration_min, 1)) %>%
  knitr::kable(col.names = c("Condition", "Duration (min)"))

cat(sprintf("\nMedian session duration: %.1f minutes\n", median(duration_by_game$duration_min)))
cat(sprintf("Range: %.1f - %.1f minutes\n",
            min(duration_by_game$duration_min), max(duration_by_game$duration_min)))
```

```{r idle-players}
# Check for idle players
idle_players <- players %>%
  filter(idleRounds > 0) %>%
  select(gameId, playerId, originalGroup, idleRounds, isActive) %>%
  left_join(games %>% select(gameId, condition), by = "gameId")

if (nrow(idle_players) > 0) {
  cat(sprintf("Players with idle rounds: %d\n", nrow(idle_players)))
  idle_players %>%
    select(condition, originalGroup, idleRounds, isActive) %>%
    knitr::kable()
} else {
  cat("No idle players detected.\n")
}
```

# 2. Outcome-Neutral Criteria

## Prep data

```{r prep-data}
# Phase 1 speaker utterances
p1_utts <- speaker_utts %>%
  filter(phaseNum == 1) %>%
  mutate(
    blockNum_c = blockNum - mean(blockNum, na.rm = TRUE),
    log_uttLength = log(uttLength + 1),
    participant = playerId,
    group = originalGroup,
    tangram = target
  )

# Phase 1 listener accuracy
p1_listener <- trials %>%
  filter(phaseNum == 1, role == "listener") %>%
  mutate(
    correct = as.numeric(clickedCorrect == TRUE | clickedCorrect == "true"),
    group = originalGroup,
    tangram = target
  )

p1_ref_accuracy <- p1_listener %>%
  group_by(gameId, roundId, blockNum, tangram, group) %>%
  summarise(
    ref_accuracy = mean(correct, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    blockNum_c = blockNum - mean(blockNum, na.rm = TRUE)
  )

# Phase 1 adjacent similarities
if (has_adjacent) {
  p1_adjacent <- adjacent_sim %>%
    filter(phaseNum == 1) %>%
    mutate(
      blockNum_c = blockNum - mean(blockNum, na.rm = TRUE),
      participant = playerId,
      group = originalGroup,
      tangram = target
    )
}
```

## 2a. Description Length Decreases

```{r model-length}
m_length <- lmer(
  log_uttLength ~ blockNum_c + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
  data = p1_utts,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

s <- summary(m_length)
coefs <- coef(s)
cat("=== Description Length Model ===\n")
print(s)
cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
            coefs["blockNum_c", "Estimate"],
            coefs["blockNum_c", "Std. Error"],
            coefs["blockNum_c", "t value"]))
```

```{r descriptive-length}
# Descriptive stats for text (by blockNum, not repNum)
length_first <- p1_utts %>% filter(blockNum == min(blockNum)) %>% pull(uttLength)
length_last <- p1_utts %>% filter(blockNum == max(blockNum)) %>% pull(uttLength)

cat(sprintf("First block: M = %.1f words (SD = %.1f)\n",
            mean(length_first), sd(length_first)))
cat(sprintf("Last block: M = %.1f words (SD = %.1f)\n",
            mean(length_last), sd(length_last)))
```

## 2b. Listener Accuracy Increases

```{r model-accuracy}
m_accuracy <- lmer(
  ref_accuracy ~ blockNum_c + (blockNum_c | group) + (blockNum_c | tangram),
  data = p1_ref_accuracy,
  control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
)

s <- summary(m_accuracy)
coefs <- coef(s)
cat("=== Listener Accuracy Model ===\n")
print(s)
cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
            coefs["blockNum_c", "Estimate"],
            coefs["blockNum_c", "Std. Error"],
            coefs["blockNum_c", "t value"]))
```

```{r descriptive-accuracy}
acc_first <- p1_ref_accuracy %>% filter(blockNum == min(blockNum)) %>% pull(ref_accuracy)
acc_last <- p1_ref_accuracy %>% filter(blockNum == max(blockNum)) %>% pull(ref_accuracy)

cat(sprintf("First block accuracy: M = %.2f (SD = %.2f)\n",
            mean(acc_first, na.rm = TRUE), sd(acc_first, na.rm = TRUE)))
cat(sprintf("Last block accuracy: M = %.2f (SD = %.2f)\n",
            mean(acc_last, na.rm = TRUE), sd(acc_last, na.rm = TRUE)))
cat(sprintf("Chance level: %.3f (1/6)\n", 1/6))
```

## 2c. Convention Stability Increases

```{r model-adjacent}
if (has_adjacent) {
  p1_adjacent_valid <- p1_adjacent %>% filter(!is.na(simAdjacent))

  m_adjacent <- lmer(
    simAdjacent ~ blockNum_c + (blockNum_c | participant) + (blockNum_c | group) + (blockNum_c | tangram),
    data = p1_adjacent_valid,
    control = lmerControl(optimizer = "bobyqa", optCtrl = list(maxfun = 100000))
  )

  s <- summary(m_adjacent)
  coefs <- coef(s)
  cat("=== Adjacent Similarity Model ===\n")
  print(s)
  cat(sprintf("\nFor LaTeX: β = %.3f, SE = %.3f, t = %.2f\n",
              coefs["blockNum_c", "Estimate"],
              coefs["blockNum_c", "Std. Error"],
              coefs["blockNum_c", "t value"]))
} else {
  cat("Skipping: adjacent_similarities.csv not found.\n")
}
```

## 2d. Group-Specificity Exceeds Chance

```{r compute-group-specificity, warning=FALSE}
source(here::here("analysis", "group_specificity.R"))

if (!has_pairwise) {
  cat("Skipping: pairwise_similarities.csv not found.\n")
} else {
  p1_pairwise <- pairwise_sim %>% filter(window == "phase1_final")

  if (nrow(p1_pairwise) > 0) {
    # Uses cached RDS if available; pass force = TRUE to recompute
    gs_cached <- compute_group_specificity(p1_pairwise, data_dir)
    gs_results <- gs_cached$gs_results
    perm_results <- gs_cached$perm_results

    cat("Group-specificity coefficients (Phase 1 final):\n")
    gs_results %>%
      left_join(games %>% select(gameId, condition), by = "gameId") %>%
      select(condition, coefficient, std_error, t_value) %>%
      mutate(across(where(is.numeric), ~round(., 3))) %>%
      knitr::kable()

    cat("\nPermutation test results:\n")
    perm_results %>%
      left_join(games %>% select(gameId, condition), by = "gameId") %>%
      select(condition, obs_coefficient, perm_mean, p_value) %>%
      mutate(across(where(is.numeric), ~round(., 4))) %>%
      knitr::kable()

    prop_significant <- mean(perm_results$p_value < 0.05)
    cat(sprintf("\nGames with significant group-specificity: %d / %d (%.0f%%)\n",
                sum(perm_results$p_value < 0.05), nrow(perm_results),
                prop_significant * 100))
    cat(sprintf("Criterion (≥80%%): %s\n",
                ifelse(prop_significant >= 0.8, "PASS", "FAIL")))
  }
}
```

# 3. Combined Figure: Convention Formation (SI Figure 1)

Three-panel figure showing description length, listener accuracy, and convention
stability over Phase 1 blocks.

```{r si-fig-convention, fig.width=14, fig.height=4.5}
library(patchwork)

# Panel A: Description length (by blockNum, not repNum)
length_by_block <- p1_utts %>%
  group_by(blockNum) %>%
  tidyboot_mean(uttLength, nboot = 1000)

p_length <- ggplot(length_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  labs(x = "Block", y = "Words per description", title = "Description length") +
  scale_x_continuous(breaks = 1:6)

# Panel B: Listener accuracy
acc_by_block <- p1_ref_accuracy %>%
  filter(!is.na(ref_accuracy), !is.na(blockNum)) %>%
  group_by(blockNum) %>%
  tidyboot_mean(ref_accuracy, nboot = 1000)

p_accuracy <- ggplot(acc_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
  geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
  geom_line(linewidth = 1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 1/6, linetype = "dashed", color = "gray50") +
  annotate("text", x = 6, y = 1/6 + 0.03,
           label = "Chance (1/6)", hjust = 1, color = "gray50", size = 3.5) +
  labs(x = "Block", y = "Proportion correct", title = "Listener accuracy") +
  scale_x_continuous(breaks = 1:6) +
  scale_y_continuous(limits = c(0, 1))

# Panel C: Adjacent similarity
if (has_adjacent && exists("p1_adjacent")) {
  p1_adjacent_valid <- p1_adjacent %>% filter(!is.na(simAdjacent))
  sim_by_block <- p1_adjacent_valid %>%
    group_by(blockNum) %>%
    tidyboot_mean(simAdjacent, nboot = 1000)

  p_sim <- ggplot(sim_by_block, aes(x = blockNum + 1, y = empirical_stat)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.15) +
    geom_line(linewidth = 1) +
    geom_point(size = 2.5) +
    labs(x = "Block", y = "Cosine similarity", title = "Convention stability") +
    scale_x_continuous(breaks = 1:6)

  fig1 <- p_length + p_accuracy + p_sim +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 16))
} else {
  fig1 <- p_length + p_accuracy +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 16))
}

fig1

save_fig(fig1, "SI_convention_formation.pdf", width = 14, height = 4.5)
```

# 4. Permutation Test Figure (SI Figure 2)

```{r si-fig-permutation, fig.width=10, fig.height=6}
if (exists("perm_results") && nrow(perm_results) > 0) {
  perm_plot_data <- perm_results %>%
    left_join(games %>% select(gameId, condition), by = "gameId") %>%
    unnest(perm_coefs) %>%
    rename(perm_coef = perm_coefs) %>%
    mutate(condition_label = format_condition(condition))

  obs_labels <- perm_results %>%
    left_join(games %>% select(gameId, condition), by = "gameId") %>%
    mutate(
      condition_label = format_condition(condition),
      sig_label = ifelse(p_value < 0.05,
                         sprintf("p = %.3f *", p_value),
                         sprintf("p = %.3f", p_value))
    )

  fig2 <- ggplot(perm_plot_data, aes(x = perm_coef)) +
    geom_histogram(bins = 40, fill = "gray75", color = "white") +
    geom_vline(data = obs_labels,
               aes(xintercept = obs_coefficient),
               color = "#ce3045", linewidth = 1) +
    geom_text(data = obs_labels,
              aes(x = obs_coefficient, y = Inf, label = sig_label),
              vjust = 1.5, hjust = -0.1, color = "#ce3045", size = 4) +
    facet_wrap(~condition_label, scales = "free_y") +
    labs(
      x = "Group-specificity coefficient (permuted)",
      y = "Count",
      title = "Permutation test for group-specificity"
    )

  print(fig2)
  save_fig(fig2, "SI_permutation_test.pdf", width = 10, height = 6)
}
```

# 5. Group-Specificity Trajectory (SI Figure 3)

Within- vs between-group similarity over all blocks, showing how group-specificity
emerges in Phase 1 and evolves in Phase 2.

```{r load-block-pairwise}
has_block_pw <- file.exists(file.path(data_dir, "block_pairwise_similarities.csv"))
if (has_block_pw) {
  block_pw <- read_csv(file.path(data_dir, "block_pairwise_similarities.csv"),
                       show_col_types = FALSE) %>%
    left_join(games %>% select(gameId, condition), by = "gameId") %>%
    continuous_block() %>%
    mutate(
      pair_type = ifelse(sameGroup == 1, "Within group", "Between group"),
      condition = factor(condition, levels = CONDITION_ORDER)
    )
} else {
  cat("Skipping: block_pairwise_similarities.csv not found.\n")
}
```

```{r si-fig-within-between, fig.width=9, fig.height=4}
if (has_block_pw) {
  # Bootstrapped means by block, pair type, and condition
  wb_summary <- block_pw %>%
    group_by(condition, block, pair_type) %>%
    tidyboot_mean(similarity, nboot = 1000) %>%
    ungroup()

  fig3a <- ggplot(wb_summary, aes(x = block, y = empirical_stat,
                                   color = pair_type, fill = pair_type)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.2,
                color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 1.8) +
    geom_vline(xintercept = PHASE_BOUNDARY, color = "gray", linetype = "dotted",
               alpha = 0.6) +
    facet_wrap(~condition, nrow = 1,
               labeller = labeller(condition = function(x) format_condition(x))) +
    scale_fill_brewer(name = "Pair type", palette = "Accent") +
    scale_color_brewer(name = "Pair type", palette = "Accent") +
    labs(x = "Block", y = "Pairwise cosine similarity",
         title = "Within- vs between-group similarity") +
    coord_cartesian(ylim = c(0, 1)) +
    theme(legend.position = "bottom")

  print(fig3a)
  save_fig(fig3a, "SI_within_between_similarity.pdf", width = 12, height = 4)
}
```

```{r si-fig-specificity-trajectory, fig.width=8, fig.height=5}
if (has_block_pw) {
  # Compute group-specificity = within - between per block, condition, tangram
  within_means <- block_pw %>%
    filter(sameGroup == 1) %>%
    group_by(condition, block, target) %>%
    summarise(sim_within = mean(similarity, na.rm = TRUE), .groups = "drop")

  between_means <- block_pw %>%
    filter(sameGroup == 0) %>%
    group_by(condition, block, target) %>%
    summarise(sim_between = mean(similarity, na.rm = TRUE), .groups = "drop")

  specificity_by_tangram <- inner_join(within_means, between_means,
                                       by = c("condition", "block", "target")) %>%
    mutate(specificity = sim_within - sim_between)

  # Bootstrapped means of the tangram-level specificity
  spec_summary <- specificity_by_tangram %>%
    group_by(condition, block) %>%
    tidyboot_mean(specificity, nboot = 1000) %>%
    ungroup() %>%
    mutate(condition = factor(condition, levels = CONDITION_ORDER))

  fig3b <- ggplot(spec_summary, aes(x = block, y = empirical_stat,
                                     color = condition, fill = condition)) +
    geom_ribbon(aes(ymin = ci_lower, ymax = ci_upper), alpha = 0.12,
                color = NA) +
    geom_line(linewidth = 0.8) +
    geom_point(size = 2) +
    geom_hline(yintercept = 0, color = "gray", linetype = "dotted", alpha = 0.5) +
    geom_vline(xintercept = PHASE_BOUNDARY, color = "gray", linetype = "dotted",
               alpha = 0.6) +
    scale_color_condition(labels = function(x) format_condition(x), name = NULL) +
    scale_fill_condition(labels = function(x) format_condition(x), name = NULL) +
    labs(x = "Block",
         y = "Group specificity\n(within \u2212 between similarity)",
         title = "Group-specificity trajectory across phases") +
    theme(legend.position = "bottom")

  print(fig3b)
  save_fig(fig3b, "SI_specificity_trajectory.pdf", width = 8, height = 5)
}
```

```{r si-fig-specificity-combined, fig.width=14, fig.height=5}
if (has_block_pw && exists("fig3a") && exists("fig3b")) {
  fig3 <- fig3a + fig3b +
    plot_layout(widths = c(3, 2)) +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold", size = 16))

  print(fig3)
  save_fig(fig3, "SI_group_specificity_trajectory.pdf", width = 14, height = 5)
}
```

# 6. Summary Table

```{r summary-table}
summary_rows <- list()

# Length
s_len <- coef(summary(m_length))
summary_rows[[1]] <- tibble(
  Criterion = "Description length decreases",
  `β` = round(s_len["blockNum_c", "Estimate"], 3),
  SE = round(s_len["blockNum_c", "Std. Error"], 3),
  `t` = round(s_len["blockNum_c", "t value"], 2),
  Result = ifelse(s_len["blockNum_c", "Estimate"] < 0, "PASS", "FAIL")
)

# Accuracy
s_acc <- coef(summary(m_accuracy))
summary_rows[[2]] <- tibble(
  Criterion = "Listener accuracy increases",
  `β` = round(s_acc["blockNum_c", "Estimate"], 3),
  SE = round(s_acc["blockNum_c", "Std. Error"], 3),
  `t` = round(s_acc["blockNum_c", "t value"], 2),
  Result = ifelse(s_acc["blockNum_c", "Estimate"] > 0, "PASS", "FAIL")
)

# Adjacent similarity
if (has_adjacent && exists("m_adjacent")) {
  s_adj <- coef(summary(m_adjacent))
  summary_rows[[3]] <- tibble(
    Criterion = "Convention stability increases",
    `β` = round(s_adj["blockNum_c", "Estimate"], 3),
    SE = round(s_adj["blockNum_c", "Std. Error"], 3),
    `t` = round(s_adj["blockNum_c", "t value"], 2),
    Result = ifelse(s_adj["blockNum_c", "Estimate"] > 0, "PASS", "FAIL")
  )
}

# Group-specificity
if (exists("perm_results")) {
  prop_sig <- mean(perm_results$p_value < 0.05)
  summary_rows[[length(summary_rows) + 1]] <- tibble(
    Criterion = "Group-specificity (≥80% significant)",
    `β` = NA,
    SE = NA,
    `t` = NA,
    Result = sprintf("%d/%d games sig. — %s",
                     sum(perm_results$p_value < 0.05),
                     nrow(perm_results),
                     ifelse(prop_sig >= 0.8, "PASS", "FAIL"))
  )
}

bind_rows(summary_rows) %>%
  knitr::kable(caption = "Outcome-neutral criteria summary")
```

# 7. Descriptive Statistics for LaTeX

Collect all numbers needed for the SI text in one place.

```{r latex-numbers}
cat("====== Numbers for LaTeX SI section ======\n\n")

# Sample
cat(sprintf("Games: %d, Players: %d\n", n_games, n_players))
cat(sprintf("Conditions: %s\n", paste(games$condition, collapse = ", ")))

# Duration
cat(sprintf("\nMedian session duration: %.1f min (range: %.1f–%.1f)\n",
            median(duration_by_game$duration_min),
            min(duration_by_game$duration_min),
            max(duration_by_game$duration_min)))

# Completion
n_complete <- sum(games$activeGroups == 3)
cat(sprintf("Games with all groups active: %d/%d\n", n_complete, n_games))

# Description length
s_len <- coef(summary(m_length))
cat(sprintf("\nDescription length: β = %.3f, SE = %.3f, t = %.2f\n",
            s_len["blockNum_c", "Estimate"],
            s_len["blockNum_c", "Std. Error"],
            s_len["blockNum_c", "t value"]))
cat(sprintf("  Block 1: M = %.1f words; Block 6: M = %.1f words\n",
            mean(length_first), mean(length_last)))

# Accuracy
s_acc <- coef(summary(m_accuracy))
cat(sprintf("\nAccuracy: β = %.3f, SE = %.3f, t = %.2f\n",
            s_acc["blockNum_c", "Estimate"],
            s_acc["blockNum_c", "Std. Error"],
            s_acc["blockNum_c", "t value"]))
cat(sprintf("  Block 1: M = %.2f; Block 6: M = %.2f\n",
            mean(acc_first, na.rm = TRUE),
            mean(acc_last, na.rm = TRUE)))

# Adjacent similarity
if (has_adjacent && exists("m_adjacent")) {
  s_adj <- coef(summary(m_adjacent))
  cat(sprintf("\nAdjacent similarity: β = %.3f, SE = %.3f, t = %.2f\n",
              s_adj["blockNum_c", "Estimate"],
              s_adj["blockNum_c", "Std. Error"],
              s_adj["blockNum_c", "t value"]))
}

# Group-specificity
if (exists("perm_results")) {
  cat(sprintf("\nGroup-specificity: %d/%d games significant (p < .05)\n",
              sum(perm_results$p_value < 0.05), nrow(perm_results)))
  perm_results %>%
    left_join(games %>% select(gameId, condition), by = "gameId") %>%
    select(condition, obs_coefficient, p_value) %>%
    mutate(across(where(is.numeric), ~round(., 4))) %>%
    print()
}

# Idle / dropout
n_idle <- nrow(players %>% filter(idleRounds > 0))
n_inactive <- nrow(players %>% filter(isActive == FALSE | isActive == "false"))
cat(sprintf("\nPlayers with idle rounds: %d/%d\n", n_idle, n_players))
cat(sprintf("Players marked inactive: %d/%d\n", n_inactive, n_players))

# Total trials and messages
cat(sprintf("\nTotal trials: %d\n", nrow(trials)))
cat(sprintf("Total messages: %d\n", nrow(messages)))
cat(sprintf("Speaker utterances: %d\n", nrow(speaker_utts)))
```
